### 2014-02-23 pituba/article/article.tex
Added pituba solver for material balance equations.

% Review
\documentclass[authoryear,preprint,review,12pt]{elsarticle}
%\documentclass[final,authoryear,5p,times,twocolumn,11pt]{elsarticle}

% Review topics:

% % TERMINOLOGY
% % fluid saturations or phase saturations?
% % Block-Oil, Black-oil or black-oil?

%% for a journal layout:
%% \documentclass[final,authoryear,1p,times]{elsarticle}
%% \documentclass[final,authoryear,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,authoryear,3p,times]{elsarticle}
%% \documentclass[final,authoryear,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,authoryear,5p,times]{elsarticle}

% Pending modifications:
% - Add bubble-point pressure to equations after volume and time integration.
% - A different fluid distribution DOES NOT lead to a different value of averaged saturation.
% - CORRECTION: saturations need not be constant in order to Bubble-point pressure be constant.

%\documentclass[final,authoryear,5p,times,twocolumn,11pt]{elsarticle}

\usepackage{lineno}
\usepackage{amsmath}
\usepackage{natbib}
\usepackage{booktabs}
%opening

\newtheorem{prop}{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{step}{Step}

\begin{document}
\begin{frontmatter}
\title{Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equations}
%\title{Obtaining the classical material balance equations from black-oil equations}
%\journal{Journal of Petroleum Science and Engineering}
%\author{R.B. Piccinini}\ead{rbpiccinini@gmail.com}
%\address{Petrobras - UO-BA/ATP-N/RES. Salvador, Bahia, Brazil.}

\begin{abstract}
This work presents a derivation of the classical material balance equations starting from black-oil equations in integral form. The existing link between both approaches is discussed and an interpretation for pressure in material balance calculations is provided. It is shown that a condition stronger than hydrostatic equilibrium is needed to reduce black-oil equations to the classical approach. The classical equations are formulated for a variable bubble-point pressure, extending their applicability to reservoirs under pressure maintenance. Sufficient conditions for solution uniqueness are presented and a scheme for numerical solution is proposed.
\end{abstract}

\begin{keyword}
material balance \sep black-oil equations \sep reservoir engineering.
\end{keyword}
\end{frontmatter}

\section{Introduction}
The so called material balance equations are widely employed in reservoir engineering and have become a classical tool in oil reservoir management. The equations were originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from phase equilibrium and balance of fluid production and volumetric expansion. \cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability, where most of the produced volume would come from and where pressure would present a more equal behavior. Such regions were said to contain an \textit{active oil}.

The equations developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium. Also, the zero-dimensional approach excluded the possibility of mechanical equilibrium under gravity and capillary forces, which produces a variable fluid distribution in space. An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}. The novelty on hydrodynamics was the introduction of a macroscopic equation of motion employing Darcy's law.

%Black-oil equations arise from the generalization of the zero dimensional concept to a multidimensional macroscopic model which incorporates hydrodynamics via Darcy's law and entails exactly the same thermodynamic approach for phase equilibrium \citep{aziz1979petroleum}.

%Although many applications of material balance equations have been reported, few explore their connection.

The term \textit{black-oil} refers to specific hydrocarbon compositions whose volume and phase equilibria can be well described by a two pseudo-component system for isothermal changes in pressure. A general restriction is that reservoir temperature must be sufficiently below the mixture critical temperature so that different fluid phases have clearly distinguishable properties \citep{danesh1998pvt,firoozabadi1999thermodynamics}.

Since black-oil and material balance equations differ only in the number of spatial dimensions and the existence of local pressure gradients, there must exist conditions under which material balance equations can be recovered from the black-oil approach. This existing link was briefly addressed in \cite{ertekin2001basic}, where both were argued to be equivalent if capillarity and gradients of flow potential $\Phi$ could be neglected, i.e.:
\begin{subequations}
\begin{equation}
p_\alpha = p \, ,
\end{equation}
\begin{equation}
\nabla \Phi_\alpha = \nabla p_\alpha - \rho_\alpha g \nabla z = 0 \, ,
\end{equation}
\end{subequations}
for $\alpha=o,g,w$ referring to oil, gas and water phases.

It is shown in this work that a stronger assumption is actually needed: mechanical equilibrium in the presence of gravity ($\nabla p = \rho g \nabla z$) is not a sufficient condition, pressure gradients must identically vanish instead ($\nabla p_\alpha \equiv 0$).
%\begin{equation}
%\nabla p \equiv 0 \, .
%\end{equation}

The restriction of a spatially constant pressure throughout the reservoir does not require PVT properties to be averaged during the integration procedure. Such averaging process is not straightforward, if not impractical, as it is shown later. For the same reason, bubble-point pressure must also be constant and it is shown that this latter assumption is equivalent to having constant fluid saturations in space.

%The requirement of a spatially constant pressure makes the material balance equations more suitable for shallow reservoirs, where gravity field is less important in phase equilibrium.

The present derivation of material balance equations allows bubble-point pressure to vary with time accordingly to the same concept widely employed in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}. The employed approach extends material balance equations to saturated reservoirs under pressure maintenance (e.g. fluid injection). The applicability, however, depend on how important non-equilibrium effects are to the phase equilibrium.

%It has already been shown the importance of not having strong flow near interface that prevents mixing and causes pressure to drop locally \citep{steffensen1973reservoir,utseth1981numerical}.

After presenting material balance equations, sufficient conditions for solution uniqueness are imposed on PVT properties. The conditions simply require fluids to be compressible and that gas dissolution occurs with reduction of system volume.

A simple scheme for numerical solution is also proposed based on classical root-finding algorithms for non-linear equations.

%\section{Black-Oil Equations}
%
%Here it is used a thermodynamic fluid model known as Black-Oil \cite{blackoil}. It consists of a simplified treatment of hydrocarbon composition and phase equilibrium. Hydrocarbon species are grouped into two pseudo-components called oil and gas, being a pseudo-component one which consists of more than one chemical molecule.
%
%Oil is composed by hydrocarbon species that remain in liquid phase for standard surface pressure and temperature. The other hydrocarbon species in gaseous phase for the same surface conditions are called gas.
%
%The mixture of reservoir fluid is then composed of oil, gas and water components and can present three fluid phases: oil, gas and water. Oil phase may be composed of oil and gas components, gaseous phase may be composed uniquely by gas component and water phase uniquely by water component.
%
%In addition to the three fluid components and phases, there is a solid phase composed by the rock.
%
%The fluid phases may be defined in a more general framework so as to account for dissolution of gas in water or water vapour, for instance. Such possibilities are not dealt with in this work.
%
%For some volume containing one of the three fluid phases, the saturation of the fluid phase $k$ is defined as
%\begin{equation}\label{eq: saturation}
%S_{k}=\frac{\phi_k}{\sum_{i \neq solid} \phi_i} \, ,
%\end{equation}
%where $\phi_i$ is the volume fraction of the fluid phase $i$.
%
%In order to apply the volume balance, a control volume ($\Omega$) is defined containing all the hydrocarbon zone. It consists of the region where hydrocarbon saturation is greater than zero for some arbitrarily small volume. This reservoir definition excludes the pure water zone, which is referred to as the aquifer.
%
%
%Right hand side of Equation \eqref{eq: matbal1}, \textit{rock and fluid expansion}, is a function of phase mass transfers and compressibilities. Both should be obtained from a thermodynamic model accounting for mass transfer among phases and the relation between phase volumes and pressure. Temperature is seldom considered to vary during reservoir production and it is held constant and equal in all phases, not being an independent property of the control volume state.
%
%Such thermodynamic model is presented in detail in \cite{blackoil} and it is commonly referred to as Black-Oil Equations. The main assumption behind them is that phase and pressure equilibria are instantaneously achieved inside the control volume, i.e., mass transfer among phases reaches equilibrium and internal forces equilibrate external ones fast enough in comparison to the interval of time between two observations of control volume properties.
%
%Pressure in each phase may differ if capillary forces are taken into account, for what extra constitutive equations relating each phase pressures would be needed.
%Capillary pressure would be function of porous medium properties given by some porous medium model, e.g., permeability and porosity; phase properties, e.g., surface tension and wettability, and the thermodynamic state.
%
%Capillary pressure is very often assumed to be function of phase saturations only and, hence, the thermodynamic state of the control volume would be completely determined by the pressure of one phase (e.g. the wetting phase) and phase saturations.
%
%Black-Oil Equations do not assume water and oil phases becoming miscible for any thermodynamic state.
%
%Capillary pressure dependence on phase saturation only is often considered as a weak assumption in contradiction with experimental observations. The hysteresis of capillary pressure exhibited when either imbibition or drainage occurs is not theoretically explained and it is accounted for by empirical relations or it is rather simply neglected.
%
%\cite{hilfer1} has proposed the inclusion of an energy equation.

\section{Derivation of Equations}

This section starts presenting the more general black-oil equations. It then proceeds by imposing the more restrictive conditions for wich material balance equations arise.

\subsection{Black-Oil Equations}

As previously mentioned, black-oil is a designation to a simplified treatment of hydrocarbon composition and phase equilibrium. Hydrocarbon species are grouped into two pseudo-components called oil and gas, being a pseudo-component one that consists of more than one chemical molecule.

Oil component is composed by hydrocarbon species that remain in liquid phase for standard surface pressure and temperature (stock tank oil). In contrast, gas component is composed by hydrocarbon species in gaseous phase for the same surface conditions.

Reservoir fluids are then composed of the two pseudo-components (oil and gas) plus the water component. There can exist three fluid phases: oil, gas and (liquid) water. Oil phase may be composed of oil and gas components, gas phase may be composed uniquely by the gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases.

In addition to the three fluid components and phases, there is the solid phase composed by the rock.

Mass balance equations for each component in fluid phase:

\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV + \oint_{\partial \Omega} \rho_w \left( \mathbf{u}_w \cdot \mathbf{dS} \right) = \rho_w^{ST}Q_w^{ST} \, ;
\end{equation}

\item oil component:
\begin{equation}\label{eq: So1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV + \oint_{\partial \Omega} \rho_o Y_o \left(\mathbf{u}_o \cdot \mathbf{dS} \right)= \rho_o^{ST}Q_o^{ST} \, ;
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg1}
\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV + \oint_{\partial \Omega} \left( \rho_o Y_g \mathbf{u}_o + \rho_g \mathbf{u}_g\right) \cdot \mathbf{dS}= \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}
Mass concentration of oil and gas components in oil phases are denoted as $Y_o$ and $Y_g$, respectively.

Phase saturation is defined as its volumetric fraction of the pore space:
\begin{equation}
S_\alpha = \frac{\phi_\alpha}{\phi} \, ,
\end{equation}
where $\phi_\alpha$ is the volumetric fraction of a fluid phase $\alpha$ and $\phi$ is the porosity, i.e., the volumetric fraction of the pore volume. Since we consider that pore volume is fully saturated with fluid,
\begin{equation}
S_o + S_g +S_w = 1 \, .
\end{equation}

A common simplification in black-oil formulations is to restrict mass transport to macroscopic advection. No diffusive or dispersive term is included in the mass balance equations. This simplification must be looked with care when gradients of species concentration are important and the following assumption does not hold:

\begin{equation}
\nabla Y_o = \nabla Y_g = 0 \, .
\end{equation}

For instance, when gas is being dissolved in oil phase in a reservoir under repressurization, this condition is not generally true.

%Mass balance equations for each fluid phase are \citep{ertekin2001basic}:
%% Oleo
%\begin{equation}\label{eq: So1}
%\frac{\partial}{\partial t}\left(\frac{\phi(p_P) S_o}{B_o(p_o,P_b)}\right)+\nabla\cdot\left(\frac{\mathbf{u}_o}{B_o (p_o,P_b)}\right)=q_{os} \, ,
%\end{equation}
%% Gas
%\begin{equation}
%\begin{split}\label{eq: Sg1}
%\frac{\partial}{\partial t}\left[\phi(p_P)\left(\frac{S_g}{B_g (p_g)} +\frac{S_o R_s (p_o,P_b)}{B_o (p_o,P_b)}\right)\right] \\
%+\nabla\cdot\left(\frac{\mathbf{u}_o R_s (p_o,P_b)}{B_o (p_o,P_b)}+\frac{\mathbf{u}_g}{B_g (p_g)}\right)=q_{gs} \, ,
%\end{split}
%\end{equation}
%% Agua
%\begin{equation}\label{eq: Sw1}
%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)+\nabla\cdot\left(\frac{\mathbf{u}_w}{B_w(p_w)}\right)=q_{ws} \, ,
%\end{equation}
%
%where saturations must satisfy the constraint
%\begin{equation}\label{eq: S11}
%S_o+S_w+S_g=1 \, .
%\end{equation}
%
%Bubble-point pressure ($P_b$) is here defined as the minimum pressure for which gas phase is locally absent, i.e., $S_g=0$. Its local value is implicitly given by the solubility ratio ($R_s$) satisfying the equation:
%\begin{equation}\label{eq: Pb}
%R_s(p_o=P_b,P_b) = \frac{B_g(p_g)}{B_o(p_o,P_b)}\frac{S_g}{S_o} +R_s(p_o,P_b) \, .
%\end{equation}
%It must be remarked that bubble-point pressure is sometimes defined in a different manner as the bubble-point pressure for the hydrocarbon composition in oil phase only. According to this latter definition, $P_b = p_o$ for any saturated reservoir.
%
%Black-oil equations make the assumption of instantaneous mass transfer between oil and gas phases, what is only reasonable for a reservoir under depletion, when oil and gas components are homonegeously distributed in oil phase. Once constant temperature and fluid composition are assumed, diffusion and advection of species and energy become unimportante for phase equilibrium calculations.
%
%If gradients of species concentration are present within oil phase, however, as it occurs in saturated reservoirs under pressure maintenance, a combined mechanism of advection and diffusion exists and those are absent of black-oil equations. Some heuristic models have been proposed for problems where this issue is critical, as for gas injection projects. \cite{utseth1981numerical} has proposed to limit $dR_s/dt$ and neglect transport of gas component in oil phase for a simplified treatment.

Momentum balance for each phase is given by an extension of Darcy's law to multiphase flow. This extension consists in multiplying absolute (or single-phase) permeability ($\mathbf{k}$) by a scalar function called phase relative permeability ($k_{r,\alpha}$):

\begin{equation}\label{eq: exdarcy}
\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right) \, , \quad \alpha=o,w,g \, .
\end{equation}

Phase densities are functions of the local thermodynamic state:
\begin{equation}
\rho_w = \rho_w \left(p_w, T_w\right)
\end{equation}
\begin{equation}
\rho_o = \rho_o \left(p_o, T_o,Y_g\right)
\end{equation}
\begin{equation}
\rho_g = \rho_g \left(p_g, T_g\right)
\end{equation}

Another simplification is now made assuming thermal equilibrium in the entire reservoir volume:
\begin{equation}
T_w = T_o = T_g = T
\end{equation}
and
\begin{equation}
\nabla T \equiv 0 \, .
\end{equation}

Temperature is thus a constant prescribed value and no energy balance equation is needed. Also, it follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom. As temperature is assumed constant, oil phase density in saturated states is a function of pressure or composition only.

Porosity ($\phi$) is assumed a function of pore pressure ($p_P$). Since three phases are present in general, such quantity is defined by some relation of type \citep{kim2011rigorous}:
\begin{equation}
p_P=p_P\left(p_o,p_w,p_g,S_o,S_w,S_g\right) \, .
\end{equation}

In order to close the system of equations, a connection among different phase pressures must be provided. Such expressions are referred to as capillary pressures and are usually employed in the following forms \citep{aziz1979petroleum}:
%capillary
\begin{equation}\label{eq: capillary_ow}
p_{c,ow}\left(\mathbf{k},\phi,S_w\right)=p_o-p_w \, ,
\end{equation}
\begin{equation}\label{eq: capillary_og}
p_{c,go}\left(\mathbf{k},\phi,S_g\right)=p_g-p_o \, .
\end{equation}

%Capillary pressure and relative permeability are controversial concepts in the sense that they have not been demonstrated to arise from theoretical foundations. For practical applications, both functions are usually assumed in the forms presented above and their properties are argued to be empirically established.

%It is well known that black-oil equations possess discontinuous solutions in saturation, bubble pressure and solubility ratio fields. Discontinuities (or shocks) are introduced by initial or boundary conditions and are propagated throughout flow domain due to the hyperbolicity of equations.
%
%In initial conditions, if hydrostatic equilibrium is imposed and no capillary pressure is included, that is,
%\begin{equation}
%p_{c,ow}=p_{c,og}=0 \quad \longrightarrow \quad p_o=p_w=p_g=p\, ,
%\end{equation}
%saturation discontinuities will be present due to gravity segregation.
%
%When capillary pressure is present and it is function of phase saturations, a diffusive term is introduced in mass balance equations and shocks originated by phase displacement become smoothed:
%\begin{equation}
%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)-\nabla\cdot\left[\frac{k_{r,w} \mathbf{k}}{B_w(p_w)\mu_w} \left(\nabla p_o - \rho_w g\nabla z -\underbrace{\frac{d p_c}{d S_w}\nabla S_w}_{\text{water diffusion}}\right)\right]=q_{ws} \, .
%\end{equation}
%
%Discontinuities in saturation may still occur, however, if capillary pressures are functions of discontinuous porous medium properties ($phi$ and $\mathbf{k}$) or if phase change takes place, as diffusion of gas dissolved in oil phase is not allowed.
%Cite Bear and Herrera.

\subsection{Material Balance Equations}

Derivation of the classical material balance equations consists in assuming no fluid motion and in integrating black-oil equations in the entire reservoir volume. The integration process, however, requires all quantities to be either averaged or assumed constant in space.

%Defining averages of most quantities is difficulty, though, as they appear in equations as products of up to three factors. Being the quantities self-correlated, the averaging process would involve unclosed terms. Quantities are thus required to be spatially constant wherever needed.

%A second problem is that averaging requires the knowledge of a spatial distribution of quantities for it be uniquely specified, as it can be seen for an average saturation depending on spatial distributions of local saturation and porosity,
%\begin{equation}\label{eq: avS}
%\bar{S}_\alpha = \int_\Omega S_\alpha \phi dV \, .
%\end{equation}

We start assuming mechanical equilibrium, i.e., no bulk motion:
\begin{equation}
\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right)\equiv 0 \, .
\end{equation}
This condition implies that pressure gradients must equal gravity force per unit volume:
\begin{equation}\label{eq: p=rhogh}
\nabla p_\alpha = \rho_\alpha g\nabla z
\end{equation}
and, consequently, pressure is a function of depth only:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha^0}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'},T\right)} = g\left(z - z_0\right) \, .
\end{equation}

Black-oil equations then become:
\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw2}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV = \rho_w^{ST}Q_w^{ST} \, ;
\end{equation}

\item oil component:
\begin{equation}\label{eq: So2}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV = \rho_o^{ST}Q_o^{ST} \, ;
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg2}
\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV = \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}
%

We now define space averaged quantities for porosity ($\phi$), saturation ($S_\alpha$) and density ($\rho_\alpha$) of a phase $\alpha$ and mass concentration of a component $k$ in phase $\alpha$ ($Y_{\alpha,k}$).

All quantities are locally functions of space and time, but, after averaged in space, they become function of time only:

\begin{equation}
\bar{\phi} (t) = \frac{1}{\Omega}\int_{\Omega}\phi \left(\mathbf{x},t\right) dV \, .
\end{equation}

We proceed omitting function arguments:
\begin{equation}
\bar{S_\alpha}\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}S_\alpha \phi dV \, ,
\end{equation}

\begin{equation}
\bar{\rho_\alpha}\bar{S_\alpha}\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha S_\alpha \phi dV \, ,
\end{equation}

\begin{equation}
\bar{\rho_\alpha}\bar{Y_{\alpha,k}}\bar{S_\alpha}\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha Y_{\alpha,k} S_\alpha \phi dV \, .
\end{equation}

Eqs. \eqref{eq: Sw2}, \eqref{eq: So2} and \eqref{eq: Sg2} rewritten in terms of averages read:

\begin{itemize}
\item water component:
\begin{equation}\label{eq: Sw3}
\frac{\partial}{\partial t} \left( \bar{\rho_w}\bar{S_w}\bar{\phi} \Omega \right) = \rho_w^{ST}Q_w^{ST}
\end{equation}

\item oil component:
\begin{equation}\label{eq: So3}
\frac{\partial}{\partial t} \left( \bar{\rho_o}\bar{Y_o}\bar{S_o}\bar{\phi} \Omega \right) = \rho_o^{ST}Q_o^{ST}
\end{equation}

\item gas component:
\begin{equation}\label{eq: Sg3}
\frac{\partial}{\partial t} \left[\left( \bar{\rho_o}\bar{Y_g} \bar{S_o} + \bar{\rho_g} \bar{S_g}  \right) \bar{\phi} \Omega\right] = \rho_g^{ST}Q_g^{ST} \, .
\end{equation}
\end{itemize}

We shall now introduce very common definitions coming from petroleum engineering for the description of volumetric and solubility changes in fluid phases. In concerning volumetric changes, $B_w$, $B_o$ and $B_g$, also called formation volume factors of water, oil and gas phases, respectively; express the ratio of a phase volume in some given pressure and temperature conditions to the volume of this same phase in standard pressure and temperature conditions, i.e.:

\begin{equation}
B_w = \frac{\rho_w^{ST}}{\bar{\rho_w}} \, ,
\end{equation}

\begin{equation}
B_o = \frac{\rho_o^{ST}}{\bar{Y_o} \bar{\rho_o}} \, ,
\end{equation}

\begin{equation}
B_g = \frac{\rho_g^{ST}}{\bar{\rho_g}} \, .
\end{equation}

In concerning changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between dissolved gas component volume and oil component volume both measured in standard conditions:

\begin{equation}
R_s = \frac{\rho_o^{ST} / \bar{Y_o}}{\rho_g^{ST} / \bar{Y_g}} \, .
\end{equation}

Equations can now be rewritten using the previous definitions and porous volume given by $V_p = \bar{\phi}\Omega$.

\begin{equation}\label{eq: Sw4}
\frac{\partial}{\partial t} \left(\frac{\bar{S_w} V_p}{B_w} \right) = Q_w^{ST} \, ,
\end{equation}

\begin{equation}\label{eq: So4}
\frac{\partial}{\partial t} \left(\frac{\bar{S_o} V_p}{B_o} \right) = Q_o^{ST} \, ,
\end{equation}

\begin{equation}\label{eq: Sg4}
\frac{\partial}{\partial t} \left[ V_p \left(\frac{\bar{S_g}}{B_g} + \frac{\bar{S_o} R_s}{B_o} \right) \right] = Q_g^{ST} \, ,
\end{equation}

\begin{equation}\label{eq: S14}
\bar{S_w}+\bar{S_o}+\bar{S_g}=1 \, .
\end{equation}

Equations from \eqref{eq: Sw4} to \eqref{eq: S14} constitute the most general form of material balance equations. So far, the only simplification made to black-oil equations was the absence of fluid motion, allowing pressure to be computed from Eqs.\eqref{eq: capillary_ow}, \eqref{eq: capillary_og} and \eqref{eq: p=rhogh}, and equations to be integrated in the entire reservoir volume.

We now look to solve Eqs. \eqref{eq: Sw3}, \eqref{eq: So3} and \eqref{eq: Sg3} for the averaged quantities instead of solving them for the original space dependent quantities.

Eq.~\eqref{eq: p=rhogh} gives an expression of how pressure, and hence PVT properties, depends on position. If gravity and capillarity are also neglected, fluid pressure becomes independent of position and phase:
\begin{subequations}\label{eq: fullsimp}
\begin{equation}
p_o=p_w=p_g=p_P=p \, ,
\end{equation}
\begin{equation}
\nabla p \equiv 0 \, .
\end{equation}
\end{subequations}

%In order to apply the volume balance, a control volume ($\Omega$) is defined containing all the hydrocarbon zone. It consists of the region where hydrocarbon saturation is greater than zero for some arbitrarily small volume. This reservoir definition excludes the pure water zone, which is referred to as the aquifer.
Bubble-point pressure is also a function of position as it depends on local values of oil and gas saturations. Assuming a spatially constant bubble-point pressure is equivalent to assuming constant saturations, as the quantities are related by Eqs.~\eqref{eq: S11} and \eqref{eq: Pb}.

%Saturations are thus assumed constant in space so that bubble-point pressure becomes constant and an average saturation can be uniquely defined by Eq.~\eqref{eq: avS}.

For spatially constant fluid and bubble-point pressures, Eqs.~\eqref{eq: matbal1a}, \eqref{eq: matbal1b} and \eqref{eq: matbal1c} can be integrated over the entire reservoir domain $\Omega$, which is defined as the region where hydrocarbon saturation is greater than zero for some arbitrary volume (provided that the restriction of macroscopic scale is observed). This reservoir definition excludes any surrounding aquifer zone.

The interpretation of pressure computed from material balance equations becomes clear at this point: it is the reservoir pressure in the absence of pressure gradients, phase segregation, capillary forces and gravity.

Performing volume and then time integration of Eq.~\eqref{eq: matbal1a} between times $t_1$ and $t_2$, material balance for oil phase is obtained:
\begin{equation}
\int_{t_1}^{t_2}\int_{\Omega} \frac{\partial}{\partial t}\left(\frac{\phi(p) S_o}{B_o(p,P_b)}\right) dV dt = \int_{t_1}^{t_2}\int_{\Omega} q_{os} dV dt\, ,
\end{equation}
\begin{equation}
\left[ \frac{S_o}{B_o(p,P_b)} \int_{\Omega}\phi dV \right]^{t_2}_{t_1} = -N_p \, ,
\end{equation}
\begin{equation}\label{eq: So5}
\frac{V_{p1} S_{o1}}{B_{o1}}-\frac{V_{p2}S_{o2}}{B_{o2}} = N_p \, .
\end{equation}

Equations for gas and water phases are obtained following the same procedure:
% Gas
\begin{equation}\label{eq: Sg5}
V_{p2}\left(\frac{S_{g2}}{B_{g2}} +\frac{S_{o2} R_{s2}}{B_{o2}}\right) - N\left(m\frac{B_{o1}}{B_{g1}} + R_{s1}\right)=G_i-G_p \, ,
\end{equation}
% Agua
\begin{equation}\label{eq: Sw5}
\frac{V_{p2} S_{w2}}{B_{w2}} - W=W_i+W_e-W_p \, ,
\end{equation}
%
\begin{equation}\label{eq: S15}
S_o+S_w+S_g=1 \, ,
\end{equation}
where
\begin{equation}
N=\frac{V_{p1} S_{o1}}{B_{o1}} \, ,
\end{equation}
\begin{equation}
W=\frac{V_{p1} S_{w1}}{B_{w1}} \, ,
\end{equation}
\begin{equation}
m=\frac{V_{p1} S_{g1}}{B_{o1} N} \, .
\end{equation}

An equation for bubble-point pressure in integrated form can be obtained by setting $S_g=0$ in Eq.~\eqref{eq: Sg2} and solving for the solubility ratio at bubble-point pressure $R_s(p=P_b)$:
\begin{equation}
\begin{split}
R_{s2}\left(p=P_b\right) = &\Big[N\left(m B_{1}/B_{g1} +R_{s1}\right)\\
&+G_i-G_p\Big]/\left(N-N_p\right) \, .
\end{split}
\end{equation}

A single equation for pressure (without fluid saturations) can be obtained by substituting Eqs.~\eqref{eq: So2}, \eqref{eq: Sg2} and \eqref{eq: Sw2} into \eqref{eq: S12}. The resulting equation allows pressure to be calculated from given production ($F_p$) and injection ($F_i$) volumes:
\begin{equation}\label{eq: MBE}
F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{porous volume contraction } (C)}
\end{equation}
or
\begin{equation}\label{eq: MBE_short}
F_p - F_i= E + C \, ,
\end{equation}

where newly introduced terms are defined in Table~\ref{tab: MBE}. Eq.~\eqref{eq: MBE} is commonly referred to as the Material Balance Equation (MBE).

\section{Solution of Material Balance Equations}
A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We shall start defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
\begin{equation}\label{eq: residual}
\varepsilon(p_2) = \frac{E+F_i+C-F_p}{V_{p1}} \, .
\end{equation}
We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that, as a consequence, only one root of $\varepsilon$ exists at most. Equivalently, we may say that only one solution exists for material balance equation.

%The conditions imposed on fluid PVT properties are the mathematical equivalent of requiring a far from critical thermodynamic state.

\subsection{Mathematical Aspects of the Residual Function}

It is a known result of real analysis that a monotonic function $f:D\subset\mathcal{R}\rightarrow\mathcal{R}$ can have at most one root in $D$. Further, if $f$ is continuous in the closed interval $I=[a,b] \subset D$ and $f(a)f(b)<0$, then $f$ has one (and only one) root in $I$.

We now state without proof that the following affirmatives are sufficient conditions to classify a function $f$ as monotonic:
%We now show that $\varepsilon$ is a monotonic function since it is:
\begin{itemize}
\item $f$ is continuous;
\item $f$ is differentiable everywhere except for a countable set of points;
\item $f'<0$ everywhere.
\end{itemize}

Residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except at bubble-point pressure.
The last affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels corresponding each one to a different phase. Each phase parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
\begin{equation}
\begin{split}
&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right)}_\text{oil phase}\\
&+\underbrace{\left[N R_{s1}+mN\frac{B_{o1}}{B_{g1}}+G_i-\left(N-N_p\right)R_{s2}-G_p\right] \frac{dB_g}{dp}}_\text{gas phase}\\
&+\underbrace{\left(W + W_i+W_e-W_p\right)\frac{dB_w}{dp}}_\text{water phase}\\
&\underbrace{-\frac{dV_p}{dp}}_\text{porous volume}\, .
\end{split}
\end{equation}

\textit{In-situ} volumes (e.g. $N-N_p$ for oil phase) are clearly non-negative values. Pressure derivatives of fluid volumes are all negative values (as fluid pressure increases, fluid volume contracts) and pressure derivative of rock porous volume is a positive value (as fluid pressure increases, porous volume expands).

The value of $d\varepsilon/dp$ is thus negative because each term is a product of a negative and a positive value, being the sum of all the four terms also a negative value.

In summary, the following four relations constitute sufficient conditions for $\varepsilon(p)$ be a strictly decreasing function and for material balance equations to have only one solution:
\begin{equation}\label{eq: PVTH1}
\left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right) <0 \, ,
\end{equation}
\begin{equation}\label{eq: PVTH2}
\frac{dB_g}{dp} < 0 \, ,
\end{equation}
\begin{equation}\label{eq: PVTH3}
\frac{dB_w}{dp} < 0 \, ,
\end{equation}
\begin{equation}\label{eq: PVTH4}
-\frac{dV_p}{dp}<0 \, .
\end{equation}

Switching inequalities to equalities in Eqs.~\eqref{eq: PVTH1}, \eqref{eq: PVTH2} and \eqref{eq: PVTH3} is equivalent to enforcing system incompressibility. Eq.~\eqref{eq: PVTH1}, for instance, would require a volume increase in oil phase to be equal to gas volume in gas phase for any additional gas dissolution.

\subsection{Solution Algorithm}
Several methods exist to find roots of nonlinear continuous functions \citep{hamming1987numerical}. One possible choice is Newton's method, which linearizes the function using either its gradient or a secant line. The secant line approach introduces an artificial smoothness to the gradient direction near continuous but non-differentiable points, for what it is the approach employed here to deal with the bubble-point pressure.

An open source implementation of secant method is available in SciPy library \citep{scipy}. This work used this library version $0.12$.

%A more simple and robust option is the bisection method, which does not require the function to be differentiable anywhere. The number of iterations $n$ required to reduce search interval by a factor of $\delta$ is \citep{hamming1987numerical}:
%\begin{equation}
%n=\log_{2} (\delta)+1 \, .
%\end{equation}

%If the inital search interval is $[a;b]$, an error tolerance $|p^{n}-p^{n-1}|\leq (b-a)/\delta $ can be achieved in $21$ iterations for $\delta=10^{6}$.

%Either methods were successfully employed to obtain solution of equations.

The steps to solve all material balance equations are summarized bellow:
\begin{step}
Compute bubble-point pressure for instant $t_2$.
\end{step}

Bubble-point pressure $P_{b2}$ is only a function of the values in initial instant $t_1$ and produced and injected volumes.

\begin{step}
Compute fluid pressure for instant $t_2$ finding the root of residual function ($\varepsilon$).
\end{step}

Pressure in Eq.~\eqref{eq: MBE} is only a function of fluid properties in instant $t_1$, produced and injected values and bubble-point pressure at instant $t_2$, $P_{b2}$, which was already computed in step 1.

\begin{step}
Compute fluid saturations $t_2$.
\end{step}

Once $S_{o2}$, $S_{g2}$, $S_{w2}$, $p_2$ and $P_{b2}$ are known, solution for instant $t_2$ is complete.

\section{Results}
Material balance equations were solved for a black-oil reservoir initially producing in depletion drive mechanism and later subjected to water injection. Fig.~\ref{fig: matbal_p} shows in the solid line curve the severe depletion during initial phases of production. In the dashed curve, Fig.~\ref{fig: matbal_p} shows the bubble-point pressure, indicating how far the saturated reservoir is from undersaturation and allowing for planning of pressure maintenance projects.

Water injection starts when oil recovery factor $(RF)$ is about $RF=0.08$ and it is maintained until the reservoir reaches a novel undersaturated state. Initially, the high compressibility of gas prevents pressure from raising and free gas phase is reduced due to redissolution. Finally, the reservoir reaches the undersaturated state and a steep pressure increase is seen in Fig.~\ref{fig: matbal_p} for a recovery factor about $RF=0.15$.

It must be observed that the result of gas reentering oil phase completely neglects non-equilibrium effects such as compositional gradients and thus must be used with care. For instance, it may be in strong disagreement with real reservoir behavior when gravity segregation is present.

Figure~\ref{fig: matbal_s} shows the formation of a free gas phase with growing $S_g$ and then its disappearance. Depending on the ratio of vertical and horizontal permeabilities or reservoir dip angle, it may also suggest the formation of a secondary gas cap.

Figure \ref{fig: epsilon} shows the shape of residual function ($\epsilon$) with respect to pressure, i.e, holding all produced or injected volumes constant and allowing only pressure to change. Each different solid line corresponds to a different volume of fluid production and injection. The function root is the solution of material balance equation, Eq.~\eqref{eq: MBE_short}, and the non-differentiable point is the bubble-point pressure. Also, it is seen that function $\epsilon$ monotonically decreases with pressure, as theoretically shown before.

Figure \ref{fig: iter} shows the count of newtonian iterations required to compute pressure with an accuracy within $0.001$ of initial bubble-point pressure. Typically, $12$ iterations are required.

Figure \ref{fig: residual} shows the residual of numerical solution as defined in Eq.~\eqref{eq: residual}. Material balance error is bellow $10^{-7}$ units of porous volume for all time steps.

\section{Conclusions}
\begin{enumerate}[1.]
\item Material balance equations can be obtained from black-oil equations under the assumptions of:
\begin{enumerate}[(a)]
\item mechanical equilibrium;
\item absence of gravity and capillary forces;
\item spatially constant bubble-point pressure or, equivalently, fluid saturations.
\end{enumerate}
\item One unique solution of pressure equation exists for a compressible system;
\item A residual function can be defined such that its root solves pressure equation. A numerical solution can be obtained with nonlinear root-finding algorithms.
\end{enumerate}

\section*{Nomenclature}
\begin{tabular}{ll}
$B_g$ & gas formation volume factor\\
$B_o$ & oil formation volume factor\\
$B_w$ & water formation volume factor\\
$R_s$ & solubility ratio of gas in oil phase\\
$S_g$ & gas saturation \\
$S_o$ & oil saturation \\
$S_w$ & water saturation \\
$G_p$ & produced volume of gas \\
$m$ & ratio of original free gas volume to original oil phase volume in reservoir conditions \\
$N$ & original volume of oil in place \\
$N_p$ & produced volume of oil \\
$P_b$ & bubble-point pressure \\
$V_p$ & porous volume \\
$W_i$ & injected volume of water \\
$W_p$ & produced volume of water
\end{tabular}
\section*{References}
\bibliographystyle{elsarticle-harv}
\bibliography{refs}
\cite{*}

\section*{Highlights}
\begin{enumerate}[1.]
\item The existing link between black-oil and material balance equations is examined in detail.
\item It is shown that gravity and capillary forces must be neglected and saturations must be constant in space so one can integrate black-oil equations in reservoir volume and obtain material balance equations.
\item Material balance equations are shown to have a unique solution for a compressible system.
\item Material balance equations are formulated for a variable bubble-point and they are employed to a reservoir under pressure maintenance.
\end{enumerate}

\pagebreak

\begin{figure}
\centering
\includegraphics[width=\linewidth]{./python/matbal_p}
\caption{Reservoir pressure and bubble-point pressure as functions of recovery factor. Once depletion goes below bubble-point, pressure is hardly recovered for a saturated reservoir due to high system compressibility. Pressure is here given in units of the original bubble-point pressure.}
\label{fig: matbal_p}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{./python/matbal_water}
\caption{Volume of water production and injection and aquifer influx.}
\label{fig: matbal_water}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{./python/matbal_gas}
\caption{Volume of gas production.}
\label{fig: matbal_gas}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{./python/matbal_S}
\caption{Phase saturations. Gas saturation is reduced as water injection progresses.}
\label{fig: matbal_s}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{./python/matbal_res}
\caption{Shape of residual function ($\varepsilon$). Each curve corresponds to different volumes of production and injection. Residual function monotonically decreases with pressure and it has no derivative at the bubble-point.}
\label{fig: epsilon}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{./python/matbal_iter}
\caption{Count of newtonian iterations required to compute pressure with accuracy within $0.001$ units of initial bubble-point presure.}
\label{fig: iter}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{./python/matbal_residual}
\caption{Residual of numerical solution in units of porous volume.}
\label{fig: residual}
\end{figure}

\begin{table*}
\small
\centering
\begin{tabular}{ll}
\toprule
\multicolumn{2}{l}{\textit{Material Balance Equations}}\\
\midrule
oil & $S_{o2} = \left(N-N_p\right) B_{o2}/V_{p2}$ \label{eq: oiltab}
\\
water & $S_{w2} = \left(W+W_i+W_e-W_p\right) B_{w2}/V_{p2}$\\
gas & $S_{g2}=\left\lbrace N\left[\left(R_{s1}-R_{s2}\right)+m B_{o1}/B_{g1}\right]+N_p R_{s2}+G_i-G_p\right\rbrace Bg_2/V_{p2}$ \\
bubble-point pressure & $R_{s2}\left(P_b\right) = \left[N\left(m B_{1}/B_{g1} +R_{s1}\right)+G_i-G_p\right]/\left(N-N_p\right)$ \\
pressure & $F_p - F_i= N E_o + mN E_g + W E_w +C$ \\
\midrule
\multicolumn{2}{l}{\textit{Definition of terms in pressure equation}} \\
\midrule
fluid production & $F_p=N_p \left[B_{o2} + \left(G_p/N_p -R_{s2}\right)B_{g2}\right]+W_p B_{w2}$ \\
fluid injection and influx & $F_i=\left(W_i +W_e\right) B_{w2}+G_{i}B_{g2}$ \\
oil phase expansion & $E_o=\left(B_{o2}-B_{o1}\right) - \left(R_{so2}-R_{so1}\right)B_{g2}$ \\
gas phase expansion & $E_g = B_{o1}\left(B_{g2}/B_{g1}-1\right)$ \\
water volume expansion & $E_w = B_{w2}-B_{w1}$\\
porous volume contraction & $C=V_{p2}- V_{p1}$\\ \bottomrule
\end{tabular}
\caption{Complete set of material balance equations.}
\label{tab: MBE}
\end{table*}

% % %
% Calcular derivada da função resíduo com relação à pressão.
% % %

\end{document}

### 2014-02-23 pituba/pitubalib.py

@@ -1,0 +1,399 @@
+from pylab import *
+import scipy.optimize as sp
+
+class classPVT:
+	pe=[]
+	Boe=[]
+	Bge=[]
+	Bwe=[]
+	Rse=[]
+	uoe=[]
+	uae=[]
+	Bwe=[]
+	viscoe=[]
+	viscwe=[]
+	Pb=0.
+	co=0.
+	T=0.
+	Ze=0.
+
+	def Bo(self,p,Pb):
+		if p >= Pb:
+			return interp(Pb,self.pe[::-1],self.Boe[::-1])*(1.-self.co*(p-Pb))
+		else:
+			return interp(p,self.pe[::-1],self.Boe[::-1])
+
+	def Bg(self,p):
+		if type(self.Ze)==type(1.0):
+			Bgi=1.033/self.pe*(self.T)/293.
+			self.Ze=self.Bge/Bgi
+
+		Z=interp(p,self.pe[::-1],self.Ze[::-1])
+		return 1.033/p*(self.T)/293.*Z
+
+	def Bw(self,p):
+		return interp(p,self.pe[::-1],self.Bwe[::-1])
+
+	def Rs(self,p,Pb):
+		if p >= Pb:
+			return interp(Pb,self.pe[::-1],self.Rse[::-1])
+		else:
+			return interp(p,self.pe[::-1],self.Rse[::-1])
+
+	def visco(self,p):
+		return interp(p,self.pe[::-1],self.viscoe[::-1])
+
+	def viscw(self,p):
+		return interp(p,self.pe[::-1],self.viscwe[::-1])
+
+	def cw(self,p,pref):
+		return -(self.Bw(p)-self.Bw(pref))/(p-pref+0.01)/self.Bw(pref)
+
+
+
+class classAquifero:
+	WW=0.
+	p0=0.
+	dpc=0.
+	time=0 # time is given in days
+
+	cw, cr=0.,0.
+	k,L=0.,0.
+
+	pvt=classPVT()
+
+	pee=0.
+	Wee=0.
+
+	model='tank'
+
+	def We(self,p,dt,Np=0):
+		if self.model == 'tank':
+			return self.tank(p)
+		elif self.model == 'FetPcte':
+			return self.FetPcte(p,dt)
+		elif self.model == 'user':
[…]

### 2014-02-23 reservoirFoam/darcyTwoPhase/readGravitationalAcceleration.H
Added OpenFOAM code for petroleum reservoir simulation (still a very simple solver).

@@ -1,0 +1,17 @@
+Info<< "\nReading gravitationalProperties" << endl;
+
+IOdictionary gravitationalProperties
+(
+    IOobject
+    (
+        "gravitationalProperties",
+        runTime.constant(),
+        mesh,
+        IOobject::MUST_READ_IF_MODIFIED,
+        IOobject::NO_WRITE
+    )
+);
+
+const dimensionedVector g(gravitationalProperties.lookup("g"));
+const dimensionedScalar magg = mag(g);
+const dimensionedVector gHat = g/magg;

### 2014-02-23 reservoirFoam/darcyTwoPhase/tutorial/clean.sh

@@ -1,0 +1,2 @@
+#!/bin/bash
+rm -r 1* 2* 3* 4* 5* 6* 7* 8* 9*

### 2014-02-23 reservoirFoam/rockFluidModels/coreyModel/coreyModel.H

@@ -1,0 +1,233 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 2011 OpenFOAM Foundation
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::rockFluidModels::coreyModel
+
+Description
+     Standard power-law  model of relative permeability.
+
+SourceFiles
+    coreyModel.C
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef coreyModel_H
+#define coreyModel_H
+
+#include "rockFluidModel.H"
+#include "volFields.H"
+#include "surfaceFields.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+namespace rockFluidModels
+{
+
+/*---------------------------------------------------------------------------*\
+                           Class coreyModel Declaration
+\*---------------------------------------------------------------------------*/
+
+class coreyModel
+:
+    public rockFluidModel
+{
+    // Private data
+
+        dictionary coreyModelCoeffs_;
+
+/*
+        dimensionedScalar krww_("krww", dimensionSet(0,0,0,0,0,0,0), scalar(0.0)); 
+        dimensionedScalar nw_("nw", dimensionSet(0,0,0,0,0,0,0), scalar(0.0));
+        dimensionedScalar krnn_("krnn", dimensionSet(0,0,0,0,0,0,0), scalar(0.0));
[…]

### 2014-02-23 reservoirFoam/rockFluidModels/oldCoreyModel/coreyModel.H

@@ -1,0 +1,133 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 2011 OpenFOAM Foundation
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::rockFluidModels::coreyModel
+
+Description
+     Standard power-law non-Newtonian viscosity model.
+
+SourceFiles
+    coreyModel.C
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef coreyModel_H
+#define coreyModel_H
+
+#include "rockFluidModel.H"
+#include "volFields.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+namespace rockFluidModels
+{
+
+/*---------------------------------------------------------------------------*\
+                           Class coreyModel Declaration
+\*---------------------------------------------------------------------------*/
+
+class coreyModel
+:
+    public rockFluidModel
+{
+    // Private data
+
+        dictionary coreyModelCoeffs_;
+
+        dimensionedScalar krww_;
+        dimensionedScalar nw_;
+        dimensionedScalar krnn_;
+        dimensionedScalar nn_;
+
+        volScalarField krw_;
+        volScalarField krn_;
+
+
+    // Private Member Functions
+
+        //- Calculate and return the laminar viscosity
+        tmp<volScalarField> calcKrw() const;
+        
+        tmp<volScalarField> calcKrn() const;
+
+
+public:
+
+    //- Runtime type information
+    TypeName("coreyModel");
+
+
+    // Constructors
+
[…]

### 2014-02-23 reservoirFoam/rockFluidModels/rockFluidModel/rockFluidModel.C

@@ -1,0 +1,70 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 2011 OpenFOAM Foundation
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "rockFluidModel.H"
+#include "volFields.H"
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+namespace Foam
+{
+    defineTypeNameAndDebug(rockFluidModel, 0);
+    defineRunTimeSelectionTable(rockFluidModel, dictionary);
+}
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+Foam::rockFluidModel::rockFluidModel
+(
+    const word& name,
+    const dictionary& rockFluidProperties,
+    const surfaceScalarField& Swf,
+    const surfaceScalarField& rhowf,
+    const surfaceScalarField& rhonf,
+    const surfaceScalarField& muwf,
+    const surfaceScalarField& munf
+)
+:
+    name_(name),
+    rockFluidProperties_(rockFluidProperties),
+    Swf_(Swf),
+    rhowf_(rhowf),
+    rhonf_(rhonf),
+    muwf_(muwf),
+    munf_(munf)
+{}
+
+
+// * * * * * * * * * * * * * * Member Functions  * * * * * * * * * * * * * * //
+
+bool Foam::rockFluidModel::read(const dictionary& rockFluidProperties)
+{
+    rockFluidProperties_ = rockFluidProperties;
+
+    return true;
+}
+
+
+// ************************************************************************* //

### 2014-03-06 pituba/article/article.tex
inclusão do gradiente de pressão com a profundidade.

@@
 No diffusive or dispersive term is included in the mass balance equations.
-This simplification must be looked with care when gradients of species concentration are important and the following assumption does not hold:
-
-\begin{equation}
-\nabla Y_o = \nabla Y_g = 0 \, .
-\end{equation}
-
+This simplification must be looked with care when gradients of species concentration are important and $\nabla Y_i \approx 0$ does not hold for some species $i$.
 For instance, when gas is being dissolved in oil phase in a reservoir under repressurization, this condition is not generally true.
@@
 As temperature is assumed constant, oil phase density in saturated states is a function of pressure or composition only.
+The same is valid for mass concentrations of gas and oil components in oil phase.
+
+\begin{equation}
+Y_g = Y_g (p) \, ,
+\end{equation}
+\begin{equation}
+Y_o = Y_o (p) \, .
+\end{equation}
 
@@
 
-We shall now introduce very common definitions coming from petroleum engineering for the description of volumetric and solubility changes in fluid phases.
-In concerning volumetric changes, $B_w$, $B_o$ and $B_g$, also called formation volume factors of water, oil and gas phases, respectively; express the ratio of a phase volume in some given pressure and temperature conditions to the volume of this same phase in standard pressure and temperature conditions, i.e.:
+We shall now introduce very common definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases.
+In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, also known as formation volume factors of water, oil and gas phases, respectively; express the ratio of a phase volume in some given pressure and temperature conditions to the volume of this same phase in standard pressure and temperature condition, i.e.:
 
@@
 
-In concerning changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between dissolved gas component volume and oil component volume both measured in standard conditions:
+In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between dissolved gas component volume and oil component volume both measured in standard condition:
 
 \begin{equation}
-R_s = \frac{\rho_o^{ST} / \bar{Y_o}}{\rho_g^{ST} / \bar{Y_g}} \, .
+R_s = \frac{\rho_o^{ST} / \bar{Y_o}}{\rho_g^{ST} / \bar{Y_g}} \, ,
 \end{equation}
 
-Equations can now be rewritten using the previous definitions and porous volume given by $V_p = \bar{\phi}\Omega$.
+or using $\bar{Y}_o + \bar{Y}_g = 1$:
+\begin{equation}
+R_s = \frac{\rho_o^{ST}}{\rho_g^{ST}}\frac{\bar{Y}_g}{1-\bar{Y}_g} \, .
+\end{equation}
+
+Equations can now be rewritten using the above definitions and denoting porous volume by $V_p = \bar{\phi}\Omega$.
 
@@
 
-We now look to solve Eqs. \eqref{eq: Sw3}, \eqref{eq: So3} and \eqref{eq: Sg3} for the averaged quantities instead of solving them for the original space dependent quantities.
+We now look to solve Eqs. \eqref{eq: Sw3}, \eqref{eq: So3} and \eqref{eq: Sg3} for the averaged quantities.
+
+Furthermore, we look for the implications of neglecting gravity and capillarity and assuming a constant pressure throughout the reservoir.
+This additional simplification provides a direct relation between produced volumes and reservoir pressure.
+
+Assuming a constant isothermal compressibility to oil phase
+\begin{equation}
+\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o
+\end{equation}
+and using hydrostatic equilibrium $dp_o = \rho_o g dz$, we obtain the expression for oil density variation with depth:
+\begin{equation}
+\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
+\end{equation}
+
+For the gas phase, we obtain its isothermal compressibility from gas state equation
+\begin{equation}
+\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial p_g}\right)_{T,Z} = \frac{\rho_g}{p_g} = \frac{\mathcal{M}}{ZRT}
+\end{equation}
+
+And again we use hydrostatic equilibrium to obtain gas density variation with depth:
+\begin{equation}
+\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \left(\frac{\mathcal{M}}{ Z R T}\right) g
+\end{equation}
+
+%In oil zone:
+%\begin{equation}
+%\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g
+%\end{equation}
+
+\begin{equation}
+\left(\frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 10^{-6}\ m^{-1}
+\end{equation}
+
+\begin{equation}
+\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} \approx 10^{-5}\ m^{-1}
+\end{equation}
+If fluid densities do not vary appreciately with depth.
 
@@
 
-Figure \ref{fig: epsilon} shows the shape of residual function ($\epsilon$) with respect to pressure, i.e, holding all produced or injected volumes constant and allowing only pressure to change.
+Figure~\ref{fig: epsilon} shows the shape of residual function ($\epsilon$) with respect to pressure, i.e, holding all produced or injected volumes constant and allowing only pressure to change.
 Each different solid line corresponds to a different volume of fluid production and injection.
@@
 
-Figure \ref{fig: iter} shows the count of newtonian iterations required to compute pressure with an accuracy within $0.001$ of initial bubble-point pressure. Typically, $12$ iterations are required.
+Figure~\ref{fig: iter} shows the count of newtonian iterations required to compute pressure with an accuracy within $0.001$ of initial bubble-point pressure. Typically, $12$ iterations are required.
 
-Figure \ref{fig: residual} shows the residual of numerical solution as defined in Eq.~\eqref{eq: residual}. Material balance error is bellow $10^{-7}$ units of porous volume for all time steps.
+Figure~\ref{fig: residual} shows the residual of numerical solution as defined in Eq.~\eqref{eq: residual}. Material balance error is bellow $10^{-7}$ units of porous volume for all time steps.
 

### 2014-03-27 pituba/article/article.tex
Changes in text

@@
 
-Oil component is composed by hydrocarbon species that remain in liquid phase for standard surface pressure and temperature (stock tank oil).
-In contrast, gas component is composed by hydrocarbon species in gaseous phase for the same surface conditions.
+Oil component, also called stock tank oil, is defined as the hydrocarbon composition in liquid phase for standard surface pressure and temperature.
+Similarly, gas component is the hydrocarbon composition in gaseous phase for the same surface conditions.
+All species are generally present in both phases although the equilibrium composition is shifted to liquid phase (or gaseous phase) for components of greater (lower) molecular mass.
 
 Reservoir fluids are then composed of the two pseudo-components (oil and gas) plus the water component.
-There can exist three fluid phases: oil, gas and (liquid) water.
+There can exist three fluid phases: oil, gas and liquid water.
 Oil phase may be composed of oil and gas components, gas phase may be composed uniquely by the gas component and, water phase, uniquely by water component.
@@
 We shall now introduce very common definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases.
-In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, also known as formation volume factors of water, oil and gas phases, respectively; express the ratio of a phase volume in some given pressure and temperature conditions to the volume of this same phase in standard pressure and temperature condition, i.e.:
+In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, also known as formation volume factors of water, oil and gas phases, respectively; express the ratio of a phase volume in some given pressure and temperature condition to the volume of this same phase in standard pressure and temperature, i.e.:
 
@@
 
-Equations from \eqref{eq: Sw4} to \eqref{eq: S14} constitute the most general form of material balance equations.
+Performing time integration, balance equations for water and oil are:
+\begin{equation}\label{eq: So5}
+\frac{\bar{S}_{w1} V_{p1}}{B_{w1}} - \frac{\bar{S}_{w2} V_{p2}}{B_{w2}} = W_p \, ,
+\end{equation}
+\begin{equation}\label{eq: So5}
+\frac{\bar{S}_{o2} V_{p2}}{B_{o2}} = \left(N-N_p\right) \, ,
+\end{equation}
+where $N=\bar{S_{o1}} V_{p1} / B_{o1}$ is the initial volume of oil measured in standard conditions.
+
+Balance of gas component is:
+\begin{equation}\label{eq: Sg5}
+V_{p2}\left(\frac{S_{g2}}{B_{g2}} +\frac{S_{o2} R_{s2}}{B_{o2}}\right) - N\left(m\frac{B_{o1}}{B_{g1}} + R_{s1}\right)=G_i-G_p \, ,
+\end{equation}
+where $m=V_{p1} S_{g1}/ B_{o1} N$ is the ratio of initial volumes of gas and oil phases in reservoir conditions.
+
+%Equations from \eqref{eq: Sw4} to \eqref{eq: S14} constitute the most general form of material balance equations.
 So far, the only simplification made to black-oil equations was the absence of fluid motion, allowing pressure to be computed from Eqs.\eqref{eq: capillary_ow}, \eqref{eq: capillary_og} and \eqref{eq: p=rhogh}, and equations to be integrated in the entire reservoir volume.
@@
 \end{equation}
-If fluid densities do not vary appreciately with depth.
+If fluid densities do not vary appreciably with depth.
 
@@
 
-A single equation for pressure (without fluid saturations) can be obtained by substituting Eqs.~\eqref{eq: So2}, \eqref{eq: Sg2} and \eqref{eq: Sw2} into \eqref{eq: S12}.
+A single equation for pressure may be obtained by substituting Eqs.~\eqref{eq: So2}, \eqref{eq: Sg2} and \eqref{eq: Sw2} into \eqref{eq: S12}.
 The resulting equation allows pressure to be calculated from given production ($F_p$) and injection ($F_i$) volumes:

### 2014-03-31 pituba/article/article.tex
Eternal rewritting...

@@
 The equations were originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from phase equilibrium and balance of fluid production and volumetric expansion.
-\cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability, where most of the produced volume would come from and where pressure would present a more equal behavior.
+\cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the produced volume would come from and where pressure would present a more equal behavior.
 Such regions were said to contain an \textit{active oil}.
@@
 Since black-oil and material balance equations differ only in the number of spatial dimensions and the existence of local pressure gradients, there must exist conditions under which material balance equations can be recovered from the black-oil approach.
-This existing link was briefly addressed in \cite{ertekin2001basic}, where both were argued to be equivalent if capillarity and gradients of flow potential $\Phi$ could be neglected, i.e.:
+This existing link was briefly addressed in \cite{ertekin2001basic}, where both approaches were argued to be equivalent if capillarity and gradients of flow potential $\Phi$ could be neglected, i.e.:
 \begin{subequations}
@@
 We shall now introduce very common definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases.
-In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, also known as formation volume factors of water, oil and gas phases, respectively; express the ratio of a phase volume in some given pressure and temperature condition to the volume of this same phase in standard pressure and temperature, i.e.:
+In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at some given pressure and temperature condition to the volume of this same phase at standard pressure and temperature, i.e.:
 
-\begin{equation}
+\begin{equation}\label{eq: Bw}
 B_w = \frac{\rho_w^{ST}}{\bar{\rho_w}} \, ,
@@
 
-\begin{equation}
+\begin{equation}\label{eq: Bo}
 B_o = \frac{\rho_o^{ST}}{\bar{Y_o} \bar{\rho_o}} \, ,
@@
 
-\begin{equation}
+\begin{equation}\label{eq: Bg}
 B_g = \frac{\rho_g^{ST}}{\bar{\rho_g}} \, .
 \end{equation}
+
+The above averaged fluid densities, $\bar{\rho}_w$, $\bar{\rho}_o$ and $\bar{\rho}_g$, may be approximated to a constant value measured at a reference depth, if effects of temperature and compositional gradients are neglected.
+If we put together the expressions for fluid compressibility and hydrostatic equilibrium, we conclude that density variation with depth is negligible.
+
+For showing this, we assume a constant isothermal compressibility to oil phase
+\begin{equation}\label{eq: grado}
+\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o 
+\end{equation}
+and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
+\begin{equation}
+\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
+\end{equation}
+
+Similarly for the gas phase, we obtain its isothermal compressibility from the gas state equation
+\begin{equation}
+\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial p_g}\right)_{T,Z} = \frac{1}{p_g} = \frac{\mathcal{M}}{\rho_g ZRT}
+\end{equation}
+and again we use hydrostatic equilibrium to obtain the derivative of gas density with depth:
+\begin{equation}\label{eq: gradg}
+\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
+\end{equation}
+
+Substituting typical values in Eqs.~\eqref{eq: grado} and \eqref{eq: gradg}, we see that oil and gas densities are rather constant with depth:
+\begin{equation}
+\left(\frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 10^{-6}\ m^{-1} \, ,
+\end{equation}
+
+\begin{equation}
+\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} \approx 10^{-5}\ m^{-1} \, .
+\end{equation}
+
+The reference depth for computing fluid density (or formation volume factors) may be defined at gas-oil contact for a reservoir with a gas cap.
+At the fluid contact, both gas and oil densities may be computed using the corresponding local pressure.
+If the reference depth for oil phase is otherwise defined below gas-oil contact, gas density must be computed at some different depth still inside gas zone.
+This is to avoid the effect of oil pressure gradient in the computation of gas density, i.e.,
+\begin{equation}
+\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g
+\end{equation}
+what is $\rho_o / \rho_g$ times greater than the pure effect of gas hydrostatic pressure on its density.
 
@@
 
-Performing time integration, balance equations for water and oil are:
+Performing time integration, equations for water and oil are:
 \begin{equation}\label{eq: So5}
-\frac{\bar{S}_{w1} V_{p1}}{B_{w1}} - \frac{\bar{S}_{w2} V_{p2}}{B_{w2}} = W_p \, ,
+\frac{\bar{S}_{w1} V_{p1}}{B_{w1}} - \frac{\bar{S}_{w2} V_{p2}}{B_{w2}} = W_p - W_i - W_e\, ,
 \end{equation}
@@
 This additional simplification provides a direct relation between produced volumes and reservoir pressure.
-
-Assuming a constant isothermal compressibility to oil phase
-\begin{equation}
-\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o
-\end{equation}
-and using hydrostatic equilibrium $dp_o = \rho_o g dz$, we obtain the expression for oil density variation with depth:
-\begin{equation}
-\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
-\end{equation}
-
-For the gas phase, we obtain its isothermal compressibility from gas state equation
-\begin{equation}
-\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial p_g}\right)_{T,Z} = \frac{\rho_g}{p_g} = \frac{\mathcal{M}}{ZRT}
-\end{equation}
-
-And again we use hydrostatic equilibrium to obtain gas density variation with depth:
-\begin{equation}
-\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \left(\frac{\mathcal{M}}{ Z R T}\right) g
-\end{equation}
-
-%In oil zone:
-%\begin{equation}
-%\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g
-%\end{equation}
-
-\begin{equation}
-\left(\frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 10^{-6}\ m^{-1}
-\end{equation}
-
-\begin{equation}
-\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} \approx 10^{-5}\ m^{-1}
-\end{equation}
-If fluid densities do not vary appreciably with depth.
 
@@
 
-An open source implementation of secant method is available in SciPy library \citep{scipy}. This work used this library version $0.12$.
+An open source implementation of secant method is available in SciPy library \citep{scipy}. This work used library version $0.12$.
 
@@
 
-It must be observed that the result of gas reentering oil phase completely neglects non-equilibrium effects such as compositional gradients and thus must be used with care.
+It must be observed that the result of gas reentering oil phase completely neglects non-equilibrium effects, e.g.
+compositional gradients, and must be used with care.
 For instance, it may be in strong disagreement with real reservoir behavior when gravity segregation is present.

### 2014-04-02 pituba/article/article.tex
slow progress.

@@
 
-As previously mentioned, black-oil is a designation to a simplified treatment of hydrocarbon composition and phase equilibrium.
+As already mentioned, black-oil is a designation to a simplified treatment of hydrocarbon composition and phase equilibrium.
 Hydrocarbon species are grouped into two pseudo-components called oil and gas, being a pseudo-component one that consists of more than one chemical molecule.
@@
 Temperature is thus a constant prescribed value and no energy balance equation is needed.
-Also, it follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom.
+Also, it follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom (restricting component number to the two pseudo-components).
 As temperature is assumed constant, oil phase density in saturated states is a function of pressure or composition only.
@@
 %
+\subsubsection{Averaging Quantities}
 
@@
 
-The above averaged fluid densities, $\bar{\rho}_w$, $\bar{\rho}_o$ and $\bar{\rho}_g$, may be approximated to a constant value measured at a reference depth, if effects of temperature and compositional gradients are neglected.
+The above averaged fluid densities, $\bar{\rho}_w$, $\bar{\rho}_o$ and $\bar{\rho}_g$, may be approximated to a constant value measured at a reference depth.
 If we put together the expressions for fluid compressibility and hydrostatic equilibrium, we conclude that density variation with depth is negligible.

### 2014-04-04 pituba/article/article.tex
zzz

@@
 The so called material balance equations are widely employed in reservoir engineering and have become a classical tool in oil reservoir management.
-The equations were originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from phase equilibrium and balance of fluid production and volumetric expansion.
-\cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the produced volume would come from and where pressure would present a more equal behavior.
+The equations were originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from a balance of fluid production and injection.
+\cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the produced volume would come from and where pressure would present a similar behavior.
 Such regions were said to contain an \textit{active oil}.
@@
 
-The term \textit{black-oil} refers to specific hydrocarbon compositions whose volume and phase equilibria can be well described by a two pseudo-component system for isothermal changes in pressure.
-A general restriction is that reservoir temperature must be sufficiently below the mixture critical temperature so that different fluid phases have clearly distinguishable properties \citep{danesh1998pvt,firoozabadi1999thermodynamics}.
+The term \textit{black-oil} refers to specific hydrocarbon compositions whose phase equilibrium for isothermal changes in pressure may be well described by a two pseudo-component system.
+For such reservoir fluids, gas and oil compositions remain essentially constant in the entire range of pressure the reservoir observes during production life.
 
-Since black-oil and material balance equations differ only in the number of spatial dimensions and the existence of local pressure gradients, there must exist conditions under which material balance equations can be recovered from the black-oil approach.
+A general restriction of this approach is that reservoir temperature must be sufficiently below the mixture critical temperature so that different fluid phases have clearly distinguishable properties \citep{danesh1998pvt,firoozabadi1999thermodynamics}.
+
+Since black-oil and material balance equations differ only on spatial scale and the existence of local pressure gradients, there must exist conditions under which material balance equations might be recovered from the black-oil approach.
+This work examines such conditions.
+
 This existing link was briefly addressed in \cite{ertekin2001basic}, where both approaches were argued to be equivalent if capillarity and gradients of flow potential $\Phi$ could be neglected, i.e.:
@@
 
-The restriction of a spatially constant pressure throughout the reservoir does not require PVT properties to be averaged during the integration procedure.
+%The restriction of a spatially constant pressure throughout the reservoir does not require PVT properties to be averaged during the integration procedure.
 Such averaging process is not straightforward, if not impractical, as it is shown later.
@@
 After presenting material balance equations, sufficient conditions for solution uniqueness are imposed on PVT properties.
-The conditions simply require fluids to be compressible and that gas dissolution occurs with reduction of system volume.
+These conditions require fluids to be compressible and that gas dissolution occurs with reduction of the overall system volume.
 
@@
 
-All quantities are locally functions of space and time, but, after averaged in space, they become function of time only:
+All quantities are locally functions of space and time, but, after averaged in space, they become functions of time only:
 
@@
 
-The above averaged fluid densities, $\bar{\rho}_w$, $\bar{\rho}_o$ and $\bar{\rho}_g$, may be approximated to a constant value measured at a reference depth.
+The above averaged fluid densities, $\bar{\rho}_w$, $\bar{\rho}_o$ and $\bar{\rho}_g$, may be approximated to a constant value measured at a reference depth (also called \textit{datum}).
 If we put together the expressions for fluid compressibility and hydrostatic equilibrium, we conclude that density variation with depth is negligible.

### 2014-04-04 pituba/article/article.tex
zzzz

@@
 
-%Black-oil equations arise from the generalization of the zero dimensional concept to a multidimensional macroscopic model which incorporates hydrodynamics via Darcy's law and entails exactly the same thermodynamic approach for phase equilibrium \citep{aziz1979petroleum}.
+Black-oil equations arise from the generalization of the zero dimensional concept to a multidimensional macroscopic model which incorporates hydrodynamics via Darcy's law and entails exactly the same thermodynamic approach for phase equilibrium \citep{aziz1979petroleum}.
 
@@
 The term \textit{black-oil} refers to specific hydrocarbon compositions whose phase equilibrium for isothermal changes in pressure may be well described by a two pseudo-component system.
-For such reservoir fluids, gas and oil compositions remain essentially constant in the entire range of pressure the reservoir observes during production life.
+For such reservoir fluids, gas and oil compositions remain essentially constant in the entire range of pressure observed during reservoir production life.
 
-A general restriction of this approach is that reservoir temperature must be sufficiently below the mixture critical temperature so that different fluid phases have clearly distinguishable properties \citep{danesh1998pvt,firoozabadi1999thermodynamics}.
+%A general restriction of this approach is that reservoir temperature must be sufficiently below the mixture critical temperature so that different fluid phases have clearly distinguishable properties \citep{danesh1998pvt,firoozabadi1999thermodynamics}.
 
@@
 
-It is shown in this work that a stronger assumption is actually needed: mechanical equilibrium in the presence of gravity ($\nabla p = \rho g \nabla z$) is not a sufficient condition, pressure gradients must identically vanish instead ($\nabla p_\alpha \equiv 0$).
+%It is shown in this work that a stronger assumption is actually needed: mechanical equilibrium in the presence of gravity ($\nabla p = \rho g \nabla z$) is not a sufficient condition, pressure gradients must identically vanish instead ($\nabla p_\alpha \equiv 0$).
 %\begin{equation}
@@
 
-This section starts presenting the more general black-oil equations. It then proceeds by imposing the more restrictive conditions for wich material balance equations arise.
+This section starts presenting black-oil equations and it proceeds averaging quantities in space and examining simplifying assumptions for wich material balance equations are obtained.
 
@@
 
-Mass balance equations for each component in fluid phase:
+Mass balance equations for each component in fluid phase are:
 
@@
 Temperature is thus a constant prescribed value and no energy balance equation is needed.
-Also, it follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom (restricting component number to the two pseudo-components).
+Also, it follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if we restrict component number to the two pseudo-components.
 As temperature is assumed constant, oil phase density in saturated states is a function of pressure or composition only.
@@
 We shall now introduce very common definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases.
-In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at some given pressure and temperature condition to the volume of this same phase at standard pressure and temperature, i.e.:
+In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at some given pressure and temperature condition to the volume of the same phase at standard pressure and temperature, i.e.:
 
@@
 
-The above averaged fluid densities, $\bar{\rho}_w$, $\bar{\rho}_o$ and $\bar{\rho}_g$, may be approximated to a constant value measured at a reference depth (also called \textit{datum}).
-If we put together the expressions for fluid compressibility and hydrostatic equilibrium, we conclude that density variation with depth is negligible.
+\subsubsection{Definition of a Reference Depth for Pressure}
+
+Phase pressure increases with depth according to the hydrostatic gradient.
+This pressure increase, however, is not significant to produce appreciable variation in phase density and it may be approximated to a constant density measured at some reference depth.
+
+We may quantify this approximation by putting together the expressions for fluid compressibility and hydrostatic equilibrium, we conclude that density variation with depth is negligible.
 

### 2014-04-07 pituba/article/article.tex
reference depth on the way

@@
 
-The term \textit{black-oil} refers to specific hydrocarbon compositions whose phase equilibrium for isothermal changes in pressure may be well described by a two pseudo-component system.
-For such reservoir fluids, gas and oil compositions remain essentially constant in the entire range of pressure observed during reservoir production life.
+The term black-oil refers to specific hydrocarbon compositions whose phase equilibrium for isothermal changes in pressure may be well described by a two pseudo-component system.
+For such reservoir fluids, gas and oil compositions remain essentially constant in the entire range of pressure observed during reservoir productive life.
 
@@
 
-\section{Derivation of Equations}
+\section{Derivation of equations}
 
@@
 
-\subsection{Black-Oil Equations}
+\subsection{Black-Oil equations}
 
@@
 
-\subsection{Material Balance Equations}
+\subsection{Material balance equations}
 
@@
 %
-\subsubsection{Averaging Quantities}
+\subsubsection{Averaging quantities}
 
@@
 
-\subsubsection{Definition of a Reference Depth for Pressure}
+In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between dissolved gas component volume and oil component volume both measured in standard condition:
 
+\begin{equation}
+R_s = \frac{\rho_o^{ST} / \bar{Y_o}}{\rho_g^{ST} / \bar{Y_g}} \, ,
+\end{equation}
+
+or using $\bar{Y}_o + \bar{Y}_g = 1$:
+\begin{equation}
+R_s = \frac{\rho_o^{ST}}{\rho_g^{ST}}\frac{\bar{Y}_g}{1-\bar{Y}_g} \, .
+\end{equation}
+
+\subsubsection{Definition of a reference depth for pressure}
+
+Applications of material balance equations are widely employed to stablish a relation between production (and injection) and reservoir pressure.
+It turns out, as we know, that pressure is different for different locations, mostly for different values of depth, and also for different phases.
+
+What is then the significance of a single value of pressure for the entire reservoir, i.e., the reservoir pressure?
+
+We shall see that for the purpose of computing phase densities in the material balance, if we define a reference depth and use the corresponding local pressure, no important error is introduced in computing density.
 Phase pressure increases with depth according to the hydrostatic gradient.
-This pressure increase, however, is not significant to produce appreciable variation in phase density and it may be approximated to a constant density measured at some reference depth.
+The pressure increase, however, is not significant to produce appreciable variation in phase density.
 
-We may quantify this approximation by putting together the expressions for fluid compressibility and hydrostatic equilibrium, we conclude that density variation with depth is negligible.
+%and it may be approximated to a constant density measured at some reference depth.
 
-For showing this, we assume a constant isothermal compressibility to oil phase
+We may quantify this approximation by putting together the expressions for fluid compressibility and hydrostatic equilibrium.
+For this, we assume a constant isothermal compressibility to oil phase
 \begin{equation}\label{eq: grado}
@@
 
-The reference depth for computing fluid density (or formation volume factors) may be defined at gas-oil contact for a reservoir with a gas cap.
-At the fluid contact, both gas and oil densities may be computed using the corresponding local pressure.
-If the reference depth for oil phase is otherwise defined below gas-oil contact, gas density must be computed at some different depth still inside gas zone.
-This is to avoid the effect of oil pressure gradient in the computation of gas density, i.e.,
+The reference depth for computing fluid density (or formation volume factors) is then any arbitrary depth inside phase volume.
+
+For the case of an oil reservoir covered by a gas cap, the reference depth may be defined with advantage at the gas-oil contact.
+The point is that at the contact, both gas and oil densities may be computed using the same pressure, i.e., the corresponding local value.
+
+If the reference depth for oil phase is otherwise defined below gas-oil contact, gas density must be computed at some different depth still inside gas phase.
+This is to avoid the effect of oil hydrostatic gradient in the computation of gas density,
 \begin{equation}
-\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g
+\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
 \end{equation}
@@
 
-In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between dissolved gas component volume and oil component volume both measured in standard condition:
+\subsubsection{Material balance as a pressure equation}
 
-\begin{equation}
-R_s = \frac{\rho_o^{ST} / \bar{Y_o}}{\rho_g^{ST} / \bar{Y_g}} \, ,
-\end{equation}
-
-or using $\bar{Y}_o + \bar{Y}_g = 1$:
-\begin{equation}
-R_s = \frac{\rho_o^{ST}}{\rho_g^{ST}}\frac{\bar{Y}_g}{1-\bar{Y}_g} \, .
-\end{equation}
-
-Equations can now be rewritten using the above definitions and denoting porous volume by $V_p = \bar{\phi}\Omega$.
+Equations may now be rewritten using the above definitions and denoting porous volume by $V_p = \bar{\phi}\Omega$.
 
@@
 
-Performing time integration, equations for water and oil are:
-\begin{equation}\label{eq: So5}
-\frac{\bar{S}_{w1} V_{p1}}{B_{w1}} - \frac{\bar{S}_{w2} V_{p2}}{B_{w2}} = W_p - W_i - W_e\, ,
-\end{equation}
-\begin{equation}\label{eq: So5}
-\frac{\bar{S}_{o2} V_{p2}}{B_{o2}} = \left(N-N_p\right) \, ,
-\end{equation}
-where $N=\bar{S_{o1}} V_{p1} / B_{o1}$ is the initial volume of oil measured in standard conditions.
-
-Balance of gas component is:
-\begin{equation}\label{eq: Sg5}
-V_{p2}\left(\frac{S_{g2}}{B_{g2}} +\frac{S_{o2} R_{s2}}{B_{o2}}\right) - N\left(m\frac{B_{o1}}{B_{g1}} + R_{s1}\right)=G_i-G_p \, ,
-\end{equation}
-where $m=V_{p1} S_{g1}/ B_{o1} N$ is the ratio of initial volumes of gas and oil phases in reservoir conditions.
-
-%Equations from \eqref{eq: Sw4} to \eqref{eq: S14} constitute the most general form of material balance equations.
-So far, the only simplification made to black-oil equations was the absence of fluid motion, allowing pressure to be computed from Eqs.\eqref{eq: capillary_ow}, \eqref{eq: capillary_og} and \eqref{eq: p=rhogh}, and equations to be integrated in the entire reservoir volume.
-
-We now look to solve Eqs. \eqref{eq: Sw3}, \eqref{eq: So3} and \eqref{eq: Sg3} for the averaged quantities.
-
-Furthermore, we look for the implications of neglecting gravity and capillarity and assuming a constant pressure throughout the reservoir.
-This additional simplification provides a direct relation between produced volumes and reservoir pressure.
-
-Eq.~\eqref{eq: p=rhogh} gives an expression of how pressure, and hence PVT properties, depends on position.
-If gravity and capillarity are also neglected, fluid pressure becomes independent of position and phase:
-\begin{subequations}\label{eq: fullsimp}
-\begin{equation}
-p_o=p_w=p_g=p_P=p \, ,
-\end{equation}
-\begin{equation}
-\nabla p \equiv 0 \, .
-\end{equation}
-\end{subequations}
-
-%In order to apply the volume balance, a control volume ($\Omega$) is defined containing all the hydrocarbon zone.
-It consists of the region where hydrocarbon saturation is greater than zero for some arbitrarily small volume.
-This reservoir definition excludes the pure water zone, which is referred to as the aquifer.
-Bubble-point pressure is also a function of position as it depends on local values of oil and gas saturations.
-Assuming a spatially constant bubble-point pressure is equivalent to assuming constant saturations, as the quantities are related by Eqs.~\eqref{eq: S11} and \eqref{eq: Pb}.
-
-%Saturations are thus assumed constant in space so that bubble-point pressure becomes constant and an average saturation can be uniquely defined by Eq.~\eqref{eq: avS}.
-
-For spatially constant fluid and bubble-point pressures, Eqs.~\eqref{eq: matbal1a}, \eqref{eq: matbal1b} and \eqref{eq: matbal1c} can be integrated over the entire reservoir domain $\Omega$, which is defined as the region where hydrocarbon saturation is greater than zero for some arbitrary volume (provided that the restriction of macroscopic scale is observed).
-This reservoir definition excludes any surrounding aquifer zone.
-
-The interpretation of pressure computed from material balance equations becomes clear at this point: it is the reservoir pressure in the absence of pressure gradients, phase segregation, capillary forces and gravity.
-
-Performing volume and then time integration of Eq.~\eqref{eq: matbal1a} between times $t_1$ and $t_2$, material balance for oil phase is obtained:
-\begin{equation}
-\int_{t_1}^{t_2}\int_{\Omega} \frac{\partial}{\partial t}\left(\frac{\phi(p) S_o}{B_o(p,P_b)}\right) dV dt = \int_{t_1}^{t_2}\int_{\Omega} q_{os} dV dt\, ,
-\end{equation}
-\begin{equation}
-\left[ \frac{S_o}{B_o(p,P_b)} \int_{\Omega}\phi dV \right]^{t_2}_{t_1} = -N_p \, ,
-\end{equation}
-\begin{equation}\label{eq: So5}
-\frac{V_{p1} S_{o1}}{B_{o1}}-\frac{V_{p2}S_{o2}}{B_{o2}} = N_p \, .
+Integrating above equations between times $t_1$ and $t_2$ gives:
+% Water
+\begin{equation}\label{eq: Sw5}
+\frac{V_{p2} \bar{S}_{w2}}{B_{w2}}=W+\left(W_i+W_e-W_p\right) \, ,
 \end{equation}
 
-Equations for gas and water phases are obtained following the same procedure:
+% Oil
+\begin{equation}\label{eq: So5}
+\frac{\bar{S}_{o2} V_{p2}}{B_{o2}} = N-N_p \, ,
+\end{equation}
+
 % Gas
 \begin{equation}\label{eq: Sg5}
-V_{p2}\left(\frac{S_{g2}}{B_{g2}} +\frac{S_{o2} R_{s2}}{B_{o2}}\right) - N\left(m\frac{B_{o1}}{B_{g1}} + R_{s1}\right)=G_i-G_p \, ,
-\end{equation}
-% Agua
-\begin{equation}\label{eq: Sw5}
-\frac{V_{p2} S_{w2}}{B_{w2}} - W=W_i+W_e-W_p \, ,
+V_{p2}\left(\frac{\bar{S}_{g2}}{B_{g2}} +\frac{\bar{S}_{o2} R_{s2}}{B_{o2}}\right) = N\left(m\frac{B_{o1}}{B_{g1}} + R_{s1}\right) +\left(G_i-G_p\right) \, ,
 \end{equation}
@@
 \begin{equation}\label{eq: S15}
-S_o+S_w+S_g=1 \, ,
-\end{equation}
-where
-\begin{equation}
-N=\frac{V_{p1} S_{o1}}{B_{o1}} \, ,
-\end{equation}
-\begin{equation}
-W=\frac{V_{p1} S_{w1}}{B_{w1}} \, ,
-\end{equation}
-\begin{equation}
-m=\frac{V_{p1} S_{g1}}{B_{o1} N} \, .
+S_o+S_w+S_g=1 \, .
 \end{equation}
 
-An equation for bubble-point pressure in integrated form can be obtained by setting $S_g=0$ in Eq.~\eqref{eq: Sg2} and solving for the solubility ratio at bubble-point pressure $R_s(p=P_b)$:
+Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at standard conditions and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil measured in reservoir conditions.
+
+An equation for bubble-point may be obtained by setting $S_g=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure $R_s(p=P_b)$:
 \begin{equation}
@@
 \end{equation}
+
+%Equations from \eqref{eq: Sw4} to \eqref{eq: S14} constitute the most general form of material balance equations.
+So far, the only simplification made to black-oil equations was the absence of fluid motion, allowing pressure to be computed from Eqs.\eqref{eq: capillary_ow}, \eqref{eq: capillary_og} and \eqref{eq: p=rhogh}, and equations to be integrated in the entire reservoir volume.
+
+We now look to solve Eqs. \eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5} for the averaged quantities.
 

### 2014-04-10 pituba/article/article.tex
more and more

@@
 We may quantify this approximation by putting together the expressions for fluid compressibility and hydrostatic equilibrium.
-For this, we assume a constant isothermal compressibility to oil phase
+For this, we assume a constant isothermal compressibility for oil phase
 \begin{equation}\label{eq: grado}
@@
 \end{equation}
-what is $\rho_o / \rho_g$ times greater than the pure effect of gas hydrostatic pressure on its density.
+what is $\rho_o / \rho_g$ times greater than the sole effect of gas hydrostatic pressure on its density.
 
@@
 \begin{equation}\label{eq: S15}
-S_o+S_w+S_g=1 \, .
+\bar{S}_o+\bar{S}_w+\bar{S}_g=1 \, .
 \end{equation}
 
-Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at standard conditions and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil measured in reservoir conditions.
+Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at standard conditions and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil phases measured in reservoir conditions.
 
@@
 
-A single equation for pressure may be obtained by substituting Eqs.~\eqref{eq: So2}, \eqref{eq: Sg2} and \eqref{eq: Sw2} into \eqref{eq: S12}.
+A single equation for pressure may be obtained by substituting Eqs.~\eqref{eq: So5}, \eqref{eq: Sg5} and \eqref{eq: Sw5} into \eqref{eq: S15}.
 The resulting equation allows pressure to be calculated from given production ($F_p$) and injection ($F_i$) volumes:

### 2014-04-11 pituba/article/article.tex
more and more and more.

@@
 \begin{frontmatter}
-\title{Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equations}
+\title{Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equation}
 %\title{Obtaining the classical material balance equations from black-oil equations}
@@
 \begin{abstract}
-This work presents a derivation of the classical material balance equations starting from black-oil equations in integral form.
-The existing link between both approaches is discussed and an interpretation for pressure in material balance calculations is provided.
-It is shown that a condition stronger than hydrostatic equilibrium is needed to reduce black-oil equations to the classical approach.
-The classical equations are formulated for a variable bubble-point pressure, extending their applicability to reservoirs under pressure maintenance.
-Sufficient conditions for solution uniqueness are presented and a scheme for numerical solution is proposed.
+This work presents a derivation of the classical material balance equation starting from black-oil equations in integral form.
+The existing link between both approaches is examined in detail, including the necessary step of averaging PVT properties.
+An interpretation for pressure in material balance calculations is provided and the choice of a reference depth (datum) is discussed.
+The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented.
+A scheme for numerical solution is also proposed.
 \end{abstract}
@@
 \section{Introduction}
-The so called material balance equations are widely employed in reservoir engineering and have become a classical tool in oil reservoir management.
-The equations were originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from a balance of fluid production and injection.
-\cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the produced volume would come from and where pressure would present a similar behavior.
+The so called material balance equation is widely employed in reservoir engineering and has become a classical tool in oil reservoir management.
+The equation was originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from a balance of fluid production and injection.
+\cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the produced volume would come from and where pressure would present a similar trend.
 Such regions were said to contain an \textit{active oil}.
@@
 The equations developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium.
-Also, the zero-dimensional approach excluded the possibility of mechanical equilibrium under gravity and capillary forces, which produces a variable fluid distribution in space.
+Also, the zero-dimensional approach excluded the effects of mechanical equilibrium under gravity and capillary forces and the resulting phase segregation.
 An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}.
@@
 
-Black-oil equations arise from the generalization of the zero dimensional concept to a multidimensional macroscopic model which incorporates hydrodynamics via Darcy's law and entails exactly the same thermodynamic approach for phase equilibrium \citep{aziz1979petroleum}.
-
-%Although many applications of material balance equations have been reported, few explore their connection.
+Black-oil equations arise from the generalization of the zero dimensional concept to a three-dimensional macroscopic model which incorporates Darcy's law while entailing the same approach for phase equilibrium \citep{aziz1979petroleum}.
 
 The term black-oil refers to specific hydrocarbon compositions whose phase equilibrium for isothermal changes in pressure may be well described by a two pseudo-component system.
-For such reservoir fluids, gas and oil compositions remain essentially constant in the entire range of pressure observed during reservoir productive life.
+For such reservoir fluids, compositions of gas and oil phases remain essentially constant in the entire range of pressure observed during production life.
 
@@
 
-Since black-oil and material balance equations differ only on spatial scale and the existence of local pressure gradients, there must exist conditions under which material balance equations might be recovered from the black-oil approach.
+%Since black-oil and material balance equations differ only on spatial scale and the existence of local pressure gradients, there must exist conditions under which material balance equations might be recovered from the black-oil approach.
 This work examines such conditions.
 
-This existing link was briefly addressed in \cite{ertekin2001basic}, where both approaches were argued to be equivalent if capillarity and gradients of flow potential $\Phi$ could be neglected, i.e.:
-\begin{subequations}
-\begin{equation}
-p_\alpha = p \, ,
-\end{equation}
-\begin{equation}
-\nabla \Phi_\alpha = \nabla p_\alpha - \rho_\alpha g \nabla z = 0 \, ,
-\end{equation}
-\end{subequations}
-for $\alpha=o,g,w$ referring to oil, gas and water phases.
+Although many applications of material balance equation have been reported in the literature, few explore its connection with black-oil equations.
+\cite{ertekin2001basic} briefly addressed this issue arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \mathbf{g}$) could be neglected and if PVT properties were averaged.
+
+%\begin{subequations}
+%\begin{equation}
+%p_\alpha = p \, ,
+%\end{equation}
+%\begin{equation}
+%\nabla \Phi_\alpha = \nabla p_\alpha - \rho_\alpha g \nabla z = 0 \, ,
+%\end{equation}
+%\end{subequations}
+%for $\alpha=o,g,w$ referring to oil, gas and water phases.
 
@@
 
-The present derivation of material balance equations allows bubble-point pressure to vary with time accordingly to the same concept widely employed in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}.
+The present derivation of material balance equations allows bubble-point pressure to vary accordingly to the same concept widely employed in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}.
 The employed approach extends material balance equations to saturated reservoirs under pressure maintenance (e.g.
@@
 
-This section starts presenting black-oil equations and it proceeds averaging quantities in space and examining simplifying assumptions for wich material balance equations are obtained.
+This section starts presenting black-oil equations and it proceeds averaging quantities in space and examining simplifying assumptions for which material balance equations are obtained.
 
@@
 \begin{equation}
-T_w = T_o = T_g = T
+T_w = T_o = T_g = T_{rock} = T
 \end{equation}
@@
 
-Derivation of the classical material balance equations consists in assuming no fluid motion and in integrating black-oil equations in the entire reservoir volume.
-The integration process, however, requires all quantities to be either averaged or assumed constant in space.
+Derivation of the material balance equation consists in assuming no fluid motion and in integrating black-oil equations in reservoir volume.
+The integration process, however, requires all quantities to be either spatially averaged or assumed constant.
 
@@
 
-We now define space averaged quantities for porosity ($\phi$), saturation ($S_\alpha$) and density ($\rho_\alpha$) of a phase $\alpha$ and mass concentration of a component $k$ in phase $\alpha$ ($Y_{\alpha,k}$).
+We now define spatially averaged quantities for porosity ($\phi$), saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration of a component $k$ in phase $\alpha$ ($Y_{\alpha,k}$).
 
@@
 
-Applications of material balance equations are widely employed to stablish a relation between production (and injection) and reservoir pressure.
+Material balance equation is widely employed to stablish a relation between production (and injection) and reservoir pressure.
 It turns out, as we know, that pressure is different for different locations, mostly for different values of depth, and also for different phases.
 
-What is then the significance of a single value of pressure for the entire reservoir, i.e., the reservoir pressure?
+What is then the significance of a single value of pressure for the entire reservoir?
 
-We shall see that for the purpose of computing phase densities in the material balance, if we define a reference depth and use the corresponding local pressure, no important error is introduced in computing density.
+We shall see that for the purpose of computing the averaged value of phase densities in material balance, we may define some reference depth and use the corresponding local pressure without introducing any important error.
 Phase pressure increases with depth according to the hydrostatic gradient.
-The pressure increase, however, is not significant to produce appreciable variation in phase density.
+The pressure increase, however, is not significant to produce appreciable variation in density.
 
@@
 
-We may quantify this approximation by putting together the expressions for fluid compressibility and hydrostatic equilibrium.
-For this, we assume a constant isothermal compressibility for oil phase
+We may quantify this approximation by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase
 \begin{equation}\label{eq: grado}
@@
 
-For the case of an oil reservoir covered by a gas cap, the reference depth may be defined with advantage at the gas-oil contact.
-The point is that at the contact, both gas and oil densities may be computed using the same pressure, i.e., the corresponding local value.
+For the case of an oil reservoir covered by a gas cap, the reference depth may be defined, with advantage, at the gas-oil contact.
+The point is that at the contact both gas and oil densities may be computed using the same pressure, i.e., the corresponding local value.
 
@@
 
-\subsubsection{Material balance as a pressure equation}
+\subsubsection{Material balance as an equation for reservoir pressure}
 
@@
 
-Figure~\ref{fig: epsilon} shows the shape of residual function ($\epsilon$) with respect to pressure, i.e, holding all produced or injected volumes constant and allowing only pressure to change.
+Figure~\ref{fig: epsilon} shows the shape of residual function ($\epsilon$) with respect to pressure, i.e., holding all produced or injected volumes constant and allowing only pressure to change.
 Each different solid line corresponds to a different volume of fluid production and injection.

### 2014-04-17 pituba/article/article.tex
more and more

@@
 Derivation of the material balance equation consists in assuming no fluid motion and in integrating black-oil equations in reservoir volume.
-The integration process, however, requires all quantities to be either spatially averaged or assumed constant.
+The integration process requires all quantities to be either spatially averaged or assumed constant.
 
@@
 
-We now define spatially averaged quantities for porosity ($\phi$), saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration of a component $k$ in phase $\alpha$ ($Y_{\alpha,k}$).
+We now define spatially averaged quantities for porosity ($\phi$), saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration ($Y_{\alpha,k}$) of a component $k$ in phase $\alpha$.
 
-All quantities are locally functions of space and time, but, after averaged in space, they become functions of time only:
+All quantities are locally functions of space and time that, after averaged in space, become functions of time only:
 
@@
 
-Eqs. \eqref{eq: Sw2}, \eqref{eq: So2} and \eqref{eq: Sg2} rewritten in terms of averages read:
+Eqs. \eqref{eq: Sw2}, \eqref{eq: So2} and \eqref{eq: Sg2} rewritten in terms of averaged quantities read:
 
@@
 
-We shall now introduce very common definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases.
+We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases.
 In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at some given pressure and temperature condition to the volume of the same phase at standard pressure and temperature, i.e.:
@@
 
-In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between dissolved gas component volume and oil component volume both measured in standard condition:
+In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between volumes of dissolved gas and oil both measured in standard condition:
 
@@
 
+Equations may now be rewritten using the above definitions and denoting porous volume by $V_p = \bar{\phi}\Omega$.
+
+\begin{equation}\label{eq: Sw4}
+\frac{\partial}{\partial t} \left(\frac{\bar{S_w} V_p}{B_w} \right) = Q_w^{ST} \, ,
+\end{equation}
+
+\begin{equation}\label{eq: So4}
+\frac{\partial}{\partial t} \left(\frac{\bar{S_o} V_p}{B_o} \right) = Q_o^{ST} \, ,
+\end{equation}
+
+\begin{equation}\label{eq: Sg4}
+\frac{\partial}{\partial t} \left[ V_p \left(\frac{\bar{S_g}}{B_g} + \frac{\bar{S_o} R_s}{B_o} \right) \right] = Q_g^{ST} \, ,
+\end{equation}
+
+\begin{equation}\label{eq: S14}
+\bar{S_w}+\bar{S_o}+\bar{S_g}=1 \, .
+\end{equation}
+
+Integrating Eqs.~\eqref{eq: Sw4}, \eqref{eq: So4} and \eqref{eq: Sg4} between times $t_1$ and $t_2$ gives:
+% Water
+\begin{equation}\label{eq: Sw5}
+\frac{V_{p2} \bar{S}_{w2}}{B_{w2}}=W+\left(W_i+W_e-W_p\right) \, ,
+\end{equation}
+
+% Oil
+\begin{equation}\label{eq: So5}
+\frac{\bar{S}_{o2} V_{p2}}{B_{o2}} = N-N_p \, ,
+\end{equation}
+
+% Gas
+\begin{equation}\label{eq: Sg5}
+V_{p2}\left(\frac{\bar{S}_{g2}}{B_{g2}} +\frac{\bar{S}_{o2} R_{s2}}{B_{o2}}\right) = N\left(m\frac{B_{o1}}{B_{g1}} + R_{s1}\right) +\left(G_i-G_p\right) \, ,
+\end{equation}
+%
+\begin{equation}\label{eq: S15}
+\bar{S}_o+\bar{S}_w+\bar{S}_g=1 \, .
+\end{equation}
+
+Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at standard conditions and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil phases measured in reservoir conditions.
+
+An equation for bubble-point may be obtained by setting $S_g=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure $R_s(p=P_b)$:
+\begin{equation}
+\begin{split}
+R_{s2}\left(p=P_b\right) = &\Big[N\left(m B_{1}/B_{g1} +R_{s1}\right)\\
+&+G_i-G_p\Big]/\left(N-N_p\right) \, .
+\end{split}
+\end{equation}
+
 \subsubsection{Definition of a reference depth for pressure}
 
-Material balance equation is widely employed to stablish a relation between production (and injection) and reservoir pressure.
-It turns out, as we know, that pressure is different for different locations, mostly for different values of depth, and also for different phases.
+Material balance equation is widely employed to stablish a relationship between production (and injection) and reservoir pressure.
+It turns out, as we know, that pressure is different for different locations, mostly for different depths, and also for different phases.
 
@@
 
-We shall see that for the purpose of computing the averaged value of phase densities in material balance, we may define some reference depth and use the corresponding local pressure without introducing any important error.
-Phase pressure increases with depth according to the hydrostatic gradient.
-The pressure increase, however, is not significant to produce appreciable variation in density.
+We shall see that for the purpose of computing the averaged value of phase density in material balance, we may define some reference depth and use the corresponding local pressure as an approximation to the averaged density without introducing any important error.
+Although phase pressure increases with depth according to the hydrostatic gradient, this increase is not significant to produce appreciable variation in density.
 
@@
 
-We may quantify this approximation by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase
+This approximation may be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase
 \begin{equation}\label{eq: grado}
@@
 \subsubsection{Material balance as an equation for reservoir pressure}
-
-Equations may now be rewritten using the above definitions and denoting porous volume by $V_p = \bar{\phi}\Omega$.
-
-\begin{equation}\label{eq: Sw4}
-\frac{\partial}{\partial t} \left(\frac{\bar{S_w} V_p}{B_w} \right) = Q_w^{ST} \, ,
-\end{equation}
-
-\begin{equation}\label{eq: So4}
-\frac{\partial}{\partial t} \left(\frac{\bar{S_o} V_p}{B_o} \right) = Q_o^{ST} \, ,
-\end{equation}
-
-\begin{equation}\label{eq: Sg4}
-\frac{\partial}{\partial t} \left[ V_p \left(\frac{\bar{S_g}}{B_g} + \frac{\bar{S_o} R_s}{B_o} \right) \right] = Q_g^{ST} \, ,
-\end{equation}
-
-\begin{equation}\label{eq: S14}
-\bar{S_w}+\bar{S_o}+\bar{S_g}=1 \, .
-\end{equation}
-
-Integrating above equations between times $t_1$ and $t_2$ gives:
-% Water
-\begin{equation}\label{eq: Sw5}
-\frac{V_{p2} \bar{S}_{w2}}{B_{w2}}=W+\left(W_i+W_e-W_p\right) \, ,
-\end{equation}
-
-% Oil
-\begin{equation}\label{eq: So5}
-\frac{\bar{S}_{o2} V_{p2}}{B_{o2}} = N-N_p \, ,
-\end{equation}
-
-% Gas
-\begin{equation}\label{eq: Sg5}
-V_{p2}\left(\frac{\bar{S}_{g2}}{B_{g2}} +\frac{\bar{S}_{o2} R_{s2}}{B_{o2}}\right) = N\left(m\frac{B_{o1}}{B_{g1}} + R_{s1}\right) +\left(G_i-G_p\right) \, ,
-\end{equation}
-%
-\begin{equation}\label{eq: S15}
-\bar{S}_o+\bar{S}_w+\bar{S}_g=1 \, .
-\end{equation}
-
-Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at standard conditions and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil phases measured in reservoir conditions.
-
-An equation for bubble-point may be obtained by setting $S_g=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure $R_s(p=P_b)$:
-\begin{equation}
-\begin{split}
-R_{s2}\left(p=P_b\right) = &\Big[N\left(m B_{1}/B_{g1} +R_{s1}\right)\\
-&+G_i-G_p\Big]/\left(N-N_p\right) \, .
-\end{split}
-\end{equation}
 
@@
 
-Residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except at bubble-point pressure.
+Residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
 The last affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels corresponding each one to a different phase.
@@
 \textit{In-situ} volumes (e.g.
-$N-N_p$ for oil phase) are clearly non-negative values.
+$N-N_p$ for oil phase) are non-negative values.
 Pressure derivatives of fluid volumes are all negative values (as fluid pressure increases, fluid volume contracts) and pressure derivative of rock porous volume is a positive value (as fluid pressure increases, porous volume expands).

### 2014-04-24 pituba/article/article.tex
holidays

@@
 \title{Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equation}
-%\title{Obtaining the classical material balance equations from black-oil equations}
+%\title{Obtaining the classical material balance equation from black-oil equations}
 %\journal{Journal of Petroleum Science and Engineering}
@@
 The so called material balance equation is widely employed in reservoir engineering and has become a classical tool in oil reservoir management.
-The equation was originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from a balance of fluid production and injection.
-\cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the produced volume would come from and where pressure would present a similar trend.
+It consists of an algebraic equation originally presented in \cite{schilthuis1936active} and derived from a zero dimensional model for estimating reservoir pressure from mass balance.
+\cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the production would come from and where pressure would present a similar behavior.
 Such regions were said to contain an \textit{active oil}.
@@
 The equations developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium.
-Also, the zero-dimensional approach excluded the effects of mechanical equilibrium under gravity and capillary forces and the resulting phase segregation.
+Furthermore, the zero-dimensional approach excluded the possibility of mechanical equilibrium under gravity and capillary forces as they produce a variable fluid distribution in space.
 An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}.
@@
 
-Black-oil equations arise from the generalization of the zero dimensional concept to a three-dimensional macroscopic model which incorporates Darcy's law while entailing the same approach for phase equilibrium \citep{aziz1979petroleum}.
+Black-oil equations arise from the generalization of the zero dimensional concept to a three-dimensional model that incorporates Darcy's law and entails the same approach for phase equilibrium \citep{aziz1979petroleum}.
 
-The term black-oil refers to specific hydrocarbon compositions whose phase equilibrium for isothermal changes in pressure may be well described by a two pseudo-component system.
-For such reservoir fluids, compositions of gas and oil phases remain essentially constant in the entire range of pressure observed during production life.
+The term black-oil refers to specific hydrocarbon compositions whose phase equilibrium for isothermal changes in pressure may be described by a two pseudo-component system.
+For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and we may define this composition as a pseudo gas component.
+A pseudo oil component may then be defined by subtracting the gas component from oil phase composition.
+After defining the two pseudo components, the number of transport equations needed for hydrocarbon species is reduced from usually a large number to only two.
 
@@
 
-%Since black-oil and material balance equations differ only on spatial scale and the existence of local pressure gradients, there must exist conditions under which material balance equations might be recovered from the black-oil approach.
+%Since black-oil and material balance equation differ only on spatial scale and the existence of local pressure gradients, there must exist conditions under which material balance equation might be recovered from the black-oil approach.
 This work examines such conditions.
@@
 Although many applications of material balance equation have been reported in the literature, few explore its connection with black-oil equations.
-\cite{ertekin2001basic} briefly addressed this issue arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \mathbf{g}$) could be neglected and if PVT properties were averaged.
+\cite{ertekin2001basic} briefly addressed this issue arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \mathbf{g}$) were neglected and if PVT properties were averaged.
+
+This work develops such assumptions and examines underlying details in obtaining material balance equation from the more general black-oil equations.
+We shall see that the process of integrating black-oil equations in the reservoir volume suggests the convenient averaging of PVT properties and provides a physical interpretation for pressure computed from material balance equation.
 
@@
 
-%The requirement of a spatially constant pressure makes the material balance equations more suitable for shallow reservoirs, where gravity field is less important in phase equilibrium.
+%The requirement of a spatially constant pressure makes the material balance equation more suitable for shallow reservoirs, where gravity field is less important in phase equilibrium.
 
-The present derivation of material balance equations allows bubble-point pressure to vary accordingly to the same concept widely employed in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}.
-The employed approach extends material balance equations to saturated reservoirs under pressure maintenance (e.g.
+We employ the same approach of varying bubble-point pressure widely used in reservoir simulation that is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}.
+The employed approach extends material balance equation to saturated reservoirs under pressure maintenance (e.g.
 fluid injection).
@@
 
-After presenting material balance equations, sufficient conditions for solution uniqueness are imposed on PVT properties.
+After presenting material balance equation, sufficient conditions for solution uniqueness are imposed on PVT properties.
 These conditions require fluids to be compressible and that gas dissolution occurs with reduction of the overall system volume.
-
 A simple scheme for numerical solution is also proposed based on classical root-finding algorithms for non-linear equations.
@@
 
-This section starts presenting black-oil equations and it proceeds averaging quantities in space and examining simplifying assumptions for which material balance equations are obtained.
+This section starts presenting black-oil equations and it proceeds averaging quantities in space and examining simplifying assumptions for which material balance equation are obtained.
 
@@
 
-\subsection{Material balance equations}
+\subsection{Material balance equation}
 
@@
 
-%Equations from \eqref{eq: Sw4} to \eqref{eq: S14} constitute the most general form of material balance equations.
+%Equations from \eqref{eq: Sw4} to \eqref{eq: S14} constitute the most general form of material balance equation.
 So far, the only simplification made to black-oil equations was the absence of fluid motion, allowing pressure to be computed from Eqs.\eqref{eq: capillary_ow}, \eqref{eq: capillary_og} and \eqref{eq: p=rhogh}, and equations to be integrated in the entire reservoir volume.
@@
 
-\section{Solution of Material Balance Equations}
+\section{Solution of Material balance equation}
 A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We shall start defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
@@
 
-In summary, the following four relations constitute sufficient conditions for $\varepsilon(p)$ be a strictly decreasing function and for material balance equations to have only one solution:
+In summary, the following four relations constitute sufficient conditions for $\varepsilon(p)$ be a strictly decreasing function and for material balance equation to have only one solution:
 \begin{equation}\label{eq: PVTH1}
@@
 
-The steps to solve all material balance equations are summarized bellow:
+The steps to solve all material balance equation are summarized bellow:
 \begin{step}
@@
 \section{Results}
-Material balance equations were solved for a black-oil reservoir initially producing in depletion drive mechanism and later subjected to water injection.
+Material balance equation were solved for a black-oil reservoir initially producing in depletion drive mechanism and later subjected to water injection.
 Fig.~\ref{fig: matbal_p} shows in the solid line curve the severe depletion during initial phases of production.
@@
 \begin{enumerate}[1.]
-\item Material balance equations can be obtained from black-oil equations under the assumptions of:
+\item Material balance equation can be obtained from black-oil equations under the assumptions of:
 \begin{enumerate}[(a)]
@@
 \begin{enumerate}[1.]
-\item The existing link between black-oil and material balance equations is examined in detail.
-\item It is shown that gravity and capillary forces must be neglected and saturations must be constant in space so one can integrate black-oil equations in reservoir volume and obtain material balance equations.
-\item Material balance equations are shown to have a unique solution for a compressible system.
-\item Material balance equations are formulated for a variable bubble-point and they are employed to a reservoir under pressure maintenance.
+\item The existing link between black-oil and material balance equation is examined in detail.
+\item It is shown that gravity and capillary forces must be neglected and saturations must be constant in space so one can integrate black-oil equations in reservoir volume and obtain material balance equation.
+\item Material balance equation are shown to have a unique solution for a compressible system.
+\item Material balance equation are formulated for a variable bubble-point and they are employed to a reservoir under pressure maintenance.
 \end{enumerate}
@@
 \toprule
-\multicolumn{2}{l}{\textit{Material Balance Equations}}\\
+\multicolumn{2}{l}{\textit{Material balance equation}}\\
 \midrule

### 2014-05-04 pituba/article/article.tex
minor mods.
airport changes.

@@
 No diffusive or dispersive term is included in the mass balance equations.
-This simplification must be looked with care when gradients of species concentration are important and $\nabla Y_i \approx 0$ does not hold for some species $i$.
+This simplification must be looked with care when gradients of species concentration are important.
 For instance, when gas is being dissolved in oil phase in a reservoir under repressurization, this condition is not generally true.
+
+%and $\nabla Y_i \approx 0$ does not hold for some species $i$. For instance, when gas is being dissolved in oil phase in a reservoir under repressurization, this condition is not generally true.
 
@@
 Derivation of the material balance equation consists in assuming no fluid motion and in integrating black-oil equations in reservoir volume.
-The integration process requires all quantities to be either spatially averaged or assumed constant.
+As it will be shown, the integration process requires all quantities to be either spatially averaged or assumed constant.
 
@@
 
-All quantities are locally functions of space and time that, after averaged in space, become functions of time only:
+All quantities are initially functions of space and time that become functions of time alone after being averaged in space:
 
@@
 Material balance equation is widely employed to stablish a relationship between production (and injection) and reservoir pressure.
-It turns out, as we know, that pressure is different for different locations, mostly for different depths, and also for different phases.
+It turns out, though, that pressure is different for different locations, mostly for different depths, and also for different phases.
+What is then the significance of having a single value of pressure for the entire reservoir volume?
 
-What is then the significance of a single value of pressure for the entire reservoir?
-
-We shall see that for the purpose of computing the averaged value of phase density in material balance, we may define some reference depth and use the corresponding local pressure as an approximation to the averaged density without introducing any important error.
-Although phase pressure increases with depth according to the hydrostatic gradient, this increase is not significant to produce appreciable variation in density.
+For answering this question, we examine the two roles played by pressure in material balance equation: computation of averaged phase density and phase equilibrium.
+We shall see that for the purpose of computing the averaged phase density, we may define some reference depth and use the corresponding local density as an approximation to the averaged density without introducing any important error.
+Pressure increases with depth according to the hydrostatic gradient.
+Consequently, density also increases but by a negligible amount.
 
@@
 
-The reference depth for computing fluid density (or formation volume factors) is then any arbitrary depth inside phase volume.
+The reference depth for computing phase density (or formation volume factors) is then any arbitrary depth inside phase volume.
 
 For the case of an oil reservoir covered by a gas cap, the reference depth may be defined, with advantage, at the gas-oil contact.
-The point is that at the contact both gas and oil densities may be computed using the same pressure, i.e., the corresponding local value.
+The point here is that at phase contact both gas and oil densities may be computed using the same pressure, i.e., the corresponding local value.
 
-If the reference depth for oil phase is otherwise defined below gas-oil contact, gas density must be computed at some different depth still inside gas phase.
+If the reference depth for oil phase is otherwise defined below gas-oil contact, gas density must be computed at some different depth still inside gas phase, i.e., at or above gas-oil contact.
 This is to avoid the effect of oil hydrostatic gradient in the computation of gas density,
@@
 what is $\rho_o / \rho_g$ times greater than the sole effect of gas hydrostatic pressure on its density.
+
+It has been shown that pressure variation with depth is not important for the computation of gas, oil and water densities.
+Nevertheless, it was said that pressure also plays a second role in material balance equation by defining phase equilibrium.
+For this purpose, pressure variation with depth may not be negligible in the evalutation of the solubility ratio:
+\begin{equation}\label{eq: dRsdz1}
+\frac{1}{R_s}\frac{d R_s}{dz} = \frac{1}{R_s}\frac{d R_s}{dp_o} \frac{dp_o}{dz} = \frac{\rho_o g}{R_s}\frac{dR_s}{dp_o} \, .
+\end{equation}
+
+For a non-volatile saturated oil, we may approximate $dR_s / dp_o$ to the ratio between  solubility ratio ($R_{s}$) and saturation pressure ($p_{sat}$), so that Eq.~\eqref{eq: dRsdz1} becomes:
+\begin{equation}\label{eq: dRsdz2}
+\frac{1}{R_s}\frac{d R_s}{dz} \approx \frac{\rho_o g }{p_{sat}} \approx 10^{-3}\ m^{-1} \, ,
+\end{equation}
+where the following orders of magnitudes were assumed: $p_{sat} \approx 10^{7}\ Pa$, $\rho_o \approx 10^{3}\ kg/m^3$ and $g \approx 10\ m^2/s$.
+
+From Eq.~\eqref{eq: dRsdz2}, we see that the variation of pressure with depth may not be negligible for the phase equilibrium,  specially for reservoirs of light oils with large depth ranges.
+In such cases, using an averaged value for the solubility ratio may be worth the accuracy gain in dissolved gas volume.
 

### 2014-05-08 pituba/article/article_spe.tex
america 2

--- 2014-05-04 pituba/article/article.tex
@@
 % Review
-\documentclass[authoryear,preprint,review,12pt]{elsarticle}
+\documentclass[authoryear,preprint,review,10pt]{elsarticle}
 %\documentclass[final,authoryear,5p,times,twocolumn,11pt]{elsarticle}
@@
 
+%
+%This work derives the classical material balance equation from the more general black-oil model.
+It is shown that spatially averaged PVT properties are required for the material balance approach in general, but that for many cases the averaging is simply negligible for practical purposes.
+An interpretation for pressure in material balance equation is provided and its relation to a reference depth is examined.
+The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented.
+A scheme for numerical solution is also proposed.
+
 \begin{abstract}
-This work presents a derivation of the classical material balance equation starting from black-oil equations in integral form.
-The existing link between both approaches is examined in detail, including the necessary step of averaging PVT properties.
+This work derives the classical material balance equation from the more general black-oil model.
+The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented.
+It is shown that spatially averaged PVT properties are required for material balance approach in general, though the averaging process may be neglected for many practical purposes.
+An interpretation for pressure in material balance equation is provided and its relation to a reference depth is examined.
+Finally, a scheme for numerical solution is proposed.
+
+Material balance equation and black-oil models are approaches widely employed in reservoir engineering though few works explore their connection beyond the general intuitive notion. 
+%
+%This work derives the classical material balance equation from a black-oil model.
+We examine the existing link between both approaches in detail, including the necessary step of averaging PVT properties.
 An interpretation for pressure in material balance calculations is provided and the choice of a reference depth (datum) is discussed.
@@
 
-Black-oil equations arise from the generalization of the zero dimensional concept to a three-dimensional model that incorporates Darcy's law and entails the same approach for phase equilibrium \citep{aziz1979petroleum}.
+%Black-oil equations arise from the generalization of the zero dimensional concept to a three-dimensional model that incorporates Darcy's law and entails the same approach for phase equilibrium \citep{aziz1979petroleum}.
 
@@
 Although many applications of material balance equation have been reported in the literature, few explore its connection with black-oil equations.
-\cite{ertekin2001basic} briefly addressed this issue arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \mathbf{g}$) were neglected and if PVT properties were averaged.
+\cite{ertekin2001basic} briefly addressed this issue arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \vec{g}$) were neglected and if PVT properties were averaged.
 
@@
 fluid injection).
-The applicability, however, depend on how important non-equilibrium effects are to the phase equilibrium.
+The applicability, however, depends on how important non-equilibrium effects are to the phase equilibrium.
 
@@
 
-%\section{Black-Oil Equations}
+%\section{Black-oil equations}
 %
@@
 
-\subsection{Black-Oil equations}
+\subsection{Black-oil equations}
 
@@
 \begin{equation}\label{eq: Sw1}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV + \oint_{\partial \Omega} \rho_w \left( \mathbf{u}_w \cdot \mathbf{dS} \right) = \rho_w^{ST}Q_w^{ST} \, ;
+\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV + \oint_{\partial \Omega} \rho_w \left( \vec{u}_w \cdot \vec{dS} \right) = \rho_w^{ST}Q_w^{ST} \, ;
 \end{equation}
@@
 \begin{equation}\label{eq: So1}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV + \oint_{\partial \Omega} \rho_o Y_o \left(\mathbf{u}_o \cdot \mathbf{dS} \right)= \rho_o^{ST}Q_o^{ST} \, ;
+\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV + \oint_{\partial \Omega} \rho_o Y_o \left(\vec{u}_o \cdot \vec{dS} \right)= \rho_o^{ST}Q_o^{ST} \, ;
 \end{equation}
@@
 \begin{equation}\label{eq: Sg1}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV + \oint_{\partial \Omega} \left( \rho_o Y_g \mathbf{u}_o + \rho_g \mathbf{u}_g\right) \cdot \mathbf{dS}= \rho_g^{ST}Q_g^{ST} \, .
+\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV + \oint_{\partial \Omega} \left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right) \cdot \vec{dS}= \rho_g^{ST}Q_g^{ST} \, .
 \end{equation}
@@
 
-A common simplification in black-oil formulations is to restrict mass transport to macroscopic advection.
-No diffusive or dispersive term is included in the mass balance equations.
-This simplification must be looked with care when gradients of species concentration are important.
+A common simplification in black-oil formulations is to restrict mass transport to macroscopic advection. No diffusive or dispersive term is included in the mass balance equations.
+
+%This simplification must be looked with care when gradients of species concentration are important.
 For instance, when gas is being dissolved in oil phase in a reservoir under repressurization, this condition is not generally true.
@@
 %\begin{equation}\label{eq: So1}
-%\frac{\partial}{\partial t}\left(\frac{\phi(p_P) S_o}{B_o(p_o,P_b)}\right)+\nabla\cdot\left(\frac{\mathbf{u}_o}{B_o (p_o,P_b)}\right)=q_{os} \, ,
+%\frac{\partial}{\partial t}\left(\frac{\phi(p_P) S_o}{B_o(p_o,P_b)}\right)+\nabla\cdot\left(\frac{\vec{u}_o}{B_o (p_o,P_b)}\right)=q_{os} \, ,
 %\end{equation}
@@
 %\frac{\partial}{\partial t}\left[\phi(p_P)\left(\frac{S_g}{B_g (p_g)} +\frac{S_o R_s (p_o,P_b)}{B_o (p_o,P_b)}\right)\right] \\
-%+\nabla\cdot\left(\frac{\mathbf{u}_o R_s (p_o,P_b)}{B_o (p_o,P_b)}+\frac{\mathbf{u}_g}{B_g (p_g)}\right)=q_{gs} \, ,
+%+\nabla\cdot\left(\frac{\vec{u}_o R_s (p_o,P_b)}{B_o (p_o,P_b)}+\frac{\vec{u}_g}{B_g (p_g)}\right)=q_{gs} \, ,
 %\end{split}
@@
 %\begin{equation}\label{eq: Sw1}
-%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)+\nabla\cdot\left(\frac{\mathbf{u}_w}{B_w(p_w)}\right)=q_{ws} \, ,
+%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)+\nabla\cdot\left(\frac{\vec{u}_w}{B_w(p_w)}\right)=q_{ws} \, ,
 %\end{equation}
@@
 Momentum balance for each phase is given by an extension of Darcy's law to multiphase flow.
-This extension consists in multiplying absolute (or single-phase) permeability ($\mathbf{k}$) by a scalar function called phase relative permeability ($k_{r,\alpha}$):
+This extension consists in multiplying absolute (or single-phase) permeability ($\overleftrightarrow{k}$) by a scalar function called phase relative permeability ($k_{r,\alpha}$):
 
 \begin{equation}\label{eq: exdarcy}
-\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right) \, , \quad \alpha=o,w,g \, .
+\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right) \, , \quad \alpha=o,w,g \, .
 \end{equation}
@@
 Also, it follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if we restrict component number to the two pseudo-components.
-As temperature is assumed constant, oil phase density in saturated states is a function of pressure or composition only.
-The same is valid for mass concentrations of gas and oil components in oil phase.
+As temperature is assumed constant, oil phase density in saturated states may be a function of pressure alone.
+The same is valid for mass fractions of gas and oil components in oil phase.
 
@@
 \begin{equation}\label{eq: capillary_ow}
-p_{c,ow}\left(\mathbf{k},\phi,S_w\right)=p_o-p_w \, ,
+p_{c,ow}\left(\vec{k},\phi,S_w\right)=p_o-p_w \, ,
 \end{equation}
 \begin{equation}\label{eq: capillary_og}
-p_{c,go}\left(\mathbf{k},\phi,S_g\right)=p_g-p_o \, .
+p_{c,go}\left(\vec{k},\phi,S_g\right)=p_g-p_o \, .
 \end{equation}
@@
 %\begin{equation}
-%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)-\nabla\cdot\left[\frac{k_{r,w} \mathbf{k}}{B_w(p_w)\mu_w} \left(\nabla p_o - \rho_w g\nabla z -\underbrace{\frac{d p_c}{d S_w}\nabla S_w}_{\text{water diffusion}}\right)\right]=q_{ws} \, .
+%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)-\nabla\cdot\left[\frac{k_{r,w} \vec{k}}{B_w(p_w)\mu_w} \left(\nabla p_o - \rho_w g\nabla z -\underbrace{\frac{d p_c}{d S_w}\nabla S_w}_{\text{water diffusion}}\right)\right]=q_{ws} \, .
 %\end{equation}
 %
-%Discontinuities in saturation may still occur, however, if capillary pressures are functions of discontinuous porous medium properties ($phi$ and $\mathbf{k}$) or if phase change takes place, as diffusion of gas dissolved in oil phase is not allowed.
+%Discontinuities in saturation may still occur, however, if capillary pressures are functions of discontinuous porous medium properties ($phi$ and $\vec{k}$) or if phase change takes place, as diffusion of gas dissolved in oil phase is not allowed.
 %Cite Bear and Herrera.
@@
 \begin{equation}
-\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right)\equiv 0 \, .
+\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right)\equiv 0 \, .
 \end{equation}
@@
 \begin{equation}
-\bar{\phi} (t) = \frac{1}{\Omega}\int_{\Omega}\phi \left(\mathbf{x},t\right) dV \, .
+\bar{\phi} (t) = \frac{1}{\Omega}\int_{\Omega}\phi \left(\vec{x},t\right) dV \, .
 \end{equation}
@@
 \begin{equation}
-\bar{S_\alpha}\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}S_\alpha \phi dV \, ,
+\bar{S}_\alpha\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}S_\alpha \phi dV \, ,
 \end{equation}
@@
 \begin{equation}
-\bar{\rho_\alpha}\bar{S_\alpha}\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha S_\alpha \phi dV \, ,
+\bar{\rho_\alpha}\bar{S}_\alpha \bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha S_\alpha \phi dV \, ,
 \end{equation}
@@
 \begin{equation}
-\bar{\rho_\alpha}\bar{Y_{\alpha,k}}\bar{S_\alpha}\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha Y_{\alpha,k} S_\alpha \phi dV \, .
+\bar{\rho}_\alpha \bar{Y}_{\alpha,k}\bar{S}_\alpha \bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha Y_{\alpha,k} S_\alpha \phi dV \, .
 \end{equation}
@@
 \begin{equation}\label{eq: Sw3}
-\frac{\partial}{\partial t} \left( \bar{\rho_w}\bar{S_w}\bar{\phi} \Omega \right) = \rho_w^{ST}Q_w^{ST}
+\frac{\partial}{\partial t} \left( \bar{\rho_w}\bar{S}_w\bar{\phi} \Omega \right) = \rho_w^{ST}Q_w^{ST}
 \end{equation}
@@
 \begin{equation}\label{eq: So3}
-\frac{\partial}{\partial t} \left( \bar{\rho_o}\bar{Y_o}\bar{S_o}\bar{\phi} \Omega \right) = \rho_o^{ST}Q_o^{ST}
+\frac{\partial}{\partial t} \left( \bar{\rho_o}\bar{Y}_o\bar{S}_o\bar{\phi} \Omega \right) = \rho_o^{ST}Q_o^{ST}
 \end{equation}
@@
 \begin{equation}\label{eq: Sg3}
-\frac{\partial}{\partial t} \left[\left( \bar{\rho_o}\bar{Y_g} \bar{S_o} + \bar{\rho_g} \bar{S_g} \right) \bar{\phi} \Omega\right] = \rho_g^{ST}Q_g^{ST} \, .
+\frac{\partial}{\partial t} \left[\left( \bar{\rho}_o\bar{Y}_g \bar{S}_o + \bar{\rho}_g \bar{S}_g \right) \bar{\phi} \Omega\right] = \rho_g^{ST}Q_g^{ST} \, .
 \end{equation}
@@
 \begin{equation}\label{eq: Bw}
-B_w = \frac{\rho_w^{ST}}{\bar{\rho_w}} \, ,
+B_w = \frac{\rho_w^{ST}}{\bar{\rho}_w} \, ,
 \end{equation}
@@
 \begin{equation}\label{eq: Bo}
-B_o = \frac{\rho_o^{ST}}{\bar{Y_o} \bar{\rho_o}} \, ,
+B_o = \frac{\rho_o^{ST}}{\bar{Y}_o \bar{\rho}_o} \, ,
 \end{equation}
@@
 \begin{equation}\label{eq: Bg}
-B_g = \frac{\rho_g^{ST}}{\bar{\rho_g}} \, .
+B_g = \frac{\rho_g^{ST}}{\bar{\rho}_g} \, .
 \end{equation}
@@
 \begin{equation}
-R_s = \frac{\rho_o^{ST} / \bar{Y_o}}{\rho_g^{ST} / \bar{Y_g}} \, ,
+R_s = \frac{\rho_o^{ST} / \bar{Y}_o}{\rho_g^{ST} / \bar{Y}_g} \, ,
 \end{equation}
@@
 \begin{equation}\label{eq: Sw4}
-\frac{\partial}{\partial t} \left(\frac{\bar{S_w} V_p}{B_w} \right) = Q_w^{ST} \, ,
+\frac{\partial}{\partial t} \left(\frac{\bar{S}_w V_p}{B_w} \right) = Q_w^{ST} \, ,
 \end{equation}
@@
 \begin{equation}\label{eq: So4}
-\frac{\partial}{\partial t} \left(\frac{\bar{S_o} V_p}{B_o} \right) = Q_o^{ST} \, ,
+\frac{\partial}{\partial t} \left(\frac{\bar{S}_o V_p}{B_o} \right) = Q_o^{ST} \, ,
 \end{equation}
@@
 \begin{equation}\label{eq: Sg4}
-\frac{\partial}{\partial t} \left[ V_p \left(\frac{\bar{S_g}}{B_g} + \frac{\bar{S_o} R_s}{B_o} \right) \right] = Q_g^{ST} \, ,
+\frac{\partial}{\partial t} \left[ V_p \left(\frac{\bar{S}_g}{B_g} + \frac{\bar{S}_o R_s}{B_o} \right) \right] = Q_g^{ST} \, ,
 \end{equation}
@@
 \begin{equation}\label{eq: S14}
-\bar{S_w}+\bar{S_o}+\bar{S_g}=1 \, .
+\bar{S}_w+\bar{S}_o+\bar{S}_g=1 \, .
 \end{equation}
@@
 
-For answering this question, we examine the two roles played by pressure in material balance equation: computation of averaged phase density and phase equilibrium.
+For answering this question, we examine the two roles played by pressure in material balance equation: computation of phase density and phase equilibrium.
+
 We shall see that for the purpose of computing the averaged phase density, we may define some reference depth and use the corresponding local density as an approximation to the averaged density without introducing any important error.
@@
 For the case of an oil reservoir covered by a gas cap, the reference depth may be defined, with advantage, at the gas-oil contact.
-The point here is that at phase contact both gas and oil densities may be computed using the same pressure, i.e., the corresponding local value.
+The point here is that at phase contact both gas and oil densities may be computed using the same pressure, i.e., the corresponding local value (we are neglecting capillary pressure).
 
@@
 From Eq.~\eqref{eq: dRsdz2}, we see that the variation of pressure with depth may not be negligible for the phase equilibrium,  specially for reservoirs of light oils with large depth ranges.
-In such cases, using an averaged value for the solubility ratio may be worth the accuracy gain in dissolved gas volume.
+In such cases, using an averaged value for the solubility ratio may be worth the accuracy gain in computing dissolved gas volume.
 
@@
 
-\section{Solution of Material balance equation}
+\section{Solution of material balance equation}
 A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We shall start defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
@@
 
-\subsection{Mathematical Aspects of the Residual Function}
+\subsection{Mathematical aspects of the residual function}
 
@@
 
-\subsection{Solution Algorithm}
+\subsection{Solution algorithm}
 Several methods exist to find roots of nonlinear continuous functions \citep{hamming1987numerical}.

### 2014-05-11 pituba/article/article_spe.tex
hotel writings.

@@
 \begin{abstract}
-This work derives the classical material balance equation from the more general black-oil model.
-The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented.
-It is shown that spatially averaged PVT properties are required for material balance approach in general, though the averaging process may be neglected for many practical purposes.
+This work derives the classical material balance equation from a three dimensional black-oil model.
+Sufficient conditions for solution uniqueness of material balance equation are presented.
+It is shown that spatially averaged PVT properties are required for reducing a black-oil model to the material balance approach, although it is also shown that the averaging process may be neglected for many practical purposes.
 An interpretation for pressure in material balance equation is provided and its relation to a reference depth is examined.
@@
 
-Material balance equation and black-oil models are approaches widely employed in reservoir engineering though few works explore their connection beyond the general intuitive notion. 
+Material balance equation and black-oil models are widely employed in reservoir engineering, but few works have explored their connection.
 %
@@
 Momentum balance for each phase is given by an extension of Darcy's law to multiphase flow.
-This extension consists in multiplying absolute (or single-phase) permeability ($\overleftrightarrow{k}$) by a scalar function called phase relative permeability ($k_{r,\alpha}$):
+This extension consists in multiplying absolute (or single-phase) permeability ($\vec{\vec{k}}$) by a scalar function called phase relative permeability ($k_{r,\alpha}$):
 
 \begin{equation}\label{eq: exdarcy}
-\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right) \, , \quad \alpha=o,w,g \, .
+\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right) \, , \quad \alpha=o,w,g \, .
 \end{equation}
@@
 \begin{equation}\label{eq: capillary_ow}
-p_{c,ow}\left(\vec{k},\phi,S_w\right)=p_o-p_w \, ,
+p_{c,ow}\left(\vec{\vec{k}},\phi,S_w\right)=p_o-p_w \, ,
 \end{equation}
 \begin{equation}\label{eq: capillary_og}
-p_{c,go}\left(\vec{k},\phi,S_g\right)=p_g-p_o \, .
+p_{c,go}\left(\vec{\vec{k}},\phi,S_g\right)=p_g-p_o \, .
 \end{equation}
@@
 %\begin{equation}
-%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)-\nabla\cdot\left[\frac{k_{r,w} \vec{k}}{B_w(p_w)\mu_w} \left(\nabla p_o - \rho_w g\nabla z -\underbrace{\frac{d p_c}{d S_w}\nabla S_w}_{\text{water diffusion}}\right)\right]=q_{ws} \, .
+%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)-\nabla\cdot\left[\frac{k_{r,w} \vec{\vec{k}}}{B_w(p_w)\mu_w} \left(\nabla p_o - \rho_w g\nabla z -\underbrace{\frac{d p_c}{d S_w}\nabla S_w}_{\text{water diffusion}}\right)\right]=q_{ws} \, .
 %\end{equation}
 %
-%Discontinuities in saturation may still occur, however, if capillary pressures are functions of discontinuous porous medium properties ($phi$ and $\vec{k}$) or if phase change takes place, as diffusion of gas dissolved in oil phase is not allowed.
+%Discontinuities in saturation may still occur, however, if capillary pressures are functions of discontinuous porous medium properties ($phi$ and $\vec{\vec{k}}$) or if phase change takes place, as diffusion of gas dissolved in oil phase is not allowed.
 %Cite Bear and Herrera.
@@
 \begin{equation}
-\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right)\equiv 0 \, .
+\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right)\equiv 0 \, .
 \end{equation}

### 2014-05-14 pituba/article/article_spe.tex
hotel plus.

@@
 Finally, a scheme for numerical solution is proposed.
-
-Material balance equation and black-oil models are widely employed in reservoir engineering, but few works have explored their connection.
 %
@@
 \section{Introduction}
-The so called material balance equation is widely employed in reservoir engineering and has become a classical tool in oil reservoir management.
+The so called material balance equation is widely used in reservoir engineering and has become a classical tool in oil reservoir management.
 It consists of an algebraic equation originally presented in \cite{schilthuis1936active} and derived from a zero dimensional model for estimating reservoir pressure from mass balance.
 \cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the production would come from and where pressure would present a similar behavior.
-Such regions were said to contain an \textit{active oil}.
+Such regions were said to contain an active oil.
 
@@
 An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}.
-The novelty on hydrodynamics was the introduction of a macroscopic equation of motion employing Darcy's law.
+The novelty on hydrodynamics was the introduction of a macroscopic equation of motion using Darcy's law.
 
@@
 
-The term black-oil refers to specific hydrocarbon compositions whose phase equilibrium for isothermal changes in pressure may be described by a two pseudo-component system.
+The term black-oil refers to specific hydrocarbon compositions of which phase equilibrium for isothermal changes in pressure may be described by a two pseudo-component system.
 For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and we may define this composition as a pseudo gas component.
@@
 
-We employ the same approach of varying bubble-point pressure widely used in reservoir simulation that is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}.
-The employed approach extends material balance equation to saturated reservoirs under pressure maintenance (e.g.
+We use the same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}.
+This approach extends material balance equation to saturated reservoirs under pressure maintenance (e.g.
 fluid injection).
-The applicability, however, depends on how important non-equilibrium effects are to the phase equilibrium.
+The applicability, however, depends on how important non-equilibrium effects are to phase equilibrium.
 
@@
 where $\phi_\alpha$ is the volumetric fraction of a fluid phase $\alpha$ and $\phi$ is the porosity, i.e., the volumetric fraction of the pore volume.
-Since we consider that pore volume is fully saturated with fluid,
+Because we consider that pore volume is fully saturated with fluid,
 \begin{equation}
@@
 
-Porosity ($\phi$) is assumed a function of pore pressure ($p_P$). Since three phases are present in general, such quantity is defined by some relation of type \citep{kim2011rigorous}:
+Porosity ($\phi$) is assumed a function of pore pressure ($p_P$). Because three phases are present in general, such quantity is defined by some relation of type \citep{kim2011rigorous}:
 \begin{equation}
@@
 
-In order to close the system of equations, a connection among different phase pressures must be provided.
+To close the system of equations, a connection among different phase pressures must be provided.
 Such expressions are referred to as capillary pressures and are usually employed in the following forms \citep{aziz1979petroleum}:
@@
 
-All quantities are initially functions of space and time that become functions of time alone after being averaged in space:
+All quantities initially functions of space and time become functions only of time after being averaged in space:
 
@@
 
+\subsubsection{Material balance as an equation for reservoir pressure}
+
+%Equations from \eqref{eq: Sw4} to \eqref{eq: S14} constitute the most general form of material balance equation.
+So far, the only simplification made to black-oil equations was the absence of fluid motion, allowing pressure to be computed from Eqs.\eqref{eq: capillary_ow}, \eqref{eq: capillary_og} and \eqref{eq: p=rhogh}, and equations to be integrated in the entire reservoir volume.
+
+%We now look to solve Eqs. \eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5} for the averaged quantities.
+
+Another equation may be obtained by substituting Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5} into \eqref{eq: S15}.
+The resulting equation only involves fluid PVT properties, original fluid volumes ($N$, $W$ and $G)$, cumulative production and cumulative injection:
+\begin{equation}\label{eq: MBE}
+F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{porous volume contraction } (C)}
+\end{equation}
+or
+\begin{equation}\label{eq: MBE_short}
+F_p - F_i= E + C \, ,
+\end{equation}
+where newly introduced terms are defined in Table~\ref{tab: MBE}.
+
+Eq.~\eqref{eq: MBE} is commonly referred to as the Material Balance Equation (MBE) and it is an agelbraic equation that may be solved to any of its terms provided all other are known.
+Several techniques exist in the literature for one to estimate original volume of fluids, water influx from aquifers or reservoir pressure.
+
 \subsubsection{Definition of a reference depth for pressure}
 
-Material balance equation is widely employed to stablish a relationship between production (and injection) and reservoir pressure.
-It turns out, though, that pressure is different for different locations, mostly for different depths, and also for different phases.
-What is then the significance of having a single value of pressure for the entire reservoir volume?
+Material balance equation is widely used to stablish a relationship between production (and injection) and reservoir pressure.
+In Eq.~\eqref{eq: MBE}, however, formation volume factors were defined using averaged values of fluid properties,  which depend on pressure.
+It turns out, though, that pressure is a function of depth and phase.
+Using the average reservoir pressure ($\bar{p}$) to compute fluid properties does not solve this issue because $\bar{\rho}_o$ is not equal to $\rho_o (\bar{p})$ and thus $B_o$ is not a simple function of $p$.
+
+What is then the meaning of pressure calculatoins in Material Balance Equations?
 
@@
 
-We shall see that for the purpose of computing the averaged phase density, we may define some reference depth and use the corresponding local density as an approximation to the averaged density without introducing any important error.
+We shall see that to compute the averaged phase density, we may define some reference depth and use the corresponding local density as an approximation to the averaged density without introducing any important error.
+
 Pressure increases with depth according to the hydrostatic gradient.
 Consequently, density also increases but by a negligible amount.
-
-%and it may be approximated to a constant density measured at some reference depth.
-
-This approximation may be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase
+This approximation may be quantified by putting together the expressions for compressibility and hydrostatic equilibrium.
+For this, we assume a constant isothermal compressibility for oil phase
 \begin{equation}\label{eq: grado}
@@
 
-From Eq.~\eqref{eq: dRsdz2}, we see that the variation of pressure with depth may not be negligible for the phase equilibrium,  specially for reservoirs of light oils with large depth ranges.
+From Eq.~\eqref{eq: dRsdz2}, we see that the variation of pressure with depth may not be negligible for the phase equilibrium,  specially for thick reservoirs of light oil.
 In such cases, using an averaged value for the solubility ratio may be worth the accuracy gain in computing dissolved gas volume.
 
-\subsubsection{Material balance as an equation for reservoir pressure}
+% \subsubsection{Material balance as an equation for reservoir pressure}
 
@@
 
-We now look to solve Eqs. \eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5} for the averaged quantities.
+%We now look to solve Eqs. \eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5} for the averaged quantities.
 
-A single equation for pressure may be obtained by substituting Eqs.~\eqref{eq: So5}, \eqref{eq: Sg5} and \eqref{eq: Sw5} into \eqref{eq: S15}.
+%A single equation for pressure may be obtained by substituting Eqs.~\eqref{eq: So5}, \eqref{eq: Sg5} and \eqref{eq: Sw5} into \eqref{eq: S15}.
 The resulting equation allows pressure to be calculated from given production ($F_p$) and injection ($F_i$) volumes:
-\begin{equation}\label{eq: MBE}
-F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{porous volume contraction } (C)}
-\end{equation}
-or
-\begin{equation}\label{eq: MBE_short}
-F_p - F_i= E + C \, ,
-\end{equation}
+%\begin{equation}\label{eq: MBE}
+%F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{porous volume contraction } (C)}
+%\end{equation}
+%or
+%\begin{equation}\label{eq: MBE_short}
+%F_p - F_i= E + C \, ,
+%\end{equation}
 
-where newly introduced terms are defined in Table~\ref{tab: MBE}. Eq.~\eqref{eq: MBE} is commonly referred to as the Material Balance Equation (MBE).
+%where newly introduced terms are defined in Table~\ref{tab: MBE}. Eq.~\eqref{eq: MBE} is commonly referred to as the Material Balance Equation (MBE).
 
@@
 One possible choice is Newton's method, which linearizes the function using either its gradient or a secant line.
-The secant line approach introduces an artificial smoothness to the gradient direction near continuous but non-differentiable points, for what it is the approach employed here to deal with the bubble-point pressure.
+The secant line approach introduces an artificial smoothness to the gradient direction near continuous but non-differentiable points, for what it is the approach used here to deal with the bubble-point pressure.
 
@@
 
-Pressure in Eq.~\eqref{eq: MBE} is only a function of fluid properties in instant $t_1$, produced and injected values and bubble-point pressure at instant $t_2$, $P_{b2}$, which was already computed in step 1.
+Pressure in Eq.~\eqref{eq: MBE} is a function only of fluid properties in instant $t_1$, produced and injected values and bubble-point pressure at instant $t_2$, $P_{b2}$, which was already computed in step 1.
 
@@
 Water injection starts when oil recovery factor $(RF)$ is about $RF=0.08$ and it is maintained until the reservoir reaches a novel undersaturated state.
-Initially, the high compressibility of gas prevents pressure from raising and free gas phase is reduced due to redissolution.
+Initially, the high compressibility of gas prevents pressure from raising and free gas phase is reduced because of redissolution.
 Finally, the reservoir reaches the undersaturated state and a steep pressure increase is seen in Fig.~\ref{fig: matbal_p} for a recovery factor about $RF=0.15$.
@@
 
-Figure~\ref{fig: residual} shows the residual of numerical solution as defined in Eq.~\eqref{eq: residual}. Material balance error is bellow $10^{-7}$ units of porous volume for all time steps.
+Figure~\ref{fig: residual} shows the residual of numerical solution as defined in Eq.~\eqref{eq: residual}. Material balance error is less than $10^{-7}$ units of porous volume for all time steps.
 
@@
 \item Material balance equation are shown to have a unique solution for a compressible system.
-\item Material balance equation are formulated for a variable bubble-point and they are employed to a reservoir under pressure maintenance.
+\item Material balance equation are formulated for a variable bubble-point and they are used to a reservoir under pressure maintenance.
 \end{enumerate}
@@
 \caption{Reservoir pressure and bubble-point pressure as functions of recovery factor.
-Once depletion goes below bubble-point, pressure is hardly recovered for a saturated reservoir due to high system compressibility.
+Once depletion goes below bubble-point, pressure is hardly recovered for a saturated reservoir because of high system compressibility.
 Pressure is here given in units of the original bubble-point pressure.}

### 2014-05-18 pituba/article/article_spe.tex
hotel again

@@
 \begin{frontmatter}
-\title{Solution of thermodynamic equilibrium in black-oil reservoirs: the classical material balance equation}
+\title{Solution of hydodynamic equilibrium in black-oil reservoirs: the classical material balance equation}
 %\title{Obtaining the classical material balance equation from black-oil equations}
@@
 Sufficient conditions for solution uniqueness of material balance equation are presented.
-It is shown that spatially averaged PVT properties are required for reducing a black-oil model to the material balance approach, although it is also shown that the averaging process may be neglected for many practical purposes.
+It is shown that spatially averaged PVT properties are required for reducing a black-oil model to the material balance approach, it is shown, though, that the averaging process may be neglected for many practical purposes.
 An interpretation for pressure in material balance equation is provided and its relation to a reference depth is examined.
-Finally, a scheme for numerical solution is proposed.
+A scheme numerically solve material balance equation for pressure is also proposed.
 %
@@
 \section{Introduction}
-The so called material balance equation is widely used in reservoir engineering and has become a classical tool in oil reservoir management.
-It consists of an algebraic equation originally presented in \cite{schilthuis1936active} and derived from a zero dimensional model for estimating reservoir pressure from mass balance.
-\cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the production would come from and where pressure would present a similar behavior.
+Material balance equation is a classical tool in petroleum reservoir engineering.
+It consists of an algebraic equation originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from mass balance.
+\cite{schilthuis1936active} introduced the hypothesis of equilibrium attained at all times not necessarily in the entire reservoir volume, but in the regions of high permeability where most of the production would come from and where pressure would present a similar behavior.
 Such regions were said to contain an active oil.
 
-The equations developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium.
-Furthermore, the zero-dimensional approach excluded the possibility of mechanical equilibrium under gravity and capillary forces as they produce a variable fluid distribution in space.
+As mentioned, the equations developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium.
 An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}.
@@
 
-The term black-oil refers to specific hydrocarbon compositions of which phase equilibrium for isothermal changes in pressure may be described by a two pseudo-component system.
-For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and we may define this composition as a pseudo gas component.
-A pseudo oil component may then be defined by subtracting the gas component from oil phase composition.
-After defining the two pseudo components, the number of transport equations needed for hydrocarbon species is reduced from usually a large number to only two.
+The term black-oil refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure may be described by a two pseudo-component system.
+For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component may be defined for a simplified description.
+A pseudo oil component may also be defined by subtracting the pseudo gas component from oil phase composition.
+After defining these two pseudo-components, the number of transport equations needed for hydrocarbon species is reduced from a usually large number to only two.
 
@@
 
-Although many applications of material balance equation have been reported in the literature, few explore its connection with black-oil equations.
-\cite{ertekin2001basic} briefly addressed this issue arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \vec{g}$) were neglected and if PVT properties were averaged.
-
-This work develops such assumptions and examines underlying details in obtaining material balance equation from the more general black-oil equations.
-We shall see that the process of integrating black-oil equations in the reservoir volume suggests the convenient averaging of PVT properties and provides a physical interpretation for pressure computed from material balance equation.
+A great number of applications of material balance equation have been reported in the literature, but few of them explore its connection with black-oil equations.
+\cite{ertekin2001basic} briefly addressed this subject arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \vec{g}$) were neglected and if PVT properties were averaged in reservoir volume.
+The intuitive notion is that material balance equation is obtained if fluid motion is neglected and black-oil equations are integrated in reservoir volume.
+This work develops this process examining underlying details.
 
@@
 
-We use the same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}.
+The same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}, is also applied to material balance.
 This approach extends material balance equation to saturated reservoirs under pressure maintenance (e.g.
 fluid injection).
-The applicability, however, depends on how important non-equilibrium effects are to phase equilibrium.
+The applicability, however, depends on how important are non-equilibrium effects to phase equilibrium.
 
@@
 
-After presenting material balance equation, sufficient conditions for solution uniqueness are imposed on PVT properties.
+After presenting material balance equation, sufficient conditions are shown for which material balance equation has only one solution for pressure.
 These conditions require fluids to be compressible and that gas dissolution occurs with reduction of the overall system volume.
@@
 
-This section starts presenting black-oil equations and it proceeds averaging quantities in space and examining simplifying assumptions for which material balance equation are obtained.
+This section starts presenting black-oil equations and it proceeds averaging quantities in space and examining simplifying assumptions for which material balance equation is obtained.
 
@@
 
-As already mentioned, black-oil is a designation to a simplified treatment of hydrocarbon composition and phase equilibrium.
+% As already mentioned, black-oil is a designation to a simplified treatment of hydrocarbon composition and phase equilibrium.
 Hydrocarbon species are grouped into two pseudo-components called oil and gas, being a pseudo-component one that consists of more than one chemical molecule.
 
-Oil component, also called stock tank oil, is defined as the hydrocarbon composition in liquid phase for standard surface pressure and temperature.
+% Oil component, also called stock tank oil, is defined as the hydrocarbon composition in liquid phase for standard surface pressure and temperature.
 Similarly, gas component is the hydrocarbon composition in gaseous phase for the same surface conditions.
@@
 
-Reservoir fluids are then composed of the two pseudo-components (oil and gas) plus the water component.
-There can exist three fluid phases: oil, gas and liquid water.
-Oil phase may be composed of oil and gas components, gas phase may be composed uniquely by the gas component and, water phase, uniquely by water component.
+Reservoir fluids are composed of oil and gas pseudo-components and water component.
+There can be three fluid phases: oil, gas and liquid water.
+Oil phase may be composed of oil and gas components, gas phase may be composed uniquely by gas component and, water phase, uniquely by water component.
 The only allowed mass transfer is of gas component between oil and gas phases.
@@
 
-Mass balance equations for each component in fluid phase are:
+Mass balance equations for each component are:
 
@@
 \end{equation}
-where $\phi_\alpha$ is the volumetric fraction of a fluid phase $\alpha$ and $\phi$ is the porosity, i.e., the volumetric fraction of the pore volume.
-Because we consider that pore volume is fully saturated with fluid,
+where $\phi_\alpha$ is the volumetric fraction of a fluid phase $\alpha$ and $\phi$ is the porosity.
 \begin{equation}
@@
 Temperature is thus a constant prescribed value and no energy balance equation is needed.
-Also, it follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if we restrict component number to the two pseudo-components.
+It follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if we restrict component number to the two pseudo-components.
 As temperature is assumed constant, oil phase density in saturated states may be a function of pressure alone.
@@
 \begin{equation}
-Y_g = Y_g (p) \, ,
+Y_g = Y_g (p_o) \, ,
 \end{equation}
 \begin{equation}
-Y_o = Y_o (p) \, .
+Y_o = Y_o (p_o) \, .
 \end{equation}
@@
 
-To close the system of equations, a connection among different phase pressures must be provided.
-Such expressions are referred to as capillary pressures and are usually employed in the following forms \citep{aziz1979petroleum}:
+To close the system of equations, a connection among different phase pressures is provided by capillary pressure relations:
 %capillary
@@
 
-Black-oil equations then become:
+Black-oil equations become:
 \begin{itemize}
@@
 
-We now define spatially averaged quantities for porosity ($\phi$), saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration ($Y_{\alpha,k}$) of a component $k$ in phase $\alpha$.
+Spatially averaged quantities are defined for porosity ($\phi$), saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration ($Y_{\alpha,k}$) of a component $k$ in phase $\alpha$.
 
-All quantities initially functions of space and time become functions only of time after being averaged in space:
+All quantities are initially functions of space and time and they become functions only of time after being averaged in space:
 
@@
 
-We proceed omitting function arguments:
+Omitting function arguments:
 \begin{equation}
@@
 
-Eqs. \eqref{eq: Sw2}, \eqref{eq: So2} and \eqref{eq: Sg2} rewritten in terms of averaged quantities read:
+Eqs. \eqref{eq: Sw2}, \eqref{eq: So2} and \eqref{eq: Sg2} are rewritten in terms of averaged quantities:
 
@@
 
-\subsubsection{Material balance as an equation for reservoir pressure}
+\subsubsection{Material balance equation}
 
-%Equations from \eqref{eq: Sw4} to \eqref{eq: S14} constitute the most general form of material balance equation.
-So far, the only simplification made to black-oil equations was the absence of fluid motion, allowing pressure to be computed from Eqs.\eqref{eq: capillary_ow}, \eqref{eq: capillary_og} and \eqref{eq: p=rhogh}, and equations to be integrated in the entire reservoir volume.
+It is now possible to obtain the classical material balance equation substituting the equations for phase saturation, Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5}, into Eq.~\eqref{eq: S15}:
 
-%We now look to solve Eqs. \eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5} for the averaged quantities.
-
-Another equation may be obtained by substituting Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5} into \eqref{eq: S15}.
-The resulting equation only involves fluid PVT properties, original fluid volumes ($N$, $W$ and $G)$, cumulative production and cumulative injection:
 \begin{equation}\label{eq: MBE}
@@
 \end{equation}
-where newly introduced terms are defined in Table~\ref{tab: MBE}.
+where new terms $C$, $E$, $F_i$ and $F_p$ are defined in Table~\ref{tab: MBE}.
 
-Eq.~\eqref{eq: MBE} is commonly referred to as the Material Balance Equation (MBE) and it is an agelbraic equation that may be solved to any of its terms provided all other are known.
-Several techniques exist in the literature for one to estimate original volume of fluids, water influx from aquifers or reservoir pressure.
+Eq~\eqref{eq: MBE} only involves fluid properties, original fluid volumes ($N$, $W$ and $G)$ and cumulative production and injection.
+It is an agelbraic equation that may be solved to any of its terms provided all remaining are known.
+Several techniques exist in the literature to estimate original volume of fluids, water influx from aquifers or reservoir pressure.
 
@@
 It turns out, though, that pressure is a function of depth and phase.
-Using the average reservoir pressure ($\bar{p}$) to compute fluid properties does not solve this issue because $\bar{\rho}_o$ is not equal to $\rho_o (\bar{p})$ and thus $B_o$ is not a simple function of $p$.
 
-What is then the meaning of pressure calculatoins in Material Balance Equations?
+% ERRADO: Using the average reservoir pressure ($\bar{p}$) to compute fluid properties does not solve this issue because $\bar{\rho}_o$ is not equal to $\rho_o (\bar{p})$ and thus $B_o$ is not a simple function of $p$.
+
+What is then the meaning of pressure calculations in material balance equations?
 
@@
 The reference depth for computing phase density (or formation volume factors) is then any arbitrary depth inside phase volume.
+For a negligible variation of density with depth, it is easily seen that $\bar{\rho} = \rho\left(\bar{p}\right)$ and $\bar{p}=p\left(z^0\right)$ where $z^0$ is the depth of phase gravity center.
 
@@
 %We now show that $\varepsilon$ is a monotonic function since it is:
-\begin{itemize}
+\begin{enumerate}
 \item $f$ is continuous;
@@
 \item $f'<0$ everywhere.
-\end{itemize}
+\end{enumerate}
 
 Residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
-The last affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels corresponding each one to a different phase.
-Each phase parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
+The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels corresponding each one to a different phase.
+Each parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
 \begin{equation}
@@
 $N-N_p$ for oil phase) are non-negative values.
-Pressure derivatives of fluid volumes are all negative values (as fluid pressure increases, fluid volume contracts) and pressure derivative of rock porous volume is a positive value (as fluid pressure increases, porous volume expands).
+Pressure derivatives of fluid volumes are non-positive values (fluid volume cannot increase as pressure increases) and pressure derivative of porous volume is a positive value (as fluid pressure increases, porous volume expands owing to rock compression).
 
-The value of $d\varepsilon/dp$ is thus negative because each term is a product of a negative and a positive value, being the sum of all the four terms also a negative value.
+The value of $d\varepsilon/dp$ is thus negative if at least one phase is compressible because it would be a sum of products between negative and positive factors.
 
-In summary, the following four relations constitute sufficient conditions for $\varepsilon(p)$ be a strictly decreasing function and for material balance equation to have only one solution:
+In summary, the following four relations constitute sufficient conditions for $\varepsilon(p)$ be a strictly decreasing function and, therefore, for material balance equation to have only one solution:
 \begin{equation}\label{eq: PVTH1}
-\left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right) <0 \, ,
+\left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right) \leq 0 \, ,
 \end{equation}
 \begin{equation}\label{eq: PVTH2}
-\frac{dB_g}{dp} < 0 \, ,
+\frac{dB_g}{dp} \leq 0 \, ,
 \end{equation}
 \begin{equation}\label{eq: PVTH3}
-\frac{dB_w}{dp} < 0 \, ,
+\frac{dB_w}{dp} \leq 0 \, ,
 \end{equation}
 \begin{equation}\label{eq: PVTH4}
--\frac{dV_p}{dp}<0 \, .
+\frac{dV_p}{dp} \geq 0 \, .
 \end{equation}
 
-Switching inequalities to equalities in Eqs.~\eqref{eq: PVTH1}, \eqref{eq: PVTH2} and \eqref{eq: PVTH3} is equivalent to enforcing system incompressibility.
+Above relations, specially Eq.~\eqref{eq: PVTH1}, provide a test to check whether PVT data obtained from laboratory or empirical correlations is physically correct.
+
+%Switching inequalities to equalities in Eqs.~\eqref{eq: PVTH1}, \eqref{eq: PVTH2} and \eqref{eq: PVTH3} is equivalent to enforcing system incompressibility.
 Eq.~\eqref{eq: PVTH1}, for instance, would require a volume increase in oil phase to be equal to gas volume in gas phase for any additional gas dissolution.
@@
 \midrule
+pressure & $F_p - F_i= N E_o + mN E_g + W E_w +C$ \\
+\midrule
+\multicolumn{2}{l}{\textit{Phase saturations}}\\
+\midrule
 oil & $S_{o2} = \left(N-N_p\right) B_{o2}/V_{p2}$ \label{eq: oiltab}
@@
 gas & $S_{g2}=\left\lbrace N\left[\left(R_{s1}-R_{s2}\right)+m B_{o1}/B_{g1}\right]+N_p R_{s2}+G_i-G_p\right\rbrace Bg_2/V_{p2}$ \\
-bubble-point pressure & $R_{s2}\left(P_b\right) = \left[N\left(m B_{1}/B_{g1} +R_{s1}\right)+G_i-G_p\right]/\left(N-N_p\right)$ \\
-pressure & $F_p - F_i= N E_o + mN E_g + W E_w +C$ \\
 \midrule
-\multicolumn{2}{l}{\textit{Definition of terms in pressure equation}} \\
+\multicolumn{2}{l}{\textit{List of terms}} \\
 \midrule
@@
 fluid injection and influx & $F_i=\left(W_i +W_e\right) B_{w2}+G_{i}B_{g2}$ \\
+fluid expansion & $E=N E_o + mN E_g + W E_w$ \\
 oil phase expansion & $E_o=\left(B_{o2}-B_{o1}\right) - \left(R_{so2}-R_{so1}\right)B_{g2}$ \\
@@
 water volume expansion & $E_w = B_{w2}-B_{w1}$\\
-porous volume contraction & $C=V_{p2}- V_{p1}$\\ \bottomrule
+porous volume contraction & $C=V_{p2}- V_{p1}$\\
+bubble-point pressure & $R_{s2}\left(P_b\right) = \left[N\left(m B_{1}/B_{g1} +R_{s1}\right)+G_i-G_p\right]/\left(N-N_p\right)$ \\ \bottomrule
 \end{tabular}

### 2014-05-30 pituba/article/article_spe.tex
hotel attempt

@@
 \begin{frontmatter}
-\title{Solution of hydodynamic equilibrium in black-oil reservoirs: the classical material balance equation}
+\title{Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation}
 %\title{Obtaining the classical material balance equation from black-oil equations}
@@
 This work derives the classical material balance equation from a three dimensional black-oil model.
-Sufficient conditions for solution uniqueness of material balance equation are presented.
-It is shown that spatially averaged PVT properties are required for reducing a black-oil model to the material balance approach, it is shown, though, that the averaging process may be neglected for many practical purposes.
+Sufficient conditions for solution uniqueness of material balance equation are presented for the case that pressure is the unknown quantity.
+It is shown that spatially averaged PVT properties are required for reducing a black-oil model to the material balance approach.
+It is also shown that the averaging process may be neglected for many practical purposes.
 An interpretation for pressure in material balance equation is provided and its relation to a reference depth is examined.
@@
 
-The term black-oil refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure may be described by a two pseudo-component system.
-For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component may be defined for a simplified description.
-A pseudo oil component may also be defined by subtracting the pseudo gas component from oil phase composition.
-After defining these two pseudo-components, the number of transport equations needed for hydrocarbon species is reduced from a usually large number to only two.
+The term black-oil refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system.
+For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component can be defined for a simplified description gas phase composition.
+A pseudo oil component can also be defined by subtracting the pseudo gas component from oil phase composition.
+After defining these two pseudo-components, the number of transport equations needed for hydrocarbon species is reduced from usually a large number to only two.
 
@@
 
-The same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}, is also applied to material balance.
+The same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}, is also applied here to material balance.
 This approach extends material balance equation to saturated reservoirs under pressure maintenance (e.g.
 fluid injection).
-The applicability, however, depends on how important are non-equilibrium effects to phase equilibrium.
+The applicability, however, depends on how important non-equilibrium effects are to phase equilibrium.
 
@@
 %Oil is composed by hydrocarbon species that remain in liquid phase for standard surface pressure and temperature.
-The other hydrocarbon species in gaseous phase for the same surface conditions are called gas.
+The other hydrocarbon species in gaseous phase for the same surface condition are called gas.
 %
@@
 % Oil component, also called stock tank oil, is defined as the hydrocarbon composition in liquid phase for standard surface pressure and temperature.
-Similarly, gas component is the hydrocarbon composition in gaseous phase for the same surface conditions.
+Similarly, gas component is the hydrocarbon composition in gaseous phase for the same surface condition.
 All species are generally present in both phases although the equilibrium composition is shifted to liquid phase (or gaseous phase) for components of greater (lower) molecular mass.
@@
 Reservoir fluids are composed of oil and gas pseudo-components and water component.
-There can be three fluid phases: oil, gas and liquid water.
+There can be three fluid phases: oil, gas and (liquid) water.
 Oil phase may be composed of oil and gas components, gas phase may be composed uniquely by gas component and, water phase, uniquely by water component.
@@
 
-In addition to the three fluid components and phases, there is the solid phase composed by the rock.
+%In addition to the three fluid components and phases, there is the solid phase composed by the rock.
 
@@
 \end{itemize}
-Mass concentration of oil and gas components in oil phases are denoted as $Y_o$ and $Y_g$, respectively.
+Mass concentrations of oil and gas components in oil phases are denoted as $Y_o$ and $Y_g$, respectively.
 
@@
 \end{equation}
-where $\phi_\alpha$ is the volumetric fraction of a fluid phase $\alpha$ and $\phi$ is the porosity.
+where $\phi_\alpha$ is the volumetric fraction of a fluid phase $\alpha$ and $\phi$ is the porosity. Clearly, all saturations sum up to unity. 
 \begin{equation}
@@
 
-A common simplification in black-oil formulations is to restrict mass transport to macroscopic advection. No diffusive or dispersive term is included in the mass balance equations.
+A common simplification in black-oil formulation is to restrict mass transport to macroscopic advection. No diffusive or dispersive term is included in mass balance equations.
 
@@
 
-Another simplification is now made assuming thermal equilibrium in the entire reservoir volume:
+Another simplification is now made assuming thermal equilibrium in the entire reservoir volume, i.e.,
 \begin{equation}
-T_w = T_o = T_g = T_{rock} = T
-\end{equation}
-and
-\begin{equation}
-\nabla T \equiv 0 \, .
+T_w = T_o = T_g = T_{rock} = T = \text{constant} \, .
 \end{equation}
 
-Temperature is thus a constant prescribed value and no energy balance equation is needed.
-It follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if we restrict component number to the two pseudo-components.
-As temperature is assumed constant, oil phase density in saturated states may be a function of pressure alone.
-The same is valid for mass fractions of gas and oil components in oil phase.
+Temperature is a constant prescribed value and thus no equation for energy balance is needed.
+It follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if component number is restricted to the two pseudo-components.
+As temperature is assumed constant, oil phase density is a function of pressure alone.
+The same is valid for mass fractions of gas and oil components in oil phase:
 
@@
 We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases.
-In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at some given pressure and temperature condition to the volume of the same phase at standard pressure and temperature, i.e.:
+In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to the volume at surface condition, i.e.:
 
@@
 
-In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between volumes of dissolved gas and oil both measured in standard condition:
+In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between volumes of dissolved gas and oil both measured in surface condition:
 
@@
 
-Equations may now be rewritten using the above definitions and denoting porous volume by $V_p = \bar{\phi}\Omega$.
+Equations can now be rewritten using the above definitions and denoting porous volume by $V_p = \bar{\phi}\Omega$.
 
@@
 
-Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at standard conditions and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil phases measured in reservoir conditions.
+Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at surface condition and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil phases measured in reservoir condition.
 
-An equation for bubble-point may be obtained by setting $S_g=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure $R_s(p=P_b)$:
+An equation for bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure $R_s(p=P_b)$:
 \begin{equation}
@@
 
-Eq~\eqref{eq: MBE} only involves fluid properties, original fluid volumes ($N$, $W$ and $G)$ and cumulative production and injection.
-It is an agelbraic equation that may be solved to any of its terms provided all remaining are known.
-Several techniques exist in the literature to estimate original volume of fluids, water influx from aquifers or reservoir pressure.
+Eq~\eqref{eq: MBE} involves fluid properties, original fluid volumes ($N$, $W$ and $G)$, cumulative production and injection and aquifer influx.
+It is an agelbraic equation that can be solved to any of its terms provided all remaining are known. This works focuses on its solution for pressure.
 
-\subsubsection{Definition of a reference depth for pressure}
+\subsubsection{An interpretation for calculated pressure and the definition of a reference depth}
 
 Material balance equation is widely used to stablish a relationship between production (and injection) and reservoir pressure.
-In Eq.~\eqref{eq: MBE}, however, formation volume factors were defined using averaged values of fluid properties,  which depend on pressure.
+In Eq.~\eqref{eq: MBE}, however, formation volume factors were defined using averaged values of fluid properties, and they are function of pressure.
 It turns out, though, that pressure is a function of depth and phase.
@@
 
+Oil and gas densities are thus rather constant with depth and a zeroth order approximation of $\bar{\rho}$ is accurate within $0.5\%$ for material balance calculations.
+
 The reference depth for computing phase density (or formation volume factors) is then any arbitrary depth inside phase volume.
@@
 For the case of an oil reservoir covered by a gas cap, the reference depth may be defined, with advantage, at the gas-oil contact.
-The point here is that at phase contact both gas and oil densities may be computed using the same pressure, i.e., the corresponding local value (we are neglecting capillary pressure).
+The point here is that at phase contact both gas and oil densities can be computed using the same pressure, i.e., the corresponding local value (we are neglecting capillary pressure).
 
@@
 
-For a non-volatile saturated oil, we may approximate $dR_s / dp_o$ to the ratio between  solubility ratio ($R_{s}$) and saturation pressure ($p_{sat}$), so that Eq.~\eqref{eq: dRsdz1} becomes:
+For a non-volatile saturated oil, we can approximate $dR_s / dp_o$ to the ratio between  solubility ratio ($R_{s}$) and saturation pressure ($p_{sat}$), so that Eq.~\eqref{eq: dRsdz1} becomes:
 \begin{equation}\label{eq: dRsdz2}
@@
 $G_p$ & produced volume of gas \\
-$m$ & ratio of original free gas volume to original oil phase volume in reservoir conditions \\
+$m$ & ratio of original free gas volume to original oil phase volume in reservoir condition \\
 $N$ & original volume of oil in place \\

### 2014-06-02 pituba/article/article_spe.tex
last hotel

@@
 
-The term black-oil refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system.
-For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component can be defined for a simplified description gas phase composition.
-A pseudo oil component can also be defined by subtracting the pseudo gas component from oil phase composition.
-After defining these two pseudo-components, the number of transport equations needed for hydrocarbon species is reduced from usually a large number to only two.
+The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system.
+For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component can be defined.
+Similarly, a pseudo oil component can also be defined by subtracting all the pseudo gas component from oil phase composition.
+After defining these two pseudo-components, the number of transport equations needed for hydrocarbon species is reduced from a usually large number to only two.
 
@@
 There can be three fluid phases: oil, gas and (liquid) water.
-Oil phase may be composed of oil and gas components, gas phase may be composed uniquely by gas component and, water phase, uniquely by water component.
-The only allowed mass transfer is of gas component between oil and gas phases.
+Oil phase may be composed of oil and gas components, gas phase can be composed uniquely by gas component and water phase uniquely by water component.
+The only allowed mass transfer is of gas component between oil and gas phases, i.e., composition of gas phase is always that of a dry gas.
 
@@
 Eq~\eqref{eq: MBE} involves fluid properties, original fluid volumes ($N$, $W$ and $G)$, cumulative production and injection and aquifer influx.
-It is an agelbraic equation that can be solved to any of its terms provided all remaining are known. This works focuses on its solution for pressure.
+It is an agelbraic equation that can be solved to any one of its terms, provided that all remaining are known. This works focuses on its solution for pressure.
 
@@
 Material balance equation is widely used to stablish a relationship between production (and injection) and reservoir pressure.
-In Eq.~\eqref{eq: MBE}, however, formation volume factors were defined using averaged values of fluid properties, and they are function of pressure.
+In Eq.~\eqref{eq: MBE}, however, formation volume factors were defined using averaged values of fluid properties, which are functions of local pressure.
 It turns out, though, that pressure is a function of depth and phase.
@@
 
-Above relations, specially Eq.~\eqref{eq: PVTH1}, provide a test to check whether PVT data obtained from laboratory or empirical correlations is physically correct.
+Eq.~\eqref{eq: PVTH1} also provides one condition to check whether PVT data obtained from laboratory or empirical correlations is physically correct.
+It can be alternatively formulated as shown below and indicates that any vaporization of oil phase is followed by an increase in its density.
+\begin{equation}
+\frac{1}{\rho_o}\frac{d\rho_o}{dY_o} \geq Y_o \left( \frac{\rho_o}{\rho_g} - 1 \right) \, .
+\end{equation}
 
@@
 One possible choice is Newton's method, which linearizes the function using either its gradient or a secant line.
-The secant line approach introduces an artificial smoothness to the gradient direction near continuous but non-differentiable points, for what it is the approach used here to deal with the bubble-point pressure.
+The secant line approach introduces an numerical smoothness to the gradient near continuous but non-differentiable points, for what it is the approach used here to deal with the bubble-point pressure.
 

### 2014-06-20 pituba/article/article_spe.tex
continua

@@
 
+%Phase saturation is defined as the fraction of the occupied pore volume:
+%\begin{equation}
+%S_\alpha = \frac{\phi_\alpha}{\phi} \, ,
+%\end{equation}
+%where $\phi_\alpha$ is the volumetric fraction of a fluid phase $\alpha$ and $\phi$ is the porosity. Clearly, all saturations sum up to unity. 
+%\begin{equation}
+%S_o + S_g +S_w = 1 \, .
+%\end{equation}
+
+We assume that the reservoir is a connected volume where hydrocarbon saturation is nonzero.
 Reservoir fluids are composed of oil and gas pseudo-components and water component.
 There can be three fluid phases: oil, gas and (liquid) water.
-Oil phase may be composed of oil and gas components, gas phase can be composed uniquely by gas component and water phase uniquely by water component.
-The only allowed mass transfer is of gas component between oil and gas phases, i.e., composition of gas phase is always that of a dry gas.
+Oil phase may be composed of oil and gas components, gas phase is composed uniquely by gas component and water phase uniquely by water component.
+The only allowed mass transfer is of gas component between oil and gas phases, i.e., gas phase composition is always that of a dry gas.
 
@@
 
-Mass balance equations for each component are:
+Mass balance equations for each component read:
 
@@
 \begin{equation}\label{eq: Sw1}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV + \oint_{\partial \Omega} \rho_w \left( \vec{u}_w \cdot \vec{dS} \right) = \rho_w^{ST}Q_w^{ST} \, ;
+\frac{\partial}{\partial t} \left( \rho_w S_w \phi \right) +\nabla \cdot \left( \rho_w \vec{u}_w \right) = 0 ;
 \end{equation}
@@
 \begin{equation}\label{eq: So1}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV + \oint_{\partial \Omega} \rho_o Y_o \left(\vec{u}_o \cdot \vec{dS} \right)= \rho_o^{ST}Q_o^{ST} \, ;
+\frac{\partial}{\partial t}\left( \rho_o Y_o S_o \phi \right) + \nabla \cdot \left( \rho_o Y_o \vec{u}_o \right) = 0 \, ;
 \end{equation}
@@
 \begin{equation}\label{eq: Sg1}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV + \oint_{\partial \Omega} \left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right) \cdot \vec{dS}= \rho_g^{ST}Q_g^{ST} \, .
+\begin{split}
+\frac{\partial}{\partial t} \left( \rho_o Y_g S_o + \rho_g S_g\right) + \nabla  \cdot \left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right) = 0\, .
+\end{split}
 \end{equation}
 \end{itemize}
-Mass concentrations of oil and gas components in oil phases are denoted as $Y_o$ and $Y_g$, respectively.
+Mass concentrations of oil and gas components in oil phase are denoted by $Y_o$ and $Y_g$, respectively.
 
-Phase saturation is defined as its volumetric fraction of the pore space:
-\begin{equation}
-S_\alpha = \frac{\phi_\alpha}{\phi} \, ,
-\end{equation}
-where $\phi_\alpha$ is the volumetric fraction of a fluid phase $\alpha$ and $\phi$ is the porosity. Clearly, all saturations sum up to unity. 
-\begin{equation}
-S_o + S_g +S_w = 1 \, .
-\end{equation}
+Initial conditions are of no fluid motion and boundary conditions are of no normal flow at all boundaries except for interfaces with wells or aquifers.
 
@@
 
-Derivation of the material balance equation consists in assuming no fluid motion and in integrating black-oil equations in reservoir volume.
-As it will be shown, the integration process requires all quantities to be either spatially averaged or assumed constant.
+Derivation of the material balance equation consists in integrating black-oil equations in reservoir volume.
 
@@
 
-We start assuming mechanical equilibrium, i.e., no bulk motion:
-\begin{equation}
-\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right)\equiv 0 \, .
-\end{equation}
-This condition implies that pressure gradients must equal gravity force per unit volume:
-\begin{equation}\label{eq: p=rhogh}
-\nabla p_\alpha = \rho_\alpha g\nabla z
-\end{equation}
-and, consequently, pressure is a function of depth only:
-\begin{equation}\label{eq: p=rhogh}
-\int_{p_\alpha^0}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'},T\right)} = g\left(z - z_0\right) \, .
-\end{equation}
-
-Black-oil equations become:
+Integrating Eqs.~\eqref{eq: Sw1}, \eqref{eq: So1} and \eqref{eq: Sg1} in volume and applying divergence theoreom gives:
 \begin{itemize}
@@
 \begin{equation}\label{eq: Sw2}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV = \rho_w^{ST}Q_w^{ST} \, ;
+\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV = - \oint_{\partial \Omega} \rho_w \left( \vec{u}_w \cdot \vec{dS} \right)= \rho_w^{ST}\left(Q_{we}^{ST} + Q_{wi}^{ST}-Q_w^{ST}\right) \, ;
 \end{equation}
@@
 \begin{equation}\label{eq: So2}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV = \rho_o^{ST}Q_o^{ST} \, ;
+\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV = -\oint_{\partial \Omega} \rho_o Y_o \left(\vec{u}_o \cdot \vec{dS} \right) = \rho_o^{ST}Q_o^{ST} \, ;
 \end{equation}
@@
 \begin{equation}\label{eq: Sg2}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV = \rho_g^{ST}Q_g^{ST} \, .
+\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV = - \oint_{\partial \Omega} \left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right) \cdot \vec{dS} 
+=  \rho_g^{ST}(Q_{gi}^{ST}-Q_g^{ST}) \, .
 \end{equation}
 \end{itemize}
+%
+%\begin{itemize}
+%\item water component:
+%\begin{equation}\label{eq: Sw2}
+%\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV = \rho_w^{ST}Q_w^{ST} \, ;
+%\end{equation}
+%
+%\item oil component:
+%\begin{equation}\label{eq: So2}
+%\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV = \rho_o^{ST}Q_o^{ST} \, ;
+%\end{equation}
+%
+%\item gas component:
+%\begin{equation}\label{eq: Sg2}
+%\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV = \rho_g^{ST}Q_g^{ST} \, .
+%\end{equation}
+%\end{itemize}
 %
@@
 
-All quantities are initially functions of space and time and they become functions only of time after being averaged in space:
+All quantities are initially functions of space and time and they become functions of time only after being averaged in space:
 
@@
 \begin{equation}\label{eq: Sw3}
-\frac{\partial}{\partial t} \left( \bar{\rho_w}\bar{S}_w\bar{\phi} \Omega \right) = \rho_w^{ST}Q_w^{ST}
+\frac{\partial}{\partial t} \left( \bar{\rho_w}\bar{S}_w\bar{\phi} \Omega \right) = \rho_w^{ST}\left(Q_{we}^{ST} + Q_{wi}^{ST}-Q_w^{ST}\right)
 \end{equation}
@@
 \begin{equation}\label{eq: Sg3}
-\frac{\partial}{\partial t} \left[\left( \bar{\rho}_o\bar{Y}_g \bar{S}_o + \bar{\rho}_g \bar{S}_g \right) \bar{\phi} \Omega\right] = \rho_g^{ST}Q_g^{ST} \, .
+\frac{\partial}{\partial t} \left[\left( \bar{\rho}_o\bar{Y}_g \bar{S}_o + \bar{\rho}_g \bar{S}_g \right) \bar{\phi} \Omega\right] =  \rho_g^{ST}(Q_{gi}^{ST}-Q_g^{ST}) \, .
 \end{equation}
@@
 \begin{equation}\label{eq: Sw4}
-\frac{\partial}{\partial t} \left(\frac{\bar{S}_w V_p}{B_w} \right) = Q_w^{ST} \, ,
+\frac{\partial}{\partial t} \left(\frac{\bar{S}_w V_p}{B_w} \right) = Q_{we}^{ST} + Q_{wi}^{ST}-Q_w^{ST}
 \end{equation}
@@
 \begin{equation}\label{eq: Sg4}
-\frac{\partial}{\partial t} \left[ V_p \left(\frac{\bar{S}_g}{B_g} + \frac{\bar{S}_o R_s}{B_o} \right) \right] = Q_g^{ST} \, ,
+\frac{\partial}{\partial t} \left[ V_p \left(\frac{\bar{S}_g}{B_g} + \frac{\bar{S}_o R_s}{B_o} \right) \right] = \left( Q_{gi}^{ST}-Q_g^{ST} \right) \, ,
 \end{equation}
@@
 We shall see that to compute the averaged phase density, we may define some reference depth and use the corresponding local density as an approximation to the averaged density without introducing any important error.
+
+We start assuming mechanical equilibrium, i.e., no bulk motion:
+\begin{equation}
+\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right)\equiv 0 \, .
+\end{equation}
+This condition implies that pressure gradients must equal gravity force per unit volume:
+\begin{equation}\label{eq: p=rhogh}
+\nabla p_\alpha = \rho_\alpha g\nabla z
+\end{equation}
+and, consequently, pressure is a function of depth only:
+\begin{equation}\label{eq: p=rhogh}
+\int_{p_\alpha^0}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'},T\right)} = g\left(z - z_0\right) \, .
+\end{equation}
 

### 2014-06-21 pituba/article/article_spe.tex
progresseing

@@
 \begin{equation}\label{eq: exdarcy}
-\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right) \, , \quad \alpha=o,w,g \, .
+\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha}\cdot \left(\nabla p_\alpha - \rho_\alpha \vec{g}\right) \, , \quad \alpha=o,w,g \, .
 \end{equation}
@@
 
-It is now possible to obtain the classical material balance equation substituting the equations for phase saturation, Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5}, into Eq.~\eqref{eq: S15}:
+It is now possible to obtain the general material balance equation substituting the equations for phase saturation, Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5}, into Eq.~\eqref{eq: S15}:
 
@@
 Eq~\eqref{eq: MBE} involves fluid properties, original fluid volumes ($N$, $W$ and $G)$, cumulative production and injection and aquifer influx.
-It is an agelbraic equation that can be solved to any one of its terms, provided that all remaining are known. This works focuses on its solution for pressure.
+It is an algebraic equation that can be solved for any one of its terms, provided that all remaining are known. This works focuses on its solution for pressure.
 
-\subsubsection{An interpretation for calculated pressure and the definition of a reference depth}
+\subsubsection{An interpretation for the calculated pressure}
 
-Material balance equation is widely used to stablish a relationship between production (and injection) and reservoir pressure.
-In Eq.~\eqref{eq: MBE}, however, formation volume factors were defined using averaged values of fluid properties, which are functions of local pressure.
-It turns out, though, that pressure is a function of depth and phase.
+% no assumption on static conditions are needed
+% pressure field is not known
+% pressure field can be assumed constant
+% pressure field can be given by hydrostatic equilibrium
+
+Although it was not needed to assume no fluid flow to obtain material balance equation, fluid motion can only be taken into account in the averaging of fluid properties if the pressure field is known.
+For any practical purpose, this is not true.
+
+A possible simplification to the pressure field is neglecting hydrodynamic gradients and assuming hydrostatic equilibrium:
+%We start assuming mechanical equilibrium, i.e., no bulk motion:
+\begin{equation}
+\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha \vec{g}\right)\equiv 0 \, .
+\end{equation}
+This assumption implies the identity of pressure gradient and gravity force per unit volume
+\begin{equation}\label{eq: p=rhogh}
+\nabla p_\alpha \equiv \rho_\alpha \vec{g} \, ,
+\end{equation}
+and, consequently, pressure becomes a function of time and depth only:
+\begin{equation}\label{eq: p=rhogh}
+\int_{p_\alpha^0 (t)}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - z_0\right) \, ,
+\end{equation}
+where $p_\alpha^0 (t)$ is the pressure of phase $\alpha$ at some reference depth $z_0$ and time $t$.
+
+Material balance equation is widely used to establish a relationship between production (and injection) and reservoir pressure.
+In Eq.~\eqref{eq: MBE}, however, formation volume factors were defined using averaged values of fluid properties that depend on the pressure field.
+
+Additional assumptions are then needed to simpli
 
@@
 We shall see that to compute the averaged phase density, we may define some reference depth and use the corresponding local density as an approximation to the averaged density without introducing any important error.
-
-We start assuming mechanical equilibrium, i.e., no bulk motion:
-\begin{equation}
-\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right)\equiv 0 \, .
-\end{equation}
-This condition implies that pressure gradients must equal gravity force per unit volume:
-\begin{equation}\label{eq: p=rhogh}
-\nabla p_\alpha = \rho_\alpha g\nabla z
-\end{equation}
-and, consequently, pressure is a function of depth only:
-\begin{equation}\label{eq: p=rhogh}
-\int_{p_\alpha^0}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'},T\right)} = g\left(z - z_0\right) \, .
-\end{equation}
 
@@
 We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that, as a consequence, only one root of $\varepsilon$ exists at most.
-Equivalently, we may say that only one solution exists for material balance equation.
+Equivalently, we can say that only one solution exists for material balance equation.
 
@@
 Residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
-The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels corresponding each one to a different phase.
+The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels corresponding each parcel to a different phase.
 Each parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
@@
 Eq.~\eqref{eq: PVTH1} also provides one condition to check whether PVT data obtained from laboratory or empirical correlations is physically correct.
-It can be alternatively formulated as shown below and indicates that any vaporization of oil phase is followed by an increase in its density.
-\begin{equation}
+It can be alternatively formulated as shown in Eq.~\eqref{eq: vaporization} and it simply indicates that any vaporization must be followed by an increase in oil phase density:
+\begin{equation}\label{eq: vaporization}
 \frac{1}{\rho_o}\frac{d\rho_o}{dY_o} \geq Y_o \left( \frac{\rho_o}{\rho_g} - 1 \right) \, .
@@
 One possible choice is Newton's method, which linearizes the function using either its gradient or a secant line.
-The secant line approach introduces an numerical smoothness to the gradient near continuous but non-differentiable points, for what it is the approach used here to deal with the bubble-point pressure.
+The secant line approach introduces a numerical smoothness to the gradient near continuous but non-differentiable points, for what it is the approach used here to deal with the bubble-point pressure.
 

### 2014-06-23 pituba/article/article_spe.tex
more and more

@@
 % Review
-\documentclass[authoryear,preprint,review,10pt]{elsarticle}
-%\documentclass[final,authoryear,5p,times,twocolumn,11pt]{elsarticle}
+%\documentclass[authoryear,preprint,review,10pt]{elsarticle}
+\documentclass[final,authoryear,5p,times,twocolumn,10pt]{elsarticle}
 
@@
 \section{Introduction}
-Material balance equation is a classical tool in petroleum reservoir engineering.
-It consists of an algebraic equation originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from mass balance.
+The material balance equation is a classical tool in petroleum reservoir engineering.
+It consists of the algebraic equation originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from mass balance.
 \cite{schilthuis1936active} introduced the hypothesis of equilibrium attained at all times not necessarily in the entire reservoir volume, but in the regions of high permeability where most of the production would come from and where pressure would present a similar behavior.
@@
 
-As mentioned, the equations developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium.
+As mentioned, the equation developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium.
 An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}.
@@
 \item water component:
-\begin{equation}\label{eq: Sw2}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV = - \oint_{\partial \Omega} \rho_w \left( \vec{u}_w \cdot \vec{dS} \right)= \rho_w^{ST}\left(Q_{we}^{ST} + Q_{wi}^{ST}-Q_w^{ST}\right) \, ;
-\end{equation}
+\begin{align}\label{eq: Sw2}
+\begin{split}
+\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV &= - \oint_{\partial \Omega} \rho_w \left( \vec{u}_w \cdot \vec{dS} \right) \\
+&= \rho_w^{ST}\left(Q_{we}^{ST} + Q_{wi}^{ST}-Q_w^{ST}\right) \, ;
+\end{split}
+\end{align}
 
 \item oil component:
-\begin{equation}\label{eq: So2}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV = -\oint_{\partial \Omega} \rho_o Y_o \left(\vec{u}_o \cdot \vec{dS} \right) = \rho_o^{ST}Q_o^{ST} \, ;
-\end{equation}
+\begin{align}\label{eq: So2}
+\begin{split}
+\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV &= -\oint_{\partial \Omega} \rho_o Y_o \left(\vec{u}_o \cdot \vec{dS} \right) \\
+&= \rho_o^{ST}Q_o^{ST} \, ;
+\end{split}
+\end{align}
 
 \item gas component:
-\begin{equation}\label{eq: Sg2}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV = - \oint_{\partial \Omega} \left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right) \cdot \vec{dS} 
-=  \rho_g^{ST}(Q_{gi}^{ST}-Q_g^{ST}) \, .
-\end{equation}
+\begin{align}\label{eq: Sg2}
+\begin{split}
+\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV &= - \oint_{\partial \Omega} \left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right) \cdot \vec{dS} \\
+&=  \rho_g^{ST}(Q_{gi}^{ST}-Q_g^{ST}) \, .
+\end{split}
+\end{align}
 \end{itemize}
@@
 
-It is now possible to obtain the general material balance equation substituting the equations for phase saturation, Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5}, into Eq.~\eqref{eq: S15}:
+It is now possible to obtain the so-called generalized material balance equation substituting the equations for phase saturation, Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5}, into Eq.~\eqref{eq: S15}:
 
 \begin{equation}\label{eq: MBE}
-F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{porous volume contraction } (C)}
+F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{pore volume reduction } (C)}
 \end{equation}
@@
 
-\subsubsection{An interpretation for the calculated pressure}
+\subsubsection{An interpretation for how material balance equation relates to reservoir pressure}
 
@@
 
+Material balance equation is widely used to establish a relationship between production (and injection) and reservoir pressure.
+In Eq.~\eqref{eq: MBE}, however, formation volume factors were defined using averaged values of fluid properties that depend on the pressure field.
+
 Although it was not needed to assume no fluid flow to obtain material balance equation, fluid motion can only be taken into account in the averaging of fluid properties if the pressure field is known.
@@
 
-A possible simplification to the pressure field is neglecting hydrodynamic gradients and assuming hydrostatic equilibrium:
-%We start assuming mechanical equilibrium, i.e., no bulk motion:
-\begin{equation}
-\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha \vec{g}\right)\equiv 0 \, .
+A possible simplification to the pressure field is neglecting hydrodynamic gradients and assuming hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$.
+This also implies the identity of pressure gradient and gravity force per unit volume, $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, consequently, pressure becomes a function of time and depth only:
+\begin{equation}\label{eq: p=rhogh}
+\int_{p_\alpha^{ref} (t)}^{p_\alpha (t,z)} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - z_{ref}\right) \, ,
 \end{equation}
-This assumption implies the identity of pressure gradient and gravity force per unit volume
-\begin{equation}\label{eq: p=rhogh}
-\nabla p_\alpha \equiv \rho_\alpha \vec{g} \, ,
-\end{equation}
-and, consequently, pressure becomes a function of time and depth only:
-\begin{equation}\label{eq: p=rhogh}
-\int_{p_\alpha^0 (t)}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - z_0\right) \, ,
-\end{equation}
-where $p_\alpha^0 (t)$ is the pressure of phase $\alpha$ at some reference depth $z_0$ and time $t$.
+where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha^{ref} (t)$ is the pressure at some reference depth $z_{ref}$.
 
-Material balance equation is widely used to establish a relationship between production (and injection) and reservoir pressure.
-In Eq.~\eqref{eq: MBE}, however, formation volume factors were defined using averaged values of fluid properties that depend on the pressure field.
+It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly in material balance equation.
+It can be shown, however, that neglecting the variation of fluid properties with depth does not introduce any important error.
+For instance, the average oil density can be approximated to the local density at the reference level, $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$, what also applies to the formation volume factor: $B_o \approx \rho_o (t, z_{ref})/ \rho_o^{ST}$.
 
-Additional assumptions are then needed to simpli
+%For liquids, a first-order approximation to the relationship between density and pressure at constant temperature and composition can be assumed:
 
-% ERRADO: Using the average reservoir pressure ($\bar{p}$) to compute fluid properties does not solve this issue because $\bar{\rho}_o$ is not equal to $\rho_o (\bar{p})$ and thus $B_o$ is not a simple function of $p$.
+%
+%
+%
+%
+%
+%Additional assumptions are then needed to simpli
+%
+%% ERRADO: Using the average reservoir pressure ($\bar{p}$) to compute fluid properties does not solve this issue because $\bar{\rho}_o$ is not equal to $\rho_o (\bar{p})$ and thus $B_o$ is not a simple function of $p$.
+%
+%What is then the meaning of pressure calculations in material balance equations?
+%
+%For answering this question, we examine the two roles played by pressure in material balance equation: computation of phase density and phase equilibrium. 
+%
+%We shall see that to compute the averaged phase density, we may define some reference depth and use the corresponding local density as an approximation to the averaged density without introducing any important error.
 
-What is then the meaning of pressure calculations in material balance equations?
-
-For answering this question, we examine the two roles played by pressure in material balance equation: computation of phase density and phase equilibrium.
-
-We shall see that to compute the averaged phase density, we may define some reference depth and use the corresponding local density as an approximation to the averaged density without introducing any important error.
-
-Pressure increases with depth according to the hydrostatic gradient.
-Consequently, density also increases but by a negligible amount.
+As pressure increases with depth according to the hydrostatic gradient, density also increases, but by a negligible amount.
 This approximation may be quantified by putting together the expressions for compressibility and hydrostatic equilibrium.
-For this, we assume a constant isothermal compressibility for oil phase
+For this, we assume a constant isothermal compressibility for oil phase,
 \begin{equation}\label{eq: grado}
-\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o 
+\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o \, ,
 \end{equation}
@@
 \begin{equation}
-\left( \frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
+ \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
 \end{equation}
 
-Similarly for the gas phase, we obtain its isothermal compressibility from the gas state equation
+Similarly for gas phase, we obtain its isothermal compressibility from the gas state equation, $pV = Zn R T$:
 \begin{equation}
-\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial p_g}\right)_{T,Z} = \frac{1}{p_g} = \frac{\mathcal{M}}{\rho_g ZRT}
+\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial p_g}\right)_{T} = \frac{1}{p_g}-\frac{1}{Z}\frac{\partial Z}{\partial p_g} \, ,
 \end{equation}
@@
 \begin{equation}\label{eq: gradg}
-\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
+\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
 \end{equation}
@@
 \begin{equation}
-\left(\frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 10^{-6}\ m^{-1} \, ,
+\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 10^{-6}\ m^{-1} \, ,
 \end{equation}
@@
 \begin{equation}
-\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} \approx 10^{-5}\ m^{-1} \, .
+\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T,Z} \approx 10^{-5}\ m^{-1} \, .
 \end{equation}
@@
 \begin{equation}
-\left(\frac{1}{\rho_g}\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
+\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
 \end{equation}
@@
 
-Eq.~\eqref{eq: PVTH1} also provides one condition to check whether PVT data obtained from laboratory or empirical correlations is physically correct.
+Eq.~\eqref{eq: PVTH1} also provides a condition to check whether PVT data obtained from laboratory or empirical correlations is physically correct.
 It can be alternatively formulated as shown in Eq.~\eqref{eq: vaporization} and it simply indicates that any vaporization must be followed by an increase in oil phase density:
@@
 $G_p$ & produced volume of gas \\
-$m$ & ratio of original free gas volume to original oil phase volume in reservoir condition \\
+$m$ & ratio of original free gas volume \\
+& to original oil volume in reservoir condition \\
 $N$ & original volume of oil in place \\
@@
 % % %
-
+\appendix
+\include{compressibility}
 \end{document}

### 2014-06-23 pituba/article/python/density.py

@@ -1,10 +1,13 @@
+#!/usr/bin/python
+
 from pylab import *
 
-rc('font',**{'family':'sans-serif','sans-serif':['Helvetica']})
+#rc('font',**{'family':'sans-serif','sans-serif':['Helvetica']})
 ## for Palatino and other serif fonts use:
 #rc('font',**{'family':'serif','serif':['Palatino']})
 ## rc('font',family='serif')
 
+rc('font',family='serif')
 rc('text', usetex=True)
 rc('font', size=16)
 rc('xtick', labelsize=16)
@@ -15,29 +18,52 @@ rc('figure',figsize=[8,6])
 fbottom=0.15
 fleft=0.15
 
-rho=1e3
+def errol(rho,c,z):
+    g=9.81
+    
+    dp0=rho*g*z
+    dp1=1/c*(exp(c*dp0)-1.)
+    return (dp0-dp1)/dp1
+
+def errog(rho,p0,z):
+    g=9.81
+    
+    dp0=rho*g*z
+    dp1=p0*(exp(dp0/p0)-1.)
+    return (dp0-dp1)/dp1
+
+
 g=9.81
-z=linspace(0,500.,100)
-c1=100e-11
-c2=50e-6/6894.75
+z=linspace(1e-3,500.,10)
+c1=3.6e-6/(g*1e4)*14.23
+c2=50e-6/(g*1e4)*14.23
 
-x1=rho*g*z*c1
-x2=rho*g*z*c2
-## x=0.1*100e-6*z
+rhoo=1e3
+l1=errol(rhoo,c1,z)
+l2=errol(rhoo,c2,z)
 
-y1=1./(1.-x1)
-y2=1./(1.-x2)
+p0=100e5
+rhog=p0/1e5
+g1=errog(rhog,p0,z)
+g2=errog(rhog,p0*5,z)
 
 figure()
 #plot(z, y2-1,'r',label='approx 0th')
-plot(z, 100*(y1-1),'k-',label=r'$\rho =1000\ kg/m^3$ and $c=10^{-9}\ Pa$')
-plot(z, 100*(y2-1),'k--',label=r'$\rho =1000\ kg/m^3$ and $c=2 \times 10^{-9}\ Pa$')
-legend(loc='upper left')
+plot(z, 100*l1,'k-',label=r'$\rho_o =1000\ kg/m^3$ and $c_o=10^{-3}\ 1/MPa$')
[…]

### 2014-06-25 pituba/article/article_spe.tex
good progress

@@
 % Review
-%\documentclass[authoryear,preprint,review,10pt]{elsarticle}
-\documentclass[final,authoryear,5p,times,twocolumn,10pt]{elsarticle}
+\documentclass[authoryear,preprint,review,11pt]{elsarticle}
+%\documentclass[final,authoryear,5p,twocolumn,10pt]{elsarticle}
 
@@
 This work derives the classical material balance equation from a three dimensional black-oil model.
-Sufficient conditions for solution uniqueness of material balance equation are presented for the case that pressure is the unknown quantity.
+Sufficient conditions for solution uniqueness of the material balance equation are presented for the case that pressure is the unknown quantity.
 It is shown that spatially averaged PVT properties are required for reducing a black-oil model to the material balance approach.
@@
 
-\subsubsection{An interpretation for how material balance equation relates to reservoir pressure}
+\subsubsection{Relationship between the material balance equation and reservoir pressure}
 
@@
 
-Material balance equation is widely used to establish a relationship between production (and injection) and reservoir pressure.
-In Eq.~\eqref{eq: MBE}, however, formation volume factors were defined using averaged values of fluid properties that depend on the pressure field.
+The material balance equation is widely used to relate production, injection and reservoir pressure.
+Actually, all averaged fluid properties in Eq.~\eqref{eq: MBE} depend on the pressure field, and so does the pore volume.
+If the reservoir is under production, though, the pressure field is not known and, for its prediction, one would have to resort to more advanced techniques such as reservoir simulation.
 
-Although it was not needed to assume no fluid flow to obtain material balance equation, fluid motion can only be taken into account in the averaging of fluid properties if the pressure field is known.
+%Although it was not needed to assume no fluid flow to obtain the material balance equation, fluid motion can only be taken into account in the averaging of fluid properties if the pressure field is known.
 For any practical purpose, this is not true.
 
-A possible simplification to the pressure field is neglecting hydrodynamic gradients and assuming hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$.
-This also implies the identity of pressure gradient and gravity force per unit volume, $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, consequently, pressure becomes a function of time and depth only:
+To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and to assume hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$.
+This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as consequence, pressure becomes a function of time and depth only given by
 \begin{equation}\label{eq: p=rhogh}
@@
 
-It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly in material balance equation.
-It can be shown, however, that neglecting the variation of fluid properties with depth does not introduce any important error.
-For instance, the average oil density can be approximated to the local density at the reference level, $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$, what also applies to the formation volume factor: $B_o \approx \rho_o (t, z_{ref})/ \rho_o^{ST}$.
+It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly in the material balance equation.
+Furthermore, it can be shown that neglecting the variation of fluid properties with depth does not introduce any important error.
+As pressure increases with depth according to the hydrostatic gradient, density also increases, but by a negligible amount.
+For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$; the same applies to the formation volume factor: $B_o \approx \rho_o (t, z_{ref})/ \rho_o^{ST}$.
 
@@
 
-As pressure increases with depth according to the hydrostatic gradient, density also increases, but by a negligible amount.
-This approximation may be quantified by putting together the expressions for compressibility and hydrostatic equilibrium.
-For this, we assume a constant isothermal compressibility for oil phase,
+This approximation can be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase,
 \begin{equation}\label{eq: grado}
@@
 
-Substituting typical values in Eqs.~\eqref{eq: grado} and \eqref{eq: gradg}, we see that oil and gas densities are rather constant with depth:
+Substituting typical values in Eqs.~\eqref{eq: grado} and \eqref{eq: gradg}, we see that oil and gas densities are almost constant with depth:
 \begin{equation}
@@
 
-Oil and gas densities are thus rather constant with depth and a zeroth order approximation of $\bar{\rho}$ is accurate within $0.5\%$ for material balance calculations.
+Oil and gas densities are thus approximately constant with depth. Fig~\ref and a zeroth order approximation of $\bar{\rho}$ is accurate within $0.5\%$ for material balance calculations.
 
-The reference depth for computing phase density (or formation volume factors) is then any arbitrary depth inside phase volume.
+The reference depth for computing phase density (or formation volume factors) is any arbitrary depth inside phase volume.
 For a negligible variation of density with depth, it is easily seen that $\bar{\rho} = \rho\left(\bar{p}\right)$ and $\bar{p}=p\left(z^0\right)$ where $z^0$ is the depth of phase gravity center.
@@
 For the case of an oil reservoir covered by a gas cap, the reference depth may be defined, with advantage, at the gas-oil contact.
-The point here is that at phase contact both gas and oil densities can be computed using the same pressure, i.e., the corresponding local value (we are neglecting capillary pressure).
+The point here is that at phase contact both gas and oil densities can be computed using the same pressure, i.e., the corresponding local value.
 
@@
 It has been shown that pressure variation with depth is not important for the computation of gas, oil and water densities.
-Nevertheless, it was said that pressure also plays a second role in material balance equation by defining phase equilibrium.
-For this purpose, pressure variation with depth may not be negligible in the evalutation of the solubility ratio:
+Nevertheless, it was said that pressure also plays a second role in the material balance equation by defining phase equilibrium.
+For this purpose, pressure variation with depth may not be negligible in the evaluation of the solubility ratio:
 \begin{equation}\label{eq: dRsdz1}

### 2014-06-25 pituba/article/article_spe.tex
good progress 2

@@
 Actually, all averaged fluid properties in Eq.~\eqref{eq: MBE} depend on the pressure field, and so does the pore volume.
-If the reservoir is under production, though, the pressure field is not known and, for its prediction, one would have to resort to more advanced techniques such as reservoir simulation.
+If the reservoir is under production, though, the pressure field is not known and for its prediction one would have to resort to more advanced techniques such as reservoir simulation.
 
@@
 
-To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and to assume hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$.
-This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as consequence, pressure becomes a function of time and depth only given by
+To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$.
+This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as a consequence, pressure becomes a function of time and depth only given by
 \begin{equation}\label{eq: p=rhogh}
@@
 Furthermore, it can be shown that neglecting the variation of fluid properties with depth does not introduce any important error.
-As pressure increases with depth according to the hydrostatic gradient, density also increases, but by a negligible amount.
+As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount.
 For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$; the same applies to the formation volume factor: $B_o \approx \rho_o (t, z_{ref})/ \rho_o^{ST}$.

### 2014-06-29 pituba/article/article_spe.tex
abstract

@@
 \begin{abstract}
-This work derives the classical material balance equation from a three dimensional black-oil model.
-Sufficient conditions for solution uniqueness of the material balance equation are presented for the case that pressure is the unknown quantity.
-It is shown that spatially averaged PVT properties are required for reducing a black-oil model to the material balance approach.
-It is also shown that the averaging process may be neglected for many practical purposes.
-An interpretation for pressure in material balance equation is provided and its relation to a reference depth is examined.
-A scheme numerically solve material balance equation for pressure is also proposed.
-%
-%This work derives the classical material balance equation from a black-oil model.
-We examine the existing link between both approaches in detail, including the necessary step of averaging PVT properties.
-An interpretation for pressure in material balance calculations is provided and the choice of a reference depth (datum) is discussed.
-The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented.
-A scheme for numerical solution is also proposed.
+This work examines the underlying details in obtaining the classical material balance equation from a three dimensional black-oil model typical of reservoir simulations.
+The three-dimensional model is integrated in reservoir volume and quantities such as fluid properties are spatially averaged.
+The material balance equation is obtained for the averaged quantities and it is applied to calculate reservoir pressure.
+It is shown that the calculated pressure corresponds to the local pressure at a defined depth and not to the average reservoir pressure in general.
+The definition of a convenient reference depth is discussed.
+Sufficient conditions for solution uniqueness of the material balance equation are provided for the case that pressure is the unknown quantity and one of such conditions is shown to check for physical correctness of PVT properties.
+Finally, a scheme for solving the material balance equation numerically is proposed.
 \end{abstract}
@@
 
-It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly in the material balance equation.
+It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.
 Furthermore, it can be shown that neglecting the variation of fluid properties with depth does not introduce any important error.
@@
  and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
-\begin{equation}
+\begin{equation}\label{eq: gradodz}
  \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
@@
 
-Similarly for gas phase, we obtain its isothermal compressibility from the gas state equation, $pV = Zn R T$:
+Similarly for the gas phase, using the state equation $pV = Zn R T$, we obtain its isothermal compressibility:
 \begin{equation}
-\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial p_g}\right)_{T} = \frac{1}{p_g}-\frac{1}{Z}\frac{\partial Z}{\partial p_g} \, ,
+\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial p_g}\right)_{T} = \frac{1}{p_g}-\frac{1}{Z}\frac{\partial Z}{\partial p_g} \, .
 \end{equation}
-and again we use hydrostatic equilibrium to obtain the derivative of gas density with depth:
-\begin{equation}\label{eq: gradg}
+Neglecting the variation of Z-factor and using hydrostatic equilibrium again, we to obtain the derivative of gas density with depth as:
+\begin{equation}\label{eq: gradgdz}
 \frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
@@
 
-Substituting typical values in Eqs.~\eqref{eq: grado} and \eqref{eq: gradg}, we see that oil and gas densities are almost constant with depth:
+For investigating the orders of magnitude of Eqs.~\eqref{eq: gradodz} and \eqref{eq: gradgdz}, we substitute in them the values of $\rho_o = 10^3\ kg/m^3$ and $c_o = 10^{-9}\ Pa^{-1}$ for the oil phase, and $\mathcal{M}=19\ kg/kmol$ and $T=400\ K$ for the gas phase:
 \begin{equation}
-\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 10^{-6}\ m^{-1} \, ,
+\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z}\right)_{T,Y_o} \approx 1\times10^{-5}\ m^{-1} \, ,
 \end{equation}
@@
 \begin{equation}
-\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T,Z} \approx 10^{-5}\ m^{-1} \, .
+\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T,Z} \approx 5\times10^{-5}\ m^{-1} \, .
 \end{equation}
 
-Oil and gas densities are thus approximately constant with depth. Fig~\ref and a zeroth order approximation of $\bar{\rho}$ is accurate within $0.5\%$ for material balance calculations.
+Oil and gas densities are thus approximately constant with depth.
+The reference depth for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume.
+Under the approximation of a constant density, it is easily seen that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$ where $z^0_\alpha$ is the depth of phase gravity center and $\alpha=w, o, g$.
 
-The reference depth for computing phase density (or formation volume factors) is any arbitrary depth inside phase volume.
-For a negligible variation of density with depth, it is easily seen that $\bar{\rho} = \rho\left(\bar{p}\right)$ and $\bar{p}=p\left(z^0\right)$ where $z^0$ is the depth of phase gravity center.
-
-For the case of an oil reservoir covered by a gas cap, the reference depth may be defined, with advantage, at the gas-oil contact.
+For the case of an oil reservoir with a gas cap, the reference depth may be defined, with advantage, at the gas-oil contact.
 The point here is that at phase contact both gas and oil densities can be computed using the same pressure, i.e., the corresponding local value.
@@
 
-For a non-volatile saturated oil, we can approximate $dR_s / dp_o$ to the ratio between  solubility ratio ($R_{s}$) and saturation pressure ($p_{sat}$), so that Eq.~\eqref{eq: dRsdz1} becomes:
+For a non-volatile oil, we can approximate $dR_s / dp_o$ to the ratio between the initial solubility ratio ($R_{s,0}$) and saturation pressure ($p_{sat}$), so that Eq.~\eqref{eq: dRsdz1} becomes:
 \begin{equation}\label{eq: dRsdz2}
@@
 Eq.~\eqref{eq: PVTH1} also provides a condition to check whether PVT data obtained from laboratory or empirical correlations is physically correct.
-It can be alternatively formulated as shown in Eq.~\eqref{eq: vaporization} and it simply indicates that any vaporization must be followed by an increase in oil phase density:
+It can be alternatively formulated as shown bellow in Eq.~\eqref{eq: vaporization} and it simply indicates that any vaporization must be followed by an increase in oil phase density:
 \begin{equation}\label{eq: vaporization}

### 2014-06-29 pituba/article/article_spe.tex
abstract2

@@
 
-As mentioned, the equation developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium.
+The equation developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium.
 An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}.
@@
 The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system.
-For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component can be defined.
+For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component can thus be defined.
 Similarly, a pseudo oil component can also be defined by subtracting all the pseudo gas component from oil phase composition.
-After defining these two pseudo-components, the number of transport equations needed for hydrocarbon species is reduced from a usually large number to only two.
+After defining these two pseudo-components, the number of transport equations needed for hydrocarbon components is reduced from a usually large number to only two.
 
@@
 \cite{ertekin2001basic} briefly addressed this subject arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \vec{g}$) were neglected and if PVT properties were averaged in reservoir volume.
-The intuitive notion is that material balance equation is obtained if fluid motion is neglected and black-oil equations are integrated in reservoir volume.
 This work develops this process examining underlying details.
@@
  The same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}, is also applied here to material balance.
-This approach extends material balance equation to saturated reservoirs under pressure maintenance (e.g.
+This approach extends the material balance equation to saturated reservoirs under pressure maintenance (e.g.
 fluid injection).
@@
 After presenting material balance equation, sufficient conditions are shown for which material balance equation has only one solution for pressure.
-These conditions require fluids to be compressible and that gas dissolution occurs with reduction of the overall system volume.
+These conditions require fluids to be compressible and that gas vaporization occurs with reduction of oil phase volume.
 A simple scheme for numerical solution is also proposed based on classical root-finding algorithms for non-linear equations.
@@
 
-This section starts presenting black-oil equations and it proceeds averaging quantities in space and examining simplifying assumptions for which material balance equation is obtained.
+This section presents black-oil equations, performs averaging of quantities in space and examines simplifying assumptions necessary to obtain the material balance equation.
 
@@
 There can be three fluid phases: oil, gas and (liquid) water.
-Oil phase may be composed of oil and gas components, gas phase is composed uniquely by gas component and water phase uniquely by water component.
-The only allowed mass transfer is of gas component between oil and gas phases, i.e., gas phase composition is always that of a dry gas.
+Oil phase is composed of oil and gas components, gas phase is composed uniquely by gas component and, water phase, uniquely by water component.
+The only allowed mass transfer is of gas component between oil and gas phases.
+Gas phase composition is always that of a dry gas.
 
@@
 
-Mass balance equations for each component read:
+Mass balance equation for each component reads:
 
@@
 
-Initial conditions are of no fluid motion and boundary conditions are of no normal flow at all boundaries except for interfaces with wells or aquifers.
+Initial conditions are of no fluid motion. Boundary conditions are of no normal flow at all boundaries except for interfaces with wells or aquifers.
 
@@
 Temperature is a constant prescribed value and thus no equation for energy balance is needed.
-It follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if component number is restricted to the two pseudo-components.
+It follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if the component number is restricted to the two pseudo-components.
 As temperature is assumed constant, oil phase density is a function of pressure alone.
@@
 
-Porosity ($\phi$) is assumed a function of pore pressure ($p_P$). Because three phases are present in general, such quantity is defined by some relation of type \citep{kim2011rigorous}:
+Porosity ($\phi$) is assumed a function of pore pressure ($p_P$). Because three phases are present in general, such quantity is defined by some relation of the type \citep{kim2011rigorous}:
 \begin{equation}
@@
 
-All quantities are initially functions of space and time and they become functions of time only after being averaged in space:
+All quantities are initially functions of space and time, but they become functions of time only after being averaged in space:
 
@@
 
-Omitting function arguments:
+We proceed omitting function arguments:
 \begin{equation}
@@
 We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases.
-In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to the volume at surface condition, i.e.:
+In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to its volume at surface condition, i.e.:
 
@@
 
-In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between volumes of dissolved gas and oil both measured in surface condition:
+In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between the volume of dissolved gas and the volume of oil both measured in surface condition:
 
@@
 
-An equation for bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure $R_s(p=P_b)$:
+An equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, or $R_s(p=P_b)$:
 \begin{equation}
@@
 
-It is now possible to obtain the so-called generalized material balance equation substituting the equations for phase saturation, Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5}, into Eq.~\eqref{eq: S15}:
+It is now possible to obtain the so-called generalized material balance equation by substituting the equations for phase saturation, Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5}, into Eq.~\eqref{eq: S15}:
 
@@
 
-\subsubsection{Relationship between the material balance equation and reservoir pressure}
+\subsubsection{Relationship between the material balance equation and the reservoir pressure}
 
@@
 To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$.
-This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as a consequence, pressure becomes a function of time and depth only given by
+This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as a consequence, pressure becomes a function of time and depth only given by the following expression:
 \begin{equation}\label{eq: p=rhogh}
@@
 \end{equation}
-where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha^{ref} (t)$ is the pressure at some reference depth $z_{ref}$.
+where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha^{ref} (t)$ is its pressure at some reference depth $z_{ref}$.
+It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.
 
-It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.
 Furthermore, it can be shown that neglecting the variation of fluid properties with depth does not introduce any important error.
@@
 \end{equation}
-Neglecting the variation of Z-factor and using hydrostatic equilibrium again, we to obtain the derivative of gas density with depth as:
+Neglecting the variation of the Z-factor and using hydrostatic equilibrium again, we obtain the derivative of gas density with depth as:
 \begin{equation}\label{eq: gradgdz}
@@
 
-For investigating the orders of magnitude of Eqs.~\eqref{eq: gradodz} and \eqref{eq: gradgdz}, we substitute in them the values of $\rho_o = 10^3\ kg/m^3$ and $c_o = 10^{-9}\ Pa^{-1}$ for the oil phase, and $\mathcal{M}=19\ kg/kmol$ and $T=400\ K$ for the gas phase:
+For investigating the order of magnitude of Eqs.~\eqref{eq: gradodz} and \eqref{eq: gradgdz}, we substitute in them the values of $\rho_o = 10^3\ kg/m^3$ and $c_o = 10^{-9}\ Pa^{-1}$ for the oil phase, and $\mathcal{M}=19\ kg/kmol$ and $T=400\ K$ for the gas phase:
 \begin{equation}
@@
 The reference depth for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume.
-Under the approximation of a constant density, it is easily seen that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$ where $z^0_\alpha$ is the depth of phase gravity center and $\alpha=w, o, g$.
+Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$ where $z^0_\alpha$ is the depth of phase gravity center and $\alpha=w, o, g$.
+The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$, without fluid motion, and only equals the average reservoir pressure $\bar{p}_\alpha$ if $z^{ref}=z^0$.
 
-For the case of an oil reservoir with a gas cap, the reference depth may be defined, with advantage, at the gas-oil contact.
-The point here is that at phase contact both gas and oil densities can be computed using the same pressure, i.e., the corresponding local value.
+For the case of an oil reservoir with a gas cap, the reference depth can be defined, with advantage, at the gas-oil contact.
+The point here is that at phase contact both gas and oil densities are computed using the same pressure, i.e., the corresponding local value.
 
@@
 \end{equation}
-what is $\rho_o / \rho_g$ times greater than the sole effect of gas hydrostatic pressure on its density.
+what is $\rho_o / \rho_g$ times greater than the sole effect of gas hydrostatic pressure that was computed in Eq.~\eqref{eq: gradgdz}.
 
@@
 Eq.~\eqref{eq: PVTH1} also provides a condition to check whether PVT data obtained from laboratory or empirical correlations is physically correct.
-It can be alternatively formulated as shown bellow in Eq.~\eqref{eq: vaporization} and it simply indicates that any vaporization must be followed by an increase in oil phase density:
+It can be alternatively formulated as shown bellow in Eq.~\eqref{eq: vaporization}.
+It simply indicates that any vaporization must be followed by an increase in oil phase density:
 \begin{equation}\label{eq: vaporization}

### 2014-06-30 pituba/article/article_spe.tex
improvs

@@
 An equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, or $R_s(p=P_b)$:
-\begin{equation}
+\begin{equation}\label{eq: eqPb}
 \begin{split}
@@
 \begin{equation}\label{eq: gradgdz}
-\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
+\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \frac{\rho_g \vec{g}}{p_g} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
 \end{equation}
@@
 The reference depth for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume.
-Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$ where $z^0_\alpha$ is the depth of phase gravity center and $\alpha=w, o, g$.
-The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$, without fluid motion, and only equals the average reservoir pressure $\bar{p}_\alpha$ if $z^{ref}=z^0$.
+Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$, where $z^0_\alpha$ is the depth of phase gravity center and $\alpha=w, o, g$.
+The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$, without fluid motion, and only equals the average reservoir pressure $\bar{p}_\alpha$ if the reference depth is the depth of the gravity center, $z^{ref}=z^0$.
 
-For the case of an oil reservoir with a gas cap, the reference depth can be defined, with advantage, at the gas-oil contact.
-The point here is that at phase contact both gas and oil densities are computed using the same pressure, i.e., the corresponding local value.
-
-If the reference depth for oil phase is otherwise defined below gas-oil contact, gas density must be computed at some different depth still inside gas phase, i.e., at or above gas-oil contact.
-This is to avoid the effect of oil hydrostatic gradient in the computation of gas density,
+For the case of an oil reservoir with a gas cap, the reference depth should be defined at the gas-oil contact.
+The point here is that at gas-oil contact both gas and oil densities are computed using the same pressure, its local value.
+If the reference depth, common to all phases, is otherwise defined below gas-oil contact, gas density will increase with depth at a higher rate because of the oil hydrostatic gradient,
 \begin{equation}
@@
 \end{equation}
-what is $\rho_o / \rho_g$ times greater than the sole effect of gas hydrostatic pressure that was computed in Eq.~\eqref{eq: gradgdz}.
+what is $\rho_o / \rho_g$ times greater than the gas hydrostatic gradient used in Eq.~\eqref{eq: gradgdz}.
 
-It has been shown that pressure variation with depth is not important for the computation of gas, oil and water densities.
-Nevertheless, it was said that pressure also plays a second role in the material balance equation by defining phase equilibrium.
-For this purpose, pressure variation with depth may not be negligible in the evaluation of the solubility ratio:
+It has been shown that the pressure variation with depth is not important for the computation of gas, oil and water densities.
+Nevertheless, pressure also plays a second role in the material balance equation by defining phase equilibrium, for what the pressure variation with depth may not be negligible in the solubility ratio:
 \begin{equation}\label{eq: dRsdz1}
@@
 We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that, as a consequence, only one root of $\varepsilon$ exists at most.
-Equivalently, we can say that only one solution exists for material balance equation.
+Equivalently, we can say that only one solution exists for the material balance equation.
 
@@
 Residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
-The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels corresponding each parcel to a different phase.
-Each parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
+The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels, each corresponding to a different phase (three fluid phases and the rock).
+For the fluid phases, each parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
 \begin{equation}
@@
 Eq.~\eqref{eq: PVTH1} also provides a condition to check whether PVT data obtained from laboratory or empirical correlations is physically correct.
-It can be alternatively formulated as shown bellow in Eq.~\eqref{eq: vaporization}.
+It can be alternatively formulated in terms of densities as shown bellow in Eq.~\eqref{eq: vaporization}.
 It simply indicates that any vaporization must be followed by an increase in oil phase density:
@@
 
-The steps to solve all material balance equation are summarized bellow:
+The steps to solve the material balance equation are summarized bellow:
 \begin{step}
-Compute bubble-point pressure for instant $t_2$.
+Compute bubble-point pressure for instant $t_2$ using Eq.~\eqref{eq: eqPb}.
 \end{step}
 
-Bubble-point pressure $P_{b2}$ is only a function of the values in initial instant $t_1$ and produced and injected volumes.
+Bubble-point pressure $P_{b2}$ is only a function of values at instant $t_1$ and produced and injected volumes.
 
 \begin{step}
-Compute fluid pressure for instant $t_2$ finding the root of residual function ($\varepsilon$).
+Compute pressure for instant $t_2$ by finding the root of residual function $\varepsilon \left(p\right)$.
 \end{step}
 
-Pressure in Eq.~\eqref{eq: MBE} is a function only of fluid properties in instant $t_1$, produced and injected values and bubble-point pressure at instant $t_2$, $P_{b2}$, which was already computed in step 1.
+Pressure in Eq.~\eqref{eq: MBE} is only a function of fluid properties in instant $t_1$, produced and injected values and bubble-point pressure at instant $t_2$, $P_{b2}$, which was already computed in step 1.
 

### 2014-07-03 pituba/article/article_spe.tex
improvs

@@
 \end{equation}
-Neglecting the variation of the Z-factor and using hydrostatic equilibrium again, we obtain the derivative of gas density with depth as:
+
+Assuming reservoir temperature to be sufficiently above gas critical temperature for negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
 \begin{equation}\label{eq: gradgdz}

### 2014-07-18 pituba/article/article_spe.tex
misc add

@@
 
-The material balance equation is widely used to relate production, injection and reservoir pressure.
+The material balance equation is widely used to relate production and injection to reservoir pressure.
 Actually, all averaged fluid properties in Eq.~\eqref{eq: MBE} depend on the pressure field, and so does the pore volume.
@@
 \end{equation}
-where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha^{ref} (t)$ is its pressure at some reference depth $z_{ref}$.
+where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha^{ref} (t)$ is the pressure at some reference depth $z_{ref}$.
 It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.
 
-Furthermore, it can be shown that neglecting the variation of fluid properties with depth does not introduce any important error.
+It can be shown that neglecting the variation of fluid properties with depth does not introduce any important error.
 As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount.
@@
 
-For investigating the order of magnitude of Eqs.~\eqref{eq: gradodz} and \eqref{eq: gradgdz}, we substitute in them the values of $\rho_o = 10^3\ kg/m^3$ and $c_o = 10^{-9}\ Pa^{-1}$ for the oil phase, and $\mathcal{M}=19\ kg/kmol$ and $T=400\ K$ for the gas phase:
+For investigating the order of magnitude of Eqs.~\eqref{eq: gradodz} and \eqref{eq: gradgdz}, we substitute into them the values of $\rho_o = 10^3\ kg/m^3$ and $c_o = 10^{-9}\ Pa^{-1}$ for the oil phase, and $\mathcal{M}=19\ kg/kmol$ and $T=400\ K$ for the gas phase:
 \begin{equation}
@@
 Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$, where $z^0_\alpha$ is the depth of phase gravity center and $\alpha=w, o, g$.
-The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$, without fluid motion, and only equals the average reservoir pressure $\bar{p}_\alpha$ if the reference depth is the depth of the gravity center, $z^{ref}=z^0$.
+The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$, without fluid motion, and only equals the average reservoir pressure $\bar{p}_\alpha$ if the reference depth is that of the gravity center, $z^{ref}=z^0$.
 
@@
 
-Residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
-The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels, each corresponding to a different phase (three fluid phases and the rock).
+The residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
+The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels, each corresponding to a different phase (three fluid phases and the solid rock).
 For the fluid phases, each parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:

### 2014-07-24 pituba/article/article_spe.tex
more and more

@@
 \begin{abstract}
-This work examines the underlying details in obtaining the classical material balance equation from a three dimensional black-oil model typical of reservoir simulations.
+This work examines underlying details in obtaining the classical material balance equation from a three dimensional black-oil model typical of reservoir simulations.
 The three-dimensional model is integrated in reservoir volume and quantities such as fluid properties are spatially averaged.
-The material balance equation is obtained for the averaged quantities and it is applied to calculate reservoir pressure.
-It is shown that the calculated pressure corresponds to the local pressure at a defined depth and not to the average reservoir pressure in general.
+The material balance equation is obtained for averaged quantities and it is applied to calculate reservoir pressure.
+It is shown that the calculated pressure corresponds to the local pressure at an arbitrary reference depth and not to the average reservoir pressure in general.
 The definition of a convenient reference depth is discussed.
-Sufficient conditions for solution uniqueness of the material balance equation are provided for the case that pressure is the unknown quantity and one of such conditions is shown to check for physical correctness of PVT properties.
-Finally, a scheme for solving the material balance equation numerically is proposed.
+Sufficient conditions for solution uniqueness of the material balance equation are provided for the case that pressure is the unknown quantity.
+One of such conditions provides a check for physical correctness of PVT properties obtained from laboratory experiments or empirical correlations.
+Finally, a numerical scheme for solving the material balance equation is proposed.
 \end{abstract}
@@
 The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system.
-For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component can thus be defined.
-Similarly, a pseudo oil component can also be defined by subtracting all the pseudo gas component from oil phase composition.
-After defining these two pseudo-components, the number of transport equations needed for hydrocarbon components is reduced from a usually large number to only two.
+For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo-gas component can be defined.
+Similarly, a pseudo-oil component can be defined by subtracting all the pseudo-gas component from oil phase composition.
+After performing this procedure, the number of transport equations needed for hydrocarbon components is reduced from a usually large number to only two.
 
@@
 Oil phase is composed of oil and gas components, gas phase is composed uniquely by gas component and, water phase, uniquely by water component.
-The only allowed mass transfer is of gas component between oil and gas phases.
-Gas phase composition is always that of a dry gas.
+The only allowed mass transfer is of gas component between oil and gas phases, i.e., gas phase composition is always that of a dry gas.
 
@@
 
-Another simplification is now made assuming thermal equilibrium in the entire reservoir volume, i.e.,
+Another simplification is now made assuming thermal equilibrium in the reservoir volume, i.e.,
 \begin{equation}
@@
 
-Temperature is a constant prescribed value and thus no equation for energy balance is needed.
+Temperature is a constant prescribed value and no equation for energy balance is needed.
 It follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if the component number is restricted to the two pseudo-components.
@@
 \begin{equation}\label{eq: capillary_ow}
-p_{c,ow}\left(\vec{\vec{k}},\phi,S_w\right)=p_o-p_w \, ,
+p_{c,ow}=p_o-p_w \, ,
 \end{equation}
 \begin{equation}\label{eq: capillary_og}
-p_{c,go}\left(\vec{\vec{k}},\phi,S_g\right)=p_g-p_o \, .
+p_{c,go}=p_g-p_o \, .
 \end{equation}
@@
 
-All quantities are initially functions of space and time, but they become functions of time only after being averaged in space:
+All quantities are initially functions of space and time, but they become functions of time alone after being averaged in space:
 
@@
 
-We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes in fluid phases.
+We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes.
 In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to its volume at surface condition, i.e.:
@@
 
-\subsubsection{Relationship between the material balance equation and the reservoir pressure}
+\subsubsection{Relationship between the material balance equation and reservoir pressure}
 
@@
 To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$.
-This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as a consequence, pressure becomes a function of time and depth only given by the following expression:
+This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as a consequence, pressure becomes a function of time and depth given by the following expression:
 \begin{equation}\label{eq: p=rhogh}
@@
 
-It can be shown that neglecting the variation of fluid properties with depth does not introduce any important error.
+It can be shown that neglecting the variation of density with depth does not introduce any important error.
 As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount.
@@
 
-An open source implementation of secant method is available in SciPy library \citep{scipy}. This work used library version $0.12$.
+An open source implementation of the secant method is available in SciPy library \citep{scipy}. This work used library version $0.12$.
 
@@
 
-Figure~\ref{fig: epsilon} shows the shape of residual function ($\epsilon$) with respect to pressure, i.e., holding all produced or injected volumes constant and allowing only pressure to change.
-Each different solid line corresponds to a different volume of fluid production and injection.
+Figure~\ref{fig: epsilon} shows the shape of residual function ($\epsilon$) when pressure is varying and all remaining terms are held constant.
+Each different solid line corresponds to a different instant of time and therefore different volumes of cumulative production and injection.
 The function root is the solution of material balance equation, Eq.~\eqref{eq: MBE_short}, and the non-differentiable point is the bubble-point pressure.
@@
 
-Figure~\ref{fig: iter} shows the count of newtonian iterations required to compute pressure with an accuracy within $0.001$ of initial bubble-point pressure. Typically, $12$ iterations are required.
+Figure~\ref{fig: iter} shows the count of newtonian iterations required to compute pressure with an accuracy within $0.1\%$ of the initial bubble-point pressure.
+Typically, $12$ iterations are required.
 

### 2014-08-28 pituba/article/article_spe.tex
fixes

@@
 \begin{split}
-\frac{\partial}{\partial t} \left( \rho_o Y_g S_o + \rho_g S_g\right) + \nabla  \cdot \left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right) = 0\, .
+\frac{\partial}{\partial t} \left[\left( \rho_o Y_g S_o + \rho_g S_g\right)\phi \right] + \nabla  \cdot \left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right) = 0\, .
 \end{split}
@@
 \frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV &= -\oint_{\partial \Omega} \rho_o Y_o \left(\vec{u}_o \cdot \vec{dS} \right) \\
-&= \rho_o^{ST}Q_o^{ST} \, ;
+&= - \rho_o^{ST}Q_o^{ST} \, ;
 \end{split}
@@
 \begin{equation}\label{eq: So3}
-\frac{\partial}{\partial t} \left( \bar{\rho_o}\bar{Y}_o\bar{S}_o\bar{\phi} \Omega \right) = \rho_o^{ST}Q_o^{ST}
+\frac{\partial}{\partial t} \left( \bar{\rho_o}\bar{Y}_o\bar{S}_o\bar{\phi} \Omega \right) = - \rho_o^{ST}Q_o^{ST}
 \end{equation}
@@
 \begin{equation}\label{eq: So4}
-\frac{\partial}{\partial t} \left(\frac{\bar{S}_o V_p}{B_o} \right) = Q_o^{ST} \, ,
+\frac{\partial}{\partial t} \left(\frac{\bar{S}_o V_p}{B_o} \right) = - Q_o^{ST} \, ,
 \end{equation}
@@
 As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount.
-For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$; the same applies to the formation volume factor: $B_o \approx \rho_o (t, z_{ref})/ \rho_o^{ST}$.
+For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$.
+
+%the same applies to the formation volume factor: $B_o \approx \rho_o (t, z_{ref})/ \rho_o^{ST}$.
 

### 2014-09-22 pituba/article/article_spe.tex
dake practice

@@
 The material balance equation is a classical tool in petroleum reservoir engineering.
-It consists of the algebraic equation originally presented in \cite{schilthuis1936active} and conceived as a zero dimensional model for estimating reservoir pressure from mass balance.
+It consists of the algebraic equation originally presented in \cite{schilthuis1936active} as a zero dimensional model for estimating reservoir pressure from mass balance.
 \cite{schilthuis1936active} introduced the hypothesis of equilibrium attained at all times not necessarily in the entire reservoir volume, but in the regions of high permeability where most of the production would come from and where pressure would present a similar behavior.
@@
 The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system.
-For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo-gas component can be defined.
-Similarly, a pseudo-oil component can be defined by subtracting all the pseudo-gas component from oil phase composition.
-After performing this procedure, the number of transport equations needed for hydrocarbon components is reduced from a usually large number to only two.
+For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production so that a pseudo-gas component can be defined.
+Similarly, a pseudo-oil component can be defined by subtracting the pseudo-gas component from oil phase composition.
+After defining the pseudo-components, the number of transport equations needed for hydrocarbon components is reduced from a usually large number to only two.
 
@@
 
-A great number of applications of material balance equation have been reported in the literature, but few of them explore its connection with black-oil equations.
+%Many applications of the material balance equation have been reported in the literature, but few of them explore its connection with black-oil equations.
 \cite{ertekin2001basic} briefly addressed this subject arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \vec{g}$) were neglected and if PVT properties were averaged in reservoir volume.
 This work develops this process examining underlying details.
+
+Many works have reported applications of the material balance equation in the literature, but few have explored its connection with black-oil equations.
+\cite{ertekin2001basic} briefly addressed this subject arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \vec{g}$) were neglected and if PVT properties were averaged in reservoir volume.
+This work develops this concept examining underlying details.
 
@@
 fluid injection).
-The applicability, however, depends on how important non-equilibrium effects are to phase equilibrium.
+The applicability, however, requires negligible non-equilibrium effects in phase equilibrium.
 
@@
 
-After presenting material balance equation, sufficient conditions are shown for which material balance equation has only one solution for pressure.
-These conditions require fluids to be compressible and that gas vaporization occurs with reduction of oil phase volume.
+After presenting the material balance equation, sufficient conditions are shown for which material balance equation has only one solution for pressure.
+These conditions require fluids to be compressible and gas vaporization to occur with increase of oil phase density.
 A simple scheme for numerical solution is also proposed based on classical root-finding algorithms for non-linear equations.
@@
 
-This section presents black-oil equations, performs averaging of quantities in space and examines simplifying assumptions necessary to obtain the material balance equation.
+This section presents black-oil equations, performs the averaging of quantities in space and examines simplifying assumptions necessary to obtain the material balance equation.
 
@@
 
-We assume that the reservoir is a connected volume where hydrocarbon saturation is nonzero.
+We assume that the reservoir is a hydraulically connected volume where hydrocarbon saturation is nonzero.
+By a hidraulically connected volume, we mean that it is possible to observe a common pressure decline trend when some region of the reservoir is under prodution, as discussed in \cite{dakepractice}.
 Reservoir fluids are composed of oil and gas pseudo-components and water component.
@@
 \end{equation}
+Temperature is a constant prescribed value and no equation for energy balance is needed.
 
-Temperature is a constant prescribed value and no equation for energy balance is needed.
+%Temperature is a constant prescribed value and no equation for energy balance is needed.
 It follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if the component number is restricted to the two pseudo-components.
@@
 
+It follows from Duhem's theorem that the equilibrium of oil and gas phases is determined if, in addition to the mass balance equations, two properties are prescribed.
+From Gibbs' phase rule, the system has two degrees of freedom.
+It is then necessary to specify two intensive properties.
+The constant temperature is one and pressure is chosen as the other.
+Oil phase density now becomes a function of pressure alone.
+The same is valid for mass fractions of gas and oil components in oil phase:
 \begin{equation}
@@
 
-Derivation of the material balance equation consists in integrating black-oil equations in reservoir volume.
+Derivation of the material balance equation consists in integrating black-oil equations in reservoir volume ($\Omega$).
 

### 2014-09-22 pituba/article/article_spe.tex
fixed compaction of porous volume

@@
 water volume expansion & $E_w = B_{w2}-B_{w1}$\\
-porous volume contraction & $C=V_{p2}- V_{p1}$\\
+porous volume contraction & $C=V_{p1}- V_{p2}$\\
 bubble-point pressure & $R_{s2}\left(P_b\right) = \left[N\left(m B_{1}/B_{g1} +R_{s1}\right)+G_i-G_p\right]/\left(N-N_p\right)$ \\ \bottomrule

### 2014-09-22 pituba/article/article_spe.tex
fixed compaction of porous volume

@@
 %\begin{equation}\label{eq: MBE}
-%F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{porous volume contraction } (C)}
+%F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{porous volume compaction } (C)}
 %\end{equation}
@@
 water volume expansion & $E_w = B_{w2}-B_{w1}$\\
-porous volume contraction & $C=V_{p1}- V_{p2}$\\
+porous volume compaction & $C=V_{p1}- V_{p2}$\\
 bubble-point pressure & $R_{s2}\left(P_b\right) = \left[N\left(m B_{1}/B_{g1} +R_{s1}\right)+G_i-G_p\right]/\left(N-N_p\right)$ \\ \bottomrule

### 2014-09-24 pituba/article/article_spe.tex
new changes

@@
+% This work is licensed under a Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License.
+
 % Review
@@
 %\author{R.B. Piccinini}\ead{rbpiccinini@gmail.com}
-%\address{Petrobras - UO-BA/ATP-N/RES. Salvador, Bahia, Brazil.}
 
@@
 
-Spatially averaged quantities are defined for porosity ($\phi$), saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration ($Y_{\alpha,k}$) of a component $k$ in phase $\alpha$.
+Spatially averaged quantities are defined for porosity ($\phi$), phase saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration ($Y_{\alpha,k}$) of a component $k$ in phase $\alpha$.
 
@@
 \begin{equation}\label{eq: MBE}
-F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{pore volume reduction } (C)}
+F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{pore volume compaction} (C)}
 \end{equation}
@@
 
-Assuming reservoir temperature to be sufficiently above gas critical temperature for negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
+Assuming reservoir temperature to be sufficiently above gas critical temperature for a negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
 \begin{equation}\label{eq: gradgdz}
@@
 The reference depth for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume.
-Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$, where $z^0_\alpha$ is the depth of phase gravity center and $\alpha=w, o, g$.
-The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$, without fluid motion, and only equals the average reservoir pressure $\bar{p}_\alpha$ if the reference depth is that of the gravity center, $z^{ref}=z^0$.
+Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$, where $z^0_\alpha$ is the depth of phase gravity center for $\alpha=w, o, g$.
+The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$ for negligible hydrodynamic gradients and only equals the average reservoir pressure $\bar{p}_\alpha$ if the reference depth is that of the gravity center, $z^{ref}=z^0$.
 
 For the case of an oil reservoir with a gas cap, the reference depth should be defined at the gas-oil contact.
-The point here is that at gas-oil contact both gas and oil densities are computed using the same pressure, its local value.
+The point here is that at gas-oil contact both gas and oil densities are computed using the same pressure its local value.
 If the reference depth, common to all phases, is otherwise defined below gas-oil contact, gas density will increase with depth at a higher rate because of the oil hydrostatic gradient,

### 2014-09-24 pituba/article/build.sh

@@ -1,0 +1,10 @@
+#!/bin/bash
+
+pdflatex article_spe.tex
+bibtex article_spe
+pdflatex article_spe.tex
+pdflatex article_spe.tex
+
+evince article_spe.pdf &
+exit 1
+

### 2014-09-25 pituba/article/article_spe.tex
news

@@
 The definition of a convenient reference depth is discussed.
-Sufficient conditions for solution uniqueness of the material balance equation are provided for the case that pressure is the unknown quantity.
-One of such conditions provides a check for physical correctness of PVT properties obtained from laboratory experiments or empirical correlations.
+It is found that the physical correctness of PVT properties obtained from laboratory experiments or empirical correlations is a sufficient condition for solution uniqueness of the material balance equation when pressure is the unknown quantity.
 Finally, a numerical scheme for solving the material balance equation is proposed.
@@
 It consists of the algebraic equation originally presented in \cite{schilthuis1936active} as a zero dimensional model for estimating reservoir pressure from mass balance.
-\cite{schilthuis1936active} introduced the hypothesis of equilibrium attained at all times not necessarily in the entire reservoir volume, but in the regions of high permeability where most of the production would come from and where pressure would present a similar behavior.
+\cite{schilthuis1936active} introduced the hypothesis of equilibrium attained at all times not necessarily in the entire reservoir volume, but in the regions of high permeability where most of the production would come from and where pressure would present a similar trend.
 Such regions were said to contain an active oil.
@@
 
- The same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}, is also applied here to material balance.
+ The same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}, is also applied here.
 This approach extends the material balance equation to saturated reservoirs under pressure maintenance (e.g.

### 2014-09-29 pituba/article/article_spe.tex
new changes

@@
 % Review
-\documentclass[authoryear,preprint,review,11pt]{elsarticle}
-%\documentclass[final,authoryear,5p,twocolumn,10pt]{elsarticle}
+%\documentclass[authoryear,preprint,review,11pt]{elsarticle}
+\documentclass[final,authoryear,5p,twocolumn,10pt]{elsarticle}
 
@@
 This work examines underlying details in obtaining the classical material balance equation from a three dimensional black-oil model typical of reservoir simulations.
-The three-dimensional model is integrated in reservoir volume and quantities such as fluid properties are spatially averaged.
+The black-oil equations are integrated in reservoir volume and quantities such as fluid properties are spatially averaged.
 The material balance equation is obtained for averaged quantities and it is applied to calculate reservoir pressure.
 It is shown that the calculated pressure corresponds to the local pressure at an arbitrary reference depth and not to the average reservoir pressure in general.
-The definition of a convenient reference depth is discussed.
-It is found that the physical correctness of PVT properties obtained from laboratory experiments or empirical correlations is a sufficient condition for solution uniqueness of the material balance equation when pressure is the unknown quantity.
-Finally, a numerical scheme for solving the material balance equation is proposed.
+The definition of a convenient reference depth is also discussed.
+It is found that the physical correctness of PVT properties
+% obtained from laboratory experiments or empirical correlations
+is a sufficient condition for solution uniqueness of the material balance equation when pressure is the unknown quantity.
+Finally, a numerical scheme for solving the material balance equation is proposed based on nonlinear root finding algorithms.
 \end{abstract}
@@
 
-It follows from Duhem's theorem that the equilibrium of oil and gas phases is determined if, in addition to the mass balance equations, two properties are prescribed.
+It follows from Duhem's theorem that the equilibrium of oil and gas phases is determined if, in addition to the mass balance equations, two other properties are prescribed.
 From Gibbs' phase rule, the system has two degrees of freedom.
@@
 
-Eq~\eqref{eq: MBE} involves fluid properties, original fluid volumes ($N$, $W$ and $G)$, cumulative production and injection and aquifer influx.
+Eq~\eqref{eq: MBE} involves fluid properties, original phase volumes ($N$, $W$ and $G)$, cumulative production and injection and aquifer influx.
 It is an algebraic equation that can be solved for any one of its terms, provided that all remaining are known. This works focuses on its solution for pressure.
@@
 The material balance equation is widely used to relate production and injection to reservoir pressure.
-Actually, all averaged fluid properties in Eq.~\eqref{eq: MBE} depend on the pressure field, and so does the pore volume.
+Actually, all averaged phase properties in Eq.~\eqref{eq: MBE} depend on the pressure field, and so does the pore volume.
 If the reservoir is under production, though, the pressure field is not known and for its prediction one would have to resort to more advanced techniques such as reservoir simulation.
@@
 
-It can be shown that neglecting the variation of density with depth does not introduce any important error.
+It can be shown that neglecting the variation of density with depth does not introduce  important error.
 As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount.
@@
 
-Assuming reservoir temperature to be sufficiently above gas critical temperature for a negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
+Assuming reservoir temperature to be sufficiently above gas critical temperature and thus a negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
 \begin{equation}\label{eq: gradgdz}
@@
 For the case of an oil reservoir with a gas cap, the reference depth should be defined at the gas-oil contact.
-The point here is that at gas-oil contact both gas and oil densities are computed using the same pressure its local value.
-If the reference depth, common to all phases, is otherwise defined below gas-oil contact, gas density will increase with depth at a higher rate because of the oil hydrostatic gradient,
+The point here is that at gas-oil contact both gas and oil densities are computed using the pressure local value.
+If the reference depth, common to all phases, is otherwise defined below gas-oil contact, a higher gas density will be computed because of the oil hydrostatic gradient:
 \begin{equation}
@@
 
-It has been shown that the pressure variation with depth is not important for the computation of gas, oil and water densities.
+It was shown that the pressure variation with depth is not important for the computation of gas, oil and water densities.
 Nevertheless, pressure also plays a second role in the material balance equation by defining phase equilibrium, for what the pressure variation with depth may not be negligible in the solubility ratio:
@@
 \begin{equation}\label{eq: PVTH1}
-\left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right) \leq 0 \, ,
+\left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right) < 0 \, ,
 \end{equation}
 \begin{equation}\label{eq: PVTH2}
-\frac{dB_g}{dp} \leq 0 \, ,
+\frac{dB_g}{dp} < 0 \, ,
 \end{equation}
 \begin{equation}\label{eq: PVTH3}
-\frac{dB_w}{dp} \leq 0 \, ,
+\frac{dB_w}{dp} < 0 \, ,
 \end{equation}
 \begin{equation}\label{eq: PVTH4}
-\frac{dV_p}{dp} \geq 0 \, .
+\frac{dV_p}{dp} > 0 \, .
 \end{equation}
@@
 \begin{equation}\label{eq: vaporization}
-\frac{1}{\rho_o}\frac{d\rho_o}{dY_o} \geq Y_o \left( \frac{\rho_o}{\rho_g} - 1 \right) \, .
+\frac{1}{\rho_o}\frac{d\rho_o}{dY_o} > Y_o \left( \frac{\rho_o}{\rho_g} - 1 \right) \, .
 \end{equation}
@@
 
-An open source implementation of the secant method is available in SciPy library \citep{scipy}. This work used library version $0.12$.
+An open source implementation of the secant method is available in SciPy library \citep{scipy}. This work used the library version $0.12$.
 
@@
 \section{Results}
-Material balance equation were solved for a black-oil reservoir initially producing in depletion drive mechanism and later subjected to water injection.
+For illustrating the numerical procedure, the material balance equation was solved for a fictitious black-oil reservoir initially producing in a pure depletion drive mechanism and later subjected to water injection.
 Fig.~\ref{fig: matbal_p} shows in the solid line curve the severe depletion during initial phases of production.
@@
 
-Water injection starts when oil recovery factor $(RF)$ is about $RF=0.08$ and it is maintained until the reservoir reaches a novel undersaturated state.
+Water injection starts when oil recovery factor $(RF)$ is $RF=0.08$ and it is maintained until the reservoir reaches a novel undersaturated state.
 Initially, the high compressibility of gas prevents pressure from raising and free gas phase is reduced because of redissolution.
@@
 Figure~\ref{fig: epsilon} shows the shape of residual function ($\epsilon$) when pressure is varying and all remaining terms are held constant.
-Each different solid line corresponds to a different instant of time and therefore different volumes of cumulative production and injection.
+Each different solid line corresponds to a different instant of time and, therefore, to different volumes of cumulative production and injection.
 The function root is the solution of material balance equation, Eq.~\eqref{eq: MBE_short}, and the non-differentiable point is the bubble-point pressure.
@@
 
-\section*{Highlights}
-\begin{enumerate}[1.]
-\item The existing link between black-oil and material balance equation is examined in detail.
-\item It is shown that gravity and capillary forces must be neglected and saturations must be constant in space so one can integrate black-oil equations in reservoir volume and obtain material balance equation.
-\item Material balance equation are shown to have a unique solution for a compressible system.
-\item Material balance equation are formulated for a variable bubble-point and they are used to a reservoir under pressure maintenance.
-\end{enumerate}
+%\section*{Highlights}
+%\begin{enumerate}[1.]
+%\item The existing link between black-oil and material balance equation is examined in detail.
+%\item It is shown that gravity and capillary forces must be neglected and saturations must be constant in space so one can integrate black-oil equations in reservoir volume and obtain material balance equation.
+%\item Material balance equation are shown to have a unique solution for a compressible system.
+%\item Material balance equation are formulated for a variable bubble-point and they are used to a reservoir under pressure maintenance.
+%\end{enumerate}
 

### 2014-10-07 pituba/article/article_spe.tex
petro changes

@@
 % Review
-%\documentclass[authoryear,preprint,review,11pt]{elsarticle}
-\documentclass[final,authoryear,5p,twocolumn,10pt]{elsarticle}
+\documentclass[authoryear,preprint,review,11pt]{elsarticle}
+%\documentclass[final,authoryear,5p,twocolumn,10pt]{elsarticle}
 
@@
 
-After presenting the material balance equation, sufficient conditions are shown for which material balance equation has only one solution for pressure.
+After presenting the material balance equation, sufficient conditions are shown for which the material balance equation has only one solution for pressure.
 These conditions require fluids to be compressible and gas vaporization to occur with increase of oil phase density.
@@
 \begin{equation}
-\rho_w = \rho_w \left(p_w, T_w\right)
+\rho_w = \rho_w \left(p_w, T_w\right) \, ,
 \end{equation}
 \begin{equation}
-\rho_o = \rho_o \left(p_o, T_o,Y_g\right)
+\rho_o = \rho_o \left(p_o, T_o,Y_g\right) \, ,
 \end{equation}
 \begin{equation}
-\rho_g = \rho_g \left(p_g, T_g\right)
+\rho_g = \rho_g \left(p_g, T_g\right) \, .
 \end{equation}

### 2014-11-03 pituba/article/article_spe.tex
corrections

@@
 
-In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between the volume of dissolved gas and the volume of oil both measured in surface condition:
+In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between the volume of dissolved gas and the volume of oil both measured at surface condition:
 
@@
 
-Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at surface condition and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil phases measured in reservoir condition.
+Where $N=V_{p1} \bar{S}_{o1}/B_{o1}$ and $W=V_{p1} \bar{S}_{w1}/B_{w1}$ are the initial volumes of oil and water measured at surface condition and $m=V_{p1} \bar{S}_{g1}/B_{o1} N$ is the ratio of the initial volumes of gas and oil phases measured at reservoir condition.
 
@@
 
-It is now possible to obtain the so-called generalized material balance equation by substituting the equations for phase saturation, Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5}, into Eq.~\eqref{eq: S15}:
+It is now possible to obtain the generalized material balance equation by substituting the equations for phase saturation, Eqs.~\eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5}, into Eq.~\eqref{eq: S15}:
 
@@
 
-It can be shown that neglecting the variation of density with depth does not introduce  important error.
+It can be shown that neglecting the variation of density with depth does not introduce  important errors.
 As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount.
@@
 
-For investigating the order of magnitude of Eqs.~\eqref{eq: gradodz} and \eqref{eq: gradgdz}, we substitute into them the values of $\rho_o = 10^3\ kg/m^3$ and $c_o = 10^{-9}\ Pa^{-1}$ for the oil phase, and $\mathcal{M}=19\ kg/kmol$ and $T=400\ K$ for the gas phase:
+To investigate the order of magnitude of Eqs.~\eqref{eq: gradodz} and \eqref{eq: gradgdz}, we substitute into them the values of $\rho_o = 10^3\ kg/m^3$ and $c_o = 10^{-9}\ Pa^{-1}$ for the oil phase, and $\mathcal{M}=19\ kg/kmol$ and $T=400\ K$ for the gas phase:
 \begin{equation}
@@
 \section{Results}
-For illustrating the numerical procedure, the material balance equation was solved for a fictitious black-oil reservoir initially producing in a pure depletion drive mechanism and later subjected to water injection.
+To illustrate the numerical procedure, the material balance equation was solved for a fictitious black-oil reservoir initially producing in a pure depletion drive mechanism and later subjected to water injection.
 Fig.~\ref{fig: matbal_p} shows in the solid line curve the severe depletion during initial phases of production.
@@
 $m$ & ratio of original free gas volume \\
-& to original oil volume in reservoir condition \\
+& to original oil volume at reservoir condition \\
 $N$ & original volume of oil in place \\

### 2014-11-17 pituba/article/article_spe.tex
datum level

@@
 It is shown that spatially averaged PVT properties are required for the material balance approach in general, but that for many cases the averaging is simply negligible for practical purposes.
-An interpretation for pressure in material balance equation is provided and its relation to a reference depth is examined.
+An interpretation for pressure in material balance equation is provided and its relation to a datum level is examined.
 The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented.
@@
 The material balance equation is obtained for averaged quantities and it is applied to calculate reservoir pressure.
-It is shown that the calculated pressure corresponds to the local pressure at an arbitrary reference depth and not to the average reservoir pressure in general.
-The definition of a convenient reference depth is also discussed.
+It is shown that the calculated pressure corresponds to the local pressure at the datum level.
+The definition of a convenient datum level is also discussed.
 It is found that the physical correctness of PVT properties
@@
 We assume that the reservoir is a hydraulically connected volume where hydrocarbon saturation is nonzero.
-By a hidraulically connected volume, we mean that it is possible to observe a common pressure decline trend when some region of the reservoir is under prodution, as discussed in \cite{dakepractice}.
+By a hidraulically connected volume, we mean that it is possible to observe a common pressure decline trend when any reservoir portion is under production \citep{dakepractice}.
 Reservoir fluids are composed of oil and gas pseudo-components and water component.
@@
 Oil phase is composed of oil and gas components, gas phase is composed uniquely by gas component and, water phase, uniquely by water component.
-The only allowed mass transfer is of gas component between oil and gas phases, i.e., gas phase composition is always that of a dry gas.
+The only allowed mass transfer is of gas component between oil and gas phases, i.e., the gas phase composition is always that of a dry gas.
 
@@
 
-Phase densities are functions of the local thermodynamic state:
-\begin{equation}
-\rho_w = \rho_w \left(p_w, T_w\right) \, ,
-\end{equation}
-\begin{equation}
-\rho_o = \rho_o \left(p_o, T_o,Y_g\right) \, ,
-\end{equation}
-\begin{equation}
-\rho_g = \rho_g \left(p_g, T_g\right) \, .
-\end{equation}
+%Phase densities are functions of the local thermodynamic state:
+%\begin{equation}
+%\rho_w = \rho_w \left(p_w, T_w\right) \, ,
+%\end{equation}
+%\begin{equation}
+%\rho_o = \rho_o \left(p_o, T_o,Y_g\right) \, ,
+%\end{equation}
+%\begin{equation}
+%\rho_g = \rho_g \left(p_g, T_g\right) \, .
+%\end{equation}
 
-Another simplification is now made assuming thermal equilibrium in the reservoir volume, i.e.,
+Another simplification is now assuming thermal equilibrium in the reservoir volume, i.e.,
 \begin{equation}
@@
 From Gibbs' phase rule, the system has two degrees of freedom.
-It is then necessary to specify two intensive properties.
-The constant temperature is one and pressure is chosen as the other.
+It is then necessary to specify two intensive properties: the constant temperature is one and the local pressure is chosen as the other.
 Oil phase density now becomes a function of pressure alone.
@@
 \begin{equation}
-Y_g = Y_g (p_o) \, ,
-\end{equation}
-\begin{equation}
-Y_o = Y_o (p_o) \, .
+\rho_o = \rho_o \left(p_o\right) \, , \quad Y_g = Y_g (p_o) \, , \quad Y_o = Y_o (p_o) \, .
 \end{equation}
@@
 \end{equation}
-where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha^{ref} (t)$ is the pressure at some reference depth $z_{ref}$.
+where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha^{ref} (t)$ is the pressure at the datum level $z_{ref}$.
 It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.
@@
 %
-%We shall see that to compute the averaged phase density, we may define some reference depth and use the corresponding local density as an approximation to the averaged density without introducing any important error.
+%We shall see that to compute the averaged phase density, we may define some datum level and use the corresponding local density as an approximation to the averaged density without introducing any important error.
 
@@
 Oil and gas densities are thus approximately constant with depth.
-The reference depth for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume.
+The datum level for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume.
 Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$, where $z^0_\alpha$ is the depth of phase gravity center for $\alpha=w, o, g$.
-The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$ for negligible hydrodynamic gradients and only equals the average reservoir pressure $\bar{p}_\alpha$ if the reference depth is that of the gravity center, $z^{ref}=z^0$.
+The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$ for negligible hydrodynamic gradients and only equals the average reservoir pressure $\bar{p}_\alpha$ if the datum level is that of the gravity center, $z^{ref}=z^0$.
 
-For the case of an oil reservoir with a gas cap, the reference depth should be defined at the gas-oil contact.
+For the case of an oil reservoir with a gas cap, the datum level can be defined at the gas-oil contact.
 The point here is that at gas-oil contact both gas and oil densities are computed using the pressure local value.
-If the reference depth, common to all phases, is otherwise defined below gas-oil contact, a higher gas density will be computed because of the oil hydrostatic gradient:
+If the datum level, common to all phases, is otherwise defined below gas-oil contact, a higher gas density will be computed because of the oil hydrostatic gradient:
 \begin{equation}

### 2014-11-24 pituba/article/article_spe.tex
go on

@@
 \begin{equation}\label{eq: p=rhogh}
-\int_{p_\alpha^{ref} (t)}^{p_\alpha (t,z)} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - z_{ref}\right) \, ,
+\int_{p_\alpha (t,z_{datum})}^{p_\alpha (t,z)} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - z_{datum}\right) \, ,
 \end{equation}
-where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha^{ref} (t)$ is the pressure at the datum level $z_{ref}$.
+where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha (t,z_{datum})$ is the pressure at the datum level $z_{datum}$.
 It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.
@@
 As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount.
-For instance, the average oil density can be approximated to the local density at the reference level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{ref})$.
+For instance, the average oil density can be approximated to the local density at the datum level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{datum})$.
 
-%the same applies to the formation volume factor: $B_o \approx \rho_o (t, z_{ref})/ \rho_o^{ST}$.
+%the same applies to the formation volume factor: $B_o \approx \rho_o (t, z_{datum})/ \rho_o^{ST}$.
 
@@
 Oil and gas densities are thus approximately constant with depth.
-The datum level for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume.
-Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$, where $z^0_\alpha$ is the depth of phase gravity center for $\alpha=w, o, g$.
-The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$ for negligible hydrodynamic gradients and only equals the average reservoir pressure $\bar{p}_\alpha$ if the datum level is that of the gravity center, $z^{ref}=z^0$.
+The datum level for computing phase density (or the formation volume factor) can be any arbitrary depth inside phase volume.
+The pressure calculated in the material balance equation corresponds to the local pressure at $z_{datum}$ for negligible hydrodynamic gradients.
+
+%Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$, where $z^0_\alpha$ is the depth of phase gravity center for $\alpha=w, o, g$.
 

### 2015-01-23 pituba/article/article_spe.tex
news

@@
 
-The equation developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium.
+The equation developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, chemical or thermal equilibrium.
 An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}.
@@
 fluid injection).
-The applicability, however, requires negligible non-equilibrium effects in phase equilibrium.
+The applicability, however, requires negligible non-equilibrium effects in chemical equilibrium.
 
@@
 The material balance equation is widely used to relate production and injection to reservoir pressure.
-Actually, all averaged phase properties in Eq.~\eqref{eq: MBE} depend on the pressure field, and so does the pore volume.
+Actually, all averaged phase properties in Eq.~\eqref{eq: MBE} depend on the pressure field.
 If the reservoir is under production, though, the pressure field is not known and for its prediction one would have to resort to more advanced techniques such as reservoir simulation.
@@
 
-To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$.
-This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as a consequence, pressure becomes a function of time and depth given by the following expression:
-\begin{equation}\label{eq: p=rhogh}
-\int_{p_\alpha (t,z_{datum})}^{p_\alpha (t,z)} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - z_{datum}\right) \, ,
+To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume mechanical equilibrium, i.e., $\vec{u}_\alpha = 0$.
+This implies that pressure at some time instant is only a function of depth: $\nabla p = \partial p_\alpha/\partial z = \rho_\alpha \vec{g}$.
+
+% so that pressure is a function of time and depth given by the following expression:
+%\begin{equation}\label{eq: p=rhogh}
+%\int_{p_\alpha (t,z_{datum})}^{p_\alpha (t,z)} %\frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - %z_{datum}\right) \, ,
+%\end{equation}
+
+%where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha (t,z_{datum})$ is the pressure at the datum level $z_{datum}$.
+If the pressure at the datum level is known, it is possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.
+
+It will be shown that the pressure field can be further simplified by neglecting the variation of density with depth.
+For instance, the average oil density can be approximated to the local density at the datum level, $\rho_o (t,z) \approx \rho_o (t, z_{datum})$, what results in the following simple expression for the pressure field:
+\begin{equation}
+p_\alpha (t,z) = p_\alpha \left(t, z_{datum}\right)+ \rho_\alpha (t, z_{datum}) g \left(z-z_{datum}\right) \, .
 \end{equation}
-where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha (t,z_{datum})$ is the pressure at the datum level $z_{datum}$.
-It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.
-
-It can be shown that neglecting the variation of density with depth does not introduce  important errors.
-As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount.
-For instance, the average oil density can be approximated to the local density at the datum level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{datum})$.
 
@@
 
-This approximation can be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase,
+ As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount.
+This approximation can be quantified by putting together the expressions for compressibility and hydrostatic equilibrium.
+For this, we assume a constant isothermal compressibility for oil phase,
 \begin{equation}\label{eq: grado}
@@
 
-Assuming reservoir temperature to be sufficiently above gas critical temperature and thus a negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
+Assuming reservoir temperature to be sufficiently above gas phase critical temperature and thus a negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
 \begin{equation}\label{eq: gradgdz}
@@
 
-Oil and gas densities are thus approximately constant with depth.
-The datum level for computing phase density (or the formation volume factor) can be any arbitrary depth inside phase volume.
-The pressure calculated in the material balance equation corresponds to the local pressure at $z_{datum}$ for negligible hydrodynamic gradients.
+Oil and gas densities are thus approximately constant with depth and their average values can be approximated to the local values at the datum level.
+The pressure calculated in the material balance equation thus corresponds to the local pressure at $z_{datum}$.
 
@@
 
-For the case of an oil reservoir with a gas cap, the datum level can be defined at the gas-oil contact.
-The point here is that at gas-oil contact both gas and oil densities are computed using the pressure local value.
-If the datum level, common to all phases, is otherwise defined below gas-oil contact, a higher gas density will be computed because of the oil hydrostatic gradient:
+For the case of an oil reservoir with a gas cap, it is more accurate to define the datum level at the gas-oil contact.
+The point is that a higher gas density will be computed for a datum level bellow gas-oil contact because of the higher hydrostatic gradient of oil phase,
 \begin{equation}
@@
 
-It was shown that the pressure variation with depth is not important for the computation of gas, oil and water densities.
-Nevertheless, pressure also plays a second role in the material balance equation by defining phase equilibrium, for what the pressure variation with depth may not be negligible in the solubility ratio:
+It was shown that the pressure variation with depth can be neglected for the computation of gas, oil and water densities. This is true provided that phase changes do not occur.
+
+ for what the pressure variation with depth may not be negligible in the solubility ratio: 
 \begin{equation}\label{eq: dRsdz1}
@@
 \section{Solution of material balance equation}
-A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We shall start defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
+A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We start by defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
 \begin{equation}\label{eq: residual}
@@
 
-The residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
+The residual function clearly satisfies the first two affirmatives because phase properties are continuous everywhere and are also differentiable everywhere except for the phase change at bubble-point pressure.
 The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels, each corresponding to a different phase (three fluid phases and the solid rock).
@@
 \textit{In-situ} volumes (e.g.
-$N-N_p$ for oil phase) are non-negative values.
-Pressure derivatives of fluid volumes are non-positive values (fluid volume cannot increase as pressure increases) and pressure derivative of porous volume is a positive value (as fluid pressure increases, porous volume expands owing to rock compression).
+$N-N_p$ for oil phase) are non-negative values,  pressure derivatives of fluid volumes are non-positive values (fluid volume cannot increase as pressure increases) and the pressure derivative of porous volume is a positive value (as fluid pressure increases, porous volume expands owing to rock compression).
 
-The value of $d\varepsilon/dp$ is thus negative if at least one phase is compressible because it would be a sum of products between negative and positive factors.
+The value of $d\varepsilon/dp$ is thus negative if at least one phase is compressible because it is a sum of products between negative and positive factors.
 
@@
 
-It must be observed that the result of gas reentering oil phase completely neglects non-equilibrium effects, e.g.
-compositional gradients, and must be used with care.
-For instance, it may be in strong disagreement with real reservoir behavior when gravity segregation is present.
+It must be observed that the result of gas reentering oil phase completely neglects non-equilibrium effects, e.g. compositional gradients and diffusion of matter, and must be used with care. .
 

### 2015-01-23 pituba/article/article_spe.tex
news+

@@
 For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production so that a pseudo-gas component can be defined.
-Similarly, a pseudo-oil component can be defined by subtracting the pseudo-gas component from oil phase composition.
+Similarly, a pseudo-oil component can be defined by subtraction of the pseudo-gas component from oil phase composition.
 After defining the pseudo-components, the number of transport equations needed for hydrocarbon components is reduced from a usually large number to only two.
@@
 fluid injection).
-The applicability, however, requires negligible non-equilibrium effects in chemical equilibrium.
+The applicability, however, requires negligible non-equilibrium effects.
 
@@
 Reservoir fluids are composed of oil and gas pseudo-components and water component.
-There can be three fluid phases: oil, gas and (liquid) water.
+There can be three fluid phases at most: oil, gas and (liquid) water.
 Oil phase is composed of oil and gas components, gas phase is composed uniquely by gas component and, water phase, uniquely by water component.
-The only allowed mass transfer is of gas component between oil and gas phases, i.e., the gas phase composition is always that of a dry gas.
+The only allowed mass transfer is of gas component between oil and gas phases.
+The gas phase composition is always that of a dry gas.
 
@@
 
-Another simplification is now assuming thermal equilibrium in the reservoir volume, i.e.,
+Another simplification is now made assuming thermal equilibrium in the reservoir volume, i.e.,
 \begin{equation}
@@
 It is then necessary to specify two intensive properties: the constant temperature is one and the local pressure is chosen as the other.
-Oil phase density now becomes a function of pressure alone.
+Since temperature was assumed constant, oil phase density becomes a function of pressure alone.
 The same is valid for mass fractions of gas and oil components in oil phase:
@@
 
-Integrating Eqs.~\eqref{eq: Sw1}, \eqref{eq: So1} and \eqref{eq: Sg1} in volume and applying divergence theoreom gives:
+Integrating Eqs.~\eqref{eq: Sw1}, \eqref{eq: So1} and \eqref{eq: Sg1} in volume and applying divergence theorem gives:
 \begin{itemize}
@@
 We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes.
-In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to its volume at surface condition, i.e.:
+In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to its volume at standard surface condition, i.e.:
 
@@
 
-or using $\bar{Y}_o + \bar{Y}_g = 1$:
+or, using $\bar{Y}_o + \bar{Y}_g = 1$:
 \begin{equation}
@@
 
-An equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, or $R_s(p=P_b)$:
+An equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, $R_s(p=P_b)$:
 \begin{equation}\label{eq: eqPb}
@@
 \end{equation}
-We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that, as a consequence, only one root of $\varepsilon$ exists at most.
+We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that only one root of $\varepsilon$ exists at most.
 Equivalently, we can say that only one solution exists for the material balance equation.
@@
 
-The residual function clearly satisfies the first two affirmatives because phase properties are continuous everywhere and are also differentiable everywhere except for the phase change at bubble-point pressure.
-The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels, each corresponding to a different phase (three fluid phases and the solid rock).
-For the fluid phases, each parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
-\begin{equation}
+The residual function clearly satisfies the first two affirmatives because phase properties are continuous everywhere and are also differentiable everywhere except  at bubble-point pressure.
+
+The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four negative terms:
+\begin{equation}\label{eq: deps}
 \begin{split}
@@
 
-\textit{In-situ} volumes (e.g.
-$N-N_p$ for oil phase) are non-negative values,  pressure derivatives of fluid volumes are non-positive values (fluid volume cannot increase as pressure increases) and the pressure derivative of porous volume is a positive value (as fluid pressure increases, porous volume expands owing to rock compression).
-
-The value of $d\varepsilon/dp$ is thus negative if at least one phase is compressible because it is a sum of products between negative and positive factors.
+Each term in Eq.
+\eqref{eq: deps} corresponds to a different phase (three fluid phases and the solid rock).
+For the fluid phases, each term is the product of the current volume \textit{in-situ} (a positive quantity) and a volume derivative with respect to pressure (a negative quantity).
+Since pore volume expands with an increase of fluid pressure, $dV_p/dp$ is positive.
+The value of $d\varepsilon/dp$ is thus necessarily negative.
 

### 2015-03-09 pituba/article/article_spe.tex
artigo

@@
 
-To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume mechanical equilibrium, i.e., $\vec{u}_\alpha = 0$.
+To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume mechanical equilibrium: $\vec{u}_\alpha = 0$.
 This implies that pressure at some time instant is only a function of depth: $\nabla p = \partial p_\alpha/\partial z = \rho_\alpha \vec{g}$.
@@
 
-It will be shown that the pressure field can be further simplified by neglecting the variation of density with depth.
-For instance, the average oil density can be approximated to the local density at the datum level, $\rho_o (t,z) \approx \rho_o (t, z_{datum})$, what results in the following simple expression for the pressure field:
+It will be shown that the pressure field can be further simplified by neglecting the variation of density with depth, what allows us to express the pressure field in the following simple expression:
 \begin{equation}
@@
 Several methods exist to find roots of nonlinear continuous functions \citep{hamming1987numerical}.
-One possible choice is Newton's method, which linearizes the function using either its gradient or a secant line.
-The secant line approach introduces a numerical smoothness to the gradient near continuous but non-differentiable points, for what it is the approach used here to deal with the bubble-point pressure.
-
-An open source implementation of the secant method is available in SciPy library \citep{scipy}. This work used the library version $0.12$.
+One possible choice is Newton's method with a switch from the tangent to the secant line near the non-differentiable bubble-point pressure.
+Both methods are available in the open source SciPy library \citep{scipy}.
+This work used the library version $0.12$.
 

### 2015-03-09 pituba/reservoir.py

@@ -19,15 +19,15 @@ fleft=0.15
 
 ## PVT
 fluido=classPVT()
-fluido.pe=array([200.,170.,140.,110.,80.,60.,48.,35.,25.,15.,1.033])
-fluido.Boe=array([1.165,1.149,1.133,1.116,1.100,1.089,1.083,1.076,1.069,1.062,1.0000])
-fluido.Rse=array([66.9,57.8,48.6,39.4,30.2,24.1,20.4,16.4,13.1,9.4,0.0000])
-fluido.Bge=array([0.0047,0.055,0.0067,0.0085,0.0117,0.0156,0.0195,0.0310,0.0435,0.0720,1.0000])
-fluido.Bwe=1.0*ones(len(fluido.pe))
+fluido.pe=array([250.,170.,139.8,109.5,79.3,49.,41.,33.,25.,17.,9.,1.])
+fluido.Boe=array([1.15,1.13,1.12,1.11,1.098,1.083,1.079,1.074,1.068,1.062,1.052,1.035])
+fluido.Rse=array([60.,47.,41.,35.,28.,20.4,18.,15.5,13.,10.,5.,0.21])
+fluido.Bge=array([0.004,0.0057,0.007,0.0093,0.0136,0.022,0.028,0.036,0.049,0.0738,0.1433,1.2443])
+fluido.Bwe=1.02379*(1-3.45e-5*(fluido.pe - 97.45))
 fluido.viscoe=array([5.81000,7.23000,8.05000,9.05000,10.23000,11.58000,12.17000,12.80000,14.12000,14.83000,15.57000,16.33000,17.12000,17.96000,18])
 fluido.viscwe=1.0*ones(len(fluido.pe))
-fluido.co=142e-6
-fluido.Pb=60.
+fluido.co=1.05e-4
+fluido.Pb=48.
 fluido.T=273.+57.
 
 #---------------------------------------------------------------------------
@@ -35,36 +35,33 @@ fluido.T=273.+57.
 # Np*(Bo2+(Rp-Rso2)*Bg2)+Wp*Bw2=N*(Eo+m*Eg+Efw)+(Wi+We)*Bw2+Gi*Bg2
 #---------------------------------------------------------------------------
 Eq=classEBM()
[…]

### 2015-03-10 pituba/article/article_spe.tex
artigo

@@
 It consists of the algebraic equation originally presented in \cite{schilthuis1936active} as a zero dimensional model for estimating reservoir pressure from mass balance.
-\cite{schilthuis1936active} introduced the hypothesis of equilibrium attained at all times not necessarily in the entire reservoir volume, but in the regions of high permeability where most of the production would come from and where pressure would present a similar trend.
+\cite{schilthuis1936active} introduced the hypothesis of mechanical equilibrium attained at all times not necessarily in the entire reservoir volume, but in the regions of high permeability where most of the production would come from and where pressure would present a similar trend.
 Such regions were said to contain an active oil.
 
-The equation developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, chemical or thermal equilibrium.
+The equation developed by \cite{schilthuis1936active} did not account for the non-existence of hydrodynamic, chemical or thermal equilibrium.
 An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}.
 The novelty on hydrodynamics was the introduction of a macroscopic equation of motion using Darcy's law.
+To account for the hydrodynamic non-equilibrium, black-oil equations are usually solved for in three dimensional models that aim to predict the pressure field responsible for reservoir fluid flow.
+
+Many works have reported applications of the material balance equation in the literature, but few have explored its connection with black-oil equations.
+\cite{ertekin2001basic} briefly addressed this subject arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \vec{g}$) were neglected and if PVT properties were averaged in reservoir volume.
+This work develops the concept of negligible flow potential and proposes an averaging scheme for flow properties.
+Subsequently, we suggest how the pressure computed from material balance equation can be compared to the pressure field in 3D black-oil models.
 
@@
 
-The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system.
-For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production so that a pseudo-gas component can be defined.
+The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium can be described by a two pseudo-component system.
+For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures (and temperatures) observed during production so that a pseudo-gas component can be defined.
 Similarly, a pseudo-oil component can be defined by subtraction of the pseudo-gas component from oil phase composition.
@@
 This work develops this process examining underlying details.
-
-Many works have reported applications of the material balance equation in the literature, but few have explored its connection with black-oil equations.
-\cite{ertekin2001basic} briefly addressed this subject arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \vec{g}$) were neglected and if PVT properties were averaged in reservoir volume.
-This work develops this concept examining underlying details.
 

### 2015-04-06 pituba/article/article_spe.tex
fixing

@@
 This work examines underlying details in obtaining the classical material balance equation from a three dimensional black-oil model typical of reservoir simulations.
-The black-oil equations are integrated in reservoir volume and quantities such as fluid properties are spatially averaged.
+The black-oil equations are integrated in reservoir volume and quantities are spatially averaged.
 The material balance equation is obtained for averaged quantities and it is applied to calculate reservoir pressure.
@@
 The novelty on hydrodynamics was the introduction of a macroscopic equation of motion using Darcy's law.
-To account for the hydrodynamic non-equilibrium, black-oil equations are usually solved for in three dimensional models that aim to predict the pressure field responsible for reservoir fluid flow.
+Black-oil equations are usually solved for in three dimensional models that aim to predict the pressure field responsible for reservoir fluid flow.
 
@@
 \cite{ertekin2001basic} briefly addressed this subject arguing both approaches to be equivalent if capillarity and gradients of flow potential ($\nabla \Phi = \nabla p - \rho \vec{g}$) were neglected and if PVT properties were averaged in reservoir volume.
-This work develops the concept of negligible flow potential and proposes an averaging scheme for flow properties.
-Subsequently, we suggest how the pressure computed from material balance equation can be compared to the pressure field in 3D black-oil models.
+This work develops on those concepts and proposes an averaging scheme for flow properties.
+Subsequently, it suggest how the pressure computed from material balance equation can be compared to the pressure field in 3D black-oil models.
 
@@
 The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium can be described by a two pseudo-component system.
-For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures (and temperatures) observed during production so that a pseudo-gas component can be defined.
-Similarly, a pseudo-oil component can be defined by subtraction of the pseudo-gas component from oil phase composition.
+For such reservoir fluids, gas phase composition remains essentially constant in the range of pressure and temperature observed during production so that a pseudo-gas component can be defined.
+Similarly, a pseudo-oil component can be defined by subtraction of the pseudo-gas component from the oil phase composition.
 After defining the pseudo-components, the number of transport equations needed for hydrocarbon components is reduced from a usually large number to only two.
@@
 Since temperature was assumed constant, oil phase density becomes a function of pressure alone.
-The same is valid for mass fractions of gas and oil components in oil phase:
+The same is valid for mass fractions of gas and oil components in the oil phase:
 \begin{equation}
@@
 \end{equation}
+
+When the oil phase is undersaturated, the system has an additional degree of freedom and both pressure and the amount of dissolved gas must be known to define its thermodynamic state, e.g., $\rho = \rho(p_o,Y_o)$.
 
@@
 
-Spatially averaged quantities are defined for porosity ($\phi$), phase saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration ($Y_{\alpha,k}$) of a component $k$ in phase $\alpha$.
+Spatially averaged quantities are defined for porosity ($\phi$), phase saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration of a component $k$ in phase $\alpha$ ($Y_{\alpha,k}$).
 
@@
 
-To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume mechanical equilibrium: $\vec{u}_\alpha = 0$.
-This implies that pressure at some time instant is only a function of depth: $\nabla p = \partial p_\alpha/\partial z = \rho_\alpha \vec{g}$.
+To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume mechanical equilibrium: $\vec{u}_\alpha \equiv 0$.
+This implies that pressure at any time instant is only a function of depth: $\nabla p = \partial p_\alpha/\partial z = \rho_\alpha \vec{g}$.
 
@@
 %We shall see that to compute the averaged phase density, we may define some datum level and use the corresponding local density as an approximation to the averaged density without introducing any important error.
+\begin{equation}
+\rho_o=\rho_o\left(p_o,Y_o\right)
+\end{equation}
 
- As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount.
+\begin{equation}
+\rho_o\left( t,z\right) \approx \rho_o\left(t,z_{datum}\right)\left[1+\frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} dz\right]
+\end{equation}
+For an undersaturated oil with no compositional gradient,
+ %As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount.
 This approximation can be quantified by putting together the expressions for compressibility and hydrostatic equilibrium.
-For this, we assume a constant isothermal compressibility for oil phase,
-\begin{equation}\label{eq: grado}
-\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o \, ,
+For an undersaturated oil with no compositional gradient,
+ \begin{equation}
+\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \right)\frac{\partial p_o}{\partial z} = c_o \rho_o g \, ,
  \end{equation}
- and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
-\begin{equation}\label{eq: gradodz}
- \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
+ where $c_o$ is the isothermal compressibility.
+
+ For a saturated oil,
+ \begin{equation}
+\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_T = \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \right)_T = 
+\frac{1}{\rho_o}\frac{d}{d p_o}\left(\frac{\rho_o^{ST}+R_s\rho_g^{ST}}{B_o}\right) 
  \end{equation}
@@
 
-Assuming reservoir temperature to be sufficiently above gas phase critical temperature and thus a negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
+Assuming reservoir temperature to be sufficiently above gas phase critical temperature, we have a negligible variation of the Z-factor with pressure, $\partial Z / \partial p_g \approx 0$.
+We use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
 \begin{equation}\label{eq: gradgdz}

### 2015-04-07 pituba/article/article_spe.tex
fixing+

@@
 \begin{abstract}
-This work examines underlying details in obtaining the classical material balance equation from a three dimensional black-oil model typical of reservoir simulations.
-The black-oil equations are integrated in reservoir volume and quantities are spatially averaged.
-The material balance equation is obtained for averaged quantities and it is applied to calculate reservoir pressure.
-It is shown that the calculated pressure corresponds to the local pressure at the datum level.
+It is generally known that the material balance equation can be derived from a 3D black-oil model typical of reservoir simulation.
+However, the existing literature is not comprehensive about what are the needed assumptions and how the pressure calculated from the material balance equation relates to the pressure field of a 3D black-oil model.
+
+This work examines underlying details in obtaining the material balance equation from a 3D black-oil model.
+The black-oil equations are integrated in reservoir volume and quantities are averaged in space.
+The material balance equation is obtained for the averaged quantities and it is applied to calculate reservoir pressure.
+It is shown that this calculated pressure corresponds to the pressure at a chosen datum level in the 3D black-oil model (and not, in general, to the average pressure in the 3D model).
 The definition of a convenient datum level is also discussed.
 It is found that the physical correctness of PVT properties
-% obtained from laboratory experiments or empirical correlations
 is a sufficient condition for solution uniqueness of the material balance equation when pressure is the unknown quantity.
@@
 Reservoir fluids are composed of oil and gas pseudo-components and water component.
-There can be three fluid phases at most: oil, gas and (liquid) water.
+There can be three fluid phases at most: oil, gas and liquid water.
 Oil phase is composed of oil and gas components, gas phase is composed uniquely by gas component and, water phase, uniquely by water component.
@@
 It follows from Duhem's theorem that the equilibrium of oil and gas phases is determined if, in addition to the mass balance equations, two other properties are prescribed.
-From Gibbs' phase rule, the system has two degrees of freedom.
-It is then necessary to specify two intensive properties: the constant temperature is one and the local pressure is chosen as the other.
+From Gibbs' phase rule, the oil an gas system has two degrees of freedom.
+It is then necessary to specify two intensive properties.
+The constant temperature is one and the local pressure is chosen as the other.
 Since temperature was assumed constant, oil phase density becomes a function of pressure alone.
@@
 
-When the oil phase is undersaturated, the system has an additional degree of freedom and both pressure and the amount of dissolved gas must be known to define its thermodynamic state, e.g., $\rho = \rho(p_o,Y_o)$.
+When the oil phase is undersaturated, the system has one additional degree of freedom and both pressure and the amount of dissolved gas must be known to define its thermodynamic state, e.g., $\rho_o = \rho_o (p_o,Y_o)$.
 
@@
 
-Eqs. \eqref{eq: Sw2}, \eqref{eq: So2} and \eqref{eq: Sg2} are rewritten in terms of averaged quantities:
+Eqs. \eqref{eq: Sw2}, \eqref{eq: So2} and \eqref{eq: Sg2} can now be rewritten in terms of averaged quantities:
 
@@
 
-We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes.
-In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to its volume at standard surface condition, i.e.:
+Usual definitions from petroleum engineering can be introduced for the description of volumetric and solubility changes.
+For volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to its volume at standard surface condition, i.e.:
 
@@
 
-In what concerns changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between the volume of dissolved gas and the volume of oil both measured at surface condition:
+For changes in gas solubility in oil phase, the solubility ratio ($R_s$) gives the ratio between the volume of dissolved gas and the volume of oil both measured at surface condition:
 
@@
 
-An equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, $R_s(p=P_b)$:
+An inplicit equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, $R_s(p=P_b)$:
 \begin{equation}\label{eq: eqPb}
@@
 Eq~\eqref{eq: MBE} involves fluid properties, original phase volumes ($N$, $W$ and $G)$, cumulative production and injection and aquifer influx.
-It is an algebraic equation that can be solved for any one of its terms, provided that all remaining are known. This works focuses on its solution for pressure.
+It is an algebraic equation that can be solved for any one of its terms, provided that all remaining are known. This work focuses on its solution for pressure.
 
@@
 If the reservoir is under production, though, the pressure field is not known and for its prediction one would have to resort to more advanced techniques such as reservoir simulation.
+
+\begin{itemize}
+\item density is approximately constant;
+\item pressure field does not deviates from hydrostatic equilibrium.
+\end{itemize}
 
@@
 \end{equation}
-For an undersaturated oil with no compositional gradient,
+For an undersaturated oil, $\rho_o$ is a function of pressure and the amount of dissolved gas:
  %As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount.
@@
  \begin{equation}
-\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \right)\frac{\partial p_o}{\partial z} = c_o \rho_o g \, ,
+\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_{T} = \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \frac{\partial p_o}{\partial z} + \frac{\partial \rho_o}{\partial Y_o}\frac{\partial Y_o}{\partial z}\right) \, .
+ \end{equation}
+Neglecting any compositional gradient, $\partial Y_o/\partial z = 0$, and using the hydrostatic gradient:
+\begin{equation}
+\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_{T} = c_o \rho_o g \, ,
 \end{equation}
@@
  \begin{equation}
-\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_T = \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \right)_T = 
-\frac{1}{\rho_o}\frac{d}{d p_o}\left(\frac{\rho_o^{ST}+R_s\rho_g^{ST}}{B_o}\right) 
+\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_T = g\left(\frac{\partial \rho_o}{\partial p_o} \right)_T = 
+g\frac{d}{d p_o}\left(\frac{\rho_o^{ST}+R_s\rho_g^{ST}}{B_o}\right) 
  \end{equation}
@@
 
-From Eq.~\eqref{eq: dRsdz2}, we see that the variation of pressure with depth may not be negligible for the phase equilibrium,  specially for thick reservoirs of light oil.
+From Eq.~\eqref{eq: dRsdz2}, we see that the variation of pressure with depth may not be negligible for the phase equilibrium,  specially for thick reservoirs of low saturation pressure.
 In such cases, using an averaged value for the solubility ratio may be worth the accuracy gain in computing dissolved gas volume.

### 2015-04-22 pituba/article/article_spe.tex
news

@@
 The material balance equation is obtained for the averaged quantities and it is applied to calculate reservoir pressure.
-It is shown that this calculated pressure corresponds to the pressure at a chosen datum level in the 3D black-oil model (and not, in general, to the average pressure in the 3D model).
+It is shown that this calculated pressure corresponds to the pressure at a chosen datum level in the 3D black-oil model (and not, in general, to the average pressure of the 3D model).
 The definition of a convenient datum level is also discussed.
@@
 
-An inplicit equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, $R_s(p=P_b)$:
+An implicit equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, $R_s(p=P_b)$:
 \begin{equation}\label{eq: eqPb}
@@
 % pressure field can be assumed constant
-% pressure field can be given by hydrostatic equilibrium
+% pressure field can be given by hydros tatic equilibrium
 
-The material balance equation is widely used to relate production and injection to reservoir pressure.
-Actually, all averaged phase properties in Eq.~\eqref{eq: MBE} depend on the pressure field.
+The material balance equation is mostly used to relate production and injection to reservoir pressure.
+In fact, all averaged phase properties in Eq.~\eqref{eq: MBE} depend on the pressure field.
 If the reservoir is under production, though, the pressure field is not known and for its prediction one would have to resort to more advanced techniques such as reservoir simulation.
+Further, even if the pressure field were known, one would still need to average fluid properties in space in order to apply Eq.~\eqref{eq: MBE}.
+In practice, the reservoir engineer is unlikely to perform any averaging of properties since the main reason for using the material balance equation is to avoid the complexities of a 3D reservoir model.
 
+It can be shown that the averaging of fluid properties can be neglected under the following assumptions:
 \begin{itemize}
-\item density is approximately constant;
-\item pressure field does not deviates from hydrostatic equilibrium.
+\item Negligible hydrodynamic gradients and capillarity. This implies that pressure at any time is only a function of depth: $\nabla p = \partial p\partial z = \rho_\alpha \vec{g}$.
+\item Phase densities and gas solubility ratio are approximately constant with depth.
 \end{itemize}
+
+To show that phase densities can be considered approximately constant with depth, a first order Taylor expansion
+
+%To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume mechanical equilibrium: $\vec{u}_\alpha \equiv 0$.
+This implies that pressure at any time instant is only a function of depth: $\nabla p = \partial p_\alpha/\partial z = \rho_\alpha \vec{g}$.
+
+\begin{equation}
+\rho_o \left(z\right) \approx \rho_o\left(z_\text{datum}\right)+\left(\frac{\partial \rho_o}{\partial Y_o}\frac{\partial Y_o}{\partial z}+\frac{\partial \rho_o}{\partial p_o}\frac{\partial p_o}{\partial z}\right) dz
+\end{equation}
 
@@
 For any practical purpose, this is not true.
-
-To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume mechanical equilibrium: $\vec{u}_\alpha \equiv 0$.
-This implies that pressure at any time instant is only a function of depth: $\nabla p = \partial p_\alpha/\partial z = \rho_\alpha \vec{g}$.
 
@@
 
-\begin{equation}
-\rho_o\left( t,z\right) \approx \rho_o\left(t,z_{datum}\right)\left[1+\frac{1}{\rho_o}\frac{\partial \rho_o}{\partial z} dz\right]
-\end{equation}
 For an undersaturated oil, $\rho_o$ is a function of pressure and the amount of dissolved gas:

### 2015-10-23 pituba/article/article_spe.tex
new changes

@@
 
-The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium can be described by a two pseudo-component system.
+The term black-oil used in this work refers to specific hydrocarbon compositions to which phase equilibrium can be described by a two pseudo-component system.
 For such reservoir fluids, gas phase composition remains essentially constant in the range of pressure and temperature observed during production so that a pseudo-gas component can be defined.
@@
 By a hidraulically connected volume, we mean that it is possible to observe a common pressure decline trend when any reservoir portion is under production \citep{dakepractice}.
+Mathematically, this is equivalent to having pressure diffusivity above certain minimal value everywhere.
 Reservoir fluids are composed of oil and gas pseudo-components and water component.
@@
 In practice, the reservoir engineer is unlikely to perform any averaging of properties since the main reason for using the material balance equation is to avoid the complexities of a 3D reservoir model.
-
 It can be shown that the averaging of fluid properties can be neglected under the following assumptions:
 \begin{itemize}
-\item Negligible hydrodynamic gradients and capillarity. This implies that pressure at any time is only a function of depth: $\nabla p = \partial p\partial z = \rho_\alpha \vec{g}$.
+\item Negligible hydrodynamic gradients and capillarity. This implies that pressure at any time is only a function of depth: $\nabla p = \partial p / \partial z = \rho_\alpha \vec{g}$.
 \item Phase densities and gas solubility ratio are approximately constant with depth.
@@
 \begin{step}
-Compute fluid saturations $t_2$.
+Compute fluid saturations  for instant $t_2$.
 \end{step}

### 2016-05-12 pituba/article/article.tex
news

@@
 % Review
-\documentclass[authoryear,preprint,review,12pt]{elsarticle}
-%\documentclass[final,authoryear,5p,times,twocolumn,11pt]{elsarticle}
+%\documentclass[authoryear,preprint,review,10pt]{elsarticle}
+\documentclass[final,authoryear,5p,times,twocolumn,11pt]{elsarticle}
 
@@
 This work presents a derivation of the classical material balance equation starting from black-oil equations in integral form.
-The existing link between both approaches is examined in detail, including the necessary step of averaging PVT properties.
+The existing link between both approaches is examined, including the step of averaging PVT properties.
 An interpretation for pressure in material balance calculations is provided and the choice of a reference depth (datum) is discussed.
 The material balance equation is formulated for a varying bubble-point pressure and sufficient conditions for solution uniqueness are presented.
-A scheme for numerical solution is also proposed.
+A scheme for numerical solution is also proposed for the case when pressure is the unknown quantity.
 \end{abstract}
@@
 It consists of an algebraic equation originally presented in \cite{schilthuis1936active} and derived from a zero dimensional model for estimating reservoir pressure from mass balance.
-\cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the production would come from and where pressure would present a similar behavior.
+\cite{schilthuis1936active} introduced the hypothesis of thermodynamic equilibrium occurring not in the entire reservoir volume, but in the regions of high permeability where most of the production would come from and where pressure would present a similar trend.
 Such regions were said to contain an \textit{active oil}.
@@
 
-The term black-oil refers to specific hydrocarbon compositions whose phase equilibrium for isothermal changes in pressure may be described by a two pseudo-component system.
+The term black-oil refers to specific hydrocarbon compositions whose phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system.
 For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and we may define this composition as a pseudo gas component.
@@
 
-\subsection{Black-Oil equations}
+\subsection{Black-oil equations}
 
-As already mentioned, black-oil is a designation to a simplified treatment of hydrocarbon composition and phase equilibrium.
+As mentioned, black-oil is a designation to a simplified treatment of hydrocarbon composition and phase equilibrium.
 Hydrocarbon species are grouped into two pseudo-components called oil and gas, being a pseudo-component one that consists of more than one chemical molecule.
@@
 Similarly, gas component is the hydrocarbon composition in gaseous phase for the same surface conditions.
-All species are generally present in both phases although the equilibrium composition is shifted to liquid phase (or gaseous phase) for components of greater (lower) molecular mass.
-
 Reservoir fluids are then composed of the two pseudo-components (oil and gas) plus the water component.
@@
 
-A common simplification in black-oil formulations is to restrict mass transport to macroscopic advection.
-No diffusive or dispersive term is included in the mass balance equations.
+A common simplification in black-oil formulations is to restrict mass transport to advection.
+No diffusive term is included in the mass balance equations.
 This simplification must be looked with care when gradients of species concentration are important.
@@
 \begin{equation}\label{eq: exdarcy}
-\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right) \, , \quad \alpha=o,w,g \, .
+\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha \mathbf{g}\right) \, , \quad \alpha=o,w,g \, .
 \end{equation}
@@
 \begin{equation}
-T_w = T_o = T_g = T_{rock} = T
+T_w = T_o = T_g = T_{rock} = T\, , 
 \end{equation}
@@
 \begin{equation}
-\nabla T \equiv 0 \, .
+dT \equiv 0 \, .
 \end{equation}
@@
 
-Porosity ($\phi$) is assumed a function of pore pressure ($p_P$). Since three phases are present in general, such quantity is defined by some relation of type \citep{kim2011rigorous}:
+Porosity ($\phi$) is assumed a function of pore pressure ($p_P$). Since three phases are present in general, such quantity is defined by some relationship as follows \citep{kim2011rigorous}:
 \begin{equation}
@@
 \begin{equation}
-\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha g\nabla z\right)\equiv 0 \, .
+\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha \mathbf{g}\right)\equiv 0 \, .
 \end{equation}
@@
 \begin{equation}\label{eq: p=rhogh}
-\nabla p_\alpha = \rho_\alpha g\nabla z
+\nabla p_\alpha = \rho_\alpha \mathbf{g} \, .
 \end{equation}
-and, consequently, pressure is a function of depth only:
+Consequently, pressure is a function of depth only:
 \begin{equation}\label{eq: p=rhogh}
@@
 \begin{equation}
-\bar{\rho_\alpha}\bar{S_\alpha}\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha S_\alpha \phi dV \, ,
+\bar{\rho_\alpha}\bar{S}_\alpha\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha S_\alpha \phi dV \, ,
 \end{equation}
@@
 \begin{equation}
-\bar{\rho_\alpha}\bar{Y_{\alpha,k}}\bar{S_\alpha}\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha Y_{\alpha,k} S_\alpha \phi dV \, .
+\bar{\rho}_\alpha\bar{Y}_{\alpha,k}\bar{S}_\alpha\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}\rho_\alpha Y_{\alpha,k} S_\alpha \phi dV \, .
 \end{equation}
@@
 \begin{equation}\label{eq: Sw3}
-\frac{\partial}{\partial t} \left( \bar{\rho_w}\bar{S_w}\bar{\phi} \Omega \right) = \rho_w^{ST}Q_w^{ST}
+\frac{\partial}{\partial t} \left( \bar{\rho}_w\bar{S}_w\bar{\phi} \Omega \right) = \rho_w^{ST}Q_w^{ST}
 \end{equation}
@@
 \begin{equation}\label{eq: So3}
-\frac{\partial}{\partial t} \left( \bar{\rho_o}\bar{Y_o}\bar{S_o}\bar{\phi} \Omega \right) = \rho_o^{ST}Q_o^{ST}
+\frac{\partial}{\partial t} \left( \bar{\rho}_o\bar{Y}_o\bar{S}_o\bar{\phi} \Omega \right) = \rho_o^{ST}Q_o^{ST}
 \end{equation}
@@
 Material balance equation is widely employed to stablish a relationship between production (and injection) and reservoir pressure.
-It turns out, though, that pressure is different for different locations, mostly for different depths, and also for different phases.
+It turns out, though, that pressure is different for different locations and also for different phases.
+
 What is then the significance of having a single value of pressure for the entire reservoir volume?
 
-For answering this question, we examine the two roles played by pressure in material balance equation: computation of averaged phase density and phase equilibrium.
+For answering this question, we examine the two roles played by pressure in material balance equation: computation of phase density and phase equilibrium.
+
 We shall see that for the purpose of computing the averaged phase density, we may define some reference depth and use the corresponding local density as an approximation to the averaged density without introducing any important error.
@@
 \begin{equation}\label{eq: MBE}
-F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{porous volume contraction } (C)}
+F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p2}-V_{p1}}_{\text{porous volume change } (\Delta V_p)}
 \end{equation}
@@
 \begin{equation}\label{eq: MBE_short}
-F_p - F_i= E + C \, ,
+F_p - F_i= E - \Delta V_p \, ,
 \end{equation}
@@
 
-\section{Solution of Material balance equation}
-A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We shall start defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
+\section{Solution of material balance equation}
+A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We start defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
 \begin{equation}\label{eq: residual}
-\varepsilon(p_2) = \frac{E+F_i+C-F_p}{V_{p1}} \, .
+\varepsilon(p_2) = \frac{E+F_i-\Delta V_p-F_p}{V_{p1}} \, .
 \end{equation}
@@
 \begin{split}
-&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right)}_\text{oil phase}\\
+&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_{g2} \frac{dR_s}{dp} \right)}_\text{oil phase}\\
 &+\underbrace{\left[N R_{s1}+mN\frac{B_{o1}}{B_{g1}}+G_i-\left(N-N_p\right)R_{s2}-G_p\right] \frac{dB_g}{dp}}_\text{gas phase}\\
@@
 
-\textit{In-situ} volumes (e.g.
+Remaining volumes in surface condition (e.g.
 $N-N_p$ for oil phase) are non-negative values.
@@
 bubble-point pressure & $R_{s2}\left(P_b\right) = \left[N\left(m B_{1}/B_{g1} +R_{s1}\right)+G_i-G_p\right]/\left(N-N_p\right)$ \\
-pressure & $F_p - F_i= N E_o + mN E_g + W E_w +C$ \\
+pressure & $F_p - F_i= N E_o + mN E_g + W E_w - \Delta V_p$ \\
 \midrule
@@
 water volume expansion & $E_w = B_{w2}-B_{w1}$\\
-porous volume contraction & $C=V_{p2}- V_{p1}$\\ \bottomrule
+porous volume contraction & $\Delta V_p=V_{p2}- V_{p1}$\\ \bottomrule
 \end{tabular}

### 2016-05-12 pituba/pitubalib.py

@@ -274,7 +274,24 @@ class classEBM:
 			Fp=Np[i]*(pvt.Bo(p[i],pvt.Pb)+(Gp[i]/Np[i]-pvt.Rs(p[i],pvt.Pb))*pvt.Bg(p[i]))+Wp[i]*pvt.Bw(p[i])
 			N.append((Fp-Fi)/E)
 
-		return array(N)
+	def VOIPW(self):
+		p=self.res.teste.p
+		Np=self.res.teste.Np
+		Wp=self.res.teste.Wp
+		Wi=self.res.teste.Wi
+		We=zeros(len(Wi))
+		Gi=self.res.teste.Gi
+		Gp=self.res.teste.Gp
+		res=self.res
+		pvt=self.pvt
+		A=zeros(len(p))
+		B=zeros(len(p))
+		C=zeros(len(p))
+		for i in range(len(p)):
+			A[i]=self.Eo(p[i])+res.m*self.Eg(p[i])+self.Efw(p[i])
+			B[i]=(self.res.cr+self.pvt.cw)*pvt.Bw(p[i])*(self.res.p0-p[i])
+			C[i]=Np[i]*(pvt.Bo(p[i],pvt.Pb)+(Gp[i]/Np[i]-pvt.Rs(p[i],pvt.Pb))*pvt.Bg(p[i]))+Wp[i]*pvt.Bw(p[i])-Wi[i]*pvt.Bw(p[i])-Gi[i]*pvt.Bg(p[i])
+		# least square regression of Ai*N+Bi*We-Ci=0
 
 	def runHist(self,pmin,pmax):
 		out=self.out

### 2016-05-12 simple_codes/python/simple_codes/diffusion_eq_1D.py

@@ -1,77 +1,0 @@
-#-------------------------------------------------------------------------------
-# Name:        Solve diffusion Eq. in 1D (linear solution)
-#              SI units are used
-# Purpose:
-#
-# Author:      Rodrigo, bfv8
-#
-# Created:     15/02/2013
-#-------------------------------------------------------------------------------
-#!/usr/bin/env python
-
-from pylab import *
-
-# Define domain size
-np=250
-
-# Define diffusion coefficient
-k=100e-15
-phi=0.2
-ct=1e-2/1e5
-mu=1e-3
-eta=k/mu/phi/ct
-print 'eta = ', eta
-
-
-# Define reservoir limit
-R2=2000.
-R1=0.
-
-# Define points
-#r=logspace(log10(R1),log10(R2),np)
-r=linspace(R1,R2,np)
-dr=r[2]-r[1]
-
-# Define time
-dt=3600.
-nt=12
-
-# Define Initial conditions
-p=400e5*r/R2
-p0=0.
-
-p[0]=p0
-
-# Build linear system
-C=eta*dt/dr**2
-a=zeros([np,np])
-for i in range(1,np-1):
-	a[i,i]=-(1.+2*C)
-	a[i,i-1]=C
-	a[i,i+1]=C
-
-# Apply boundary conditions
-
-# Internal BC of fixed value
-a[0,0]=-1. ##C-1. ##C/2-1.
-a[0,1]=0. ##-2*C ##-C
-a[0,2]=0. ## C    ##C/2
-
-# External BC of fixed value
-a[np-1,np-1]=-1. ##C-1. ##C/2-1.
-a[np-1,np-2]=0. ##-2*C ##-C
-a[np-2,np-3]=0. ## C    ##C/2
-
-# External BC of zero gradient
-# a[np-1,np-2]=2.*C
-# a[np-1,np-1]=-(1.+2.*C)
-
-for tt in range(nt):
-	p=dot(inv(a),-1.*p)
-
-	figure(1)
-	plot(r,p/1e5, 'k-')
-#	ylim([1e-6,1.])
-
-grid(1)
-show()

### 2016-07-14 pituba/article/article.tex
added new object oriented version of GITT.

@@
 %\documentclass[authoryear,preprint,review,10pt]{elsarticle}
-\documentclass[final,authoryear,5p,times,twocolumn,11pt]{elsarticle}
+\documentclass[final,authoryear,5p,times,twocolumn,10pt]{elsarticle}
 
@@
 
-Mass balance equations for each component in fluid phase are:
+Mass balance equations for each component are:
 
-\begin{itemize}
-\item water component:
 \begin{equation}\label{eq: Sw1}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV + \oint_{\partial \Omega} \rho_w \left( \mathbf{u}_w \cdot \mathbf{dS} \right) = \rho_w^{ST}Q_w^{ST} \, ;
+\begin{split}
+\frac{\partial}{\partial t} \left( \rho_w S_w \phi \right) + \nabla\cdot\left(\rho_w \vec{u}_w \right) = \\ \rho_w^{ST}\left(Q_{wi}^{ST} - Q_w^{ST} \right) \, ;
+\end{split}
 \end{equation}
 
-\item oil component:
 \begin{equation}\label{eq: So1}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV + \oint_{\partial \Omega} \rho_o Y_o \left(\mathbf{u}_o \cdot \mathbf{dS} \right)= \rho_o^{ST}Q_o^{ST} \, ;
+\begin{split}
+\frac{\partial}{\partial t}\left( \rho_o Y_o S_o \phi \right) + \nabla\cdot\left( \rho_o Y_o \vec{u}_o \right)= - \rho_o^{ST}Q_o^{ST} \, ;
+\end{split}
 \end{equation}
 
-\item gas component:
 \begin{equation}\label{eq: Sg1}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV + \oint_{\partial \Omega} \left( \rho_o Y_g \mathbf{u}_o + \rho_g \mathbf{u}_g\right) \cdot \mathbf{dS}= \rho_g^{ST}Q_g^{ST} \, .
+\begin{split}
+\frac{\partial}{\partial t}  \left[ \phi\left( \rho_o Y_g S_o + \rho_g S_g\right) \right]  + \nabla\cdot\left( \rho_o Y_g \vec{u}_o + \rho_g \vec{u}_g\right)=\\ \rho_g^{ST}\left(Q_{gi}^{ST}-Q_{g}^{ST}\right) \, .
+\end{split}
 \end{equation}
-\end{itemize}
 Mass concentration of oil and gas components in oil phases are denoted as $Y_o$ and $Y_g$, respectively.
-
-Phase saturation is defined as its volumetric fraction of the pore space:
-\begin{equation}
-S_\alpha = \frac{\phi_\alpha}{\phi} \, ,
-\end{equation}
-where $\phi_\alpha$ is the volumetric fraction of a fluid phase $\alpha$ and $\phi$ is the porosity, i.e., the volumetric fraction of the pore volume.
-Since we consider that pore volume is fully saturated with fluid,
+Source terms above account for well production or injection.
+Pore volume is assumed to be fully saturated, thus:
 \begin{equation}
@@
 
-A common simplification in black-oil formulations is to restrict mass transport to advection.
-No diffusive term is included in the mass balance equations.
-This simplification must be looked with care when gradients of species concentration are important.
-For instance, when gas is being dissolved in oil phase in a reservoir under repressurization, this condition is not generally true.
+Integrating Eqs.~\eqref{eq: Sw1}, \eqref{eq: So1} and \eqref{eq: Sg1} in reservoir volume ($\Omega$):
+
+\begin{equation}\label{eq: Sw2}
+\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV = \rho_w^{ST}\left(Q_{wi}^{ST} - Q_w^{ST} \right) - \rho_w Q_{w,aq} \, ;
+\end{equation}
+
+\begin{equation}\label{eq: So2}
+\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV = - \rho_o^{ST}Q_o^{ST} \, ;
+\end{equation}
+
+\begin{equation}\label{eq: Sg2}
+\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV = \rho_g^{ST}\left(Q_{gi}^{ST}-Q_{g}^{ST}\right) \, .
+\end{equation}
+It was assumed that aquifer influx is the only allowed flow across reservoir boundaries.
 
@@
 
-Momentum balance for each phase is given by an extension of Darcy's law to multiphase flow.
-This extension consists in multiplying absolute (or single-phase) permeability ($\mathbf{k}$) by a scalar function called phase relative permeability ($k_{r,\alpha}$):
-
-\begin{equation}\label{eq: exdarcy}
-\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha \mathbf{g}\right) \, , \quad \alpha=o,w,g \, .
-\end{equation}
-
-Phase densities are functions of the local thermodynamic state:
-\begin{equation}
-\rho_w = \rho_w \left(p_w, T_w\right)
-\end{equation}
-\begin{equation}
-\rho_o = \rho_o \left(p_o, T_o,Y_g\right)
-\end{equation}
-\begin{equation}
-\rho_g = \rho_g \left(p_g, T_g\right)
-\end{equation}
-
-Another simplification is now made assuming thermal equilibrium in the entire reservoir volume:
-\begin{equation}
-T_w = T_o = T_g = T_{rock} = T\, , 
-\end{equation}
-and
-\begin{equation}
-dT \equiv 0 \, .
-\end{equation}
-
-Temperature is thus a constant prescribed value and no energy balance equation is needed.
-Also, it follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if we restrict component number to the two pseudo-components.
-As temperature is assumed constant, oil phase density in saturated states is a function of pressure or composition only.
-The same is valid for mass concentrations of gas and oil components in oil phase.
-
-\begin{equation}
-Y_g = Y_g (p) \, ,
-\end{equation}
-\begin{equation}
-Y_o = Y_o (p) \, .
-\end{equation}
-
-Porosity ($\phi$) is assumed a function of pore pressure ($p_P$). Since three phases are present in general, such quantity is defined by some relationship as follows \citep{kim2011rigorous}:
-\begin{equation}
-p_P=p_P\left(p_o,p_w,p_g,S_o,S_w,S_g\right) \, .
-\end{equation}
-
-In order to close the system of equations, a connection among different phase pressures must be provided.
-Such expressions are referred to as capillary pressures and are usually employed in the following forms \citep{aziz1979petroleum}:
-%capillary
-\begin{equation}\label{eq: capillary_ow}
-p_{c,ow}\left(\mathbf{k},\phi,S_w\right)=p_o-p_w \, ,
-\end{equation}
-\begin{equation}\label{eq: capillary_og}
-p_{c,go}\left(\mathbf{k},\phi,S_g\right)=p_g-p_o \, .
-\end{equation}
-
-%Capillary pressure and relative permeability are controversial concepts in the sense that they have not been demonstrated to arise from theoretical foundations.
-For practical applications, both functions are usually assumed in the forms presented above and their properties are argued to be empirically established.
-
-%It is well known that black-oil equations possess discontinuous solutions in saturation, bubble pressure and solubility ratio fields.
-Discontinuities (or shocks) are introduced by initial or boundary conditions and are propagated throughout flow domain due to the hyperbolicity of equations.
-%
-%In initial conditions, if hydrostatic equilibrium is imposed and no capillary pressure is included, that is,
-%\begin{equation}
-%p_{c,ow}=p_{c,og}=0 \quad \longrightarrow \quad p_o=p_w=p_g=p\, ,
-%\end{equation}
-%saturation discontinuities will be present due to gravity segregation.
-%
-%When capillary pressure is present and it is function of phase saturations, a diffusive term is introduced in mass balance equations and shocks originated by phase displacement become smoothed:
-%\begin{equation}
-%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)-\nabla\cdot\left[\frac{k_{r,w} \mathbf{k}}{B_w(p_w)\mu_w} \left(\nabla p_o - \rho_w g\nabla z -\underbrace{\frac{d p_c}{d S_w}\nabla S_w}_{\text{water diffusion}}\right)\right]=q_{ws} \, .
-%\end{equation}
-%
-%Discontinuities in saturation may still occur, however, if capillary pressures are functions of discontinuous porous medium properties ($phi$ and $\mathbf{k}$) or if phase change takes place, as diffusion of gas dissolved in oil phase is not allowed.
-%Cite Bear and Herrera.
-
-\subsection{Material balance equation}
-
-Derivation of the material balance equation consists in assuming no fluid motion and in integrating black-oil equations in reservoir volume.
-As it will be shown, the integration process requires all quantities to be either spatially averaged or assumed constant.
-
-%Defining averages of most quantities is difficulty, though, as they appear in equations as products of up to three factors.
-Being the quantities self-correlated, the averaging process would involve unclosed terms.
-Quantities are thus required to be spatially constant wherever needed.
-
-%A second problem is that averaging requires the knowledge of a spatial distribution of quantities for it be uniquely specified, as it can be seen for an average saturation depending on spatial distributions of local saturation and porosity,
-%\begin{equation}\label{eq: avS}
-%\bar{S}_\alpha = \int_\Omega S_\alpha \phi dV \, .
-%\end{equation}
-
-We start assuming mechanical equilibrium, i.e., no bulk motion:
-\begin{equation}
-\mathbf{u}_\alpha = -\frac{k_{r,\alpha} \mathbf{k}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha \mathbf{g}\right)\equiv 0 \, .
-\end{equation}
-This condition implies that pressure gradients must equal gravity force per unit volume:
-\begin{equation}\label{eq: p=rhogh}
-\nabla p_\alpha = \rho_\alpha \mathbf{g} \, .
-\end{equation}
-Consequently, pressure is a function of depth only:
-\begin{equation}\label{eq: p=rhogh}
-\int_{p_\alpha^0}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'},T\right)} = g\left(z - z_0\right) \, .
-\end{equation}
-
-Black-oil equations then become:
-\begin{itemize}
-\item water component:
-\begin{equation}\label{eq: Sw2}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV = \rho_w^{ST}Q_w^{ST} \, ;
-\end{equation}
-
-\item oil component:
-\begin{equation}\label{eq: So2}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV = \rho_o^{ST}Q_o^{ST} \, ;
-\end{equation}
-
-\item gas component:
-\begin{equation}\label{eq: Sg2}
-\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV = \rho_g^{ST}Q_g^{ST} \, .
-\end{equation}
-\end{itemize}
-%
-\subsubsection{Averaging quantities}
+\subsection{Averaging quantities}
 
@@
 \begin{equation}
-\bar{S}_\alpha\bar{\phi} = \frac{1}{\Omega}\int_{\Omega}S_\alpha \phi dV \, ,
+\bar{S}_\alpha\bar{\phi} = \frac{1}{\Omega}\int_{ \Omega}S_\alpha \phi dV \, ,
 \end{equation}
@@
 
-\begin{itemize}
-\item water component:
 \begin{equation}\label{eq: Sw3}
-\frac{\partial}{\partial t} \left( \bar{\rho}_w\bar{S}_w\bar{\phi} \Omega \right) = \rho_w^{ST}Q_w^{ST}
+\frac{\partial}{\partial t} \left( \bar{\rho}_w\bar{S}_w\bar{\phi} \Omega \right) = \rho_w^{ST}\left(Q_{wi}^{ST} - Q_w^{ST} \right) - \rho_w Q_{w,aq} \, ;
 \end{equation}
 
-\item oil component:
 \begin{equation}\label{eq: So3}
-\frac{\partial}{\partial t} \left( \bar{\rho}_o\bar{Y}_o\bar{S}_o\bar{\phi} \Omega \right) = \rho_o^{ST}Q_o^{ST}
+\frac{\partial}{\partial t} \left( \bar{\rho}_o\bar{Y}_o\bar{S}_o\bar{\phi} \Omega \right) = - \rho_o^{ST}Q_o^{ST} \, ;
 \end{equation}
 
-\item gas component:
 \begin{equation}\label{eq: Sg3}
-\frac{\partial}{\partial t} \left[\left( \bar{\rho}_o\bar{Y}_g \bar{S}_o + \bar{\rho}_g \bar{S}_g \right) \bar{\phi} \Omega\right] = \rho_g^{ST}Q_g^{ST} \, .
+\frac{\partial}{\partial t} \left[\left( \bar{\rho}_o\bar{Y}_g \bar{S}_o + \bar{\rho}_g \bar{S}_g \right) \bar{\phi} \Omega\right] = \rho_g^{ST}\left(Q_{gi}^{ST}-Q_{g}^{ST}\right) \, .
 \end{equation}
-\end{itemize}
 
@@
 
+Phase densities are functions of the local thermodynamic state:
+\begin{equation}
+\rho_w = \rho_w \left(p_w, T_w\right)
+\end{equation}
+\begin{equation}
+\rho_o = \rho_o \left(p_o, T_o,Y_g\right)
+\end{equation}
+\begin{equation}
+\rho_g = \rho_g \left(p_g, T_g\right)
+\end{equation}
+
+Another simplification is now made assuming thermal equilibrium in the entire reservoir volume:
+\begin{equation}
+T_w = T_o = T_g = T_{rock} = T\, , 
+\end{equation}
+and
+\begin{equation}
+dT \equiv 0 \, .
+\end{equation}
+
+Temperature is thus a constant prescribed value and no energy balance equation is needed.
+Also, it follows from Gibbs' phase rule that the equilibrium of oil and gas phases has only two degrees of freedom if we restrict component number to the two pseudo-components.
+As temperature is assumed constant, oil phase density in saturated states is a function of pressure or composition only.
+The same is valid for mass concentrations of gas and oil components in oil phase:
+\begin{equation}
+Y_g = Y_g (p_o) \, ,
+\end{equation}
+\begin{equation}
+Y_o = Y_o (p_o) \, .
+\end{equation}
+
+Porosity ($\phi$) is assumed a function of pore pressure ($p_P$). Since three phases are present in general, such quantity is defined by some relationship as follows \citep{kim2011rigorous}:
+\begin{equation}
+p_P=p_P\left(p_o,p_w,p_g,S_o,S_w,S_g\right) \, .
+\end{equation}
+
+In order to close the system of equations, a connection among different phase pressures must be provided.
+Such expressions are referred to as capillary pressures and are usually employed in the following forms \citep{aziz1979petroleum}:
+%capillary
+\begin{equation}\label{eq: capillary_ow}
+p_{c,ow}\left(\vec{\vec{k}},\phi,S_w\right)=p_o-p_w \, ,
+\end{equation}
+\begin{equation}\label{eq: capillary_og}
+p_{c,go}\left(\vec{\vec{k}},\phi,S_g\right)=p_g-p_o \, .
+\end{equation}
+
+So far, no simplification was made to black-oil equations.
+
+Momentum balance for each phase is given by an extension of Darcy's law to multiphase flow.
+This extension consists in multiplying absolute (or single-phase) permeability ($\vec{\vec{k}}$) by a scalar function called phase relative permeability ($k_{r,\alpha}$):
+
+\begin{equation}\label{eq: exdarcy}
+\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha \vec{g}\right) \, , \quad \alpha=o,w,g \, .
+\end{equation}
+%Capillary pressure and relative permeability are controversial concepts in the sense that they have not been demonstrated to arise from theoretical foundations.
+For practical applications, both functions are usually assumed in the forms presented above and their properties are argued to be empirically established.
+
+%It is well known that black-oil equations possess discontinuous solutions in saturation, bubble pressure and solubility ratio fields.
+Discontinuities (or shocks) are introduced by initial or boundary conditions and are propagated throughout flow domain due to the hyperbolicity of equations.
+%
+%In initial conditions, if hydrostatic equilibrium is imposed and no capillary pressure is included, that is,
+%\begin{equation}
+%p_{c,ow}=p_{c,og}=0 \quad \longrightarrow \quad p_o=p_w=p_g=p\, ,
+%\end{equation}
+%saturation discontinuities will be present due to gravity segregation.
+%
+%When capillary pressure is present and it is function of phase saturations, a diffusive term is introduced in mass balance equations and shocks originated by phase displacement become smoothed:
+%\begin{equation}
+%\frac{\partial}{\partial t}\left(\frac{\phi (p_P) S_w}{B_w (p_w)}\right)-\nabla\cdot\left[\frac{k_{r,w} \vec{\vec{k}}}{B_w(p_w)\mu_w} \left(\nabla p_o - \rho_w g\nabla z -\underbrace{\frac{d p_c}{d S_w}\nabla S_w}_{\text{water diffusion}}\right)\right]=q_{ws} \, .
+%\end{equation}
+%
+%Discontinuities in saturation may still occur, however, if capillary pressures are functions of discontinuous porous medium properties ($phi$ and $\vec{\vec{k}}$) or if phase change takes place, as diffusion of gas dissolved in oil phase is not allowed.
+%Cite Bear and Herrera.
+
+\subsection{Material balance equation}
+
+%Derivation of the material balance equation consists in assuming no fluid motion and in integrating black-oil equations in reservoir volume.
+As it will be shown, the integration process requires all quantities to be either spatially averaged or assumed constant.
+
+%Defining averages of most quantities is difficulty, though, as they appear in equations as products of up to three factors.
+Being the quantities self-correlated, the averaging process would involve unclosed terms.
+Quantities are thus required to be spatially constant wherever needed.
+
+%A second problem is that averaging requires the knowledge of a spatial distribution of quantities for it be uniquely specified, as it can be seen for an average saturation depending on spatial distributions of local saturation and porosity,
+%\begin{equation}\label{eq: avS}
+%\bar{S}_\alpha = \int_\Omega S_\alpha \phi dV \, .
+%\end{equation}
+
+We start assuming mechanical equilibrium, i.e., no bulk motion:
+\begin{equation}
+\vec{u}_\alpha = -\frac{k_{r,\alpha} \vec{\vec{k}}}{\mu_\alpha} \left(\nabla p_\alpha - \rho_\alpha \vec{g}\right)\equiv 0 \, .
+\end{equation}
+This condition implies that pressure gradients must equal gravity force per unit volume:
+\begin{equation}\label{eq: p=rhogh}
+\nabla p_\alpha = \rho_\alpha \vec{g} \, .
+\end{equation}
+Consequently, pressure is a function of depth only:
+\begin{equation}\label{eq: p=rhogh}
+\int_{p_\alpha^0}^{p_\alpha} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'},T\right)} = g\left(z - z_0\right) \, .
+\end{equation}
+
+Black-oil equations then become:
+\begin{itemize}
+\item water component:
+\begin{equation}\label{eq: Sw2}
+\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV = \rho_w^{ST}Q_w^{ST} \, ;
+\end{equation}
+
+\item oil component:
+\begin{equation}\label{eq: So2}
+\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_o Y_o S_o \phi dV = \rho_o^{ST}Q_o^{ST} \, ;
+\end{equation}
+
+\item gas component:
+\begin{equation}\label{eq: Sg2}
+\frac{\partial}{\partial t} \int\limits_{\Omega} \left( \rho_o Y_g S_o + \rho_g S_g\right) \phi dV = \rho_g^{ST}Q_g^{ST} \, .
+\end{equation}
+\end{itemize}
+%
+
+\subsubsection{Material balance as an equation for reservoir pressure}
+
+%Equations from \eqref{eq: Sw4} to \eqref{eq: S14} constitute the most general form of material balance equation.
+So far, the only simplification made to black-oil equations was the absence of fluid motion, allowing pressure to be computed from Eqs.\eqref{eq: capillary_ow}, \eqref{eq: capillary_og} and \eqref{eq: p=rhogh}, and equations to be integrated in the entire reservoir volume.
+
+We now look to solve Eqs. \eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5} for the averaged quantities.
+
+A single equation for pressure may be obtained by substituting Eqs.~\eqref{eq: So5}, \eqref{eq: Sg5} and \eqref{eq: Sw5} into \eqref{eq: S15}.
+The resulting equation allows pressure to be calculated from given production ($F_p$) and injection ($F_i$) volumes:
+\begin{equation}\label{eq: MBE}
+F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p2}-V_{p1}}_{\text{porous volume change } (\Delta V_p)}
+\end{equation}
+or
+\begin{equation}\label{eq: MBE_short}
+F_p - F_i= E - \Delta V_p \, ,
+\end{equation}
+
+where newly introduced terms are defined in Table~\ref{tab: MBE}. Eq.~\eqref{eq: MBE} is commonly referred to as the Material Balance Equation (MBE).
+
 \subsubsection{Definition of a reference depth for pressure}
@@
 
-This approximation may be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase
+This approximation can be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase
+\begin{equation}
+B_o \approx B_o^0 + \left(\frac{\partial B_o}{\partial p}\right)_T\frac{\partial p}{\partial z}dz
+\end{equation}
 \begin{equation}\label{eq: grado}
@@
 In such cases, using an averaged value for the solubility ratio may be worth the accuracy gain in computing dissolved gas volume.
-
-\subsubsection{Material balance as an equation for reservoir pressure}
-
-%Equations from \eqref{eq: Sw4} to \eqref{eq: S14} constitute the most general form of material balance equation.
-So far, the only simplification made to black-oil equations was the absence of fluid motion, allowing pressure to be computed from Eqs.\eqref{eq: capillary_ow}, \eqref{eq: capillary_og} and \eqref{eq: p=rhogh}, and equations to be integrated in the entire reservoir volume.
-
-We now look to solve Eqs. \eqref{eq: Sw5}, \eqref{eq: So5} and \eqref{eq: Sg5} for the averaged quantities.
-
-A single equation for pressure may be obtained by substituting Eqs.~\eqref{eq: So5}, \eqref{eq: Sg5} and \eqref{eq: Sw5} into \eqref{eq: S15}.
-The resulting equation allows pressure to be calculated from given production ($F_p$) and injection ($F_i$) volumes:
-\begin{equation}\label{eq: MBE}
-F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p2}-V_{p1}}_{\text{porous volume change } (\Delta V_p)}
-\end{equation}
-or
-\begin{equation}\label{eq: MBE_short}
-F_p - F_i= E - \Delta V_p \, ,
-\end{equation}
-
-where newly introduced terms are defined in Table~\ref{tab: MBE}. Eq.~\eqref{eq: MBE} is commonly referred to as the Material Balance Equation (MBE).
 

### 2016-07-14 GITT/auxiliar2.py

@@ -1,0 +1,70 @@
+##
+## Solves 1D transient heterogeneous diffusion using GITT
+##
+import scipy.special
+#from sympy import *
+import numpy as np
+import scipy as sp
+from pylab import *
+
+class classGITT():
+	def __init__(self,x,w,k,N):
+		self.x=x
+		self.dx=np.gradient(x)
+		self.w=w
+		self.k=k
+		self.N=N
+
+	def omega(self,i):
+		if i==0:
+			return np.ones(len(self.x))
+		else:
+			return np.sqrt(2)*np.cos(i*pi*self.x)
+
+	def calcA(self):
+		self.A=np.zeros([self.N,self.N])
+		for n in range(self.N):
+			omegan=self.omega(n)
+			for m in range(self.N):
+				omegam=self.omega(m)
+				domegan=np.gradient(self.k*np.gradient(omegan,self.dx),self.dx)
+				self.A[n,m]=np.trapz(omegam*domegan,x=self.x)
+	
+	def calcB(self):
+		self.B=np.zeros([self.N,self.N])
+		for n in range(self.N):
+			for m in range(self.N):
+				self.B[m,n] = -1.*np.trapz(self.omega(n)*self.w*self.omega(m),x=self.x)
+
+	def solveEigenProb(self):
+		self.calcA()
+		self.calcB()
+		self.eigenvalues, self.eigenvectors = sp.linalg.eig(self.A,b=self.B)
+		self.eigenvalues=self.eigenvalues.real
+		ind=np.argsort(self.eigenvalues)
+
+		self.eigenvalues=self.eigenvalues[ind]
+		self.eigenvectors=self.eigenvectors[ind]
+
+
+	def Psi(i):
+		Psi=0.0
+		for n in range(self.N):
+			Psi = Psi + self.omega(n)*self.eigenvectors[i][n] 
+		return Psi
+
+x=np.linspace(0.,1.,500)
+beta=1.0
+k0=1.0
+w0=10.0
+
+k=k0*exp(2.0*beta*x)
+w=w0*exp(2.0*beta*x)
+khe=k0*(5.0+4.0*sin(2*pi*x))
+N=20
+
+homog=classGITT(x=x,w=w,k=k,N=N)
+homog.solveEigenProb()
+
+print(np.sqrt(homog.eigenvalues)[1:11:1])
+

### 2016-09-11 master/ppt/slides.tex
added master files

\documentclass[bars,mathserif]{beamer}

\usetheme{Dresden}
\usepackage{graphicx}
\usepackage[utf8x]{inputenc}
\usepackage{default}

\usepackage{amssymb,amsmath}
\newcommand{\bv}[1]{\mathbf{#1}} 
\newcommand{\req}[1]{#1^{*}} 
\usepackage[brazilian]{babel}

\setbeamertemplate{footline}[page number]
\setbeamercovered{invisible}
\setbeamertemplate{navigation symbols}{}
\setbeamersize{text margin left=.5cm,text margin right=.5cm}

\title[Eulerian-Lagrangian Simulation of a Turbulent Evaporating Spray]{Eulerian-Lagrangian Simulation of a Turbulent Evaporating Spray}
\author[candidate,advisor]{Rodrigo B. Piccinini}
\institute[ITA]
{
\medskip{\emph{e-mail: rbpiccinini@gmail.com}} \\ \vspace{10pt}
  \normalsize Apresentação de Tese de Mestrado \\ \vspace{10pt}

\includegraphics[width=0.2\textwidth]{./imgs/ita_rgb.jpg} \hspace{20pt} \includegraphics[width=0.1\textwidth]{./imgs/lcft.jpg} \\ \vspace{5pt}
  \footnotesize  Instituto Tecnológico de Aeronáutica \\
  Programa de Engenharia Aeronáutica e Mecânica\\
  Área de Aerodinâmica, Propulsão e Energia \\ \vspace{10pt}
\vspace{5pt}
\footnotesize Novembro, 2011.

}
\date{}
%
% Adicionar slide com o algoritmo do PISO!
%
%
\begin{document}
%
{
%   \usebackgroundtemplate{\centering\includegraphics[width=\paperwidth]{imgs/ita_cinza.png}}
  \begin{frame}[plain]
    \titlepage
  \end{frame}
} 
%
\begin{frame}
   \frametitle{Conteúdo}
   \tableofcontents
\end{frame}

\section{Introdução}
%
\begin{frame}

\begin{block}{Definição:}
\includegraphics[width=\textwidth]{./imgs/visit/spray.png}
\end{block}

\begin{block}{Motivação:}
\begin{itemize}
 \item Aplicações em geração de energia e propulsão;
 \item Escoamento multifásico e turbulento com transferências de momento, calor e massa entre as fases.
\end{itemize}
\end{block}
\begin{block}{Tratamento das fases:}
\begin{itemize}
 \item Método Euleriano-Langragiano.
\end{itemize}
\end{block}
\end{frame}

\begin{frame}{Objetivos}
\begin{itemize}
  \item Implementar um modelo de escoamento de baixo Mach.
  \item Aplicar modelos de troca de massa, momento e energia entre as fases e um modelo de viscosidade turbulenta para descrever o jato do spray.
  \item Comparar quantidades calculadas como velocidade e diâmetro das gotas, velocidade do gás e propriedades da turbulência com experimentos relatados na literatura.
\end{itemize}

\end{frame}

%
\section{Equações}
\frame{\tableofcontents[currentsection]}
%
\subsection{A Fase Contínua}
%
\begin{frame}
\frametitle{Obtendo as equações da fase contínua:}
\begin{enumerate}
 \item<1-> Equações (na forma compressível) para uma mistura de gases ideais \pause + \alert{termos fontes do spray};
 \item<3-> Aproximação para baixo Mach;
 \item<5-> Escoamento médio (Reynolds e Favre);
 \item<7-> Modelo de turbulência;
\end{enumerate}
%
\begin{overprint}
\onslide<2>
\footnotesize
\centering
\begin{tabular}{|l|}
\hline
\textit{Two-way coupling} ou acoplamento em duas vias \\
\hline 
\end{tabular}

\onslide<4>
\footnotesize
\centering
\begin{tabular}{|l|}
\hline \\
Séries de potências em $\xi = \sqrt{\gamma_{\infty}} M$:\\ \\
$p = p_{0} + p_{1}\xi + p_{2}\xi^2 + O\left( \xi^3 \right)$\\
$Y_k = Y_{k,0} + O\left( \xi \right) $\\
$\bv{U} = \bv{U}_{0} + O\left( \xi \right)$ \\
$T = T_{0} + O\left( \xi \right)$ \\ \\
\hline 
\end{tabular}
%
\onslide<8>
\footnotesize
\centering
\vspace{10pt}
\begin{tabular}{|l|}
\hline
Modelo de viscosidade turbulenta (k-epsilon): \\
2 EDPs para $k$ e $\epsilon$ \\
1 Equação Algébrica para $\mu_t$ \\
Acoplamento com o spray em uma via (\textit{one-way coupling}) \\
\hline 
\end{tabular}
\end{overprint}
\end{frame}
%
\begin{frame}
\begin{block}{Equação da Continuidade:}
\begin{equation*}
  \frac{\partial \bar{\rho}}{\partial t} + \nabla \cdot \left( \bar{\rho}
\tilde{\bv{U}} \right) =  \alert{\bar{S}_m}
\end{equation*}
\end{block}
\end{frame}
%
\begin{frame}
\begin{block}{Equação das Espécies Químicas: Acetona e Ar.}
\begin{itemize}
 \pause
 \item<1-> Acetona - espécie escassa:
\begin{equation*}
 \frac{\partial \bar{\rho} \tilde{Y}_{ac} }{\partial t} + \nabla
\cdot\left(\bar{\rho} \tilde{\bv{U}} \tilde{Y}_{ac} \right) = \nabla \cdot \left[ \bar{\rho} \left( \alert{\nu + \nu_T} \right)
\nabla \tilde{Y}_{ac} \right] +\alert{\bar{S}_{Yac}}
\end{equation*}
\pause
\item<4-> Ar - espécie diluente:
\begin{equation*}
\tilde{Y}_{air} = 1- \tilde{Y}_{ac}
\end{equation*}
\end{itemize}
\end{block}
%
\onslide<3->
\footnotesize
\begin{tabular}{|l|}
\hline \\
Lei de Fick para difusão, $\bv{V}_{ac} Y_{ac} = -\mathcal{D} \nabla Y_{ac}$.\\
Número de Schmidt unitário, $Sc=\nu/\mathcal{D} = 1$. \\ \\
\hline 
\end{tabular}
%
%
\end{frame}
%
\begin{frame}
%\setbeamercovered{transparent}
\frametitle{Equação do Momento}
Em quantidades não dimensionais. 
%
\onslide<4->
\hspace{10pt}
\footnotesize
\centering
\begin{tabular}{|l|}
\hline \\
$p = p_{0} + p_{1}\xi + p_{2}\xi^2 + O\left( \xi^3 \right)$ \\ \\
\hline
\end{tabular}
%
\pause
\begin{block}<2->{Formulação Compressível:}
\begin{equation*}
 \frac{\partial \rho \bv{U} }{\partial t} + \nabla \cdot \left( \rho \bv{U}
\bv{U} \right) = -\alert{\frac{1}{\Gamma M^2} \nabla p} + \frac{1}{Re} \nabla \cdot
\bv{\tau} +\frac{\rho}{Fr^2} \frac{\bv{g}}{|\bv{g}|}+\alert{\bv{S}_{mom}}
\end{equation*}
\end{block}

\begin{block}<3->{Formulação de baixo Mach:}
\begin{equation*}
 \frac{\partial \rho_0 \bv{U}_0 }{\partial t} + \nabla \cdot \left( \rho_0
\bv{U}_0 \bv{U}_0 \right) = -\alert{\nabla p_2} + \frac{1}{Re} \nabla \cdot \bv{\tau}_0
+\frac{\rho_0}{Fr^2} \frac{\bv{g}}{|\bv{g}|}+\alert{\bv{S}_{mom}}
\end{equation*}
\begin{equation*}
\nabla p_0 \equiv 0 \Longrightarrow p_0 = \alert{p_0 (t)}
\end{equation*}
\end{block}
%
\end{frame}
%
\begin{frame}
\begin{block}<1->{Decomposição da pressão:}
\begin{itemize}
 \item<2-> $p_0 (t)$ é utilizada na equação de estado: 
\begin{equation*}
\alert{p_0} W = \bar{\rho} R \tilde{T}
\end{equation*}
 \item<3-> $p_2 (\bv{x},t)$ é utilizada na equação do momento:
 \begin{equation*}
  \frac{\partial \bar{\rho} \tilde{\bv{U}} }{\partial t} + \nabla \cdot \left(
\bar{\rho} \tilde{\bv{U}} \tilde{\bv{U}} \right) = -\alert{\nabla \bar{p}_2} +  
 \nabla \cdot\overline{\bv{\tau}} +\bar{\rho}\bv{g}+\bar{\bv{S}}_{mom}
 \end{equation*}
Em quantidades dimensionais e para o escoamento médio.
\end{itemize}
\end{block}
%
\onslide<4->
\footnotesize
\begin{tabular}{|l|}
\hline \\
$\bar{\bv{\tau}}$ é o tensor de tensões de cisalhamento com $\mu_{eff}=\mu+\mu_T$.\\ \\
\hline 
\end{tabular}
%
\end{frame}

\begin{frame}
\frametitle{Equação da Entalpia:}
Em quantidades não dimensionais. 
%
\pause
\begin{block}<2->{Formulação Compressível:}
\begin{equation*}
\begin{split}
\frac{\partial \rho h_s}{\partial t} +  \nabla \cdot \left(\rho \bv{U} h_s
\right) &= \frac{\Gamma -1}{\Gamma}\frac{Dp}{Dt} + \nabla \cdot \bv{J}_s \\
&+ \alert{M^2 \left( \Gamma-1 \right) \left( \frac{\bv{\tau} : \nabla\bv{U}}{Re} \right)} + \alert{S_{hs}}
\end{split}
\end{equation*}
\end{block}

\begin{block}<3->{Formulação de baixo Mach:}
\begin{equation*}
\frac{\partial \rho_0 h_{s0}}{\partial t} +  \nabla \cdot \left(\rho_0 \bv{U}_0
h_{s0} \right) = \frac{\Gamma -1}{\Gamma}\frac{Dp_0}{Dt} + \nabla \cdot
\bv{J}_{s0} + \alert{S_{hs}}
\end{equation*}
\end{block}
%
\end{frame}
%
\begin{frame}
\frametitle{Equação da Entalpia:}
Em quantidades dimensionais e para o escoamento médio.
\begin{block}<1->{}
 \begin{equation*}
\frac{\partial \bar{\rho} \tilde{h}_{s}}{\partial t} + \nabla \cdot
\left(\bar{\rho} \tilde{\bv{U}} \tilde{h}_{s} \right) = \nabla \cdot \bar{\bv{J}}_s + \bar{S}_{hs}
\end{equation*} 
\begin{equation*}
\bar{\bv{J}}_s =  \bar{\rho} \left( \alpha + \nu_T\right) \nabla \tilde{h}_s 
\end{equation*}
\end{block}

\vspace{10pt}
\onslide<2->
\begin{block}{Simplificações:}
 \begin{itemize}
 \item $Dp_0/Dt\approx 0$;
 \item $h_{s,ac} \approx h_{s,ar}$;
 \item $Pr_T = 1$.
 \end{itemize}

\end{block}

\end{frame}

%
\subsection{A Fase Dispersa}
\begin{frame}
\frametitle{A Fase Dispersa}
\begin{itemize}
 \item<2-> As gotas são partículas de diâmetro D;
 \item<3-> Equações em referencial Lagrangiano (3 EDOs para cada gota);
 \item<4-> Termodinâmica: $m_d,\ p_0,\ T_d$;
 \item<5-> Cinemática: posição ($\bv{x}_d$), velocidade translacional ($\bv{U}_d$).
\end{itemize}

\end{frame}

%
\begin{frame}
\frametitle{A Fase Dispersa}
\begin{block}<1->{Equação do Momento:}
\begin{equation*}
m_d \frac{d\bv{U}_d}{dt} =-\rho\frac{\pi D^2}{8}C_D | \bv{U}_{slip}| \bv{U}_{slip} + m_d \bv{g}
\end{equation*} 
\end{block}
\begin{block}<2->{Dispersão das Gotas:}
 \begin{equation*}
 \bv{U}_{slip} = \bv{U_d} - \underbrace{(\tilde{\bv{U}}+\bv{U}'')}_{\text{g\'as}}
\end{equation*}
\end{block}

\onslide<3->
\footnotesize
\begin{tabular}{|l|}
\hline \\
$|\bv{U}''|$ é uma variável aleatória dada por $\mathcal{N} (0,2/3 k^2)$.\\ \\
\hline 
\end{tabular}
\end{frame}
%
\begin{frame}
\begin{columns}
 \begin{column}{0.5\textwidth}
  \begin{block}<1->{Equação da Energia:}
 \begin{equation*}
 m_d \frac{dh_d}{dt} = \frac{d m_d}{dt} L_v \left(T_d\right)+Q_d 
\end{equation*}
\end{block}
\begin{block}<2->{Transferência de Calor:}
Equação da energia para o vapor na vizinhança da gota:
\begin{equation*}
Q_d = \pi D \kappa Nu \left( \tilde{T}-T_d \right)  \alert{\frac{z}{e^z-1}}
\end{equation*}
\end{block}
 \end{column}
\begin{column}{0.5\textwidth}
\begin{block}<2->
\centering
 \includegraphics[width=\textwidth]{./imgs/spray_energy.png}
\end{block}  
 \end{column}
\end{columns}
%
\onslide<3->
\footnotesize
\begin{tabular}{|l|}
\hline \\
Correlação de Ranz \& Marshall:  $Nu= 2.0 +0.6 Re^{1/2} Pr^{1/3}$.\\ \\
\hline 
\end{tabular}
%
\end{frame}
%
\begin{frame}
\begin{block}{Transferência de Massa}
 \begin{itemize}
  \item<2-> Líquido e vapor em equilíbrio na superfície da gota.
  \begin{equation*}
  Y_{ac,s} = \frac{p_v \left(T_d\right)}{p_0}\frac{W_{ac}}{W} 
  \end{equation*}
  \item<3-> Equações da continuidade e das espécies para o vapor na vizinhança da gota:
\begin{equation*}
 \frac{dm_d}{dt} = -2\pi D \mathcal{D}_{ac} \rho_{ac} ln \left(
\frac{1-Y_{ac,\infty}}{1-Y_{ac,s}} \right) Sh 
\end{equation*}
 \end{itemize}
\end{block}

\onslide<4->
\vspace{10pt}
\footnotesize
\begin{tabular}{|l|}
\hline \\
Correlação de Ranz \& Marshall:  $ Sh = 2.0 +0.6 Re^{1/2} Sc^{1/3}$.\\ \\
\hline 
\end{tabular}
\end{frame}
%
\section{Cálculo Numérico}
\frame{\tableofcontents[currentsection]}
%
\begin{frame}[fragile] % Notice the [fragile] option %
\frametitle{Cálculo Numérico}
\begin{block}<2->{Método de Volumes Finitos}
\begin{itemize}
  \item<2-> Discretização da geometria;
  \item<2-> Discretização das equações;
  \item<2-> Algoritmo de solução do sistema de equações;
\end{itemize}
\end{block}

\onslide<3->
\begin{figure}\centering
\includegraphics[width=0.35\textwidth]{./imgs/foam_logo.png}\footnote<3->{http://www.openfoam.org}
\end{figure}

\end{frame}
%
\begin{frame}
\frametitle{Discretização do Espaço}
\begin{itemize}
  \item Malha ortogonal 2D (axissimétrica);
  \item Malha fina: $14.800$ células ($74 \times 200$);
  \item Malha grossa: $5.460$ células ($42 \times 130$);
\end{itemize}
\centering \includegraphics[width=\textwidth]{./imgs/grid.png}
\end{frame}
%
\begin{frame}
\frametitle{Discretização das Equações}
\begin{block}<1->{}
\begin{equation*}
 \underbrace{\frac{\partial \rho \phi}{\partial t}}_{\text{derivada temporal}} + \underbrace{\nabla \cdot \left(\rho \bv{U} \phi \right)}_{\text{advecção}} - \underbrace{\nabla \cdot \left( \rho \Gamma_{\phi} \nabla \phi \right)}_{\text{difusão}} = \underbrace{S\left(\phi\right)}_{\text{fontes}}
\end{equation*}
\end{block}
\begin{block}<2->{}
\begin{itemize}
  \item<2-> Tempo: Euler implícito;
  \item<3-> Advecção: teorema de Gauss + interpolação \textit{upwind};
  \item<4-> Difusão: teorema de Gauss + interpolação linear + diferença centrada;
  \item<5-> Fontes: tratamento explícito.
\end{itemize}
\end{block}
\end{frame}
%
\begin{frame}
\frametitle{O Algoritmo de Solução}
\begin{block}<1->{}
\begin{itemize}
  \item<2-> Sistemas lineares segregados;
  \item<3-> Acoplamento das variáveis: algoritmo PISO;
\end{itemize}
\end{block}
\end{frame}
%
\section{Resultados}
\frame{\tableofcontents[currentsection]}

\subsection{O Jato Turbulento}
\frametitle{Perfis Auto-Similares do Jato Turbulento (fase gasosa)}
\begin{frame}
\begin{overprint}
\onslide<1>
\begin{figure}
\centering
\includegraphics[width=0.85\textwidth]{../thesis/figuras/selfsimilar/selfsimilar_U.png}
\caption{Perfis auto-similares de velocidade adimensional ($\hat{U}$).}
\end{figure}

\onslide<2>
\begin{figure}
\centering
\includegraphics[width=0.85\textwidth]{../thesis/figuras/selfsimilar/selfsimilar_k.png}
\caption{Perfis auto-similares de energia cinética turbulenta adimensional ($\hat{k}$).}
\end{figure}

\onslide<3>
\begin{figure}
\centering
\includegraphics[width=0.85\textwidth]{../thesis/figuras/selfsimilar/selfsimilar_nut.png}
\caption{Perfis auto-similares da viscosidade turbulenta adimensional ($\hat{\nu}_T$). Superestimação: $\hat{\nu}_T=0.047 > \underbrace{0.029}_{\text{(POPE,2000)}}$.}
\end{figure}
\end{overprint}
\end{frame}
%
\begin{frame}
 \begin{figure}[!htb]
 \centering
\begin{tabular}{c}
 \includegraphics[width=0.85\textwidth]{../thesis/figuras/selfsimilar/selfsimilar_num_exp.png}
\end{tabular}
 \caption{Comparação do perfis numérico e experimental (Chen et al., 2006) de $\hat{U}$.}
\end{figure}
\end{frame}
%
\begin{frame}
 \begin{figure}[!htb]
 \centering
\begin{tabular}{c}
 \includegraphics[width=0.75\textwidth]{../thesis/figuras/selfsimilar/selfsimilar_spread.png}
\end{tabular}
 \caption{Raio médio numérico e experimental (Chen et al., 2006).}
\end{figure}

%\vspace{5pt}
\footnotesize
\begin{tabular}{|l|}
\hline \\
$(d y_{1/2}/dx)_{\text{num}}$ $\sim 20\%$ maior do que $(d y_{1/2}/dx)_{\text{exp}}$.\\ \\
\hline 
\end{tabular}
\end{frame}
%
\subsection{Velocidade das Fases}
\begin{frame}[plain]{Velocidade Axial Média da Fase Gasosa ($\tilde{U}_x$)}
 \begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.45\textwidth]{../thesis/figuras/Ux/Ux_gas5.png} & \includegraphics[width=0.45\textwidth]{../thesis/figuras//Ux/Ux_gas10.png} \\
\includegraphics[width=0.45\textwidth]{../thesis/figuras/Ux/Ux_gas15.png} & \includegraphics[width=0.45\textwidth]{../thesis/figuras/Ux/Ux_gas25.png}  
\end{tabular}
 \caption{Velocidade da fase gasosa. Numérico e experimental (Chen et al., 2006).}
 \label{fig: Ux_gas}
\end{figure}
\end{frame}

\begin{frame}[plain]{{Flutuação da Velocidade da Fase Gasosa ($< U''>$)}}
 \begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.5\textwidth]{../thesis/figuras/UU/UUx5.png} & \includegraphics[width=0.5\textwidth]{../thesis/figuras/UU/UUx10.png} \\
  \includegraphics[width=0.5\textwidth]{../thesis/figuras//UU/UUx20.png} & \includegraphics[width=0.5\textwidth]{../thesis/figuras/UU/UUx25.png}
\end{tabular}
 \caption{\footnotesize Numérico $<U''>$ e experimental $<U''_x>$ e $<U''_y>$ (Chen et al., 2006).}
\end{figure}
\end{frame}

\begin{frame}[plain]{Velocidade das Gotas}
 \begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.45\textwidth]{../thesis/figuras/Ux/Ux5.png} & \includegraphics[width=0.45\textwidth]{../thesis/figuras//Ux/Ux10.png} \\
\includegraphics[width=0.45\textwidth]{../thesis/figuras//Ux/Ux20.png} & \includegraphics[width=0.45\textwidth]{../thesis/figuras/Ux/Ux25.png}
\end{tabular}
 \caption{ \footnotesize Velocidade das gotas da classe ($D \in [10,20] \mu m$). Numérico e experimental (Chen et al., 2006).}
\end{figure}
\end{frame}

% \begin{frame}
% \begin{figure}[!htb]
%  \centering
% \begin{tabular}{cc}
%  \includegraphics[width=0.5\textwidth]{../thesis/figuras/dispersion/class0_5.png} & \includegraphics[width=0.5\textwidth]{../thesis/figuras/dispersion/class4_5.png} \\
%  \includegraphics[width=0.5\textwidth]{../thesis/figuras/dispersion/class0_10.png} & \includegraphics[width=0.5\textwidth]{../thesis/figuras/dispersion/class4_10.png} \\
%  \includegraphics[width=0.5\textwidth]{../thesis/figuras/dispersion/class0_25.png} & \includegraphics[width=0.5\textwidth]{../thesis/figuras/dispersion/class4_25.png} \\
% \end{tabular}
%  \caption{Scatter plot of the numerical radial fluctuating velocity of droplets ($U''_{y}$) in two different size classes: $D< 10\ \mu m$ on the left or (a), (c) and (e) labels; and $D> 30\ \mu m$ on the right or (b), (d) and (f) labels. The dispersion in the radial velocity is higher for the smaller droplet class and in the vicinity of the nozzle exit.}
% \label{fig: jointUV}
% \end{figure}
% \end{frame}

\subsection{Vazão Mássica das Gotas e do Vapor}
\begin{frame}{Vazão Mássica de Líquido}
\begin{figure}[!htb]
\centering
  \includegraphics[width=0.8\textwidth]{../thesis/figuras/drop_mflux.png}
\caption{Vazão de massa na fase líquida. Numérico e experimental (Chen et al., 2006).}
\label{fig: droplet_flux}
\end{figure} 
\end{frame}

\begin{frame}[plain]{Fluxo Mássico de Vapor}  % notice the fragile option, since the body
\begin{figure}[!htb]
 \centering
\begin{tabular}{ccc}
 \includegraphics[width=0.45\textwidth]{../thesis/figuras/mvapor/mvapor5.png} & \includegraphics[width=0.45\textwidth]{../thesis/figuras/mvapor/mvapor10.png} \\
\includegraphics[width=0.45\textwidth]{../thesis/figuras/mvapor/mvapor15.png} & \includegraphics[width=0.45\textwidth]{../thesis/figuras//mvapor/mvapor20.png}
\end{tabular}
 \caption{\footnotesize Fluxo mássico de vapor: $\dot{m}''_{ac}=\rho \bar{Y}_{ac} \tilde{U}_x$. Numérico e experimental (Chen et al., 2006).}
\end{figure}
\end{frame}

\begin{frame}[plain]{Diâmetro de Sauter}  % notice the fragile option, since the body
\begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.5\textwidth]{../thesis/figuras/smd/smd5.png} &
 \includegraphics[width=0.5\textwidth]{../thesis/figuras/smd/smd15.png} \\
 \includegraphics[width=0.5\textwidth]{../thesis/figuras/smd/smd25.png} \\
\end{tabular}
 \caption{\footnotesize Diâmetro de Sauter (SMD). Numérico e experimental (Chen et al., 2006).}
\end{figure}
\end{frame}
%
\begin{frame}[plain]
\frametitle{Temperatura da Fase Gasosa}
\includegraphics[width=\textwidth]{./imgs/visit/T.png}
\end{frame}
%
\begin{frame}[plain]
\frametitle{Densidade da Fase Gasosa}
\includegraphics[width=\textwidth]{./imgs/visit/rho.png}
\end{frame}
%
\begin{frame}[plain]
\frametitle{Energia Cinética Turbulenta}
\includegraphics[width=\textwidth]{./imgs/visit/k.png}
\end{frame}
%
\begin{frame}[plain]
\frametitle{Viscosidade Turbulenta}
\includegraphics[width=\textwidth]{./imgs/visit/mut.png}
\end{frame}
%
\begin{frame}[plain]
\frametitle{Velocidade das Gotas}
\begin{center}
\includegraphics[height=\textheight]{./imgs/visit/drop_U.png}
\end{center}
\end{frame}
%
\begin{frame}[plain]
\frametitle{Temperatura das Gotas}
\begin{center}
\includegraphics[height=\textheight]{./imgs/visit/drop_T.png}
\end{center}
\end{frame}
%

\section{Conclusões}
\frame{\frametitle{Resumo} \tableofcontents}
\frame{\tableofcontents[currentsection]}

\begin{frame}

\begin{block}{Formulação}
  \begin{itemize}
  \item $M \rightarrow  0 \Longrightarrow$  métodos numéricos para escoamento incompressível;
  \item O método Euleriano-Lagrangiano é uma \textit{boa} descrição computacional para a fase dispersa.
 \end{itemize}
\end{block}

\begin{block}{Numérico $\times$ Experimental:}
 \begin{itemize}
  \item $\bv{\tilde{U}}$, $\bv{U}_d$ e $\dot{m}_d$ \textbf{sub}estimados;
  \item $\text{k-epsilon} \Longrightarrow \mu_T$ \textbf{super}estimado.
 \end{itemize}
\end{block}

\begin{block}{Trabalhos Futuros:}
 \begin{itemize}
  \item $\dot{h}_{d}$, $\dot{m}_d \Longleftarrow \text{turbulência}$;
  \item $Sc,\ Pr \neq 1$;
  \item Tratar termos fontes de forma \textit{mais implícita}.
 \end{itemize}
\end{block}
\end{frame}

\begin{frame}
\centerline{Obrigado pela atenção.}
\end{frame}
% End of slides
\end{document}

### 2016-09-11 master/thesis/abstract.tex
added master files

This work consists in a numerical simulation of a spray jet: a two-phase flow composed by a gaseous and a liquid phase. The gaseous phase is treated as a continuous medium, and the liquid phase is treated as a dispersed phase. An asymptotic approximation of zero Mach number was applied to the gaseous phase in order to account for density variations due to temperature gradients without dealing with extra complexities of the fully compressible flow formulation. The effects of turbulence on the gas flow were modeled using the concept of turbulent viscosity determined by a system of two partial differential equations, the so called k-epsilon model. The liquid dispersed phase was modeled as point-particles, to whom thermodynamic properties and 0-d models for momentum, heat and mass transfers were assigned. The gaseous phase was discretized and numerically solved using the finite volume method. The numerical results were compared to measurements and a reasonable prediction was found. It was verified, though, underestimations of droplet velocity and evaporation rate.

### 2016-09-11 master/thesis/acknowledgements.tex
added master files

Eu agredeço à minha família pelo apoio e pelas oportunidades que tive.

Agradeço ao Prof. Marcelo J.S. de Lemos pela orientação e pela ajuda com os recursos computacionais. Para mim, foi um privilégio trabalhar sob orientação de um pesquisador de sucesso exemplar.

Agradeço à Sygma Motors, nominalmente a Marcos Langeani, onde iniciei minha vida profissional e a quem sou grato pelas oportunidades e pelo aprendizado oferecidos.

Agradeço ao Prof. Francisco José de Souza pelo aprendizado, pelo incentivo ao estudo da mecânica dos fluidos e pelo exemplo profissional.

Agradeço aos ex-colegas Carla Fernandes, Clayton Zabeu, Edilson Viana, Felipe Tannus Dorea, Gustavo Hindi, José Eduardo de Oliveira, Marcelo Andreotti e Oswaldo França Jr pelo companheirismo e pelo aprendizado.

Agradeço ao Prof. Flávio Luiz Cardoso Ribeiro, procurador para os trâmites burocráticos do ITA, colega no curso de graduação em engenharia aeronáutica e um dos meus melhores amigos, pelo apoio à conclusão dos estudos do mestrado.

Agradeço à Blandina Oliveira pela ajuda com a impressão e com a distribuição das cópias da dissertação.

Agradeço às comunidades de código aberto que mantêm softwares de qualidade disponibilizados sem custo pelo simples espírito de colaboração.

Este trabalho começou na quase sub-tropical São José dos Campos e foi finalizado em Salvador, no coração do Brasil.

### 2016-09-11 master/thesis/apendiceB.tex
added master files

\section{Extra Figures}\label{appendix 2}
\begin{figure}[h]
 \centering
 \includegraphics[height=0.7\textheight]{./figuras/appA2/visit_Td.png}
 % k_bc.png: 800x493 pixel, 100dpi, 20.32x12.52 cm, bb=
 \caption{Temperature of droplets. The computational domain is only half of the shown above, it is here mirrored for visualization purpose.}
 \label{fig: dropT}
\end{figure}

\clearpage
\begin{figure}
 \centering
 \includegraphics[height=0.9\textheight]{./figuras/appA2/visit_U.png}
 \caption{Magnitude of droplet velocities. The computational domain is only half of the shown above, it is here mirrored for visualization purpose.}
 \label{fig: dropU}
\end{figure}

\clearpage
\begin{figure}
\begin{center}
  \includegraphics[height=0.9\textheight]{./figuras/appA2/visit_T.png}
 \end{center}
 \caption{Gas mean temperature field $\tilde{T}$. The computational domain is only half of the shown above, it is here mirrored for visualization purpose.}
 \label{fig: field_T}
\end{figure}

\clearpage
\begin{figure}
\begin{center}
  \includegraphics[height=0.9\textheight]{./figuras/appA2/visit_Ux.png}
 \end{center}
\caption{Gas mean velocity magnitude $|\tilde{\bv{U}}|$. The computational domain is only half of the shown above, it is here mirrored for visualization purpose.}
 \label{fig: field_U}
\end{figure}

\clearpage
\begin{figure}
\begin{center}
  \includegraphics[height=0.9\textheight]{./figuras/appA2/visit_k.png}
 \end{center}
 \caption{Gas mean turbulent kinetic energy. The computational domain is only half of the shown above, it is here mirrored for visualization purpose.}
 \label{fig: field_k}
\end{figure}

\clearpage
\begin{figure}
\begin{center}
  \includegraphics[height=0.9\textheight]{./figuras/appA2/visit_aC3H6O.png}
 \end{center}
 \caption{Mean of acetone vapor mass concentration - $\tilde{Y}_{ac}$. The computational domain is only half of the shown above, it is here mirrored for visualization purpose.}
 \label{fig: aa}
\end{figure}

\FloatBarrier

### 2016-09-11 master/thesis/apendiceC.tex
added master files

\section{Equation Discretization and PISO Algorithm}\label{appendix 3}

\begin{figure}[h]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.55\textwidth]{./figuras/appA3/finite_volume.png}
\end{tabular}
 \caption{Finite volume discretization, reproduced from \cite{foamDev}.}
 \label{fig: fvm}
\end{figure}

A brief explanation of PISO algorithm and equation discretization as used in this work is here presented. The notation of equations follow OpenFOAM formulation explained in \cite{jasak}.

Superscripts \textbf{o} and \textbf{n} denote values from previous and current time-steps, respectively. \textbf{*} and \textbf{**} subscripts are used to denote intermediate fields computed before the final value of the time-step.

Subscripts \textbf{P} and \textbf{N} denote values stored in the center of owner and neighbor cells.  See Figure \ref{fig: fvm} for the geometric representation of two neighboring cells.

Subscript \textbf{f} indicates values interpolated from cell center to a face center. Advective terms were interpolated using the simple upwind scheme. The linear scheme was used for remaining interpolations. Setup of \verb|fvSchemes| input file is presented after PISO algorithm.

\clearpage
\textbf{Beginning of a new time-step:}
\begin{itemize}
  \item Evolve spray: move droplets and compute new spray source terms using old gas properties.

  \item Solve continuity equation for first density estimate $\bar{\rho}^{*}$:
    \begin{equation}
      \frac{\bar{\rho}^{*}-\bar{\rho}^{o}}{\Delta t} V_P +\sum_f F^{o} = S_m V_P \, ,
    \end{equation}
    where $F^{o}$ is the mass flow through cell faces from previous time-step:
  \begin{equation}
    F^{o}=\left(\bar{\rho}^{o}\tilde{\bv{U}}^{o}\right)_f\cdot \bv{S}_f \, .
  \end{equation} 
  This equation provides an explicit formulation for $\bar{\rho}^{*}$.

  \item Solve momentum equation for first velocity estimate $\bv{U}^{*}$ (momentum predictor):
    \begin{equation}
      \frac{\bar{\rho}^{*}\tilde{\bv{U}}^{*}-\bar{\rho}^{o}\tilde{\bv{U}}^{o}}{\Delta t} V_P + \sum_f F^{o}\tilde{\bv{U}}_f^{*} = \sum_f \bv{\tilde{\tau}}_f \cdot \bv{S}_f+\left(\bar{\rho}^{*}\bv{g}+\bv{S}_{mom} \right)V_P - \sum_f \bar{p}_{2,f}^{o}\bv{S}_{f}
    \end{equation}
    where 
    \begin{equation}
    \tilde{\bv{\tau}}=\underbrace{\sum_f \left( \mu_{eff}\nabla\tilde{\bv{U}}^{*} \right)_f \cdot \bv{S}_f}_{\text{implicit}} - \underbrace{\sum_f \mu_{eff,f}\left[\nabla \bv{\tilde{U}}^{o,T} -\frac{2}{3}\left(\nabla\cdot\tilde{\bv{U}}^{o}\right)\bv{I}\right]_f \cdot \bv{S}_f}_{\text{explicit}}
    \end{equation}
    The equation may be rewritten in a reduced semi-discretized form as:
    \begin{equation}
      \tilde{\bv{U}}_P^{*} = \frac{1}{a_P}\bv{H}(\tilde{\bv{U}}^{o})-\frac{1}{a_P}\nabla \bar{p}_2 \, .
    \end{equation}

  \item Solve species equation for $\tilde{Y}^{n}_k$:
  \begin{equation}
    \frac{\bar{\rho}^{*}\tilde{Y}_k^{n}-\bar{\rho}^{o}\tilde{Y}_k^o}{\Delta t} V_P +\sum_f F^{o}\tilde{Y}_{k,f}^{n} = \sum_f \left(\mu_{eff}\nabla\tilde{Y}_{k}^{n}\right)_f \cdot \bv{S}_f + S_{Yk} V_P
  \end{equation}

  \item Solve enthalpy equation for $\tilde{h}_{s}^{n}$:
  \begin{equation}
    \frac{\bar{\rho}^{*}\tilde{h}_s^{n}-\bar{\rho}^{o}\tilde{h}_s^o}{\Delta t} V_P +\sum_f F^{o}\tilde{h}_{s,f}^{n} = \sum_f \left(\mu_{eff}\nabla\tilde{h}_{s}^{n}\right)_f \cdot \bv{S}_{f}  + S_{hs} V_P
  \end{equation}
\end{itemize}

The computation of new pressure ($\bar{p}^n$) and velocity ($\bv{\tilde{U}}^n$) fields is performed in a iterative manner shown in next page.

\begin{tabular}{|p{0.95\textwidth}|} \hline
\textbf{PISO Loop (pressure-velocity coupling)} \\
\hline
Update $\rho$ with state equation using new temperature: $\bar{\rho}^{**} = p_0 \tilde{W} / R \tilde{T}^{n}$ \, ;\\
Compute second estimate of velocity ($\bv{U}^{**}$) without pressure contribution:
\begin{equation}
\tilde{\bv{U}}_P^{**} = \frac{1}{a_P}\bv{H}(\tilde{\bv{U}}^{*}) \, ;
\end{equation}

Update mass flow through cell faces using new density and velocity:
\begin{equation}
F^{*}=\left(\bar{\rho}^{**}\tilde{\bv{U}}^{**}\right)_f\cdot \bv{S}_f \, ;
\end{equation}

Solve pressure equation for the new dynamic pressure field $\bar{p}_2^n$:
\begin{equation}
  \sum_f \left[ F^{*} - \bv{S}_f\cdot\left(\frac{\bar{\rho}^{**}}{a_P}\nabla \bar{p}_2^{n} \right)_f \right] = S_m V_P \, ;
\end{equation}

Solve continuity equation for new density ($\bar{\rho}^n$) with $F^{*}$:
\begin{equation}
\frac{\bar{\rho}^{n}-\bar{\rho}^{o}}{\Delta t} V_P +\sum_f F^{*} = S_m V_P \, ;
\end{equation}

Compute the new velocity ($\tilde{\bv{U}}^{n}$) adding pressure contribution (momentum-corrector) and update mass fluxes:
\begin{equation}
\tilde{\bv{U}}^{n} = \tilde{\bv{U}}^{**} - \frac{1}{a_P}\sum_f \bar{p}_{2,f}^{n}\bv{S}_f \, ,
\end{equation}
\begin{equation}
 F^{n}=\left(\bar{\rho}^{n}\tilde{\bv{U}}^{n}\right)_f\cdot \bv{S}_f \, .
\end{equation}\\
\hline
\end{tabular}
PISO loop may be repeated inside the same time-step by setting $\tilde{\bv{U}}^{n}=\tilde{\bv{U}}^{o}$ and performing all steps again.

\clearpage
\begin{singlespace}
Setup of discretization schemes in \verb|fvSchemes| input file:
\begin{verbatim}
/*--------------------------------*- C++ -*----------------------------------*\
| =========                 |                                                 |
| \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox           |
|  \\    /   O peration     | Version:  1.7.1                                 |
|   \\  /    A nd           | Web:      www.OpenFOAM.com                      |
|    \\/     M anipulation  |                                                 |
\*---------------------------------------------------------------------------*/
FoamFile
{
    version     2.0;
    format      binary;
    class       dictionary;
    location    "system";
    object      fvSchemes;
}
// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //

ddtSchemes
{
    default         Euler;
}

gradSchemes
{
    default         none;

    //pEqn.H, velocity correction in PISO.
    grad(pd)         Gauss linear;

    // strain tensor
    grad(U)			Gauss linear;
}

divSchemes
{
    default         none;
    div(phi,rho)    Gauss upwind;
    div(phi,U)      Gauss upwind;
    div(phi,k)      Gauss upwind;
    div(phi,epsilon) Gauss upwind;
    div(phi,Yi_h)   Gauss upwind; // divergent scheme for Yk and hs
    div((muEff*dev2(grad(U).T()))) Gauss linear;
}

laplacianSchemes
{
    default         none;

    //UEq.H
    laplacian(muEff,U) Gauss linear uncorrected;

    //YEq.H
    laplacian(muEff,aC3H6O) Gauss linear uncorrected;
    laplacian(muEff,N2) Gauss linear uncorrected;
    laplacian(muEff,O2) Gauss linear uncorrected;
    laplacian(muEff,CO2) Gauss linear uncorrected;
    laplacian(muEff,H2O) Gauss linear uncorrected;

    //hsEq.H
    laplacian(alphaEff,hs) Gauss linear uncorrected;

    //pEq.H
    laplacian((rho*(1|A(U))),pd) Gauss linear uncorrected;

    //epsEq <kEpsilon.C>
    laplacian(DepsilonEff,epsilon) Gauss linear uncorrected;

    //kEq <kEpsilon.C>
    laplacian(DkEff,k) Gauss linear uncorrected;
}

interpolationSchemes
{
    default         linear;
    interpolate(HbyA) linear;
}

snGradSchemes
{
    default         uncorrected;
}

fluxRequired
{
    default         no;
    pd;
}
// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
\end{verbatim}
\end{singlespace}

### 2016-09-11 master/thesis/chap1.tex
added master files

\section{Motivation}

The study of spray jets is important for several technical applications: gas turbines, furnaces, automotive and rocket engines.  An extensive presentation of spray applications may be found in \cite{liu2000science}, but any equipment or transportation powered by the combustion of a liquid fuel has its design, efficiency and environmental impact challenged by the proper description of processes such as: liquid atomization, droplet transport, inter-phase transfers and vapor mixing with oxygen. Notwithstanding the expected complexity of all said, spray jets are often turbulent.

Studying the spray jet by means of computer modeling alongside with experimentation is still a growing activity in industry, specially in what concerns brazilian activities. One notable difficulty is the requirement of expensive computational and laboratorial resources as well as capable professionals to operate them, and there is always doubt about how predictive/descriptive computations/experiments may be.

Undoubtful are, though, the benefits that society might receive from the evolution of technical applications: fuel flexibility, lower emission levels, higher work efficiency, safer operability.

\section{Objective}
This thesis has the following objectives:
\begin{itemize}
  \item Implement a low Mach number formulation to allow modeling of gas density variations due to high temperature gradients while neglecting fluid compressibility.
  \item Review some existent models for droplet-gas heat, mass and momentum transfers under the framework of RANS\footnote{RANS: Reynolds-averaged Navier-Stokes equations.} modeling for gas turbulence;
  \item Compare computed quantities to measurements reported in the literature such as droplet velocity and diameter, vapor and liquid mass flux, gas velocity and turbulence properties.
\end{itemize}

\section{The Spray Jet and the Lagrangian Point-Particle Method}

A spray jet is a particular case of a dispersed multiphase flow originated by the instabilization of a liquid jet emerging in a gaseous atmosphere.  The spray is composed by a continuous gaseous phase and a dispersed phase of liquid droplets.

Differently from many multiphase flows where the description of phase interfaces is of great importance, in the dispersed regime it is the volume fraction of each phase the determinant factor of their interaction. The liquid volume fraction in spray jets may span from low or dilute ($\phi_{v,l} < 10^{-3}$) to high or dense ($\phi_{v,l} > 10^{-3}$).

In dilute sprays, the dynamics of liquid droplets is mainly governed by the gas turbulence. Modeling such sprays may be concentrated on modeling the gas effect on droplets and neglecting the droplet effect on the gas (one-way coupling) or taking both ways into account (two-way coupling). For dense sprays, however, the interaction between droplets (collision and coalescence) become important and must also be modeled (four-way coupling). A review on the computational approaches for different volume fractions is given in \cite{balachandar2010turbulent}.

This work deals with a dilute spray ($\phi_{v,l} \approx 2.1 \times 10^{-5}$). A two-way coupling modeling was used for averaged flow properties and a one-way coupling was used for turbulence modeling; that is, no direct influence of droplets is present in the turbulence model equations and they are identical to those for a pure gaseous flow.

In the modeling formulation used in this work, the droplets were treated as point-particles with assigned properties and instead of tracking an interface, the problem is to track the droplet position in the domain. The Lagrangian point-particle nomenclature comes from the fact that droplet motion is described by ordinary differential equations written in a Lagrangian reference frame. This, coupled with the gaseous phase being modeled by partial differential equations in an Eulerian refence frame is what is called the Eulerian-Lagrangian description of the spray.

The interaction droplet-gas is done by computing source/sink terms to the gaseous equations and the correct droplet position must be known to correctly distribute the spray source terms in the gas flow. The more droplets are represented in the domain, smoother are the computed source terms.

\section{Low Mach Number Approximation}

In a subsonic flow at low Mach number, pressure variations affect the continuity equation much more by changing the velocity field than by changing the fluid density. If heat transfer is present, though, temperature variations may cause considerable variations in fluid density; and continuity, momentum and energy equations must be solved in a coupled manner. Solving the equations in the original compressible formulation adds unwanted difficulties to the numerical solver and requires extra care with the boundary conditions to avoid wave interferences in the solution \cite{poinsot2005theoretical}.

One possible approach is the Boussinesq simplification, which neglects density variations everywhere except in the buoyancy term in the momentum equation. This decouples the energy equation from both the momentum and continuity equations, but it is limited to low temperature differences of about $15K$ \cite{ferziger}.

In the case of a spray jet, temperature variations may be much higher and mass source terms are present due to droplet evaporation. Neglecting local changes in density may thus cause significant errors in the solution.

A possible way to overcome the limitations of the Boussinesq approximation and the complexity of the fully compressible formulation is the asymptotic approximation of zero Mach number for the Navier-Stokes equations. The simplified equations have been formally derived in \cite{majda} and are further discussed in \cite{muller1999low} and \cite{viozat1997implicit}.

For the spray jet studied in this thesis, the maximum Mach number is found in the nozzle exit and it is less than $0.1$.

\section{Droplet Evaporation}

A comprehensive text on droplet evaporation and heating is presented in \cite{sirignano}, where some covered topics are the case of a single and spherically symmetric droplet in a quiescent gas, the effect of gas convection around the droplet, the extra complexities of a multi-component liquid droplet, droplet behavior near critical conditions and the heating and evaporation of a group of droplets nearly located.

As previously mentioned, this thesis deals with a dilute spray and thus droplets are considered to heat and vaporize similarly to a single and isolated droplet. The effects of gas convection are taken into account using empirically established laws. The liquid phase has one single component (acetone) and the droplets are far from critical conditions.

Some more simplifications are made regarding the radial transport of heat inside the droplet. They will be discussed as the heating and the evaporation models are presented, in Chapter \ref{chap: equations}.

\section{Turbulence and Droplet Submodels}

\begin{figure}[h]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.75\textwidth]{./figuras/chap1/stokes.png}
\end{tabular}
 \caption{Effect of Stokes number on particle dispersion in large-scale turbulent structures, reproduced from \cite{crowe1988particle}.}
 \label{fig: intro_stokes}
\end{figure}

Apart all the implications concerning a multiphase flow, a spray jet is often turbulent, and the correct prediction of its evolution is dependent on how turbulence is modeled. It was said that the choice of the Lagrangian point-particle method to model the dispersed phase was based on the low liquid volume fraction. A second precaution is also very important when choosing the turbulence modeling for the gaseous flow and the dispersion modeling for the droplets. For spray jets where droplet Stokes number is very low ($St << 1$), see \cite{crowe1988particle} for the definition, the droplets are likely to completely follow oscillations in gas velocity and the interaction with turbulence is strong.

For such sprays, RANS modeling is likely to fail in predicting droplet motion and dispersion\footnote{Dispersion in the context of a spray is used in a similar sense to that in waves. It means that droplet velocity will be significantly dependent on its diameter as wave velocity would be dependent on frequency or wavenumber. } because the interaction with eddies will be restricted to some submodel.  The subject of turbulent dispersed multiphase flow is reviewed by \cite{balachandar2010turbulent} and the most important implications in droplet motion and evaporation are discussed in \cite{sirignano}.

In this work, the jet Reynolds number is not high ($Re=16,300$) and the Stokes number is $\left( St \in [1,100] \right)$ with an average of $10$. It is verified in Chapter \ref{chap: results} that the stochastic subgrid model used together with the traditional k-epsilon model was able to predict some dispersion in droplet velocities.

For the gas phase, it is a known fact that the k-epsilon model must have its coefficients tunned for improved accuracy in predicting the velocity field of a turbulent round jet. However, the exact modifications are not known \textit{a priori} and this work has used the usual coefficients. The consequences of this setup are discussed in Chapter \ref{chap: results}.

% \section{The OpenFOAM code}

% This thesis has made use of moderate computational resources and the open source software OpenFOAM\footnote{OpenFOAM is managed and distributed by the OpenFOAM Foundation and is freely available and open source, licensed under the GNU General Public Licence.}.

\section{Thesis Outline}

Chapter \ref{chap: equations} presents the governing equations for both the continuous and dispersed phases and the type of boundary conditions.
The continuous phase section starts with the fully compressible flow formulation, proceeds with the low Mach number approximation and then the averaging process of the turbulence modeling. The dispersed phase section describes the Lagrangian point-particle method and derives the models for momentum, heat and mass transfer between the droplets and the gas phase.

Chapter \ref{chap: exp} briefly describes the experiment of \cite{chen}, which was used to evaluate the numerical results. It also presents the boundary conditions for the simulation obtained from the available experimental data.

Chapter \ref{chap: numerical} presents some aspects of the numerical solution such as the equation discretization schemes, the algorithms for solving linear systems and the mesh grid used for the domain discretization.

Chapter \ref{chap: results} presents the results and discussions. The first section presents the results for the properties of the gas phase turbulent jet: the spread rate and the self-similar profiles. The second section presents the results for droplet and gas velocities. The third section presents the mass fluxes of liquid and vapor and the droplet Sauter mean diameter.

Chapter \ref{chap: conclusion} summarizes the conclusions and suggestions for future work.

Appendix \ref{appendix 1} shows the derivation of the conservation equation of sensible enthalpy starting from the total energy equation.

Appendix \ref{appendix 2} show 2D Figures of droplet and gas properties.

Appendix \ref{appendix 3} briefly describes equation discretization and PISO algorithm.

### 2016-09-11 master/thesis/chap2.tex
added master files

This chapter presents the governing equations for continuous and dispersed phases.

Continuous phase equations start from the compressible formulation found in \cite{poinsot2005theoretical} with the additional spray source terms. It then proceeds with the low Mach number approximation proposed in \cite{majda}. Finally, the turbulence model is presented with the extra assumptions concerning the turbulent fluxes.

Dispersed phase equations are basically those presented in \cite{nordin} and further discussed in \cite{vonkarman}, \cite{naca} and \cite{baumgarten2006mixture}.

\section{Continuous (Gaseous) Phase}

The continuous phase is the denomination for all gaseous phase in the flow
domain. The set of governing equations is composed by mass, species, momentum, energy
 and state equations. The dispersed phase will
contribute with source/sink terms for each of the gaseous phase equations and extra closure relations
will be used to account for all the phenomena complexity e.g. turbulence and
species diffusion.

All unknowns $(\rho, Y_k, \bv{U}, T)$ are functions of position and time, e.g.
$\bv{U}=\bv{U}(\bv{x},t)$. For notation reduction, however, the function
arguments will be omitted.

\subsection{Fully Compressible Formulation}

The mass (or continuity) equation for a compressible flow with the
spray source term $S_m$ is
\begin{equation}\label{ns: mass}
\frac{\partial \rho}{\partial t} + \nabla \cdot \left( \rho \bv{U} \right) =
S_m \, .
\end{equation}

\begin{remark}Since mass equation is not homogeneous due to the presence
of the spray source term, the traditional
non-conservative form of the equations must be modified. For an intensive
property $\eta$,
\begin{equation}
\begin{split}
 \frac{\partial \rho \eta}{\partial t} + \nabla \cdot \left( \rho \bv{U} \eta
\right) &= \frac{\partial \rho}{\partial t}\eta + \rho\frac{\partial
\eta}{\partial t} + \rho \bv{U} \cdot \nabla \eta + \eta \nabla \cdot \left(
\rho \bv{U} \right) \\
&= \rho \left( \frac{\partial \eta}{\partial t} + \bv{U} \cdot \nabla \eta
\right) + \eta \left[ \frac{\partial \rho}{\partial t} + \nabla \cdot \left(
\rho \bv{U} \right) \right]\\
&= \rho \left( \frac{\partial \eta}{\partial t} + \bv{U} \cdot \nabla \eta
\right) + \eta S_m \, .
\end{split}
\end{equation}
\end{remark}

Species equation with the spray source term $S_{Yk}$ is
\begin{equation}\label{ns: species}
\frac{\partial \rho Y_k }{\partial t} + \nabla \cdot \left( \rho \left( \bv{U} +
\bv{V}_{k}\right) Y_k \right) = S_{Yk} \, ,
\end{equation} where $V_k$ is the diffusion velocity of species $k$ satisfying
$\sum_{k} V_k Y_k = 0$.

The diffusion velocities will be obtained under the approximation of a dilute
mixture in which one of the species has a considerably higher concentration
relative to the others. This is the case for the spray evaporation into an air
atmosphere, where $76.7\%$ of air mass composition is nitrogen.

If $M-1$ species are present in scarce quantities relative to the last species
denoted by $M$, then these $M-1$ species diffuse as in a binary mixture:
\begin{subequations}\label{eq: byndiff}
\begin{align}
V_k Y_k &= -\mathcal{D}_k \nabla Y_k \, , \quad k=1,2,...,M-1 \, ,\\
V_M Y_M &=  \sum_{k=1}^{M-1} \mathcal{D}_k \nabla Y_k \, ,
\end{align}
\end{subequations}
where $\mathcal{D}_k$ is the binary diffusion coefficient of species $k$ and does not need
to be the same for all species.

The scarce species transport equation is then:
\begin{equation}
 \frac{\partial \rho Y_k }{\partial t} + \nabla \cdot\left( \rho Y_k \bv{U}
\right) = \nabla \cdot \left( \rho \mathcal{D}_k\nabla Y_k \right) +S_{Yk}
\, , \quad k=1,2,...,M-1.
\end{equation}

$Y_M$ is obtained from the fact that $\sum_k Y_k =1$,
\begin{equation}
Y_M = 1-\sum_{k=1}^{M-1} Y_k \, .
\end{equation}

The momentum equation is written with the assumption of a compressible Newtonian
fluid \cite{batchelor2000introduction} and includes the gravitational field
$\bv{g}$ and the spray source term $\bv{S}_{mom}$.
\begin{equation}\label{ns: momentum}
\frac{\partial \rho \bv{U} }{\partial t} + \nabla \cdot \left(\rho \bv{U}\bv{U}
\right) = -\nabla p + \nabla \cdot \bv{\tau} +\rho \bv{g}+\bv{S}_{mom} \, ,
\end{equation} where $\bv{\tau}$ is the corresponding viscous stress tensor
\begin{equation}
\bv{\tau} =  2\mu  \left[ \bv{S} -\frac{1}{3} \left( \nabla \cdot \bv{U}
\right)\bv{I} \right]\, ,
\end{equation}
and 
$\bv{S}$ is the strain tensor:
\begin{equation}
\bv{S} = \frac{1}{2} \left( \nabla\bv{U} + \nabla \bv{U}^{T} \right) \, .
\end{equation}

\begin{notation}
 $\nabla \bv{U} = \partial u_i/\partial x_j$ where $u_i$ denotes the
components of $\bv{U}$. Similarly, $\nabla \bv{U}^{T} = \partial
u_j/\partial x_i$. $\bv{I}$ denotes the identity tensor.
\end{notation}

The energy equation might be written in several different forms
(see \cite{poinsot2005theoretical}, chapter 1). Here, the sensible enthalpy
formulation will be preferred. The sensible enthalpy is defined as $h_s = h -
\sum_{k} \Delta h^{0}_{f,k} Y_k = \int_{T_{0}}^{T} c_{p} dT$, where $h$ is the
enthalpy and $\Delta h^{0}_{f,k}$ is the enthalpy of formation of species $k$ at
the reference temperature $T_{0}$.

\begin{equation}\label{ns: energy}
 \frac{\partial \rho h_s}{\partial t} +  \nabla \cdot \left( \rho h_s \bv{U}
\right) = \frac{Dp}{Dt} + \nabla \cdot \bv{J}_s +  \bv{\tau} : \nabla\bv{U} +
 S_{hs}\, ,
 \end{equation}
$S_{hs}$ is the spray enthalpy source term. $\bv{J}_s$ includes heat conduction and the enthalpy
diffusion vector:
\begin{equation}
\bv{J}_s =  \kappa \nabla T - \rho \sum_{k=1}^{M} h_{s,k} \bv{V}_k Y_k \, .
\end{equation}

With the assumptions made for species diffusion in \eqref{eq: byndiff},
$\bv{J}_s$ becomes
\begin{equation}
\bv{J}_s =  \kappa \nabla T + \rho \sum_{k=1}^{M-1} \left( h_{s,k}
-h_{s,M}\right) \mathcal{D}_k \nabla Y_k \, .
\end{equation}

\begin{notation}
 $\tau : \nabla \bv{U}$ reads in tensorial notation as following:
\begin{equation}
 \tau : \nabla \bv{U} = 2\mu\left[ \frac{1}{2} \left( \frac{\partial
u_i}{\partial x_j}+\frac{\partial u_j}{\partial x_i}\right)-\frac{1}{3}\left(
\frac{\partial u_i}{\partial x_i} \right) \delta_{ij} \right]  \frac{\partial
u_i}{\partial x_j} \, ,
\end{equation} $\delta_{ij}$ is the Kronecker delta.
\end{notation}

Constant pressure specific heat capacity for species $k$ is denoted by
$c_{p,k}$. Similarly, constant volume specific heat capacity for species
$k$ is $c_{v,k}$. The same quantities for the gaseous mixture and the
corresponding $\gamma$ constant are then
\begin{subequations}
 \begin{align}
  c_{p} &= \sum_{k} c_{p,k} Y_k \, , \\
  c_{v} &= \sum_{k} c_{p,k} Y_k \, , \\
  \gamma &= c_{p} / c_{v} \, .
 \end{align}
\end{subequations}
The gas mixture is assumed to be an ideal gas mixture and the state equation is
given by
\begin{equation}
 p =  \frac{\gamma -1}{\gamma} \rho c_p T = \frac{1}{W}\rho R T \, ,
\end{equation} where

\begin{equation}
 \frac{1}{W}=\left( \sum_k \frac{Y_k}{W_k} \right) \, .
\end{equation}

\subsection{Non-Dimensional Equations}

All equations may be written in non-dimensional form once
reference quantities are defined.
\begin{itemize}
 \item Distance is given in units of nozzle diameter $D_{jet}$;
 \item Velocity is given in units of average axial velocity
at nozzle exit $\req{U}$;
 \item Time is given in units of $ D_{jet} / \req{U}$;      
 \item Temperature is given in units of ambient temperature $T_{\infty}$;
 \item Pressure is given in units of ambient pressure $p_{\infty}$;
 \item Mass fractions are already non-dimensional;
 \item Molar mass of species k, $W_k$, is given in units of ambient air
molar mass $W_{\infty}$;
 \item Density is given in units of $\rho_{\infty}=p_{\infty} W_{\infty} / R T_{\infty}$;
 \item $c_p$ is given in units of air $c_p$ at conditions $(p_{\infty},T_{\infty})$, or
$c_{p,\infty}$. $\gamma_{\infty}$ is $c_p / c_v$ for the air at the same conditions;
 \item $h_s$ is given in units of $c_{p,\infty} T_{\infty}$.
 \item $\mathcal{D}_k$, $\mu$ and $\kappa$ are given in units of their values at ambient conditions: $\mathcal{D}_{k,\infty}$, $\mu_{\infty}$ and $\kappa_{\infty}$.
 \item $g$ is one unit of the constant gravitational field.
\end{itemize}

The following non-dimensional parameters are also defined:
\begin{subequations}
 \begin{align}
  Re &= \frac{\rho_{\infty} \req{U} D_{jet}}{\mu_{\infty}} \, ,\\
  M  &= \frac{\req{U}}{\sqrt{\gamma_{\infty} R T_{\infty}}}= \frac{\req{U}}{\sqrt{\gamma_{\infty}
  p_{\infty} /\rho_{\infty}}} \, , \\
  Pr &= \frac{\mu_{\infty} c_{p,\infty}}{\kappa_{\infty}} \, , \\
  Le_k &= \frac{\kappa_{\infty}}{\rho_{\infty} c_{p,_{\infty}} \mathcal{D}_{k,\infty}} \, , \\
  Sc_k &= \frac{\mu_{\infty}}{\rho_{\infty} \mathcal{D}_{k,\infty}} = Pr Le_k \, , \\
  Fr &= \frac{\req{U}}{\sqrt{g D_{jet}}} \, .
 \end{align}
\end{subequations}

Equations are then rewritten in terms of non-dimensional quantities.
\begin{description}
\item[State equation:]
\begin{equation}\label{nd: state}
p=\rho T \left( \sum_k \frac{Y_k}{W_k}\right) = \frac{\rho T}{W} \, .
\end{equation}
\item[Mass equation:]
\begin{equation}\label{nd: mass}
\frac{\partial \rho}{\partial t} + \nabla \cdot \left( \rho \bv{U} \right) = 
S_m \, .
\end{equation}
\item[Species equation:]
\begin{subequations}\label{nd: species}
\begin{equation}
 \frac{\partial \rho Y_k }{\partial t} + \nabla \cdot\left(\rho \bv{U} Y_k
\right) = \frac{1}{Re Sc_k} \nabla \cdot \left( \rho \mathcal{D}_k \nabla Y_k
\right) +S_{Yk} \, , \quad k=1,2,...,M-1 \, .
\end{equation}
\begin{equation}
Y_M = 1- \sum_{k=1}^{M-1} Y_k \, .
\end{equation}
\end{subequations}
\item[Momentum equation:]
\begin{equation}\label{nd: momentum}
 \frac{\partial \rho \bv{U} }{\partial t} + \nabla \cdot \left( \rho \bv{U}
\bv{U} \right) = -\frac{1}{\gamma_{\infty} M^2} \nabla p + \frac{1}{Re} \nabla \cdot
\bv{\tau} +\frac{\rho}{Fr^2} \frac{\bv{g}}{|\bv{g}|}+\bv{S}_{mom} \, .
\end{equation} 
\item[Energy equation:]
\begin{equation}\label{nd: enthalpy}
\begin{split}
\frac{\partial \rho h_s}{\partial t} +  \nabla \cdot \left(\rho \bv{U} h_s
\right) &= \frac{\gamma_{\infty} -1}{\gamma_{\infty}}\frac{Dp}{Dt} + \nabla \cdot \bv{J}_s \\
&+ M^2 \left( \gamma_{\infty}-1 \right) \left( \frac{\bv{\tau} : \nabla\bv{U}}{Re}  \right) + S_{hs}\, ,
\end{split}
\end{equation} 
\end{description}

where $\bv{\tau}$ is the shear stress tensor \begin{equation}
\bv{\tau} =  2 \mu \left[ \bv{S} -\frac{1}{3} \left( \nabla \cdot \bv{U}
\right)\bv{I} \right]\, ,
\end{equation}
$\bv{I}$ is the identity tensor and $\bv{S}$ is the strain tensor
\begin{equation}
\bv{S} = \frac{1}{2} \left( \nabla\bv{U} + \nabla \bv{U}^{T} \right) \, .
\end{equation}
$\bv{J}_s$ is the heat conduction and the sensible enthalpy diffusion:
\begin{equation}
\bv{J}_s =  \frac{\kappa}{Re Pr} \nabla T + \sum_{k=1}^{M-1} \frac{\rho
\mathcal{D}_k}{Re Sc_k} \left( h_{s,k} -h_{s,M} \right) \nabla Y_k  \, .
\end{equation}

\subsection{Low Mach Number Approximation}
Consider the following definitions \cite{viozat1997implicit}:

\begin{quote}
A \textbf{compressible flow} is a flow in which density depends on pressure and
temperature.
\end{quote}

\begin{quote}
A \textbf{dilatable flow} is a flow in which density depends on temperature.
\end{quote}

The gas flow considered in this work is fairly incompressible, indeed the
local Mach number is below $0.1$ throughout the flow domain. However, the
intense heat and mass transfers between gaseous and liquid phases when
spray is present produce large temperature gradients which cause density to vary
considerably and the two-phase flow may be classified as a dilatable flow in the
sense of the definition above.

An important difference between incompressible and dilatable flows is
that $\nabla \cdot \bv{U} = 0$ does not hold for the latter.

A common treatment for dilatable flows is the low Mach number approximation, which consists in expanding unknown quantities $p$, $Y_k$, $\bv{U}$ and $T$ in Equations \eqref{ns:
mass}, \eqref{ns: momentum} and \eqref{ns: energy} into power series
of $\xi \equiv \sqrt{\gamma_{\infty}} M$:
\begin{subequations}\label{eq: expansion}
 \begin{align}
   p &= p_{0} + p_{1}\xi + p_{2}\xi^2 + O\left( \xi^3 \right) \, ,\\
   Y_k &= Y_{k,0} + O\left( \xi \right) \, ,\\
  \bv{U} &= \bv{U}_{0} + O\left( \xi \right) \, ,\\
  T &= T_{0} + O\left( \xi \right) \, .
 \end{align}
\end{subequations}

The reason for expanding pressure up to the second order while keeping only the
leading (or zeroth) order for the other variables is discussed in
\cite{muller1999low}, but it is essentially because of the $M^{-2}$ factor in momentum equation \eqref{nd: momentum}.

The dependence of the spray source terms on the flow variables is also considered only for the leading order.

$\rho_0$ is obtained by substitution of \eqref{eq: expansion} into the
non-dimensional state equation. $\rho_0$ is such that
\begin{equation}\label{lm: state}
 p_0 = \rho_0 T_0 \left( \sum_k \frac{Y_{k,0}}{W_k} \right) =  \frac{ \rho_0
T_0}{W_0}\, .
\end{equation}

Mass equation \eqref{nd: mass} of order $\xi^0$ becomes:
\begin{equation}
 \frac{\partial \rho_0}{\partial t} + \nabla \cdot \left( \rho_0 \bv{U}_0
\right) =  S_m \, .
\end{equation}

Species equation \eqref{nd: species} of order $\xi^0$ is:
\begin{subequations}\label{lm: species}
\begin{align}
 \frac{\partial \rho_0 Y_{k,0} }{\partial t} + \nabla \cdot\left(\rho_0 \bv{U}_0
Y_{k,0} \right) &= \frac{1}{Re Sc_k} \nabla \cdot \left( \rho \mathcal{D}_k
\nabla Y_{k,0} \right) +S_{Yk} \, , \quad k=1,2,...,M-1 \, . \\
  Y_{M,0} &= 1- \sum_{k=1}^{M-1} Y_{k,0} \, .
\end{align}
\end{subequations}

The momentum equation gives information of orders $\xi^{-2}$, $\xi^{-1}$ and
$\xi^0$ due to the presence of the coefficient $1/\xi^2$ in the pressure gradient . The
resulting equations are, respectively,
\begin{subequations}\label{lm: momentum}
\begin{align}
 \nabla p_0 &\equiv 0 \, , \\
 \nabla p_1 &\equiv 0 \, , \\
 \frac{\partial \rho_0 \bv{U}_0 }{\partial t} + \nabla \cdot \left( \rho_0
\bv{U}_0 \bv{U}_0 \right) &= -\nabla p_2 + \frac{1}{Re} \nabla \cdot \bv{\tau}_0
+\frac{\rho_0}{Fr^2} \frac{\bv{g}}{|\bv{g}|}+\bv{S}_{mom}  \, .
\end{align}
\end{subequations}

From $\nabla p_0 \equiv \nabla p_1 \equiv 0$, it is observed that $p_0$ and
$p_1$ are functions of time only, $p_0 = p_0 (t)$ and $p_1 = p_1 (t)$. The
coupling between  velocity and pressure fields is now separated into two
contributions:  $p_2$ establishes the pressure gradient and will be referred to
as the dynamic pressure; $p_0$ affects $\rho_0$ via the state equation and will
be referred to as the thermodynamic pressure.

The energy equation reads
\begin{equation}\label{lm: enthalpy}
\frac{\partial \rho_0 h_{s0}}{\partial t} +  \nabla \cdot \left(\rho_0 \bv{U}_0
h_{s0} \right) = \frac{\gamma_{\infty} -1}{\gamma_{\infty}}\frac{Dp_0}{Dt} + \nabla \cdot
\bv{J}_{s0} + S_{hs}\, ,
\end{equation} where 
\begin{equation}
\bv{J}_{s0} =  \frac{\kappa}{Re Pr} \nabla T_0 + \sum_{k=1}^{M-1}
\frac{\rho\mathcal{D}_k}{Re Sc_k} \left( h_{s0,k} -h_{s0,M} \right) \nabla
Y_{k,0} \, . 
\end{equation}

An ordinary differential equation in time may be derived for computing the thermodynamic pressure ($p_0$) by differentiating the state equation \eqref{lm: state} with respect to time and integrating in domain volume:
\begin{equation}\label{lm: dstate}
 \frac{Dp_0}{Dt}\int_{\Omega}W dV + p_0 \int_{\Omega}\frac{DW}{Dt} dV=\int_{\Omega} \frac{D\rho T}{Dt} dV\, ,
\end{equation}
where 
\begin{equation}
\frac{D}{Dt} = \frac{\partial}{\partial t}+\bv{U}\cdot\nabla
\end{equation}
is the total derivative.

Each total derivative may be computed from mixture molar mass ($W$) definition and mass, species and enthalpy equations.

For an infinite (or open) domain $\Omega$, 
\begin{equation}
\int_{\Omega}\frac{DW}{Dt} dV < \infty \, , \quad \int_{\Omega} \frac{D\rho T}{Dt} dV < \infty \quad \text{and} \quad \int_{\Omega}W dV = \infty \, .
\end{equation}
 Hence, $Dp_0/Dt = 0$ and $p_0 = 1$, the non-dimensional pressure in far-field boundary.

Herein, the subscripts denoting the order of expansion will be omitted, except
for the pressure. $\bv{U}_0$ is simply $\bv{U}$ and the same holds to $\rho$,
$Y_k$, $h_s$ and $T$. For pressure, $p_0$ is the thermodynamic pressure and
$p_2$ is the dynamic pressure.

\subsection{Turbulence Modeling}

The expression \textit{turbulent and evaporating spray} entitling this work
refers to a liquid spray evolving in a gaseous turbulent flow where mass, energy
and momentum transfers between both phases are strongly affected by turbulence.

Among all possible treatments for turbulent flows, see
\cite{pope2000turbulent}, one is perhaps the most widely used in industrial
applications: the so called \textit{standard k-epsilon model}. This two-equation
model within the RANS (Reynolds-averaged Navier-Stokes) framework provides a good
compromise between complexity and computational cost for those not using
exceptional computational resources.

In the context of RANS modeling, the governing equations are averaged and solved
for the mean values of flow properties. The effect of turbulence on the mean
field is accounted by modeling the terms depending on the fluctuations. In the
specific case of the momentum equation, the hypothesis of turbulent-viscosity
(or Boussinesq hypothesis) is used to compute an effective value of local
viscosity composed by molecular and turbulent viscosities, being the
latter an approximation of the extra momentum flux of Reynolds stresses.

For compressible flows, two averages are commonly used: the simple time average
and a mass-weighted time average (or Favre average), see \cite{favre1969statistical}.
The time average $(\bar{f})$ and the deviation $(f')$  of a quantity $f$ are
defined as
\begin{equation}
 \bar{f}(x,t) \equiv \frac{1}{T}\int_{t}^{t+T} f(x,\tau) d\tau \quad \text{and}
\quad f'(x,t)= f(x,t)-\bar{f}(x,t) \, .
\end{equation}

The mass-weighted average (or Favre average) $(\tilde{f})$ and the deviation
$(f'')$ are defined as
\begin{equation}
 \tilde{f}(x,t) = \frac{\overline{\rho f(x,t)}}{\bar{\rho}} \quad \text{and}
\quad f^{''}(x,t) = f(x,t) - \tilde{f}(x,t) \, .
\end{equation}

Following the definitions,
\begin{equation}
 \quad \overline{f'}=\bar{f}-\bar{f}=0 \quad \text{and} \quad \bar{\bar{f}}=
\bar{f}\, ,
\end{equation}
\begin{equation}
 \quad \tilde{f''}=\tilde{f}-\tilde{f}=0 \quad \text{and} \quad \tilde{\tilde{f}}=
\tilde{f}\, .
\end{equation}
Averaging and differential operators are assumed to commute, see \cite{pope2000turbulent}.

\subsubsection{Mass and Species Equations}
Averaged mass and species equations are easily obtained. The
averaged mass equation reads:
\begin{equation}\label{av: mass}
  \frac{\partial \bar{\rho}}{\partial t} + \nabla \cdot \left( \bar{\rho}
\tilde{\bv{U}} \right) =  \bar{S}_m \, .
\end{equation}

The species averaged equation reads:
\begin{subequations}\label{av: species1}
\begin{equation}
 \frac{\partial \bar{\rho} \tilde{Y}_k }{\partial t} + \nabla
\cdot\left(\bar{\rho} \tilde{\bv{U}} \tilde{Y}_k +\bar{\rho}\widetilde{\bv{U''}
Y''_{k}} \right) = \frac{1}{Re Sc_k} \nabla \cdot \left( \overline{\rho
\mathcal{D}_k \nabla Y_k} \right) +\bar{S}_{Yk} \, , \quad k=1,2,...,M-1 \, ,
\end{equation}
\begin{equation}
\tilde{Y}_M = 1- \sum_{k=1}^{M-1} \tilde{Y}_{k} \, .
\end{equation}
\end{subequations}
Equation \eqref{av: species1} shows two new terms: $\overline{\rho \mathcal{D}_k \nabla
Y_k}$ and $\bar{\rho}  \widetilde{\bv{U}'' Y_k^{''}}$. Unless new equations are
derived for them, both terms are not known and must be modeled.

For the first one, regarding the laminar diffusivity of species, the average is
commonly approximated as shown below:
\begin{equation}
\overline{\rho \mathcal{D}_k \nabla Y_k} \approx \bar{\rho}\mathcal{D}_k \nabla
\tilde{Y}_k \, .
\end{equation}

For the second term, which deals with turbulent flux, an approach analogous
to turbulent-viscosity is used and a turbulent coefficient for species
diffusion $(\mathcal{D}_{k,t})$ is used:
\begin{equation}
\bar{\rho}  \widetilde{\bv{U}'' Y_k^{''}} \approx -\frac{\bar{\rho}
\mathcal{D}_{k,t}}{Re Sc_k} \nabla \tilde{Y}_k \, ,
\end{equation}
$\mathcal{D}_{k,t}$ is computed by the turbulence model.

The species equation then becomes:
\begin{subequations}\label{av: species}
\begin{equation}
 \frac{\partial \bar{\rho} \tilde{Y}_k }{\partial t} + \nabla
\cdot\left(\bar{\rho} \tilde{\bv{U}} \tilde{Y}_k \right) = \frac{1}{Re Sc_k}
\nabla \cdot \left[ \bar{\rho} \left( \mathcal{D}_k + \mathcal{D}_{k,t} \right)
\nabla \tilde{Y}_k \right] +\bar{S}_{Yk} \, , \quad k=1,2,...,M-1 \, .
\end{equation}
\begin{equation}
\tilde{Y}_M = 1- \sum_{k=1}^{M-1} \tilde{Y}_{k} \, .
\end{equation}
\end{subequations}

\subsubsection{Momentum Equation}
Treatment of momentum equation is more complicated and it is shown in more
steps here. The averaged momentum equation reads:
\begin{equation}
 \frac{\partial \bar{\rho} \tilde{\bv{U}} }{\partial t} + \nabla \cdot \left(
\bar{\rho} \tilde{\bv{U}} \tilde{\bv{U}} \right) = -\nabla \bar{p}_2 +  \nabla
\cdot \left(  \frac{\bar{\bv{\tau}}}{Re} - \bar{\rho}\widetilde{\bv{U}''
\bv{U}''} \right) +\frac{\bar{\rho}}{Fr^2}
\frac{\bv{g}}{|\bv{g}|}+\bar{\bv{S}}_{mom}  \, .
\end{equation}

Again, two terms are new and unknown: $\bar{\tau}$ and $-
\bar{\rho}\widetilde{\bv{U}'' \bv{U}''}$. From the definitions of averages, 
\begin{equation}
 \tau = \tilde{\tau}+\tau'' \quad \Rightarrow \quad \bar{\tau} =
\tilde{\tau}+\overline{\tau''} \, ,
\end{equation}
and assuming $\tilde{\tau} \gg \overline{\tau''}$, we have:
\begin{equation}
 \bar{\tau} \approx  2\mu\left[ \tilde{S}-\frac{I}{3}\left(\nabla \cdot
\tilde{\bv{U}} \right)  \right] \, .
\end{equation}

The modeling of the turbulent momentum flux is made by the following
expression (known as Boussinesq hypothesis):
\begin{equation}\label{eq: boussinesq}
 - \bar{\rho}\widetilde{\bv{U}'' \bv{U}''} =  \frac{2 \mu_t}{Re} \left[
\tilde{S}-\frac{I}{3} \left( \nabla \cdot \tilde{\bv{U}} \right)  \right] -
\frac{2}{3}\bar{\rho}kI \, .
\end{equation}

$\mu_t$ is the turbulent-viscosity accounting for the extra momentum flux due to the Reynolds stresses ($-
\bar{\rho}\widetilde{\bv{U}'' \bv{U}''}$). $k$ is turbulent kinetic energy defined as:
\begin{equation}
  k = \frac{1}{2} \widetilde{\left(\bv{U}'' \cdot \bv{U}'' \right) \, .}
\end{equation}
Both $\mu_t$ and $k$ are computed by the turbulence model.

$\tau_{tot}$ is defined as a ``new viscous stress tensor'' composed by the
laminar and turbulent part:
\begin{equation}
 \overline{\tau_{tot}} \equiv 2 (\mu + \mu_t) \left[
\tilde{S}-\frac{I}{3}\left(\nabla \cdot \tilde{\bv{U}} \right)  \right]
-\frac{2}{3}\bar{\rho}k I \, ,
\end{equation}
and the momentum equation finally becomes:
\begin{equation}\label{av: momentum}
 \frac{\partial \bar{\rho} \tilde{\bv{U}} }{\partial t} + \nabla \cdot \left(
\bar{\rho} \tilde{\bv{U}} \tilde{\bv{U}} \right) = -\nabla \bar{p}_2 +  
\frac{1}{Re}  \nabla \cdot\overline{\bv{\tau}_{tot}} +\frac{\bar{\rho}}{Fr^2}
\frac{\bv{g}}{|\bv{g}|}+\bar{\bv{S}}_{mom}  \, .
\end{equation}

\subsubsection{Sensible Enthalpy Equation}
The averaged sensible enthalpy equation reads:
\begin{equation}
\frac{\partial \bar{\rho} \tilde{h}_{s}}{\partial t} + \nabla \cdot
\left(\bar{\rho} \tilde{\bv{U}} \tilde{h}_{s} + \bar{\rho} \widetilde{\bv{U}''
h_s^{''}} \right) = \frac{\gamma_{\infty} -1}{\gamma_{\infty}}\overline{\frac{Dp_0}{Dt}} + \nabla
\cdot \bar{\bv{J}}_s + \bar{S}_{hs}\, ,
\end{equation} where 
\begin{equation}
\bv{J}_s =  \overline{\frac{\kappa}{Re Pr} \nabla T} + \sum_{k=1}^{M-1}
\overline{\frac{\rho\mathcal{D}_k}{Re Sc_k} \left( h_{s,k} -h_{s,M} \right)
\nabla Y_{k}} \, . 
\end{equation}

Four new terms are unknown and must be modeled as well:\\  $\bar{\rho}
\widetilde{\bv{U}'' h_s^{''}} $ , $\overline{\frac{Dp}{Dt}} $, 
$\overline{\frac{\kappa}{Re Pr} \nabla T}$ and 
$\overline{\frac{\rho\mathcal{D}_k}{Re Sc_k} \left( h_{s,k} -h_{s,M} \right)
\nabla Y_{k}} $.

Turbulent flux of enthalpy is modeled similarly to species and momentum
turbulent fluxes:
\begin{equation}
 \bar{\rho} \widetilde{\bv{U}'' h_s^{''}} \approx -\frac{\bar{\rho} \alpha_t
}{Re Pr} \nabla \tilde{h}_s \, ,
\end{equation}
where $\alpha_t$ is the turbulent thermal diffusivity that is also computed by the
turbulence model.

The average of the total derivative $\overline{Dp/Dt}$ reads:
\begin{equation}
 \overline{\frac{D p_0}{D t}} = \frac{\partial \bar{p}_0}{\partial
t}+\overline{\bv{U} \cdot \nabla p_0} = \frac{\partial \bar{p}_0}{\partial t} + \left( \tilde{\bv{U}} \cdot \nabla \bar{p}_0 \right) + \overline{\bv{U}'' \cdot \nabla p_0}
\end{equation}
Neglecting the last term ($\overline{\bv{U}'' \cdot \nabla p_0}$) gives:
\begin{equation}
 \overline{\frac{D p_0}{D t}}  \approx \frac{\partial \bar{p}_0}{\partial t} +
\left( \tilde{\bv{U}} \cdot \nabla \bar{p}_0 \right)\, .
\end{equation}

For laminar heat diffusion, the effect of fluctuations is neglected and only
averaged terms are considered:
\begin{equation}
 \overline{k \nabla T} = \bar{k} \nabla \tilde{T} + \overline{k\nabla T''}
\approx \bar{k} \nabla \tilde{T} \, .
\end{equation}

Under the same assumption:
\begin{equation}
 \overline{\frac{\rho \mathcal{D}_k}{Re Sc_k}\left( h_{s,k} - h_{s,M} \right)
\nabla Y_k} \approx \frac{\bar{\rho} \mathcal{D}_k}{Re Sc_k} \left[ \left( \tilde{h}_{s,k} -
\tilde{h}_{s,M }\right) \nabla \tilde{Y}_k \right] \, .
\end{equation}

Sensible enthalpy equation finally becomes:
\begin{equation}\label{av: enthalpy}
\frac{\partial \bar{\rho} \tilde{h}_{s}}{\partial t} + \nabla \cdot
\left(\bar{\rho} \tilde{\bv{U}} \tilde{h}_{s} \right) = \frac{\gamma_{\infty} -1}{\gamma_{\infty}}
\left[ \frac{\partial \bar{p}_0}{\partial t} + \left( \tilde{\bv{U}} \cdot \nabla
\bar{p}_0 \right) \right]  + \nabla \cdot \bar{\bv{J}}_s + \bar{S}_{hs}\, ,
\end{equation} where 
\begin{equation}
\bar{\bv{J}}_s =  \frac{\bar{\rho} \left( \alpha + \alpha_t \right)}{Re Pr}
\nabla \tilde{h}_s + \sum_{k=1}^{M-1} \frac{\rho \mathcal{D}_k}{Re Sc_k} \left[
\left( \tilde{h}_{s,k} - \tilde{h}_{s,M }\right) \nabla \tilde{Y}_k \right] \, .
\end{equation}

All unclosed terms are now specified if the following parameters are given by
the turbulence model:
\begin{equation*}
 k, \quad \mu_t , \quad \alpha_t , \quad \mathcal{D}_{k,t} \, .
\end{equation*}

Further assumptions are made:

\begin{itemize}
 \item Unity Schmidt and Prandtl numbers ($Sc_k = Pr = 1$) everywhere in the
flow domain for laminar and turbulent fluxes:
\begin{equation}\label{eq: sc1}
 \bar{\rho} \left(\mathcal{D}_k +\mathcal{D}_{k,t} \right) = \bar{\rho} \left(\alpha +\alpha_{k,t} \right) = \mu+\mu_t \, .
\end{equation} 
Now, the turbulence model only needs to provide $k$ and $\mu_t$.

 \item All species have the same sensible enthalpy:

\begin{equation}
 \sum_{k=1}^{M-1} \frac{\rho \mathcal{D}_k}{Re Sc_k} \left[ \underbrace{\left(
\tilde{h}_{s,k} - \tilde{h}_{s,M }\right)}_{\approx 0} \nabla \tilde{Y}_k
\right] =  0 \, .
\end{equation}

\end{itemize}

\subsubsection{State Equation}

Averaged state equation reads:
\begin{equation}
\overline{p W} = R \overline{\left( \rho T \right)} = R \bar{\rho} \tilde{T} \, ,
\end{equation}
where $1/W = \sum_k Y_k/W_k$.

The term on the rhs - $\overline{p W}$ - is simply approximated as
\begin{equation}
\overline{p W} = \bar{p}\tilde{W}+\overline{p W''} \approx \bar{p}\tilde{W} \, .
\end{equation}

\subsubsection{Relation of Averaged Enthalpy and Temperature}

An expression is needed to relate $\tilde{h}_s$, which is being solved for in Equation \eqref{av: enthalpy}, to Favre-averaged temperature ($\tilde{T}$). From the definition of $h_s$:
\begin{equation}
\tilde{h}_s =\frac{\overline{\rho h_s}}{\bar{\rho}} = \frac{1}{\bar{\rho}}\overline{\left[ \rho \left( \int_{T_0}^{T} \sum_k c_{p,k}(T) dT \right)\right] }.
\end{equation}

Using $c_p=\sum_k c_{p,k} Y_k$, $\ Y_k = \tilde{Y}_k + Y_{k}^{''}$ and $T = \tilde{T}+ T^{''}$, a tricky expression is obtained:
\begin{equation}
\tilde{h}_s = \frac{1}{\bar{\rho}} \overline{\left(\bar{\rho}+\rho' \right) \int_{T_0}^{\tilde{T}+T''} \sum_k c_{p,k} \left(\tilde{T}+ T^{''}\right) \left(\tilde{Y}_k+Y^{''}_k \right) dT} \, .
\end{equation}

A convenient simplification is to simply neglect all fluctations and consider the following approximation:
\begin{equation}
\tilde{h}_s \approx \int_{T_0}^{\tilde{T}} \sum_k c_{p,k}(\tilde{T})\ \tilde{Y}_k\ dT \, .
\end{equation}

\subsubsection{Standard k-epsilon Turbulence Model}

According to the simplification of unity Schmidt and Prandtl numbers made in Equation \eqref{eq: sc1}, the turbulence modeling
is complete once $k$ and $\mu_t$ are specified. The \textit{standard k-epsilon}
model provides one algebraic equation for $\mu_t$ and two differential equations
for $k$ and $\epsilon$, the dissipation rate of $k$.

\begin{equation}
 \mu_t = \bar{\rho} C_{\mu} \frac{k^2}{\epsilon} \, ,
\end{equation}

\begin{equation}
 \frac{\partial \bar{\rho}k}{\partial t} + \nabla \cdot \left( \bar{\rho}
\tilde{\bv{U}} k \right) = \frac{1}{Re}\nabla \cdot \left[ \left( \mu +
\frac{\mu_t}{\sigma_k} \right) \nabla k \right] -\frac{2}{3} \bar{\rho} k \left(
\nabla \cdot \tilde{\bv{U}} \right) + P_k -\bar{\rho} \epsilon \, ,
\end{equation}

\begin{equation}\begin{split}
 \frac{\partial \bar{\rho}\epsilon}{\partial t} + \nabla \cdot \left( \bar{\rho}
\tilde{\bv{U}} \epsilon \right) = \frac{1}{Re} \nabla \cdot \left[ \left( \mu +
\frac{\mu_t}{\sigma_{\epsilon}} \right) \nabla \epsilon \right]  - \left(
\frac{2}{3}C_{\epsilon1}+C_{\epsilon3}\right)\bar{\rho}\epsilon \left( \nabla
\cdot \tilde{\bv{U}}\right) \\
+ C_{\epsilon 1} \frac{\epsilon}{k} P_k - C_{\epsilon 2} \bar{\rho}
\frac{\epsilon^2}{k} \, ,
\end{split}\end{equation}
where $P_k = - \bar{\rho} \widetilde{\bv{U}'' \bv{U}''}: \nabla \tilde{\bv{U}}$.
The approximation for the term $- \bar{\rho} \widetilde{\bv{U}'' \bv{U}''}$ was
already defined in \eqref{eq: boussinesq} and is repeated here for convenience:

\begin{equation}
 - \bar{\rho}\widetilde{\bv{U}'' \bv{U}''} =  \frac{2 \mu_t}{Re} \left[
\tilde{S}-\frac{I}{3} \left( \nabla \cdot \tilde{\bv{U}} \right)  \right] -
\frac{2}{3}\bar{\rho}kI \, .
\tag{\ref{eq: boussinesq}}
\end{equation}

The model has six parameters whose values are shown in Table \ref{table:
kepscoeff}.
\begin{table}
\centering
 \begin{tabular}{cccccc}
  \hline
  $C_{\mu}$ & $C_{\epsilon 1}$ & $C_{\epsilon 2}$ & $C_{\epsilon 3}$  & $\sigma_k$ & $\sigma_{\epsilon}$\\
  $0.09$ & $1.44$ & $1.92$ & $-0.33$ & $1$ & $1.3$ \\ \hline
 \end{tabular}
\caption{k-epsilon model constants.}
\label{table: kepscoeff}
\end{table}

It must be pointed out that the droplet dispersion due to turbulence will be modeled in the dispersed phase description. 
However, no sink of turbulent kinetic energy was defined here due to the work done by the eddies on the droplets.

\subsection{Boundary Conditions}

Four distinct boundary regions are present in the experiment of Chen, St\aa{}rner and Masri:
 nozzle inlet ($\partial \Omega_{nozzle}$), co-flow inlet ($\partial \Omega_{\text{co-flow}}$), far-field boundaries ($\partial \Omega_{\infty}$) and outlet ($\partial \Omega_{outlet}$), see Figure \ref{fig: jet_bc}.

% The prescribed conditions are zero gradient for pressure and fixed values for the remaining unknowns. In the far-field boundaries, pressure is the atmosphere and the other unknowns have a zero gradient condition. In the outlet, zero gradient is specified for all variables.

\begin{description}
 \item[Nozzle inlet conditions:]
\begin{subequations}
 \begin{align}
  \frac{\partial p}{\partial \bv{n}}(\partial \Omega_{nozzle},t)  = 0 \\
  \bv{U}(\partial \Omega_{nozzle},t) = \bv{U}_{nozzle}\left(\bv{x}\right) \\
  T(\partial \Omega_{nozzle},t) = T_{nozzle} \\
  Y_k(\partial \Omega_{nozzle},t) = Y_{k,nozzle}
 \end{align}
\end{subequations}

\item[Coflow conditions:]
\begin{subequations}
 \begin{align}
  \frac{\partial p}{\partial n}(\partial \Omega_{\text{co-flow}},t)  = 0 \\
  \bv{U}(\partial \Omega_{\text{co-flow}},t) = \bv{U}_{\text{co-flow}} \\
  T(\partial \Omega_{\text{co-flow}},t) = T_{\text{co-flow}} \\
  Y_k(\partial \Omega_{\text{co-flow}},t) = Y_{k,\text{co-flow}}
 \end{align}
\end{subequations}

\item[Far-field boundaries:]
\begin{subequations}
 \begin{align}
  p \left( \Omega_b,t \right) &= 1 \\ % non-dimensional value
  \frac{\partial \bv{U}}{\partial \bv{n}} (\partial \Omega_b,t) &= 0 \\
  \frac{\partial T}{\partial \bv{n}} (\partial \Omega_b,t) &= 0 \\
  \frac{\partial Y_k}{\partial \bv{n}}(\partial \Omega_b,t) &= 0 
 \end{align}
\end{subequations}

\item[Outlet conditions:]
\begin{subequations}
 \begin{align}
  \frac{\partial p}{\partial n}(\partial \Omega_{outlet},t)  &= 0 \\
  \frac{\partial \bv{U}}{\partial \bv{n}} (\partial \Omega_{outlet},t) &= 0 \\
  \frac{\partial T}{\partial \bv{n}} (\partial \Omega_{outlet},t) &= 0 \\
  \frac{\partial Y_k}{\partial \bv{n}}(\partial \Omega_{outlet},t) &= 0 
 \end{align}
\end{subequations}
 \end{description}
where $\bv{n}$ is the unit vector normal to the boundary.

Function $\bv{U}_{nozzle}(\bv{x})$ and scalars $T_{nozzle}$, $Y_{k,nozzle}$, $\bv{U}_{\text{co-flow}}$, $T_{\text{co-flow}}$ and $Y_{k,\text{co-flow}}$ are determined in Chapter \ref{chap: exp} from available experimental data.

\begin{figure}
\begin{center}
\includegraphics[width=0.75\textwidth]{./figuras/chap2/jet.png}
\end{center}
\caption{Definition of axis orientation and domain boundaries for the emerging round jet flow. Adapted from \cite{luppes}.}
\label{fig: jet_bc}
\end{figure}

Numerical solution of the equations will be performed time accurately due to the stochastic modeling of liquid atomization, as explained later in the dispersed phase section of this chapter. The initial conditions would then be the steady state solution of a pure gas jet.

The time-dependence of thermodynamic pressure $p_0$ is neglected and it is assumed constantly equal to ambient pressure, or an unity value in dimensionless form,
\begin{equation}
 p_0 (t) = 1 \, .
\end{equation}

\subsubsection{Boundary Conditions of Turbulent Quantities}

Introduction of two new variables in the turbulence model, namely $k$ and $\epsilon$, requires the definition of their boundary conditions.

\begin{description}
 \item[Nozzle inlet conditions:]
\begin{subequations}
 \begin{align}
  k(\partial \Omega_{nozzle},t) &= k_{nozzle}(\bv{x}) \, ,\\
  \epsilon(\partial \Omega_{nozzle},t) &= \epsilon_{nozzle}(\bv{x}) \, ,
 \end{align}
\end{subequations}

\item[Co-flow conditions:]
\begin{subequations}
 \begin{align}
  k(\partial \Omega_{\text{co-flow}},t) &= k_{\text{co-flow}} \, , \\
  \epsilon(\partial \Omega_{\text{co-flow}},t) &= \epsilon_{\text{co-flow}} \, ,
 \end{align}
\end{subequations}

\item[Far-field boundaries:]
\begin{subequations}
 \begin{align}
  \frac{\partial k}{\partial \bv{n}} (\partial \Omega_b,t) &= 0 \, , \\
  \frac{\partial \epsilon}{\partial \bv{n}} (\partial \Omega_b,t) &= 0 \, ,
 \end{align}
\end{subequations}

\item[Outlet conditions:]
\begin{subequations}
 \begin{align}
  \frac{\partial k}{\partial \bv{n}} (\partial \Omega_{outlet},t) &= 0 \, , \\
  \frac{\partial \epsilon}{\partial \bv{n}} (\partial \Omega_{outlet},t) &= 0 \, .
 \end{align}
\end{subequations}
\end{description}

Functions $k_{nozzle}(\bv{x})$ and $\epsilon_{nozzle}(\bv{x})$ and scalars $k_{\text{co-flow}}$ and $\epsilon_{\text{co-flow}}$ are also determined in Chapter \ref{chap: exp}.

\section{Dispersed (Liquid) Phase}

The modeling presented for the liquid phase here is solely intended to deal with
the situation of a narrow liquid jet emerging from a nozzle in a coaxial gaseous
flow in 
specific conditions that cause the liquid jet to atomize, that is, to
disintegrate into a large number of non-contiguous small volumes called
droplets. 
Despite that, this situation is further simplified to produce results for
engineering problems and many of the physical complexities are not addressed.

The ensemble of droplets originated from atomization is called the spray.
The spray evolution in time is a complex phenomenon as the liquid droplets exchange heat, mass
and momentum with the surrounding gas and other droplets. The droplets
also have their inner thermodynamics and might become unstable to the
point where they subdivide into smaller ones.

Here, the droplets will be treated as particles with the following properties:
\begin{itemize}
 \item liquid substance: according to which the constitutive properties will be
defined (only single-component liquids are concerned, see
\cite{baumgarten2006mixture});
 \item geometric properties: all droplets are assumed to be spheric with 
diameter $D$;
 \item thermodynamic properties: pressure, volume, mass, temperature;
 \item kinematic properties: position and velocity (translational only).
\end{itemize}

This description is often called the point-particle method, see \cite{balachandar2010turbulent}.

The next subsections will describe the processes being modeled for
computation of spray evolution (the evolution of all droplet properties in time).

The approach to the inter-phase interaction shown below is usually called the two-way coupling because it
comprises the interaction between gas and droplet but neglects the interactions between two different droplets, like
droplet collisions. It is mostly used for the case of dilute sprays, when the liquid volumetric fraction ($\phi_{v,l}$) satisfies $\phi_{v,l}<10^{-3}$.

\subsection{Atomization or Primary Break-Up}

The atomization is not modeled in a deterministic way, with the dynamics of liquid jet disintegration into droplets.

Rather, it is used a stochastic approach called the Monte Carlo's method, see \cite{baumgarten2006mixture}, that describes resulting droplets from atomization process based on available measurements near the nozzle exit plane.

Droplets are inserted into the domain in groups called parcels. A parcel is
a group of droplets with all properties being exactly the same.
This means that one single set of ordinary differential equations is solved for
the entire group (and
not one set for each droplet) the only difference being that the parcel mass ($m_{parcel}$) is
the sum of all droplet mass ($m_d$). This means that,
\begin{equation}
 m_{parcel} = N_d m_d = N_d\left( \rho_d \frac{\pi D^3}{6}\right) \, ,
\end{equation}
where $m_d$ is the droplet mass, $N_d$ is the number of droplets represented by this parcel and $D$ is the parcel (and the droplet) diameter.

For each time instant, a certain number of parcels is inserted into the domain to represent the
liquid injection. The parcel properties in the nozzle exit are specified as either constant values or random variables sampled from statistical distributions built from experimental data.

In this work, all parcel properties in the nozzle have constant values except for the parcel diameter,
which is a random variable.

\subsection{Secondary Break-Up}

The secondary break-up of droplets is neglected. It is considered that droplets
vaporize (and either disappear or have their diameter reduced) before they
become unstable.

\subsection{Momentum Transfer}

The motion and momentum equations for a particle are:
\begin{equation}\label{eq: drop_mom}
\frac{d\bv{x}_d}{dt} =\bv{U}_d  , \quad m_d \frac{d\bv{U}_d}{dt} = \bv{F} \, .
\end{equation}

The resulting force ($\bv{F}$) on the particle  proposed by \cite{vonkarman} is here
simplified for the case of a rigid spheric droplet under small pressure
gradient and in low droplet Reynolds number. The buoyancy force is
neglected because the gas density is much lower than the liquid density ($\rho \ll \rho_d $).  The effect of mass transfer on the droplet momentum equation was also neglected so that
\begin{equation}
 \frac{d \left(m_d \bv{U}_d \right) }{dt}\approx  m_d\frac{d \bv{U}_d }{dt} \, .
\end{equation}

The droplet slip velocity (the droplet velocity relative to the gas) is given by:
\begin{equation}\label{eq: Urel}
 \bv{U}_{slip} = \bv{U_d} - \underbrace{\left( \tilde{\bv{U}}+\bv{U}'' \right)}_{\text{gas}} \, ,
\end{equation}
where $\tilde{\bv{U}}$ is the Favre-averaged gas velocity computed in Equation \eqref{av: momentum} and $\bv{U}''$ is its fluctuation.
Since $\bv{U}''$ is unknown, it is modeled as explained later in the droplet dispersion subsection.

\begin{equation}\label{eq: drop_mom_F}
 \bv{F}=-\rho\frac{\pi D^2}{8}C_D | \bv{U}_{slip}| \bv{U}_{slip}
+ m_d \bv{g} \, .
\end{equation}

where the drag coefficient and the droplet Reynolds number read:
\begin{equation}
 C_D =\begin{cases}
\frac{24}{Re_d} \left( 1 + \frac{1}{6} Re^{2/3}_d \right) & Re_d < 1000,\\
0.424 & Re_d \geq 1000.
\end{cases}
\end{equation}

\begin{equation}
 Re_d = \frac{\rho | \bv{U}_{slip}| D}{\mu} \, .
\end{equation}

Rearranging \eqref{eq: drop_mom} and \eqref{eq: drop_mom_F},
\begin{equation}\label{eq: drop_dudt}
 \frac{d\bv{U}_d}{dt} = -\frac{ \bv{U}_{slip}}{\tau_{u}} + \bv{g} \, .
\end{equation}

$\tau_u$ is the momentum relaxation time,
\begin{equation}
 \tau_u = \frac{8 m_d}{\pi \rho C_D D^2 | \bv{U}_{slip}|}=\frac{4}{3}
\frac{\rho_d D}{\rho C_D | \bv{U}_{slip}|} \, .
\end{equation}

\subsection{Heat and Mass Transfer}

In the case of an evaporating spray with the droplet temperature below the air temperature,
 energy is transfered from the gas to the droplets (heat transfer) and liquid mass is evaporating if the gas is not saturated with vapor (mass transfer). The liquid temperature will either increase or decrease depending on the rate at which each process takes place.

The droplet heating and vaporization is given a simplified treatment presented in \cite{sirignano} as infinity-liquid-conductivity model, it considers spheric symmetry for the droplet and the surrounding gas with a time varying but spatially uniform temperature of liquid phase.

The droplet energy equation is written as
\begin{equation}\label{eq: drop_energy}
 m_d \frac{dh_d}{dt} = \frac{d m_d}{dt} L_v \left(T_d\right)+Q_d \, ,
\end{equation}
where $L_v\left( T_d\right)$ is the latent heat of evaporation at temperature $T_d$.

The modeling of heat ($Q_d$) and mass ($dm_d/dt$) tranfers  will be divided in two subsections. The heat transfer is discussed first and the mass transfer comes next.

\subsubsection{Gas-Droplet Heat Transfer}

The gas-droplet heat transfer model is developed from two basic assumptions:
\begin{itemize}
 \item Thermal diffusivity is larger in gas than in liquid: this means that changes in temperature of liquid surface instantaneously affect the surrounding gas phase, what is generally true for liquids far from critical conditions. Mathematically, this means that temperature derivatives with respect to time are neglected in the gas phase.
 \item The time required for heat diffusion from droplet surface to its interior is much smaller than the droplet lifetime (the time required for complete evaporation). 
\end{itemize}

Consider the heat flux scheme proposed in \cite{naca} and outlined in Figure \ref{fig: drop_heat}: heat is transfered from air to a film composed of air and vapor and then to the droplet.

The droplet has a temperature $T_d$ and is surrounded by the air-vapor film whose temperature $T_f$ varies radially until it reaches the air temperature $T$ at radius $r=D/2+\delta_f$. The following heat fluxes occur:
\begin{itemize}
  \item Total heat transfered from the air to the film and the droplet: $Q$;
  \item Heat that actually arrives at droplet surface: $Q_d$;
  \item Sensible heat that increases liquid temperature: $Q_L$;
  \item Latent heat of evaporation: $L_v$;
  \item Heat transferred from air to film: $Q_S$.
\end{itemize}

And they are related by the following equation:
\begin{equation}
\begin{split}
  Q_d = Q_L + L_v & = Q - Q_S \, .
\end{split}  
\end{equation}

The enthalpy equation for the air-vapor film surrounding the droplet in spherical coordinates is:
\begin{equation}\label{eq: Qd}
 \frac{d}{dr}\left[ 4\pi r^2 h_c \frac{dT_f}{dr} - \dot{m}_d c_{p,f} \left(T_f -T_d\right)\right] =0 \, ,
\end{equation}
where $r$ is the distance from the droplet center, $h_c$ is the coefficient of convective heat transfer through the film, $\dot{m}_d$ is the rate at which vapor mass evaporates and diffuses out and $c_{p,f}$ is the specific heat in the film. They are all assumed to be constant throughout the film.

Applying the boundary condition at the droplet surface, Equation \eqref{eq: Qd} becomes:
\begin{equation}\label{eq: Qdbc}
 Q_d = 4\pi r^2 h_c \frac{dT_f}{dr} - \dot{m}_d c_{p,f} \left(T_f -T_d\right)
\end{equation}

Assuming that the film thickness is small so that:
\begin{equation}
4\pi r^2 \approx \pi D^2 \quad \text{for} \quad r \in [D/2, D/2+\delta_f] \, ,
\end{equation}
and integrating Equation \eqref{eq: Qdbc}:
\begin{equation}
\int_{D/2}^{D/2+\delta_f} dr = \int_{T_d}^{T} \frac{\pi D^2 h_c}{Q_d +\dot{m}_d c_{p,f} \left(T_f-T_d\right)}dT_f \, ,
\end{equation}
the result is the rate of heat transfer from the air to the droplet:
\begin{equation}\label{eq: Qd_final}
Q_d = \pi D^2 h_c \left( T - T_d \right)\frac{z}{e^z-1} \, ,
\end{equation}
%h_c = [W/m/K]
where
\begin{equation*}
z= - \frac{c_{p,f} \dot{m}_d \delta_f}{ \pi D^2 h_c} \, .
\end{equation*}

The expression for modeling the heat transfer coefficient - $h_c$ - is the one used by \cite{nordin},
\begin{equation}
  \frac{\pi D^2 h_c}{\delta_f} = \pi D \kappa_f Nu_f \, .
\end{equation}

Equation \eqref{eq: Qd_final} is thus: 
\begin{equation}
Q_d = \pi D \kappa Nu \left( T-T_d \right)  \frac{z}{e^z-1} \, .
\end{equation}

The Nusselt number - $Nu$ - is given by the correlation bellow and the Prandtl number - $Pr$ - is computed directly from its definition:
\begin{equation}
 Nu= 2.0 +0.6 Re^{1/2} Pr^{1/3} \, , \quad Pr = \frac{\mu_f c_{p,f}}{\kappa_f} \, .
\end{equation}

All gas properties ($c_{p,f}$, $\mu_f$ and $\kappa_f$) in the above equation should be computed using the characteristic film temperature ($T^{*}_f$) defined by the the so called one-third rule:
\begin{equation}
 T^{*}_f = \frac{2T_d + T}{3} \, .
\end{equation}

The equation for droplet energy \eqref{eq: drop_energy} may also be written in terms of temperature and characteristic time scales for heating - $\tau_h$ - and evaporation - $\tau_e$:
\begin{equation}
 \frac{dT_d}{dt}= \frac{T-T_d}{\tau_h} f - \frac{1}{c_{p,d}} \frac{L_v\left( T_d\right) }{\tau_e} \, ,
\end{equation}
where
\begin{equation}
 \tau_h=\frac{m_d c_{l,d}}{\pi D \kappa Nu} \, , \quad f =  \frac{z}{e^z-1} \, .
\end{equation}

An expression for $\tau_e$ is given by Equation \eqref{eq: tau_e}, derived in the next subsection.

\begin{figure}
 \centering
 \includegraphics[width=0.8\textwidth]{./figuras/chap2/droplet_heat.png} 
 \caption{Scheme of heat flux from air to droplet, adapted from \cite{naca}.}
 \label{fig: drop_heat}
\end{figure}

\subsubsection{Mass Transfer}

The exposition below follows the already commented simplifications and is commonly denoted as $D^2$-law, see \cite{sirignano}.
Only mass transfer from the droplet to the gas is considered, thus condensation is not allowed.

The steady-state continuity equation for the surrounding gas around the droplet in spherical coordinates is:
\begin{equation}\label{eq: film_mass}
 \frac{d}{d r} \left( \rho u r^2 \right) = 0 \quad \rightarrow \quad \rho u r^2 = cte \, ,
\end{equation}
where $r$ is the radial coordinate and $u$ is the radial velocity.

Using the boundary condition at droplet surface,
\begin{equation}\label{eq: film_species_bc}
 \dot{m}_d = \rho u \left( 4\pi r^2  \right) \, .
\end{equation}

The steady-state species equation is:
\begin{equation}\label{eq: film_species}
\frac{d}{d r} \left[ 4\pi r^2 \left( \rho u Y_v -\rho \mathcal{D}_v  \frac{d Y_v}{d r}  \right)\right] = 0\, ,
\end{equation}
where $Y_v$ is the vapor mass fraction and $\mathcal{D}_v$ is the vapor diffusion coefficient. The boundary condition at droplet surface is:
\begin{equation}\label{eq: film_species_bc}
 \left\lbrace 4\pi r^2 \left[ \rho u Y_v -\rho \mathcal{D}_v  \frac{d Y_v}{d r} \right]\right\rbrace  _{r=D/2}  = \dot{m}_d\, . \\
\end{equation}

The integral of \eqref{eq: film_species} using \eqref{eq: film_mass} and \eqref{eq: film_species_bc} may be written as:
\begin{equation}
-\int_{D/2}^{\infty} \frac{\dot{m}_d}{4\pi \rho \mathcal{D}_v}\frac{dr}{r^2} = \int_{Y_{v,s}}^{Y_{v,\infty}} \frac{dY_v}{1-Y_v} \, ,
%- \rho D_v \frac{d Y_v}{d r} = \frac{\dot{m}_d}{4\pi r^2}\left(1-Y_v\right) \, .
\end{equation}
where $Y_{v}$ is the vapor concentration in the air-vapor film. Performing the integration, it becomes:
\begin{equation}
 \frac{dm_d}{dt} = -2\pi D \mathcal{D}_v \rho_v ln \left(
\frac{1-Y_{v,\infty}}{1-Y_{v,s}} \right) \, .
\end{equation}

The vapor mass fraction near the droplet surface ($Y_{v,s}$) is obtained assuming liquid-vapor equilibrium. In this situation, the vapor partial pressure ($p_v$) is given by Clausius-Clapeyron equation:
\begin{equation}
ln \left(\frac{p_v}{p_\infty}\right) = -\frac{L_v W_{v}}{R} \left(\frac{1}{T_d}-\frac{1}{T_b}\right) \, ,
\end{equation}
where $T_b$ is the liquid boiling temperature at pressure $p_{\infty}$ and $L_v$ is the heat of vaporization per unit mass.

Using Raoult's law and the ideal gas state equation, the vapor mass concetration near droplet surface ($Y_{v,s}$) is:
\begin{equation}
Y_{v,s} = \frac{p_v \left(T_d\right)}{p}\frac{W_v}{W} \, .
\end{equation}

The extra mass transfer due to gas motion around the droplet is accounted by the Sherwood number given by the empirical correlation of \cite{ranzmarshall}:
\begin{equation}\label{eq: dropxxxmass}
 \frac{dm_d}{dt} = -2\pi D \mathcal{D}_v \rho_v ln \left(
\frac{1-Y_{v,\infty}}{1-Y_{v,s}} \right) Sh \, ,
\end{equation}
where
\begin{equation}
 Sh = 2.0 +0.6 Re^{1/2} Sc^{1/3} \, .
\end{equation}
As for the Nusselt number, all properties here are computed with the characteristic film temperature ($T^{*}_f$) previously defined.

The droplet evaporation rate ($dm_d/dt$) may then be written as:
\begin{equation}
 \frac{dm_d}{dt}=-\frac{m_d}{\tau_e} \, ,
\end{equation}
where
\begin{equation}\label{eq: tau_e}
 \tau_e = \frac{m_d}{\pi D \mathcal{D} Sh \rho_v ln \left(
\frac{1-Y_{v,\infty}}{1-Y_{v,s}} \right)}
\end{equation}
is the characteristic time scale of droplet evaporation.

% 
% For a droplet temperature above the boiling point, the evaporation occurs at constant temperature and the mass transfer is determined by the rate at which energy is transfered from the gas to the liquid, sustaining the phase change. The equations bellow 
% 
% \begin{equation}
%  \frac{dm_d}{dt}=-\frac{\pi D \kappa Nu}{c_{p,v}} ln \left[ \frac{c_{p,v}}{L_v}
% \left( T- T_d \right) +1 \right]
% \end{equation}
% 
% 
% \begin{equation}
%  \frac{dD}{dt} = -\frac{D}{\tau_{boil}}
% \end{equation}
% 
% \begin{equation}
%  \tau_{boil} = \frac{D^2 \rho_d c_{p,v}}{2 \kappa Nu ln \left[
% \frac{c_{p,v}}{L_v} \left( T- T_d \right) +1 \right]}
% \end{equation}

\subsection{Droplets at Domain Boundaries}

When the droplet crosses the domain boundary, it is simply removed from the computation.

\subsection{Droplet Dispersion}

Although the gas phase velocity has been averaged in \eqref{av: momentum}, the droplet experiences oscillations in its slip velocity if turbulence is present.
This oscillation not only affects the droplet drag, the extra shear on droplet surface might cause deformation and instabilities leading to droplet break-up.
Furthermore, due to changes in the surrounding gas flow, mass and heat transfers are also affected.

Here, the effects of turbulence on droplet motion are simply accounted for by summing a fluctuation ($\bv{U}''$) to the gas averaged velocity ($\tilde{\bv{U}}$):
\begin{equation}
  \bv{U}_{slip} = \bv{U_d} - \left( \tilde{\bv{U}}+\bv{U}'' \right) \, . \tag{\ref{eq: Urel}}
\end{equation}

For the k-epsilon model, a characteristic magnitude for velocity fluctuation may be defined as the scalar
\begin{equation}
 <U''>= \sqrt{\frac{2}{3} k} \, .
\end{equation}

The magnitude of the velocity fluctuation ($\bv{U}''$) is then sampled from a Gaussian distribution with a variance equal to $<U''>$ and zero mean. Finally, the direction of $\bv{U}''$  is randomly chosen.

The sampling is performed once the time passed from the last sampling is greater than the characteristic time $\tau_{turb}$: the minimum between the eddy lifetime and the time taken to the droplet to cross it, given by the ratio of a turbulent length scale ($l_t = C_{\mu}^{3/4} k^{3/2} / \epsilon $) and the magnitude of the droplet slip velocity:

\begin{equation}
 \tau_{turb} = min \left[ \frac{k}{\epsilon} , \frac{k^{3/2}}{\epsilon} \frac{C_{\mu}^{3/4}}{|\bv{U}_{slip}|}\right] \, .
\end{equation}

Further details are discussed in \cite{kiva} and \cite{baumgarten2006mixture}.

### 2016-09-11 master/thesis/chap3.tex
added master files

Chen, St\aa{}rner and Masri performed at Sydney University a detailed experimental investigation of a turbulent evaporating jet of acetone, see \cite{chen}. Droplet diameter, droplet velocity, droplet number density and liquid volumetric flux were measured using a two-component phase Doppler interferometry (PDI) and acetone vapor mass flux was measured using planar laser-induced fluorescence (PLIF).

The experiment consists of a spray nozzle centered on the exit plane of a wind tunnel with 150mm by 150mm that supplies a co-flowing air stream, see Figure \ref{spray_jet}. 
Inside the nozzle, a pressurized liquid jet is surrounded by a carrier air flow until the nozzle exit. The co-flow has a low turbulence intensity of less than $2\%$, so that the effect on the spray jet turbulence is negligible. The main benefit of this setup is the avoidance of flow recirculation near the nozzle exit, what would be an extra complication for the boundary conditions of a numerical simulation.

Acetone evaporation takes place even before the nozzle exit, cooling both the gas and the droplets. The exit temperature shown in Table \ref{table: experiment} was not measured, but computed from energy conservation based on the measurement of acetone vapor mass flux and  the assumption of thermal equilibrium.

\begin{notation}
Sauter mean diameter - SMD or $D_{32}$ - is an average droplet diameter given by:
\begin{equation}
 SMD= \frac{\sum_{droplets} D^3}{\sum_{droplets} D^2} \, .
\end{equation}
\end{notation}

\begin{table}
\centering
 \begin{tabular}{lc}
\hline
Experiment  Data & \\ \hline
Liquid Phase & Acetone \\
Liquid Flow rate at nozzle exit $(g/min)$  &  $7.0$ \\
Carrier air flow rate $(g/min)$ & $135$ \\
Vapor flux at nozzle exit $(g/min)$ & $5.0$ \\
SMD at nozzle exit $(\mu m)$ & $13.7$ \\
Gas temperature at nozzle exit $(K)$ & $280$ \\ 
Gas jet Reynolds Number $(Re=4 \dot{m}_g/\pi D_{nozzle} \mu_g)$ & $16,300$ \\
\hline
 \end{tabular}
\caption{Basic information about the experiment of Chen, St\aa{}rner and Masri, \cite{chen}.}
\label{table: experiment}
\end{table}

\begin{figure}[!htb]
 \centering
 \includegraphics[width=0.4\textheight]{./figuras/chap3/setup.png}
 \caption{Configuration of the spray jet nozzle from \cite{chen}.}
 \label{spray_jet}
\end{figure}

\FloatBarrier
 \section[Boundary Conditions for the Numerical Simulation]{Determining the Boundary Conditions for the Numerical Simulation}

In this section, the boundary conditions for the numerical simulation are obtained from the evailable experimental data. The measurements provide almost complete data for establishing the boundary conditions and few assumptions had to be made.
The list of conditions to be specified for each phase is summarized below.

\subsection{Nozzle}

Below, the boundary conditions at nozzle exit.

\subsubsection{Liquid Phase:}

\begin{description}
  \item[Diameter ($D$):]

The parcel diameter is sampled from a statical distribution built from measurements. \cite{chen} does not provide such measurements at the nozzle exit plane, but only the Sauter mean diameter.  A Lognormal distribution was then specified with $\mu = 10\ \mu m$ and $\sigma=5.5\ \mu m$.

\begin{equation}
 cpf(d; \mu, \sigma) = \frac12 \operatorname{erfc}\!\left[-\frac{\ln D - \mu}{\sigma\sqrt{2}}\right] = \Phi\bigg(\frac{\ln D - \mu}{\sigma}\bigg) \, ,
\end{equation}

\begin{equation}
 pdf(D;\mu,\sigma) = \frac{1}{D \sigma \sqrt{2 \pi}}\, e^{-\frac{(\ln D - \mu)^2}{2\sigma^2}} \, .
\end{equation}

Figure \ref{fig: lognormal} shows the distribution shape. The correspondent Sauter mean diameter is $SMD =14\ \mu m$.
\begin{figure}[!htb]
\centering
  \includegraphics[width=0.75\textwidth]{./figuras/chap3/lognormal.png}
\caption{Lognormal probability density function used for sampling droplet diameter at nozzle exit.}
\label{fig: lognormal}
\end{figure}

\item[Position ($\bv{x}_d$):]
The droplets are inserted in the domain with an aleatory radial coordinate in the interval going from the nozzle axis to the nozzle external radius $(y \in [0, 4.9] \ mm )$, which is sampled for each parcel injection.

  \item[Velocity ($\bv{U}_d$):]
  The droplet injection velocity was made dependent only on the injection position according to Figure \ref{fig: bc_dropU}.

  \begin{figure}[h]
  \centering
  \includegraphics[width=0.65\textwidth]{./figuras/chap3/bc/bc_dropU.png}
  \caption{Droplet injection velocity as a function of its injection position.}
  \label{fig: bc_dropU}
  \end{figure}

  \item[Temperature ($T_d$):] The temperature was determined by assuming thermal equilibrium in the nozzle exit for given mass flow rates of liquid, air and acetone vapor:
  \begin{equation}\label{eq: bc_est_T}
  T_{nozzle}=T_{\infty} - \frac{\dot{m}_{ac} L_{v}}{\dot{m}_g c_{p,g}+\dot{m}_l c_{p,l}} = 280\ K
  \end{equation}
  where $T_{\infty}$ is the ambient temperature and $L_v$ is the latent heat of vaporization of acetone.

  The fixed value of $280\ K$ is specified for droplet and gas temperature.

  \item[Mass flow rate ($\dot{m}_d$):] According to the measurements, the liquid mass flow rate is $\dot{m}_d=2\ g/min$.  The parcel injection rate was $5\times10^5\ parcels/s$, yelding about 35,000 parcels in the computational domain.
  \end{description}

\subsubsection{Gas Phase:}

\begin{description}
  \item[Velocity ($\tilde{\bv{U}}_{nozzle}(\bv{x})$), Turbulent KE ($k_{nozzle}(\bv{x})$) and Dissipation Rate ($\epsilon_{nozzle}(\bv{x})$):] The droplet velocity was obtained from a gas phase only numerical simulation of the nozzle interior with fixed mass flow rate of $\dot{m}_g = 140 g/min$. The turbulence properties in the nozzle inlet was estimated using the following relations from \cite{luppes}:
  \begin{equation}
  k=\frac{3}{2} i^2 \bar{U}_{nozzle}^{2}\, , \quad \epsilon = \frac{c_{\mu}^{3/4} k^{3/2}}{l_m} \, ,
  \end{equation}
  where $\bar{U}_{nozzle}$ is the average velocity inside the nozzle, $i=0.03$ and $l_m = 0.07 D_{nozzle}$.
  The resulting velocity profile was then compared to the mean velocity of the smallest droplets showing good agreement in Figure \ref{fig: bc_gas}.

  \item[Temperature ($\tilde{T}_{nozzle}$):] As explained in the liquid phase section, the fixed value of $280\ K$ was used.
  \item[Species Concentration ($\tilde{Y}_{k,nozzle}$):] Given the measured mass flow rates of air ($\dot{m}_{air}$) and acetone vapor ($\dot{m}_{ac}$):
  \begin{equation}
  \begin{split}
  Y_{ac} &= \frac{\dot{m}_{ac}}{\dot{m}_{ac}+\dot{m}_{air}}=0.0357 \, , \\
  Y_{O2} &= 0.233\left( 1- Y_{ac} \right)=0.224 \, , \\
  Y_{N2} &= 1- Y_{ac}-Y_{O2}=0.740 \, .
   \end{split}
  \end{equation}
  \item[Dynamic Pressure ($p_{d,nozzle}$):] A zero gradient condition was used.
\end{description}

\begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \subfloat[]{\label{fig: bc_Ux}\includegraphics[width=0.5\textwidth]{./figuras/chap3/bc/bc_Ux.png}} &  \subfloat[]{\label{fig: bc_k}\includegraphics[width=0.5\textwidth]{./figuras/chap3/bc/bc_k.png}} \\
  \subfloat[]{\label{fig: bc_eps}\includegraphics[width=0.5\textwidth]{./figuras/chap3/bc/bc_eps.png}} 
\end{tabular}
 \caption{Radial profiles of the boundary conditions for the gas phase at the nozzle exit plane: mean axial velocity (a), turbulent kinetic energy (b) and its dissipation rate (c).}
 \label{fig: bc_gas}
\end{figure}

\FloatBarrier
\subsection{Coflow}

In the co-flow boundary, only the gas phase is present.

\begin{description}
  \item[Velocity ($\tilde{\bv{U}}_{co-flow}$), Turbulent KE ($k_{co-flow}$) and Dissipation Rate ($\epsilon_{co-flow}$):] A fixed value for velocity was used with the same relations for turbulence properties from \cite{luppes}:
  \begin{equation}
  \begin{split}
  &\tilde{U}_{co-flow} =3\ m/s\, ,\\
   & k =\frac{3}{2} i^2 \bar{U}_{nozzle}^{2}= 0.0054\ J/kg \, , \\
   &\epsilon = \frac{c_{\mu}^{3/4} k^{3/2}}{l_m}=0.0124\ J/kg/s \, ,
  \end{split}
  \end{equation}
  where $\bar{U}_{nozzle}$ is the average velocity in the co-flow region, $i=0.02$ and $l_m = 0.07 D_{co-flow}$.

  \item[Temperature ($\tilde{T}_{co-flow}$):] The fixed value of $298\ K$ was used, the ambient temperature.
  \item[Species Concentration ($\tilde{Y}_{k,co-flow}$):] Given the measured mass flow rates of air ($\dot{m}_{air}$) and acetone vapor ($\dot{m}_{ac}$):
  \begin{equation}
  Y_{ac} = 0 \, , \quad Y_{O2} = 0.233 , \quad Y_{N2} = 0.767 \, .
  \end{equation}
  \item[Dynamic Pressure ($p_{d,co-flow}$):] A zero gradient condition was used.
\end{description}

\subsection{Far-field}

Ambient pressure ($p_{\infty}=1\times 10^5\ Pa$) is set for dynamic pressure. Zero gradient is set for the remaining quantities.

\section{Species Equations}

It is only considered the presence of two species: acetone vapor and air. Equation \eqref{av: species1} for the mean mass concentration of acetone vapor ($\tilde{Y}_{ac}$) is then
\begin{equation}
 \frac{\partial \bar{\rho} \tilde{Y}_{ac} }{\partial t} + \nabla
\cdot\left(\bar{\rho} \tilde{\bv{U}} \tilde{Y}_{ac} \right) = \nabla \cdot \left[ \bar{\rho} \left( \nu + \nu_T \right)
\nabla \tilde{Y}_{ac} \right] +\bar{S}_{Yac} \, ,
\end{equation}
The acetone vapor is the scarce species relative to air,
\begin{equation}
\tilde{Y}_{air} = \tilde{Y}_{O2} + \tilde{Y}_{N2}= 1- \tilde{Y}_{ac} \, .
\end{equation}

### 2016-09-11 master/thesis/chap4.tex
added master files

\section{Numerical Solution of Equations}

\subsection{The OpenFOAM Code}

The numerical solution of the equations presented in Chapter \ref{chap: equations} was obtained using the OpenFOAM code. Details of OpenFOAM formulation are explained in \cite{jasak} and \cite{weller}.  From all the set of available solvers and libraries in OpenFOAM, the dieselFOAM solver and dieselSpray class were the major pieces of code used in this work. To my knowledge, both were written by Niklas Nordin, \cite{nordin}.

The dieselSpray class handles the modeling of lagrangian particles and their submodels. Minor modifications were made in order to have more flexibility in boundary conditions and to adapt them to the experimental conditions.

The dieselFOAM solver couples the modeling of the lagrangian particles and the gas flow solution. The spray sources are explicitly treated and the coupling among variables is solved with PISO algorithm, see \cite{jasak} and \cite{ferziger}.

Minor modifications were added to the solution of low Mach number equations instead of the fully compressible formulation.
They are briefly explained here, but the understanding requires from the reader some familiarity with OpenFOAM programming.

The thermodynamic pressure retained its original name \verb|p| and is the pressure used in the state equation:
\begin{verbatim}
<createFields.H>
volScalarField& p = thermo.p();                               
\end{verbatim} 
and in the lagrangian models:
\begin{verbatim}
<createSpray.H>
spray dieselSpray
(
    U,
    rho,
    p,
    T,
    composition,
    gasProperties,
    thermo,
    g
);\end{verbatim}

A new scalar field was assigned to the dynamic pressure, \verb|volumeScalarField pd|:
\begin{verbatim}
<createFields.H>
 volScalarField pd
(
    IOobject
    (
        "pd",
        runTime.timeName(),
        mesh,
        IOobject::MUST_READ,
        IOobject::AUTO_WRITE
    ),
    mesh
);
\end{verbatim}

The momentum equation was modified to be computed using the gradient of \verb|pd| instead of \verb|p| in the momentum predictor:
\begin{verbatim}
<UEqn.H>
     fvVectorMatrix UEqn
    (
        fvm::ddt(rho, U)
      + fvm::div(phi, U)
      + turbulence->divDevRhoReff(U)
     ==
        rho*g
      + dieselSpray.momentumSource()
    );

    if (momentumPredictor)
    {
        solve(UEqn == -fvc::grad(pd));
    }
\end{verbatim}

Finally, the the pressure equation is now a Poisson equation for the dynamic pressure and it uses the thermodynamic presure for computing the density.
\begin{verbatim}
<pEqn.H>

  fvScalarMatrix pdEqn
  (
      fvc::ddt(psi,p)
      + fvc::div(phi)
      - fvm::laplacian(rho*rUA, pd)
      ==
      Sevap
  );
\end{verbatim} 
where \verb|psi| or $\Psi$ is the isothermal compressibility. For an ideal gas:
\begin{equation}
 \rho=p\Psi=\frac{p}{RWT} \, .
\end{equation}

In this work, the time-dependence of the thermodynamic pressure is neglected and the therm \verb|fvc::ddt(psi,p)| vanishes.

\section{Aspects of the Numerical Solution}
The calculations deviate from experiments due to three groups of errors as stated in \cite{jasak} and briefly explained here:

\begin{description}
 \item[Modeling Errors:] the difference of the real flow and the exact solution of the mathematical model.
 \item[Discretization Errors:] the difference between the exact solution of the mathematical model and the exact solution of the discretized equations on the discretized domain (the numerical grid).
 \item[Iteration Convergence Errors:] the difference between the approximate and the exact solution of the discretized equations in the discretized domain. The approximate solution is obtained by iterative methods that reduce the convergence error up to a certain level known as the solver tolerance. 
 \end{description}

 The modeling errors were already discussed as the governing equations were presented in Chapter \ref{chap: equations}. The discretization of equations is treated in next section. The criterion for iteration convergence was set to $10^{-7}$ for the pressure equation and $10^{-6}$ for the remaining variables. The residual of equations were monitored during execution to ensure the validity of the solution.

The equations were solved in segregated linear systems using conjugate gradient algorithms. The solution did not present convergence difficulties as the convergence tolerances were satisfied in few iterations.

\subsection{Discretization of the Governing Equations}
The general transport equation for a scalar intensive property $\phi$ is:
\begin{equation}
 \underbrace{\frac{\partial \rho \phi}{\partial t}}_{\text{temporal derivative}} + \underbrace{\nabla \cdot \left(\rho \bv{U} \phi \right)}_{\text{advection}} - \underbrace{\nabla \cdot \left( \rho \Gamma \nabla \phi \right)}_{\text{diffusion}} = \underbrace{S\left(\phi\right)}_{\text{source}}
\end{equation}

Each term was discretized using the following schemes:
\begin{itemize}
 \item Temporal derivative: Euler implicit method;
 \item Advection term: Gauss theorem is applied to transform the divergence in surface integrals and the upwind scheme is used to interpolate cell centered values to cell faces;
 \item Diffusion term: Gauss theorem with linear interpolation from cell center to faces. Central difference is used for the gradient;
 \item Sources: all sprays sources are explicitly handled. The spray is evolved from time step "n" to "n+1" using the gas phase properties at time step "n", and all source terms are computed. The gas is then evolved to time step "n+1".
\end{itemize}

\subsection{Domain Discretization (The Numerical Grid)}

Two axisymmetric (2D) orthogonal mesh grids were built: a fine and a coarse grid. OpenFOAM uses the collocated grid arrangement. The fine grid was composed of $14,800$ cells ($74$ cells in the radial direction and $200$ in the axial direction) and the coarse grid was composed of $5,460$ cells ($42$ in the radial direction and $130$ in the axial direction). The geometries are shown in Figure \ref{fig: meshes}.

\begin{figure}[!htb]
 \centering
 \includegraphics[height=0.9\textheight]{./figuras/chap4/grid.png}
 \caption{Planar view of the fine (a) and the coarse grid (b).}
\label{fig: meshes}
\end{figure}

Solutions for both grids were obtained for a gas phase only flow using the same boundary conditions described in Chapter \ref{chap: exp}, hence the same Reynolds number, and the same discretization schemes for the equations and the same convergence tolerances previously mentioned.

Figure \ref{fig: grid_test} shows on the left the comparison of radial profiles of mean axial velocity ($\tilde{U}_x$) for the axial coordinates: $x/D=5$, $x/D=15$ and $x/D=25$. In the same Figure, but on the right, it is shown radial profiles of the turbulent kinetic energy ($k$) for the same axial coordinates. In despite of some differences, the coarse grid solution was considered enough accurate and it was used for the results presented in Chapter \ref{chap: results}.

\begin{figure}[h]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.5\textwidth]{./figuras/chap4/coarse_fine/grid_0_U.png} & \includegraphics[width=0.5\textwidth]{./figuras/chap4/coarse_fine/grid_0_k.png} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{./figuras/chap4/coarse_fine/grid_2_U.png} & \includegraphics[width=0.5\textwidth]{./figuras/chap4/coarse_fine/grid_2_k.png} \\
(c) & (d) \\
\includegraphics[width=0.5\textwidth]{./figuras/chap4/coarse_fine/grid_4_U.png} &   \includegraphics[width=0.5\textwidth]{./figuras/chap4/coarse_fine/grid_4_k.png} \\
(e) & (f)
\end{tabular}
 \caption{Comparison of radial profiles obtained for the fine and the coarse meshes for a pure gaseous jet. Mean axial velocity ($\tilde{U}_x$) on the left for axial coordinates: $x/D=5$ in (a), $x/D=15$ in (c) and $x/D=25$ in (e); and turbulent kinetic energy ($k$) on the right: $x/D=5$ in (b), $x/D=15$ in (d) and $x/D=25$ in (f).}
 \label{fig: grid_test}
\end{figure}

It is not clear how mesh refinements will necessarily improve computations for the lagrangian description of droplets. This happens because it is made the assumption that the liquid volume fraction in the computational cell is negligible. If the mesh is refined further and further, this might become untrue. For the present work, the liquid volumetric fraction in the cells near the nozzle is about $10^{-3}$ for the fine mesh and about $10^{-5}$ for the coarse mesh.

Another problem with mesh refinement using the lagrangian-eulerian approach is the differences obtained in interpolating gas properties to parcel positions from one grid to another. This problem has been discussed in \cite{nordin} and \cite{baumgarten2006mixture}.

So far, the recommendation for the mesh dependency is to refine the cells until certain accuracy for the continuous phase is satisfied and not until the eventual limitation of computational cost, what could violate the assumption of a low liquid volume fraction.

% DROPLET TRACKING
% PISO ALGORITHM

### 2016-09-11 master/thesis/chap5.tex
added master files

In this chapter, it is presented results and discussions on numerical simulation of the experiment of \cite{chen} using the methodology introduced in the preceding chapters. The first section shows results of spread rate and self-similar profiles of the gas phase turbulent jet. The second section shows results of droplet and gas velocities and droplet dispersion. The third section presents liquid mass flow rates, vapor mass fluxes and droplet Sauter mean diameter.

\section{The Turbulent Round jet}

Consider the following definitions for the turbulent round jet:
\begin{description}
 \item[Jet half-radius ($y_{1/2}(x)$):] the radial coordinate for which the mean axial velocity is half of the mean axial velocity at the centerline for some axial coordinate $x$:
\begin{equation}
\tilde{U}_x (y=y_{1/2}) = \frac{1}{2}\tilde{U}_x(y=0) \, .
\end{equation}
\item[Dimensionless radial coordinate ($\hat{y}(x,y)$):]
\begin{equation}
 \hat{y} = \frac{y}{y_{1/2}} \, .
\end{equation}

\item[Dimensionless Mean Axial Velocity ($\hat{U}$):]
\begin{equation}
 \hat{U} = \frac{\tilde{U}_x-\tilde{U}_{\text{co-flow}}}{\tilde{U}_x (y=0)-\tilde{U}_{\text{co-flow}}} \, ,
\end{equation}
where $\tilde{U}_{\text{co-flow}}$ is the mean axial velocity of the co-flow stream specified in the boundary condition.

\item[Dimensionless Turbulent Kinetic Energy $\hat{k}$:]
\begin{equation}
 \hat{k}=\frac{k}{(\tilde{U}_x (y=0)-\tilde{U}_{\text{co-flow}})^2} \, .
\end{equation}

\item[Dimensionless Turbulent Viscosity ($\hat{\nu}_T$):]
\begin{equation}
\hat{\nu}_T = \frac{\mu}{\bar{\rho}\tilde{U}_x(y=0) y_{1/2}} \, .
\end{equation}

\item[Jet Spread Rate ($S$):]
\begin{equation}
S = \frac{d y_{1/2} (x)}{dx} \, .
\end{equation}
\end{description}

An empirical observation of a gas turbulent round jet is that the profiles of $\hat{U}$, $\hat{k}$ and $\hat{\nu}_T$ as function of $\hat{y}$ becomes self-similar (i.e. independent of the axial coordinate) from some distance to the nozzle exit.
Furthermore, there is a linear relation between the half-radius ($y_{1/2}$) and the axial coordinate, that is, the jet sprad rate is constant.

\cite{chen} has experimentally shown that self-similarity also occurred for $\hat{U}$ in the presence of the spray jet for $x/D \ge 10$. For arriving at this conclusion, it assumed the velocity measured for the smaller droplets ($D<3\mu m$) as being the gas velocity.

The group of Figures \ref{fig: selfsimilar_profile} shows the same investigation for the gas velocity field obtained in the numerical simulation. The Figure \ref{ssA} shows the profiles of $\hat{U}$ for several axial coordinates. Differently than the experiment, the self-similarity only occurs for $x/D \ge 15$ (and not for $x/D \ge 10$). This suggests that the transition from the jet developing region to the turbulent region is retarded in the simulation.

\begin{notation}
$<f''>$ or $<f'>$ denotes the root mean square of the fluctuation $f''$ (for a Favre average) or $f'$ (for a time average).
\end{notation}

Figures \ref{ssB} and \ref{ssC} show that the evolution to self-similar profiles is slower for the turbulent quantities $\hat{k}$ and $\hat{\nu}_T$ and one may not say that the self-similarity was achieved inside the domain. \cite{chen} has shown that for $x/D \le 25$ it was also not achieved in the experiment for the root mean square of the axial velocity fluctuation ($<U''_x>$).

One observation concerning the turbulent viscosity ($\hat{\nu}_T$) is that the centerline value seems to evolve to a value of $\hat{\nu}_T = 0.047$, higher than values found in the literature for the turbulent round jet. \cite{pope2000turbulent} reports a self-similar turbulent viscosity at centerline of $\hat{\nu}_T = 0.029$.

\begin{figure}[!htb]
 \centering
\begin{tabular}{c}
 \subfloat[]{\label{ssA}\includegraphics[width=0.65\textwidth]{./figuras/chap5/selfsimilar/selfsimilar_U.png}} \\
 \subfloat[]{\label{ssB}\includegraphics[width=0.65\textwidth]{./figuras/chap5/selfsimilar/selfsimilar_k.png}} \\
 \subfloat[]{\label{ssC}\includegraphics[width=0.65\textwidth]{./figuras/chap5/selfsimilar/selfsimilar_nut.png}} \\
\end{tabular}
 \caption{Numerical results showing the evolution to self-similar profiles of the following gas phase properties: (a) dimensionless mean axial velocity - $\hat{U}_x$, (b) dimensionless turbulent kinetic energy - $\hat{k}$ - and (c) dimensionless turbulent viscosity - $\hat{\nu}_t$. The different curves show the profiles for the corresponding axial coordinate indicated in the legend.}
\label{fig: selfsimilar_profile}
\end{figure}

The jet spread rate measures the spread of the axial velocity in the radial direction. A value about $S=0.095$ is found on the literature for the pure gas flow \cite{pope2000turbulent}, and it is independent of Reynolds number. For the spray jet, however, \cite{chen} has reported $S=0.066$. In the simulation, it was found a higher value of $S=0.071$, a difference of $23\%$.

The Figure \ref{fig: ssS} shows the indeed different half-radius values found in the simulation and the experiment for the same axial coordinates, and Figure \ref{fig: ssP} compares the self-similar profiles of $\hat{U}_x$ from the simulation and the experiment.

The overestimation of the round jet spread rate with the default coefficients of k-epsilon turbulence model has been reported previously by different authors. \cite{luppes} has discussed four alternatives to the correction. The modifications consist not only in changing the model coefficients, but also tunning of boundary conditions.

No correction was used in this work because the exact modifications are not known in advance. The effect of this discrepancy on the droplet velocities is discussed in the next section.

\begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \subfloat[]{\label{fig: ssS}\includegraphics[width=0.5\textwidth]{./figuras/chap5/selfsimilar/selfsimilar_spread.png}} & \subfloat[]{\label{fig: ssP}\includegraphics[width=0.5\textwidth]{./figuras/chap5/selfsimilar/selfsimilar_num_exp.png}}
\end{tabular}
 \caption{(a) Half-radius as a function of the axial coordinate in the experiment and in the simulation.  (b) Self-similar radial profile of the numerical and the experimental dimensionless axial velocity - $\hat{U}_x$. The measurements were obtained from \cite{chen}.}
\label{fig: selfsimilar_spread}
\end{figure}

\FloatBarrier
\section{Gas and Droplet Velocities and Droplet Dispersion}

\cite{chen} has presented velocity measurements for droplets of four diameter classes in the spray jet: $D< 5\mu m$, $10\mu m <D<20\mu m$, $20 \mu m < D < 30 \mu m$ and $30 \mu m < D < 40 \mu m$. The importance of separating the data in droplet size classes is that the drag force is dependent on its size.

Figure \ref{fig: drop_drag} shows the time response to drag force as modeled in Equation \eqref{eq: drop_dudt} for droplets with diameter ranging from $10$ to $40\mu m$ and initial velocity of $10\ m/s$. The time for the droplet with diameter of $40\ \mu m$ comes to rest is nearly $10$ times greater than the required for the droplet with $10\ \mu m$.

\begin{figure}
 \centering
\begin{tabular}{c}
 \includegraphics[width=0.65\textwidth]{./figuras/chap5/drag/drop_drag.png}
\end{tabular}
 \caption{Solution of droplet drag model of Equation \eqref{eq: drop_dudt}. Droplet velocity as a function of time for an initial velocity of $10\ m/s$ in stagnant air at standard conditions and for droplet diameters of $10$, $20$, $30$ and $40 \mu m$. }
 \label{fig: drop_drag}
\end{figure}

% The calculation accuracy of the droplet dispersion is mainly dependent on how much of gas turbulence is resolved and the droplets Stoke number. Higher Stokes numbers ($St \ge 10$) allow less resolution of flow turbulence because droplet is more likely to follow the mean velocity field.

This fast response of smaller droplets provides a good estimative of the gas velocity. \cite{chen} has measured the velocity of droplets with $D< 5\ \mu m$ that here will be assumed to represent also the experimental gas velocity. The consequences of the larger spread rate mentioned before may then be evaluated directly in the radial profile of the mean axial velocity.

Figure \ref{fig: Ux_gas} shows in red squares the measured radial profiles of the mean axial velocity for the gas phase ($\tilde{U}_x$) in the axial coordinates of: $x/D=5$, $10$, $15$, $20$ and $25$. The radial profile obtained in the simulation is also shown in a solid black line. It is evident that the underprediction of the gas velocity in the simulation becomes higher in the downstream direction. In the axial coordinate of $x/D=25$, the centerline value in the simulation is about $20\%$ below the experimental value.

\begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.5\textwidth]{./figuras/chap5/Ux/Ux_gas/Ux_gas5.png} & \includegraphics[width=0.5\textwidth]{./figuras/chap5/Ux/Ux_gas/Ux_gas10.png} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{./figuras/chap5/Ux/Ux_gas/Ux_gas15.png} & \includegraphics[width=0.5\textwidth]{./figuras/chap5/Ux/Ux_gas/Ux_gas20.png} \\
(c) & (d) \\
\includegraphics[width=0.5\textwidth]{./figuras/chap5/Ux/Ux_gas/Ux_gas25.png} &   \\
(e) & \\
\end{tabular}
 \caption{Measured and computed radial profiles of the mean axial velocity of the gas phase ($\tilde{U}_x$). Each figure shows the profile in a different axial location: $x/D=5$ in (a), $x/D=10$ in (b), $x/D=15$ in (c), $x/D=20$ in (d) and $x/D=25$ in (e). The measurements were obtained from \cite{chen}.}
 \label{fig: Ux_gas}
\end{figure}

It may be questioned whether the flux of gas momentum in the nozzle exit were correct because the gas exiting temperature was estimated from an assumption of thermal equilibrium with the liquid phase, as explained in Equation \eqref{eq: bc_est_T} in Chapter \ref{chap: exp}. If temperature were lower than in the experiment, then the density and the momentum flux would also be lower.

However, that was not the case. Velocity agreement was good, see Figure \ref{fig: bc_Ux}, and nozzle mass flow rate matches measurements within $1.5\%$. This led to the conclusion that the momentum flux was correctly specified in the nozzle exit and that the discrepancies in velocity were due only to the turbulence modeling.

In fact, \cite{luppes} has reported an error of the same percentual magnitude using the k-epsilon model with the default coefficients. This result is reproduced in Figure \ref{fig: luppes}. The simulation with default k-epsilon coefficients is represented as the curve \textit{simulation 1}. The others curves are results obtained by tunning the model parameters and the boundary conditions.

\begin{figure}[!htb]
 \centering
\begin{tabular}{c}
 \includegraphics[width=0.6\textwidth]{./figuras/chap5/refs/luppes_vel.png} 
\end{tabular}
 \caption{The comparison of four simulations of a turbulent air jet (Re = 37,600) with measurements. The centerline axial velocity is shown. In each plot the axial distance is normalized by the nozzle diameter. Simulation 1:
standard k-epsilon model. Simulation 2: $c_{\mu} =0.06$. Simulation 3: $c_{\mu}=0.06$, $c_{\epsilon, 2} =1.87$. Simulation 4: $c_{\mu}=0.09$, $c_{\epsilon, 2}=1.87$, n =10. Figure reproduced from \cite{luppes}.}
 \label{fig: luppes}
\end{figure}

 \cite{chen} has also reported the root mean square of axial and radial velocities, respectively $<U''_x>$ and $<U''_y>$. Again, the values for the smaller droplets ($D< 5\ \mu m$) are being used to represent the gas velocity. The same quantities are not provided in the simulation for direct comparison. The turbulence model only solves for the turbulent kinetic energy and its dissipation rate. A characteristic velocity, however, may be defined as $<U''>=\sqrt{2/3 k}$. This quantity is shown in Figure \ref{fig: UUx} with $<U''_x>$ and $<U''_y>$. Again, it is shown radial profiles for different axial coordinates.

Three observations are readily made:  
\begin{itemize}
  \item The radial profile of $<U''_x>$ is similar to that of $<U''_y>$;
  \item The characteristic velocity defined for the k-epsilon model ($<U''>$) agrees well with $<U''_x>$ and $<U''_y>$;
 \item $<U''>$ declines faster than $<U''_x>$ and $<U''_y>$ as the axial coordinate increases.
\end{itemize}

The first observation indicates that the turbulent jet differs from the self-similar profiles found in the literature \cite{pope2000turbulent}, where $<U''_x>$ is about twice as big as $<U''_y>$ in the centerline. It remains the question whether this occurs because the jet is not fully developed or because of some spray influence.

The second observation suggests that the k-epsilon model is appropriate to describe the round jet turbulence, providing an appropriate characteristic velocity. The third observation, however, indicates that the turbulent dissipation is overpredicted.

\begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.5\textwidth]{./figuras/chap5/UU/UUx5.png} & \includegraphics[width=0.5\textwidth]{./figuras/chap5/UU/UUx10.png} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{./figuras/chap5/UU/UUx15.png} & \includegraphics[width=0.5\textwidth]{./figuras/chap5/UU/UUx20.png} \\
(c) & (d) \\
\includegraphics[width=0.5\textwidth]{./figuras/chap5/UU/UUx25.png} &  \\
(e) & \\
\end{tabular}
 \caption{Measured radial profiles of the gas phase velocities $<U''_x>$ and $<U''_y>$ and the characteristic velocity defined for the k-epsilon model $<U''>=\sqrt{2/3 k}$ obtained from the simulation. Each figure shows the profile in a different axial location: $x/D=5$ in (a), $x/D=10$ in (b), $x/D=15$ in (c), $x/D=20$ in (d) and $x/D=25$ in (e). The measurements were obtained from \cite{chen}.}
\label{fig: UUx}
\end{figure}

Returning to the mentioned influence of the discrepancies in the mean axial velocity of the gas phase on the droplet velocities, the lower velocity of the gas increases the droplet slip velocity and causes a larger drag to be felt by them. The obvious conclusion is that the same discrepancy in the gas velocity will be observed in the comparison of experimental and numerical values of the droplet velocities.

Figure \ref{fig: Ux} shows in red squares the radial velocity profile measured  by \cite{chen} for droplets in the class of $10\mu m <D<20\mu m$ in the axial coordinates of: $x/D=5$, $10$, $15$, $20$ and $25$.  For each radial position, the velocity value is the ensemble average of the droplets crossing the laser PDI beams. The droplet velocity obtained in the simulation is also shown as black scattered points representing each of them a different droplet in the computation, no averaging was performed.

It is observed that the computed droplet velocity also becomes systematically lower than the measurements in the downstream direction.

\begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.5\textwidth]{./figuras/chap5/Ux/Ux_drops/Ux_x5.png} & \includegraphics[width=0.5\textwidth]{./figuras/chap5/Ux/Ux_drops/Ux_x10.png} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{./figuras/chap5/Ux/Ux_drops/Ux_x15.png} & \includegraphics[width=0.5\textwidth]{./figuras/chap5/Ux/Ux_drops/Ux_x20.png} \\
(c) & (d) \\
\includegraphics[width=0.5\textwidth]{./figuras/chap5/Ux/Ux_drops/Ux_x25.png} &   \\
(e) & \\
\end{tabular}
 \caption{Red squares present measured radial profiles of mean axial velocity of droplets ($\tilde{U}_{d,x}$) in the size class of $10\mu m <D<20\mu m$. Scattered black dots are the computed velocities for each droplet in the numerical simulation. Each figure shows the velocity profile in the axial coordinate specified above the plot window. The measurements were obtained from \cite{chen}.}
 \label{fig: Ux}
\end{figure}

The last result concerning the droplet velocity is the droplet dispersion: the differences in droplet velocities found in different size classes. Figure \ref{fig: jointUV} shows the numerical instantaneous radial velocity ($U_{y}$) as function of radial coordinate for two extreme size classes: $D< 10\ \mu m$ and $D> 30\ \mu m$. It is seen that the smaller droplets are more sensitive to the gas velocity sampled in the turbulent dispersion model introduced in Chapter \ref{chap: equations}, specially near the nozzle exit. The biggest droplets show a smaller dispersion around the zero mean value because of their higher inertia. As the axial coordinate increases, both behaviors become similar.

\begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.5\textwidth]{./figuras/chap5/dispersion/class0_5.png} & \includegraphics[width=0.5\textwidth]{./figuras/chap5/dispersion/class4_5.png} \\
(a) & (b) \\
 \includegraphics[width=0.5\textwidth]{./figuras/chap5/dispersion/class0_10.png} & \includegraphics[width=0.5\textwidth]{./figuras/chap5/dispersion/class4_10.png} \\
(c) & (d) \\
 \includegraphics[width=0.5\textwidth]{./figuras/chap5/dispersion/class0_25.png} & \includegraphics[width=0.5\textwidth]{./figuras/chap5/dispersion/class4_25.png} \\
(e) & (f)
\end{tabular}
 \caption{Scatter plot of the numerical radial velocity of droplets ($U_{y}$) in two different size classes: $D< 10\ \mu m$ on the left or (a), (c) and (e) labels; and $D> 30\ \mu m$ on the right or (b), (d) and (f) labels. The dispersion in the radial velocity is higher for the smaller droplet class and in the vicinity of the nozzle exit.}
\label{fig: jointUV}
\end{figure}

\FloatBarrier
\section{Liquid Mass Flow Rate and Vapor Mass Fluxes}

Correct prediction of droplet evaporation is, together with velocity, of high importance for applications of spray simulation. Reasonably low discrepancies found for droplet velocity in Figure \ref{fig: Ux} already indicates that the evaporation model performed well. This is due to the fact that an incorrect modeling of evaporation would lead to an erroneous droplet diameter, further affecting velocity prediction.

Figure \ref{fig: droplet_flux} shows numerical and experimental liquid mass flow rate for different axial coordinates. Computed values are in good agreement with measurements for $x/D \le 10$. Further downstream, the simulation overpredicts the liquid flow rate. Clearly, this means that droplet evaporation is lower than it should be. In fact, in Figure \ref{fig: vapor_flux}, it is made the comparison of experimental and numerical radial profiles of vapor mass fluxes and the trend is exactly the same: as the axial coordinate increases, less vapor (hence more liquid) is present in simulation than in experiment.

\begin{figure}[!htb]
\centering
  \includegraphics[width=0.6\textwidth]{./figuras/chap5/massflow/drop_mflux.png}
\caption{Droplet mass flow rate at four axial locations. Measurements were obtained from \cite{chen}.}
\label{fig: droplet_flux}
\end{figure}

\begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.45\textwidth]{./figuras/chap5/massflow/mvapor5.png} & \includegraphics[width=0.45\textwidth]{./figuras/chap5/massflow/mvapor10.png} \\
(a) & (b) \\
\includegraphics[width=0.45\textwidth]{./figuras/chap5/massflow/mvapor15.png} & \includegraphics[width=0.45\textwidth]{./figuras/chap5/massflow/mvapor20.png} \\
(c) & (d)
\end{tabular}
 \caption{Radial profiles of mean axial mass flux of acetone vapor: $\dot{m}''_{ac}=\rho \bar{Y}_{ac} \tilde{U}_x$. The measurements were obtained from \cite{chen}.}
 \label{fig: vapor_flux}
\end{figure}

The underestimation of evaporation rate has also been noticed in a large-eddy simulation performed by \cite{bini}, where it was discussed the effect of a subgrid model to the evaporation. It is said that the lack of a subgrid model affects evaporation mainly in the jet core, where the mean vapor mass fraction might be saturated, but negative oscillations might allow for extra vaporization. In fact, the evaporation model presented in Chapter \ref{chap: equations} only deals with the averaged flow properties, and the temperature and mass fraction oscillations, ($T''$) and ($Y''_{ac}$), are not taken into account.

Surely, for a large-eddy simulation, the lack of a subgrid model is less severe than in RANS modeling since at least the non-filtered oscillations are present.

A stochastic model for RANS simulation was proposed in \cite{santanu}. The proposed approach was to establish stochastic behaviors for $T''$ and $Y''_{ac}$ and use the instantaneous values of them ($T$ and $Y_{ac}$) in the heating and evaporation models. The results, however, did not improve significantly in the case it studied. This approach implies an odd assumption that the experimental correlations used for the droplet models, which were developed for a non-oscillating gas field, would work for an oscillating situation by only using instantaneous flow properties. A more profound discussion is found in \cite{sirignano}, where directly modeling of the Sherwood and the Nusselt numbers is suggested. This is the same approach used in \cite{bini} with good results.

It must also be pointed that uncertainties in the gas mean velocity also affect the droplet heating and evaporation prediction again because of the different slip velocity. The correct attempt to improve accuracy of the present computation would be to adjust the turbulence model to firstly correct the gas mean velocity and see the new evaporation rates. Next, a stochastic subgrid model for the heat and mass transfer coefficients could be studied.

\begin{figure}[!htb]
 \centering
\begin{tabular}{c}
 \includegraphics[width=0.5\textwidth]{./figuras/chap5/smd/smd5.png} \\
(a) \\
 \includegraphics[width=0.5\textwidth]{./figuras/chap5/smd/smd15.png} \\
(b) \\
 \includegraphics[width=0.5\textwidth]{./figuras/chap5/smd/smd25.png} \\
(c) 
\end{tabular}
 \caption{Numerical and experimental radial profiles of Sauter mean diameter (SMD). The measurements were obtained from \cite{chen}.}
 \label{fig: SMDprofile}
\end{figure}

The last studied property of the spray is the droplet Sauter mean diameter (SMD). Figure \ref{fig: SMDprofile} shows the numerical and experimental radial profiles of SMD.

The computed Sauter mean diameter shows a good agreement with measurements, in despite of the uncertainty about the droplet size distribution in the boundary conditions and the higher liquid flow rates in Figure \ref{fig: droplet_flux}. This may also be a consequence of the low variance in droplet diameter in the nozzle injection.

This agreement also confirms that the velocity discrepancy is caused by the anomalous jet spread rate. As discussed previously, if diameters were also in disagreement, the predicted drag force on the droplets would be further incorrect.

\FloatBarrier

### 2016-09-11 master/thesis/chap6.tex
added master files

%\section{Considerações finais}

This work presented a set of equations to model the evolution of a spray jet and its numerical solution. Computations were compared to available measurements indicating a good general prediction of flow quantities with some disagreements susceptible to improvements in the droplet velocity and evaporation rate.

Velocity of both gas and liquid phases were underpredicted due to the high turbulent viscosity computed by the turbulence model. It is suggested in the literature that the discrepancies for a round jet may be diminished by changing the model coefficients and reducing the turbulent length scale in jet inlet boundary conditions.

Evaporation rate was underpredicted after the jet developing region. Turbulence effects on Nusselt and Sherwood numbers were not modeled and this is believed to decrease evaporation, specially in the jet core, as reported by \cite{bini}.

Further, disagreements in the prediction of gas velocity may also have contributed to differences in the evaporation rate. The maximum error in liquid mass flow rate at some axial location was about $20\%$.

The shape of radial profile of vapor mass fluxes was well predicted. The same can be said to the SMD radial profiles.

Future work could explore improvements in the round jet turbulence modeling and new heating and evaporation models for RANS simulations accounting for fluctuations of flow properties. Something similar to what was done in \cite{bini} could serve as a starting point.

The hypothesis of unity Prandtl and Schmidt numbers was also not tested.

Finally, the OpenFOAM source code offers powerful and flexible libraries that may be adapted to user needs. It consists of a great platform for collaborative code development in fluid dynamics. It is also an alternative for industry in the development of in-house solutions.

### 2016-09-11 master/thesis/resumo.tex
added master files

Este trabalho consistiu na simulação numérica de um spray: um escoamento bifásico composto por uma fase gasosa e uma fase líquida. A fase gasosa é tratada como um meio contínuo, e a fase líquida é tratada com um conjunto de gotas dispersas na fase gasosa. Foi aplicada a aproximação assintótica de baixo número de Mach para a fase gasosa com o objetivo de modelar variações da densidade causadas por gradientes de temperatura sem o envolvimento de complicações existentes na formulação de um escoamento compressível. Os efeitos da turbulência na fase gasosa foram modelados utilizando o conceito de viscosidade turbulenta governada por duas equações diferenciais que compõe o assim denominado modelo \textit{k-epsilon}. A fase líquida dispersa foi modelada como partículas-pontos, às quais são atribuídas propriedades termodinâmicas e modelos 0-d para cálculos de transferência de momento, energia e massa. A fase gasosa foi discretizada segundo o método de volumes finitos. Os resultados numéricos foram comparados com medições experimentais e indicaram que a metodologia utilizada descreve razoavelmente bem o spray. Foram verificadas, porém, subestimativas da velocidade e da taxa de evaporação das gotas com relação às medições experimentais.

### 2016-09-11 master/tutorial/tutorial.tex
added master files

\documentclass[a4paper,10pt]{article}
\usepackage{amssymb,amsmath}
\usepackage{graphicx}
\usepackage{html}
\usepackage[utf8x]{inputenc}
\usepackage{listings}

%opening
\title{\textit{lowmachSolver}: a low Mach number solver for spray simulation with OpenFOAM}
\author{Rodrigo B. Piccinini}

\begin{document}

\maketitle

\begin{abstract}
\textit{lowmachSolver} is an OpenFOAM solver that deals with two-phase flows composed of a continuous gaseous phase and a liquid dispersed phase.
\end{abstract}

\section{Introduction}
\textit{lowmachSolver} is an OpenFOAM solver that deals with two-phase flows composed of a continuous gaseous phase and a liquid dispersed phase. It is mainly based on \textit{dieselFOAM} and it was written only for purpose of my master thesis.

\subsection{OpenFOAM Code}

By \htmladdnormallink{OpenCFD own words}{http://www.openfoam.com/features/}:

\begin{quote}
The OpenFOAM  (Open Field Operation and Manipulation) CFD Toolbox is a free, open source CFD software package produced by OpenCFD Ltd. It has a large user base across most areas of engineering and science, from both commercial and academic organisations. OpenFOAM has an extensive range of features to solve anything from complex fluid flows involving chemical reactions, turbulence and heat transfer, to solid dynamics and electromagnetics. It includes tools for meshing, notably snappyHexMesh, a parallelised mesher for complex CAD geometries, and for pre- and post-processing. Almost everything (including meshing, and pre- and post-processing) runs in parallel as standard, enabling users to take full advantage of computer hardware at their disposal.

By being open, OpenFOAM offers users complete freedom to customise and extend its existing functionality, either by themselves or through support from OpenCFD. It follows a highly modular code design in which collections of functionality (e.g. numerical methods, meshing, physical models, …) are each compiled into their own shared library. Executable applications are then created that are simply linked to the library functionality. OpenFOAM includes over 80 solver applications that simulate specific problems in engineering mechanics and over 170 utility applications that perform pre- and post-processing tasks, e.g. meshing, data visualisation, etc. \end{quote}

Details of OpenFOAM formulation are explained in \cite{jasak} and \cite{weller}.  From all the set of available solvers and libraries in OpenFOAM, the dieselFOAM solver and dieselSpray class (as implemented in version 1.7.1 of OpenCFD release) were the major pieces of code used in this work. To my knowledge, both were written by Niklas Nordin, \cite{nordin}.

The dieselSpray class handles the modeling of lagrangian particles and their submodels. Minor modifications were made in order to have more flexibility in boundary conditions and to adapt them to the experimental conditions.

The dieselFOAM solver couples the modeling of the lagrangian particles and the gas flow solution. The spray sources are explicitly treated and the coupling among variables is solved with PISO algorithm, see \cite{jasak} and \cite{ferziger}.

Minor modifications were added to the solution of low Mach number equations instead of the fully compressible formulation. They are briefly explained here, but the understanding requires from the reader some familiarity with OpenFOAM programming.

The thermodynamic pressure retained its original name \verb|p| and is the pressure used in the state equation:
\begin{verbatim}
<createFields.H>
volScalarField& p = thermo.p();                               
\end{verbatim} 
and in the lagrangian models:
\begin{verbatim}
<createSpray.H>
spray dieselSpray
(
    U,
    rho,
    p,
    T,
    composition,
    gasProperties,
    thermo,
    g
);\end{verbatim}

A new scalar field was assigned to the dynamic pressure, \verb|volumeScalarField pd|:
\begin{verbatim}
<createFields.H>
 volScalarField pd
(
    IOobject
    (
        "pd",
        runTime.timeName(),
        mesh,
        IOobject::MUST_READ,
        IOobject::AUTO_WRITE
    ),
    mesh
);
\end{verbatim}

The momentum equation was modified to be computed using the gradient of \verb|pd| instead of \verb|p| in the momentum predictor:
\begin{verbatim}
<UEqn.H>
     fvVectorMatrix UEqn
    (
        fvm::ddt(rho, U)
      + fvm::div(phi, U)
      + turbulence->divDevRhoReff(U)
     ==
        rho*g
      + dieselSpray.momentumSource()
    );

    if (momentumPredictor)
    {
        solve(UEqn == -fvc::grad(pd));
    }
\end{verbatim}

Finally, the the pressure equation is now a Poisson equation for the dynamic pressure and it uses the thermodynamic pressure for computing the density.
\begin{verbatim}
<pEqn.H>

  fvScalarMatrix pdEqn
  (
      fvc::ddt(psi,p)
      + fvc::div(phi)
      - fvm::laplacian(rho*rUA, pd)
      ==
      Sevap
  );
\end{verbatim} 
where \verb|psi| or $\Psi$ is the isothermal compressibility. For an ideal gas:
\begin{equation}
 \rho=p\Psi=\frac{p}{RWT} \, .
\end{equation}

The time-dependence of the thermodynamic pressure was neglected (valid for open domains) and the therm \verb|fvc::ddt(psi,p)| vanishes.

\section{Folder Structure}

The thesis work is organized in a typical SVN folder structure: 
\begin{verbatim}
--- branches
--- tags
--- trunk
--- wiki
\end{verbatim}
After finishing the thesis, everything was merged to \textbf{trunk} folder and the others were left empty.

\htmladdnormallink{./trunk}{../../} folder has the following structure:
\begin{verbatim}
--- trunk
    +-- case
    |   +-- halle			test case from Sommerfeld, and H.-H. Qiu (not fully finished).
    |   |   +-- 2D				2D mesh and case setup.
    |   |   +-- exp				experimental data.
    |   |   +-- misc			miscellaneous Python scripts.
    |   |
    |   +-- sydney			test case from Chen et al (Sydney University). In the thesis.
    |       +-- case			fine mesh setup.
    |       +-- case.coarse			coarse mesh setup.
    |       +-- ic.coarse			Initial conditions for the coarse mesh case.
    |       +-- ic.fine			Initial conditions for the fine mesh case.
    |       +-- misc			miscellaneous Python scripts for plotting results and building mesh.
    |
    +-- code
    |   +-- 1.7.x			code based on 1.7.1 release from OpenCFD
    |       +-- lowmachLib			lagrangian spray lib.
    |       +-- lowmachSolver		low Mach number solver coupled with lowmachLib.
    |       +-- myLiquids			acetone liquid and vapor properties
    |       +-- mypdfs			spray probability density functions
    |
    +-- paper			thesis results in paper format
    |   +-- elsevier			Elsevier latex class
    |
    +-- ppt				presentation given in 2011/nov/25 at ITA.
    |   +-- imgs				figures in presentation.
    |       +-- visit				figures post-processed in visIT
    |
    +-- thesis			thesis .tex source files.
    |   +-- figuras			figures with .dat files for each of the subsequent thesis appendix and chapters.
    |       +-- appA1
    |       +-- appA2
    |       +-- appA3
    |       +-- chap1
    |       +-- chap2
    |       +-- chap3
    |       +-- chap4
    |       +-- chap5
    |
    +-- tutorial			this tutorial source files.
	+-- html				html files from latex2html.
\end{verbatim}

\section{Compiling}

Compiling was tested with Ubuntu Natty (numbered 11.04) and .deb package of OpenFOAM 1.7.1 released by OpenCFD.

\htmladdnormallink{http://www.openfoam.org/archive/1.7.1/download/}{http://www.openfoam.org/archive/1.7.1/download/}

Development tools for Ubuntu must be present. They can be installed by:
\begin{verbatim}
 sudo apt-get install build-essential flex bison cmake zlib1g-dev qt4-dev-tools libqt4-dev gnuplot libreadline-dev libxt-dev
\end{verbatim}
For parallel runs, installation of the following libraries is also required:
\begin{verbatim}
sudo apt-get install libscotch-dev libopenmpi-dev
\end{verbatim}

Code is located in \htmladdnormallink{./trunk/code/1.7.x/}{../../code/1.7.x}
\begin{verbatim}
--- trunk
    +-- code
        +-- 1.7.x			code based on 1.7.1 release from OpenCFD
            +-- lowmachLib			lagrangian spray lib.
            +-- lowmachSolver			low Mach number solver coupled with lowmachLib.
            +-- myLiquids			acetone liquid and vapor properties
            +-- mypdfs				spray probability density functions
\end{verbatim}

Inside \htmladdnormallink{./trunk/code/1.7.x/}{../../code/1.7.x}, the following scripts may be issued to:
 \begin{verbatim}
# ./Allwmake			compile the solver and all needed libraries.
# ./CleanSrc			clean the source files for a new compiling.
# ./CleanAll			clean both source and binary files for a new compiling.
\end{verbatim}

\textbf{Allwmake} script will make a user installation after compiling the code.

\section{Running}
lowmachSolver can be run as any OpenFOAM solver with usual options:
\begin{verbatim}
#lowmachSolver -help

Usage: lowmachSolver [-parallel] [-case dir]  [-help]
\end{verbatim}

See Test Case section \ref{sydney} for an example.

\section{Test Case}\label{sydney}

This is the test case used to provide results for the thesis. It consists of the numerical simulation of the experiment performed by \cite{chen} in Sydney University.

\htmladdnormallink{http://dx.doi.org/10.1016/j.ijmultiphaseflow.2005.09.002}{http://dx.doi.org/10.1016/j.ijmultiphaseflow.2005.09.002}

Case folder structure
\begin{verbatim}
--- case
    +-- 0			initial and boundary conditions
    +-- chemkin			chemkin files
    +-- constant		mesh and physical properties
    +-- system			solution parameters
\end{verbatim}

\paragraph{./case/0/}
Folder \textbf{case/0/} contains initial and boundary conditions. The type of each boundary conditions is presented in the thesis \cite{piccinini_msc} and may be directly read in the files as well.

Iinitial conditions of the gaseous flow were obtained by simulating only the gas flow in \htmladdnormallink{../cases/sydney/ic.fine/}{../cases/sydney/ic.fine/} and \htmladdnormallink{../cases/sydney/ic.coarse/}{../cases/sydney/ic.coarse/}.

\paragraph{./case/chemkin/}
Chemkin folder contains chemical kinetic and thermodynamic data of all present species in gas phase. No chemical reaction is present. Thermodynamic and transport properties of acetone in liquid phase are computed "inside" the code.

\paragraph{./case/constant/}
\begin{description}
 \item[turbulenceProperties] informs that the simulation is turbulent and it is using RANS modeling.

 \item[RASProperties] selects the turbulence model: standard k-epsilon.

 \item[g] defines the gravity vector field.

 \item[thermophysicalProperties] defines the thermodynamic and transport models for gas and liquid phases.
\begin{verbatim}
thermoType      hsPsiMixtureThermo<reactingMixture<gasThermoPhysics>>; 
\end{verbatim}

hsPsiMixtureThermo computes enthalpy for combustion mixture based on sensible enthalpy $h_s$ and compressibility $\Psi$.
reactingMixture  is a combustion mixture using thermodynamics and reaction schemes.
gasThermoPhysics is a type definition for \verb|sutherlandTransport<specieThermo<janafThermo<perfectGas>>>|.

 \item[combustionProperties] no combustion is present, so the important information is \verb|ignite off;|.

 \item[chemistryProperties] chemical reactions are not solved. Just like before, \verb|chemistry off;| is what matters.

 \item[injectorProperties] defines almost all droplet properties in the injector exit: position, temperature, velocity magnitude and direction, mass flow rate and parcels per second (nParcels). Droplet diameter is specified in next file.

 \item[sprayProperties] defines all droplet modeling and the diameter probability density function (pdf).

The time step for integrating droplet equations is the minimum between momentum and heat transfer characteristic times divided by \verb|subCycles|. \verb|dtInjection| was added to limit the number of parcels in the domain: only one parcel is added during the specified time interval.
\begin{verbatim}
subCycles          2;
dtInjection		1e-9;           		     
\end{verbatim}

Atomization is modeled statistically by the pdf function and no physical modeling is used.
\begin{verbatim}
atomizationModel        off;\end{verbatim}

Since acetone is a highly volatile fuel, droplets are more likely to evaporate before undergo secondary break-up instabilities.
\begin{verbatim*}
includeOscillation      off;
//breakupModel    TAB;
//breakupModel    ETAB;
//breakupModel    ReitzDiwakar;
breakupModel    off;\end{verbatim*}

The injector type is set.
\begin{verbatim}
injectorModel           hollowConeInjector;\end{verbatim}

Collision model is not used since the spray is dilute.  Evaporation, heat transfer, drag and droplet dispersion is modeled as discussed in \cite{piccinini_msc}.
\begin{verbatim}
collisionModel  off;
evaporationModel standardEvaporationModel;
heatTransferModel RanzMarshall;
dispersionModel stochasticDispersionRAS;
dragModel       standardDragModel;\end{verbatim}

Added the lognormal pdf to the available pdfs. It is used to sample droplet diameter for injection.
\begin{verbatim}
hollowConeInjectorCoeffs
{
    dropletPDF
    {
			pdfType         lognormal;
    }
}   
\end{verbatim}

 \end{description}

\subsection{Solver Output}

Basic information about the spray and execution timing followed by extremal values of temperature, density and dynamic pressure.
\begin{verbatim}
Number of parcels in system.... | 29921
Injected liquid mass........... | 6.05092 mg
Liquid Mass in system.......... | 0.522468 mg
SMD, Dmax...................... | 16.0027 mu, 42.7612 mu
ExecutionTime = 102.28 s  ClockTime = 104 s
T max/min : 298 275.949
rho max/min : 1.28461 1.1644
pd max/min : 100000 99966.9
\end{verbatim}

Mass flow rates trough defined patches (walls are being called as such, but they are actually the far boundary).
\begin{verbatim}
Courant Number mean: 0.00232359 max: 0.242796
MassFlows:   nozzle = -3.43682e-05  coflow = -0.000852618  outlet = 0.00100765  walls = -0.00012141
Time = 0.330113
\end{verbatim}

The following lines show the sequence equations are solved: first spray is updated, then all transport equations are solved. "\verb|Solving for rho|" presents zero residual and iterations because it is being explicitly solved.
\begin{verbatim}
Evolving Spray
Solving chemistry
diagonal:  Solving for rho, Initial residual = 0, Final residual = 0, No Iterations 0
DILUPBiCG:  Solving for Ux, Initial residual = 5.77748e-07, Final residual = 5.2592e-09, No Iterations 1
DILUPBiCG:  Solving for Uy, Initial residual = 2.6999e-05, Final residual = 7.62742e-11, No Iterations 2
DILUPBiCG:  Solving for Uz, Initial residual = 5.88654e-05, Final residual = 7.16722e-08, No Iterations 1
DILUPBiCG:  Solving for aC3H6O, Initial residual = 3.79856e-07, Final residual = 8.03328e-10, No Iterations 1
DILUPBiCG:  Solving for O2, Initial residual = 3.88094e-07, Final residual = 3.96676e-10, No Iterations 1
DILUPBiCG:  Solving for CO2, Initial residual = 0, Final residual = 0, No Iterations 0
DILUPBiCG:  Solving for H2O, Initial residual = 0, Final residual = 0, No Iterations 0
DILUPBiCG:  Solving for hs, Initial residual = 5.01197e-07, Final residual = 1.08758e-09, No Iterations 1
\end{verbatim}

The following step is the PISO loop, which might be repeated more than once (I used 3 to 5 loops).
\begin{verbatim}
DICPCG:  Solving for pd, Initial residual = 0.0447168, Final residual = 9.75726e-10, No Iterations 185
diagonal:  Solving for rho, Initial residual = 0, Final residual = 0, No Iterations 0
time step continuity errors : sum local = 3.66451e-16, global = -2.59971e-18, cumulative = -7.82325e-16
\end{verbatim}

For last, solution of k-epsilon equations.
\begin{verbatim}
DILUPBiCG:  Solving for epsilon, Initial residual = 5.26145e-07, Final residual = 5.43398e-10, No Iterations 1
DILUPBiCG:  Solving for k, Initial residual = 1.22238e-07, Final residual = 1.64783e-10, No Iterations 1
\end{verbatim}

\subsection{Some Results}

\begin{figure}
 \centering
 \includegraphics[angle=90,width=\textwidth]{../thesis/figuras/appA2/visit_U.png}
 \caption{Magnitude of droplet velocities. The computational domain is only half of the shown above, it is here mirrored for visualization purpose.}
 \label{fig: dropU}
\end{figure}

\begin{figure}
\begin{center}
 \includegraphics[angle=90,width=\textwidth]{../thesis/figuras/appA2/visit_k.png}
 \end{center}
 \caption{Gas mean turbulent kinetic energy. The computational domain is only half of the shown above, it is here mirrored for visualization purpose.}
 \label{fig: field_k}
\end{figure}

\section{Links}

Openfoam:
\htmladdnormallink{OpenFOAM}{http://www.openfoam.com/}
\htmladdnormallink{OpenFOAM-extend}{http://www.extend-project.de}

Academics:
\htmladdnormallink{ITA}{http://www.ita.br}
\htmladdnormallink{LCFT/ITA}{http://lcft.mec.ita.br}

\bibliographystyle{plain}
\bibliography{../thesis/references}
\end{document}

### 2016-09-11 master/cases/halle/misc/CFD-3DvEXP.py

@@ -1,0 +1,33 @@
+#!/usr/bin/env python
+
+from pylab import *
+
+pathCFD='/home/rodrigo/master/run/spraybench/3D/RANS.kepsilon/sets/4550/'
+#pathCFD='/home/rodrigo/OpenFOAM/run/spraybench/3D/rans/sets/3700/'
+pathEXP='../exp/gasflow/'
+
+
+locs=[2,25,50,100,200,300,400]
+figure(1)
+for i in locs:
+	x=loadtxt(pathEXP+'z'+str(i)+'mm.dat',skiprows=2)
+	x[:,0]=1e-3*x[:,0]
+	y=loadtxt(pathCFD+'z'+str(i)+'mm_U.xy')
+	figure()
+	plot(x[:,0],x[:,1],'ro',y[:,0],-y[:,3],'b-',-y[:,0],-y[:,3],'b-')
+	title('z = '+str(i)+'mm')
+	savefig(pathCFD+'../../imgs/'+'z'+str(i)+'mm.png')
+
+figure(2)
+for i in locs:
+	x=loadtxt(pathEXP+'z'+str(i)+'mm.dat',skiprows=2)
+	x[:,0]=1e-3*x[:,0]
+	y=loadtxt(pathCFD+'z'+str(i)+'mm_k.xy')
+	figure()
+	plot(x[:,0],x[:,2],'rx-',x[:,0],x[:,4],'gx-',y[:,0],sqrt(2./3.*y[:,1]),'k-')
+	xlim([0,0.1])
+	title('z = '+str(i)+'mm')
+	legend(('Urms','Vrms','sqrt(2/3k)'))
+	savefig(pathCFD+'../../imgs/'+'z'+str(i)+'mm_k.png')
+
+#show()

### 2016-09-11 master/cases/halle/misc/injectors.py

@@ -1,0 +1,84 @@
+#!/usr/bin/env python
+
+from pylab import *
+
+temperature=273. +35.
+massflow=array([1.6483E-05, 8.0020E-05, 1.8868E-04, 2.6372E-04, 1.6256E-04, 7.3488E-05, 2.4049E-05, 8.4273E-06, 3.1542E-06, 1.4191E-06])
+print sum(massflow)
+t0=0.
+tf=5.
+
+diameter=0.001
+
+x=array([4,5,6,7,8,9,10,11,12,13])*1e-3
+ninj=len(x)
+vel=array([12.56, 15.66, 17.39, 15.72, 12.75, 10.79, 9.89, 9.49, 9.29, 9.16])
+
+interval=tf-t0
+
+oneDrop=1000.*4./3.*pi*(30e-6)**3
+nParcels=1 #int(mass/oneDrop)
+
+f=open('injectorProperties','w')
+f.write('/*---------------------------------------------------------------------------*\ \n')
+f.write('| =========                 |                                                 | \n')
+f.write('| \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox           | \n')
+f.write('|  \\    /   O peration     | Version:  1.6                                   | \n')
+f.write('|   \\  /    A nd           | Web:      http://www.openfoam.org               | \n')
+f.write('|    \\/     M anipulation  |                                                 | \n')
+f.write('\*---------------------------------------------------------------------------*/ \n')
+f.write('\n')
+f.write('FoamFile \n')
+f.write('{')
+f.write('    version         2.0; \n')
+f.write('    format          ascii; \n')
+f.write('\n')
+f.write('    root            "";\n')
+f.write('    case            "";\n')
+f.write('    instance        "";\n')
+f.write('    local           "";\n')
+f.write('\n')
+f.write('    class           dictionary; \n')
+f.write('    object          injectorProperties;\n')
+f.write('}\n')
+f.write('\n')
+f.write('// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * // \n')
+
+f.write('(')
+
+for i in range(ninj):
+	f.write('\n\n// z = '+str(x[i])+' ------------------------------------------------------------------------- \n')
+	f.write('\n    {')
+	f.write('\n        injectorType        definedInjector;')
+	f.write('\n        definedInjectorProps')
+	f.write('\n        {')
+	f.write('\n            position        ('+str(x[i])+' 0 1e-4);')
[…]

### 2016-09-11 master/cases/halle/misc/mesh.py

@@ -1,0 +1,195 @@
+#!/usr/bin/env python
+from pylab import *
+import os
+
+f=open('/home/piccinini/master/icodieselspray/trunk/cases/2D/ico_spray/constant/polyMesh/blockMeshDict','w')
+f.write('/*---------------------------------------------------------------------------*\ \n')
+f.write('| =========                 |                                                 | \n')
+f.write('| \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox           | \n')
+f.write('|  \\    /   O peration     | Version:  1.4                                   | \n')
+f.write('|   \\  /    A nd           | Web:      http://www.openfoam.org               | \n')
+f.write('|    \\/     M anipulation  |                                                 | \n')
+f.write('\*---------------------------------------------------------------------------*/ \n')
+f.write('\n')
+f.write('FoamFile \n')
+f.write('{')
+f.write('    version         2.0; \n')
+f.write('    format          ascii; \n')
+f.write('\n')
+f.write('    root            "";\n')
+f.write('    case            "";\n')
+f.write('    instance        "";\n')
+f.write('    local           "";\n')
+f.write('\n')
+f.write('    class           dictionary; \n')
+f.write('    object          blockMeshDict;\n')
+f.write('}\n')
+f.write('\n')
+f.write('// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * // \n')
+
+f.write('\n')
+f.write('convertToMeters 1;\n')
+f.write('\n')
+f.write('vertices (')
+
+wedge=pi/180.*5.
+z0,z1,z2,z3=-0.5,0.0,0.2,1.5
+
+# fine mesh
+nr=['30','65','85','60']
+nt='1'
[…]

### 2016-09-11 master/cases/sydney/case.coarse/getlast.sh

@@ -1,0 +1,5 @@
+#!/bin/sh
+rm -rf VTK*
+HOST=ec2-46-51-159-229.eu-west-1.compute.amazonaws.com
+scp -i ~/master/amazon/piccinini.pem ubuntu@$HOST:/home/ubuntu/master/sydney/case.coarse/VTK.tar.gz .
+tar xvfz VTK.tar.gz

### 2016-09-11 master/cases/sydney/case.coarse/run.sh

@@ -1,0 +1,4 @@
+#!/bin/sh
+rm -r processor*
+decomposePar
+mpirun -np 8 lowMachDieselFoam -parallel > log &

### 2016-09-11 master/cases/sydney/ic.coarse/getlast.sh

@@ -1,0 +1,4 @@
+#!/bin/sh
+rm -rf VTK*
+scp -i ~/master/amazon/piccinini.pem ubuntu@ec2-79-125-86-170.eu-west-1.compute.amazonaws.com:/home/ubuntu/master/sydney/case/VTK.tar.gz .
+tar xvfz VTK.tar.gz

### 2016-09-11 master/cases/sydney/misc/ic/profiles.py

@@ -1,0 +1,46 @@
+#!/usr/bin/env python
+
+from pylab import *
+
+D=0.0098
+print '(x/D)max = ', 0.5/D
+print '(y/D)max = ', 0.075/D
+
+
+path='fineGrid/'
+rh=[]
+locs=arange(15,50,5) #[5,20,40]
+for n in locs:
+	file=path+'sets/5000/Profiles'+str(n)+'_U.xy'
+	y=loadtxt(file,delimiter=' ')[:,0]
+	u=loadtxt(file,delimiter=' ')[:,1]
+	ub=(u-3.)/(u[0]-3.)
+	ih=argmin(abs(u-0.5*u[0]))
+	rh.append(y[ih])
+	print '(x/D,rh,error) =', n,rh[-1],min(abs(u-0.5*u[0]))/u[0]
+	
+	figure(1)
+	plot(y/rh[-1],ub,'-',label='x/D = '+str(n))
+	xlabel('y/D')
+	ylabel('Axial Velocity - U (m/s)')
+	title('Axial Position - x/D = '+str(n))
+	legend()
+	
+	file=path+'sets/5000/Profiles'+str(n)+'_k.xy'
+	y=loadtxt(file,delimiter=' ')[:,0]
+	k=loadtxt(file,delimiter=' ')[:,1]
+	figure(2)
+	plot(y/rh[-1],2./3.*k/u[0]**2,'-',label='x/D = '+str(n))
+	xlabel('y0.5/D')
+	ylabel('Turbulent Kinetic Energy - k (m2/s2)')
+	legend()
+	title('Axial Position - x/D = '+str(n))
+xlim([0,2.2])
+
+figure(3)
+S=(0.594595-1.39189)/(8.04545-18.9545)
+B=sum(array(rh)+S*locs)/len(locs)
+
+plot(locs,array(rh)/D,'x-')
+plot(locs,S*locs+B,'r-')
+show()

### 2016-09-11 master/cases/sydney/misc/massflow/droplet_flux.py

@@ -1,0 +1,193 @@
+#!/usr/bin/env python
+
+from pylab import *
+
+def adjust_spines(ax,spines):
+    for loc, spine in ax.spines.iteritems():
+        if loc in spines:
+            spine.set_position(('outward',10)) # outward by 10 points
+            spine.set_smart_bounds(True)
+        else:
+            spine.set_color('none') # don't draw spine
+
+    # turn off ticks where there is no spine
+    if 'left' in spines:
+        ax.yaxis.set_ticks_position('left')
+    else:
+        # no yaxis ticks
+        ax.yaxis.set_ticks([])
+
+    if 'bottom' in spines:
+        ax.xaxis.set_ticks_position('bottom')
+    else:
+        # no xaxis ticks
+        ax.xaxis.set_ticks([])
+
+def reyn(x,dx,xp,yp):
+	w=[]
+	for i in range(len(xp)):
+		if xp[i] > x-dx and xp[i] < x+dx:
+			w.append(yp[i])
+	w=array(w)
+	if len(w) == 1:
+		return 0.
+	return average(w)
+
+class parcel:
+	def __init__(self,d,u,v,uu,vv,x,y,T,m):
+		self.d=d
+		self.u=u
+		self.v=v
+		self.uu=uu
+		self.vv=vv
+		self.x=x
+		self.y=y
+		self.T=T
+		self.m=m
+	
+	def reynolds(self,y,dy):
+		self.yc=array(y)
+		self.uc=zeros(len(y))
+		self.uuc=zeros(len(y))
+		self.vc=zeros(len(y))
+		self.vvc=zeros(len(y))
+		
+		for j in range(len(y)):
+			ut=[]
+			uut=[]
+			vt=[]
+			vvt=[]
+			mt=[]
+			for i in range(len(self.y)):
+				if self.y[i] > y[j]-dy and self.y[i] < y[j]+dy:
+					#mt.append(self.m[i]/(pi/6.*self.d[i]**3))
+					mt.append(self.m[i])
+					ut.append(self.u[i])
+					uut.append(self.uu[i])
+					vt.append(self.v[i])
+					vvt.append(self.vv[i])
+			ut=array(ut)
+			uut=array(uut)
[…]

### 2016-09-11 master/cases/sydney/misc/massflow/mvapor.py

@@ -1,0 +1,142 @@
+#!/usr/bin/env python
+
+from pylab import *
+
+def adjust_spines(ax,spines):
+    for loc, spine in ax.spines.iteritems():
+        if loc in spines:
+            spine.set_position(('outward',10)) # outward by 10 points
+            spine.set_smart_bounds(True)
+        else:
+            spine.set_color('none') # don't draw spine
+
+    # turn off ticks where there is no spine
+    if 'left' in spines:
+        ax.yaxis.set_ticks_position('left')
+    else:
+        # no yaxis ticks
+        ax.yaxis.set_ticks([])
+
+    if 'bottom' in spines:
+        ax.xaxis.set_ticks_position('bottom')
+    else:
+        # no xaxis ticks
+        ax.xaxis.set_ticks([])
+
+
+fig_width = 6.  # Get this from LaTeX using \showthe\columnwidth
+golden_mean = (sqrt(5)-1.0)/2.0         # Aesthetic ratio
+fig_height = fig_width*golden_mean      # height in inches
+fig_size =  [fig_width,fig_height]
+params = {'backend': 'ps',
+          'axes.labelsize': 14,
+          'text.fontsize': 14,
+          'legend.fontsize': 13,
+          'xtick.labelsize': 18,
+          'ytick.labelsize': 18,
+          'text.usetex': False,
+          'font.family': 'monospace',
+          'figure.figsize': fig_size}
+rcParams.update(params)
+
+# 0 "origId"
+# 1 "origProcId"
+# 2 "m"
+# 3 "d"
+# 4 "yDot"
+# 5 "injector"
+# 6 "ct"
+# 7 "ms"
+# 8 "aC3H6O"
+# 9 "T"
+# 10 "tTurb"
+# 11 "liquidCore"
+# 12 "y"
+# 13 "U:0"
+# 14 "U:1"
+# 15 "U:2"
+# 16 "n:0"
+# 17 "n:1"
+# 18 "n:2"
+# 19 "Uturb:0"
+# 20 "Uturb:1"
+# 21 "Uturb:2"
+# 22 "Points:0"
+# 23 "Points:1"
+# 24 "Points:2"
+
[…]

### 2016-09-11 master/cases/sydney/misc/mesh3.py

@@ -1,0 +1,146 @@
+#!/usr/bin/env python
+from pylab import *
+import os
+
+f=open('blockMeshDict','w')
+f.write('/*---------------------------------------------------------------------------*\ \n')
+f.write('| =========                 |                                                 | \n')
+f.write('| \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox           | \n')
+f.write('|  \\    /   O peration     | Version:  1.4                                   | \n')
+f.write('|   \\  /    A nd           | Web:      http://www.openfoam.org               | \n')
+f.write('|    \\/     M anipulation  |                                                 | \n')
+f.write('\*---------------------------------------------------------------------------*/ \n')
+f.write('\n')
+f.write('FoamFile \n')
+f.write('{')
+f.write('    version         2.0; \n')
+f.write('    format          ascii; \n')
+f.write('\n')
+f.write('    root            "";\n')
+f.write('    case            "";\n')
+f.write('    instance        "";\n')
+f.write('    local           "";\n')
+f.write('\n')
+f.write('    class           dictionary; \n')
+f.write('    object          blockMeshDict;\n')
+f.write('}\n')
+f.write('\n')
+f.write('// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * // \n')
+
+f.write('\n')
+f.write('convertToMeters 1;\n')
+f.write('\n')
+f.write('vertices (')
+
+wedge=pi/180.*5.
+x0=0.5
+
+# fine mesh
+nx=200
+ny=50
+nyNozzle=24
+nyWall=15
+xGrad=10
+yGrad=15
+yGradWall=0.1
+
+r1=0.0049
+r2=0.065
+r3=0.075
+
+vertices=[]
+#central point
+vertices.append([0,0,0])#0
+vertices.append([0,r1*cos(wedge/2.),r1*sin(-wedge/2.)])#1
[…]

### 2016-09-11 master/cases/sydney/misc/pdf2.py

@@ -1,0 +1,131 @@
+
+#!/usr/bin/env python
+
+from pylab import *
+import scipy.special as sp
+import numpy as np
+import scipy.integrate as spi
+import scipy.optimize as opt
+
+fig_width = 8.  # Get this from LaTeX using \showthe\columnwidth
+golden_mean = (sqrt(5)-1.0)/2.0         # Aesthetic ratio
+fig_height = fig_width*golden_mean      # height in inches
+fig_size =  [fig_width,fig_height]
+params = {'backend': 'ps',
+          'axes.labelsize': 14,
+          'text.fontsize': 14,
+          'legend.fontsize': 13,
+          'xtick.labelsize': 18,
+          'ytick.labelsize': 18,
+          'text.usetex': False,
+          'font.family': 'monospace',
+          'figure.figsize': fig_size}
+rcParams.update(params)
+
+def erfInv(y):
+	a=0.147
+	k=2./pi/a+0.5*(log(1.0-y**2))
+	h=log(1.0 - y*y)/a
+	x=sqrt(-k + sqrt(k*k - h))
+	
+	if (y < 0.0):
+		x = -1.0*x
+	return x
+	
+
+def sample(mu,sd):
+	minValue=1.e-6
+	maxValue=60.e-6
+	
+	a = sp.erf((minValue - mu)/sd)
+	b = sp.erf((maxValue - mu)/sd)
+	y=random()
+	x = erfInv(y)*sd + mu
+	x = min(max(exp(x), exp(minValue)), exp(maxValue))
+	return log(x);
+	
+def estimatelogn(a,draw):
+	de=loadtxt('exp/bc_x_0p5/pdf.csv',delimiter=',')[:,0]*1e-6
+	pe=loadtxt('exp/bc_x_0p5/pdf.csv',delimiter=',')[:,1]*1e6
+	
+	E=a[0]
+	V=a[1]**2
+	
+	mu=log(E)-0.5*log(1+V/E**2)
+	sd=sqrt(log(1.+V/E**2))
+	
+	dn=linspace(0,max(de),200)
+
+	pn=1./(dn*sd*sqrt(2*pi))*exp(-(log(dn)-mu)**2/(2.*sd**2))
[…]

### 2016-09-11 master/cases/sydney/misc/smd/SMD.py

@@ -1,0 +1,236 @@
+#!/usr/bin/env python
+
+from pylab import *
+
+def adjust_spines(ax,spines):
+    for loc, spine in ax.spines.iteritems():
+        if loc in spines:
+            spine.set_position(('outward',10)) # outward by 10 points
+            spine.set_smart_bounds(True)
+        else:
+            spine.set_color('none') # don't draw spine
+
+    # turn off ticks where there is no spine
+    if 'left' in spines:
+        ax.yaxis.set_ticks_position('left')
+    else:
+        # no yaxis ticks
+        ax.yaxis.set_ticks([])
+
+    if 'bottom' in spines:
+        ax.xaxis.set_ticks_position('bottom')
+    else:
+        # no xaxis ticks
+        ax.xaxis.set_ticks([])
+
+def reyn(x,dx,xp,yp):
+	w=[]
+	for i in range(len(xp)):
+		if xp[i] > x-dx and xp[i] < x+dx:
+			w.append(yp[i])
+	w=array(w)
+	if len(w) == 1:
+		return 0.
+	return average(w)
+
+class parcel:
+	def __init__(self,d,u,v,uu,vv,x,y,T,m):
+		self.d=d
+		self.u=u
+		self.v=v
+		self.uu=uu
+		self.vv=vv
+		self.x=x
+		self.y=y
+		self.T=T
+		self.m=m
+	
+	def smd(self,y,dy):
+		self.yc=array(y)
+		self.smd=zeros(len(y))
+		for j in range(len(y)):
+			num=[]
+			den=[]
+			for i in range(len(self.y)):
+				if self.y[i] > y[j]-dy and self.y[i] < y[j]+dy:
+					den.append(self.d[i]**2)
+					num.append(self.d[i]**3)
+			num=array(num)
+			den=array(den)
+			self.smd[j]=average(num/den)
+		
+	
+	def reynolds(self,y,dy):
+		self.yc=array(y)
+		self.uc=zeros(len(y))
+		self.uuc=zeros(len(y))
+		self.vc=zeros(len(y))
+		self.vvc=zeros(len(y))
+		
+		for j in range(len(y)):
+			ut=[]
+			uut=[]
[…]

### 2016-09-11 master/cases/sydney/misc/stat.py

@@ -1,0 +1,195 @@
+#!/usr/bin/env python
+
+from pylab import *
+
+def adjust_spines(ax,spines):
+    for loc, spine in ax.spines.iteritems():
+        if loc in spines:
+            spine.set_position(('outward',10)) # outward by 10 points
+            spine.set_smart_bounds(True)
+        else:
+            spine.set_color('none') # don't draw spine
+
+    # turn off ticks where there is no spine
+    if 'left' in spines:
+        ax.yaxis.set_ticks_position('left')
+    else:
+        # no yaxis ticks
+        ax.yaxis.set_ticks([])
+
+    if 'bottom' in spines:
+        ax.xaxis.set_ticks_position('bottom')
+    else:
+        # no xaxis ticks
+        ax.xaxis.set_ticks([])
+
+def reyn(x,dx,xp,yp):
+	w=[]
+	for i in range(len(xp)):
+		if xp[i] > x-dx and xp[i] < x+dx:
+			w.append(yp[i])
+	w=array(w)
+	if len(w) == 1:
+		return 0.
+	return average(w)
+
+class parcel:
+	def __init__(self,d,u,v,uu,vv,x,y,T,m):
+		self.d=d
+		self.u=u
+		self.v=v
+		self.uu=uu
+		self.vv=vv
+		self.x=x
+		self.y=y
+		self.T=T
+		self.m=m
+	
+	def reynolds(self,y,dy):
+		self.yc=array(y)
+		self.uc=zeros(len(y))
+		self.uuc=zeros(len(y))
+		self.vc=zeros(len(y))
+		self.vvc=zeros(len(y))
+		
+		for j in range(len(y)):
+			ut=[]
+			uut=[]
+			vt=[]
+			vvt=[]
+			mt=[]
+			for i in range(len(self.y)):
+				if self.y[i] > y[j]-dy and self.y[i] < y[j]+dy:
+					#mt.append(self.m[i]/(pi/6.*self.d[i]**3))
+					mt.append(self.m[i])
+					ut.append(self.u[i])
+					uut.append(self.uu[i])
+					vt.append(self.v[i])
+					vvt.append(self.vv[i])
+			ut=array(ut)
+			uut=array(uut)
+			vt=array(vt)
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/injector/injector/injectorI.H

@@ -1,0 +1,54 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * * Member Functions  * * * * * * * * * * * * * //
+        
+//- Return clone
+inline autoPtr<injector> injector::clone() const
+{
+    return autoPtr<injector>(new injector(*this));
+}
+
+inline autoPtr<injectorType>& injector::properties()
+{
+    return properties_;
+}
+
+inline const autoPtr<injectorType>& injector::properties() const
+{
+    return properties_;
+}
+
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+} // End namespace Foam
+
+// ************************************************************************* //

### 2016-09-11 master/code/1.7.x/lowmachLib/spray/spray.H

@@ -1,0 +1,367 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::spray
+
+Description
+    A spray is a cloud of parcels
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef spray_H
+#define spray_H
+
+#include "parcel.H"
+#include "injector.H"
+#include "IOPtrList.H"
+#include "interpolation.H"
+#include "liquid.H"
+#include "autoPtr.H"
+#include "liquidMixture.H"
+#include "Random.H"
+#include "thermoPhysicsTypes.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+class atomizationModel;
+class breakupModel;
+class collisionModel;
+class dispersionModel;
+class dragModel;
+class evaporationModel;
+class injectorModel;
+class heatTransferModel;
+class wallModel;
+
+class basicMultiComponentMixture;
+
+/*---------------------------------------------------------------------------*\
+                        Class spray Declaration
+\*---------------------------------------------------------------------------*/
+
+class spray
+:
+    public Cloud<parcel>
+{
+    // Private data
+
+        // References to the database and meshes
+
+            const Time& runTime_;
+            scalar time0_;
+            const fvMesh& mesh_;
+
+            //- Random number generator
+            Random rndGen_;
+
+        //- Acceleration due to gravity
+        const vector& g_;
+
+
+        // References to the physical fields
+
+            const volVectorField& U_;
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spray/sprayInject.C

@@ -1,0 +1,192 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "spray.H"
+#include "breakupModel.H"
+#include "collisionModel.H"
+#include "dispersionModel.H"
+#include "injectorModel.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+void spray::inject()
+{
+    scalar time = runTime_.value();
+    scalar timecheck = time0_;
+    
+    if (time - timecheck > dtinj_) {
+
+        scalar time0 = time0_;
+
+        // Inject the parcels for each injector sequentially
+        forAll(injectors_, i)
+        {
+            autoPtr<injectorType>& it = injectors()[i].properties();
+            if (!it->pressureIndependentVelocity())
+            {
+                scalar referencePressure = p().average().value();
+                it->correctProfiles(fuels(), referencePressure);
+            }
+
+            const label nHoles = it->nHoles();
+
+            // parcels have the same mass during a timestep
+            scalar mass = it->mass(time0, time, twoD_, angleOfWedge_);
+
+            label Np = it->nParcelsToInject(time0, time);
+
+            if (mass > 0)
+            {
+                Np = max(1, Np);
+                scalar mp = mass/Np/nHoles;
+
+                // constT is only larger than zero for the first
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/atomizationModel/blobsSheetAtomization/blobsSheetAtomization.H

@@ -1,0 +1,115 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::blobsSheetAtomization
+
+Description
+    Primary Breakup Model for pressure swirl atomizers.
+
+    Accurate description in
+    @verbatim
+    Z. Han, S. Parrish, P.V. Farrell, R.D. Reitz
+    "Modeling Atomization Processes Of Pressure Swirl Hollow-Cone Fuel Sprays"
+    Atomization and Sprays, vol. 7, pp. 663-684, 1997
+
+    and
+
+    L. Allocca, G. Bella, A. De Vita, L. Di Angelo
+    "Experimental Validation of a GDI Spray Model"
+    SAE Technical Paper Series, 2002-01-1137
+    @endverbatim
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef blobsSheetAtomization_H
+#define blobsSheetAtomization_H
+
+#include "atomizationModel.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+/*---------------------------------------------------------------------------*\
+                           Class blobsSheetAtomization Declaration
+\*---------------------------------------------------------------------------*/
+
+class blobsSheetAtomization
+:
+    public atomizationModel
+{
+private:
+
+    // Private data
+
+        dictionary coeffsDict_;
+        scalar B_;
+        scalar angle_;
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/breakupModel/breakupModel/breakupModel.C

@@ -1,0 +1,150 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "error.H"
+#include "breakupModel.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+defineTypeNameAndDebug(breakupModel, 0);
+
+defineRunTimeSelectionTable(breakupModel, dictionary);
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+// Construct from components
+breakupModel::breakupModel
+(
+    const dictionary& dict,
+    spray& sm
+)
+:
+    dict_(dict),
+    spray_(sm),
+    rndGen_(sm.rndGen()),
+    includeOscillation_(dict_.lookup("includeOscillation")),
+    TABcoeffsDict_(dict.subDict("TABCoeffs")),
+    y0_(0.0),
+    yDot0_(0.0),
+    TABComega_(0.0),
+    TABCmu_(0.0),
+    TABWeCrit_(0.0)
+{
+    if (includeOscillation_)
+    {
+        y0_ = readScalar(TABcoeffsDict_.lookup("y0"));
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/breakupModel/breakupModel/newBreakupModel.C

@@ -1,0 +1,76 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "error.H"
+#include "breakupModel.H"
+#include "noBreakup.H"
+#include "reitzDiwakar.H"
+#include "reitzKHRT.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+autoPtr<breakupModel> breakupModel::New
+(
+    const dictionary& dict,
+    spray& sm
+)
+{
+    word breakupModelType
+    (
+        dict.lookup("breakupModel")
+    );
+
+    Info<< "Selecting breakupModel "
+         << breakupModelType << endl;
+
+    dictionaryConstructorTable::iterator cstrIter =
+        dictionaryConstructorTablePtr_->find(breakupModelType);
+
+    if (cstrIter == dictionaryConstructorTablePtr_->end())
+    {
+        FatalError
+            << "breakupModel::New(const dictionary&, const spray&) : " << nl
+            << "    unknown breakupModelType type "
+            << breakupModelType
+            << ", constructor not in hash table" << endl << nl
+            << "    Valid breakupModel types are :" << nl
+            << dictionaryConstructorTablePtr_->sortedToc()
+            << abort(FatalError);
+    }
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/breakupModel/noBreakup/noBreakup.C

@@ -1,0 +1,83 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "error.H"
+
+#include "noBreakup.H"
+#include "addToRunTimeSelectionTable.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+defineTypeNameAndDebug(noBreakup, 0);
+
+addToRunTimeSelectionTable
+(
+    breakupModel,
+    noBreakup,
+    dictionary
+);
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+// Construct from components
+noBreakup::noBreakup
+(
+    const dictionary& dict,
+    spray& sm
+)
+:
+    breakupModel(dict, sm)
+{}
+
+
+// * * * * * * * * * * * * * * * * Destructor  * * * * * * * * * * * * * * * //
+
+noBreakup::~noBreakup()
+{}
+
+
+// * * * * * * * * * * * * * * * Member Functions  * * * * * * * * * * * * * //
+
+void noBreakup::breakupParcel
+(
+    parcel&,
+    const scalar,
+    const vector&,
+    const liquidMixture&
+) const
+{
+    // Do nothing
+}
+
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/breakupModel/noBreakup/noBreakup.H

@@ -1,0 +1,98 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::noBreakup
+
+Description
+    Spray breakup "off"
+
+SourceFiles
+    noBreakupI.H
+    noBreakup.C
+    noBreakupIO.C
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef noBreakup_H
+#define noBreakup_H
+
+#include "breakupModel.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+/*---------------------------------------------------------------------------*\
+                           Class noBreakup Declaration
+\*---------------------------------------------------------------------------*/
+
+class noBreakup
+:
+    public breakupModel
+{
+
+public:
+
+    //- Runtime type information
+    TypeName("off");
+
+
+    // Constructors
+
+        //- Construct from components
+        noBreakup
+        (
+            const dictionary& dict,
+            spray& sm
+        );
+
+
+    // Destructor
+
+        ~noBreakup();
+
+
+    // Member Functions
+
+        void breakupParcel
+        (
+            parcel& parcel,
+            const scalar deltaT,
+            const vector& vel,
+            const liquidMixture& fuels
+        ) const;
+
+};
+
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+} // End namespace Foam
+
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/breakupModel/reitzKHRT/reitzKHRT.C

@@ -1,0 +1,239 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "reitzKHRT.H"
+#include "addToRunTimeSelectionTable.H"
+#include "mathematicalConstants.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+defineTypeNameAndDebug(reitzKHRT, 0);
+
+addToRunTimeSelectionTable
+(
+    breakupModel,
+    reitzKHRT,
+    dictionary
+);
+
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+// Construct from components
+reitzKHRT::reitzKHRT
+(
+    const dictionary& dict,
+    spray& sm
+)
+:
+    breakupModel(dict, sm),
+    coeffsDict_(dict.subDict(typeName + "Coeffs")),
+    g_(sm.g()),
+    b0_(readScalar(coeffsDict_.lookup("B0"))),
+    b1_(readScalar(coeffsDict_.lookup("B1"))),
+    cTau_(readScalar(coeffsDict_.lookup("Ctau"))),
+    cRT_(readScalar(coeffsDict_.lookup("CRT"))),
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/dispersionModel/dispersionLESModel/dispersionLESModel.H

@@ -1,0 +1,96 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::dispersionLESModel
+
+Description
+    Foam::dispersionLESModel
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef dispersionLESModel_H
+#define dispersionLESModel_H
+
+#include "dispersionModel.H"
+#include "LESModel.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+/*---------------------------------------------------------------------------*\
+                       Class dispersionLESModel Declaration
+\*---------------------------------------------------------------------------*/
+
+class dispersionLESModel
+:
+    public dispersionModel
+{
+    // Private data
+
+        const compressible::LESModel& turbulence_;
+
+
+public:
+
+    //- Runtime type information
+    TypeName("dispersionLESModel");
+
+
+    // Constructors
+
+        //- Construct from components
+        dispersionLESModel
+        (
+            const dictionary& dict,
+            spray& sm
+        );
+
+
+    // Destructor
+
+        virtual ~dispersionLESModel();
+
+
+    // Member Functions
+
+        const compressible::LESModel& turbulence() const
+        {
+            return turbulence_;
+        }
+
+        virtual void disperseParcels() const = 0;
+};
+
+
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/dragModel/noDragModel/noDragModel.C

@@ -1,0 +1,93 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "error.H"
+
+#include "noDragModel.H"
+#include "addToRunTimeSelectionTable.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+defineTypeNameAndDebug(noDragModel, 0);
+
+addToRunTimeSelectionTable
+(
+    dragModel,
+    noDragModel,
+    dictionary
+);
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+// Construct from components
+noDragModel::noDragModel
+(
+    const dictionary& dict
+)
+:
+    dragModel(dict)
+{}
+
+// * * * * * * * * * * * * * * * * Destructor  * * * * * * * * * * * * * * * //
+
+noDragModel::~noDragModel()
+{}
+
+
+// * * * * * * * * * * * * * * * Member Functions  * * * * * * * * * * * * * //
+
+scalar noDragModel::Cd
+(
+    const scalar,
+    const scalar
+) const
+{
+    return 0.0;
+}
+
+
+scalar noDragModel::relaxationTime
+(
+    const vector&,
+    const scalar,
+    const scalar,
+    const scalar,
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/evaporationModel/evaporationModel/evaporationModel.C

@@ -1,0 +1,58 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "error.H"
+
+#include "evaporationModel.H"
+#include "fvCFD.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+defineTypeNameAndDebug(evaporationModel, 0);
+
+defineRunTimeSelectionTable(evaporationModel, dictionary);
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+// Construct from dictionary
+evaporationModel::evaporationModel
+(
+    const dictionary& dict
+)
+:
+    dict_(dict)
+{}
+
+// * * * * * * * * * * * * * * * * Destructor  * * * * * * * * * * * * * * * //
+
+evaporationModel::~evaporationModel()
+{}
+
+} // End namespace Foam
+
+// ************************************************************************* //

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/injectorModel/blobsSwirl/blobsSwirlInjector.H

@@ -1,0 +1,168 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::blobsSwirlInjector
+
+Description
+    Model for parcels injection in pressure Swirl Atomizers.
+
+    Accurate description in:
+    @verbatim
+        Z. Han, S. Parrish, P.V. Farrell, R.D. Reitz
+        "Modeling Atomization Processes Of Pressure Swirl Hollow-Cone Fuel Sprays"
+        Atomization and Sprays, vol. 7, pp. 663-684, 1997
+    @endverbatim
+    and
+    @verbatim
+        L. Allocca, G. Bella, A. De Vita, L. Di Angelo
+        "Experimental Validation of a GDI Spray Model"
+        SAE Technical Paper Series, 2002-01-1137
+    @endverbatim
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef blobsSwirlInjector_H
+#define blobsSwirlInjector_H
+
+#include "injectorModel.H"
+#include "scalarList.H"
+
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+/*---------------------------------------------------------------------------*\
+                           Class blobsSwirlInjector Declaration
+\*---------------------------------------------------------------------------*/
+
+class blobsSwirlInjector
+:
+    public injectorModel
+{
+
+private:
+
+    // Private data
+
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/injectorModel/definedPressureSwirl/definedPressureSwirl.C

@@ -1,0 +1,328 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "definedPressureSwirl.H"
+#include "addToRunTimeSelectionTable.H"
+#include "mathematicalConstants.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+defineTypeNameAndDebug(definedPressureSwirlInjector, 0);
+
+addToRunTimeSelectionTable
+(
+    injectorModel,
+    definedPressureSwirlInjector,
+    dictionary
+);
+
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+// Construct from components
+definedPressureSwirlInjector::definedPressureSwirlInjector
+(
+    const dictionary& dict,
+    spray& sm
+)
+:
+    injectorModel(dict, sm),
+    definedPressureSwirlInjectorDict_(dict.subDict(typeName + "Coeffs")),
+
+    coneAngle_(definedPressureSwirlInjectorDict_.lookup("ConeAngle")),
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/injectorModel/hollowCone/hollowCone.C

@@ -1,0 +1,196 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "hollowCone.H"
+#include "addToRunTimeSelectionTable.H"
+#include "mathematicalConstants.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+defineTypeNameAndDebug(hollowConeInjector, 0);
+
+addToRunTimeSelectionTable
+(
+    injectorModel,
+    hollowConeInjector,
+    dictionary
+);
+
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+// Construct from components
+hollowConeInjector::hollowConeInjector
+(
+    const dictionary& dict,
+    spray& sm
+)
+:
+    injectorModel(dict, sm),
+    hollowConeDict_(dict.subDict(typeName + "Coeffs")),
+    dropletPDF_
+    (
+        pdfs::pdf::New
+        (
+            hollowConeDict_.subDict("dropletPDF"),
+            sm.rndGen()
+        )
+    ),
+    innerAngle_(hollowConeDict_.lookup("innerConeAngle")),
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/injectorModel/injectorModel/newInjectorModel.C

@@ -1,0 +1,75 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "error.H"
+
+#include "injectorModel.H"
+#include "volFields.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+autoPtr<injectorModel> injectorModel::New
+(
+    const dictionary& dict,
+    spray& sm
+)
+{
+    word injectorModelType
+    (
+        dict.lookup("injectorModel")
+    );
+
+    Info<< "Selecting injectorModel "
+         << injectorModelType << endl;
+
+    dictionaryConstructorTable::iterator cstrIter =
+        dictionaryConstructorTablePtr_->find(injectorModelType);
+
+    if (cstrIter == dictionaryConstructorTablePtr_->end())
+    {
+        FatalError
+            << "injectorModel::New(const dictionary&, spray&) : " << nl
+            << "    unknown injectorModelType type "
+            << injectorModelType
+            << ", constructor not in hash table" << nl << nl
+            << "    Valid injectorModel types are :" << nl
+            << dictionaryConstructorTablePtr_->sortedToc()
+            << abort(FatalError);
+    }
+
+    return autoPtr<injectorModel>(cstrIter()(dict, sm));
+}
+
+
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/wallModel/removeParcel/removeParcel.H

@@ -1,0 +1,94 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::removeParcel
+
+Description
+    remove parcel, regardless of boundary type.
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef removeParcel_H
+#define removeParcel_H
+
+#include "wallModel.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+/*---------------------------------------------------------------------------*\
+                           Class removeParcel Declaration
+\*---------------------------------------------------------------------------*/
+
+class removeParcel
+:
+    public wallModel
+{
+
+public:
+
+    //- Runtime type information
+    TypeName("remove");
+
+
+    // Constructors
+
+        //- Construct from components
+        removeParcel
+        (
+            const dictionary& dict,
+            const volVectorField& U,
+            spray& sm
+        );
+
+
+    // Destructor
+
+        virtual ~removeParcel();
+
+
+    // Member Functions
+
+        //- Return true if parcel is to be kept, and false if it is to be
+        // removed
+        bool wallTreatment
+        (
+            parcel& parcel,
+            const label facei
+        ) const;
+
+};
+
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+} // End namespace Foam
+
[…]

### 2016-09-11 master/code/1.7.x/lowmachLib/spraySubModels/wallModel/wallModel/wallModel.C

@@ -1,0 +1,68 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "error.H"
+#include "wallModel.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+defineTypeNameAndDebug(wallModel, 0);
+
+defineRunTimeSelectionTable(wallModel, dictionary);
+
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+
+// Construct from components
+
+wallModel::wallModel
+(
+    const dictionary& dict,
+    const volVectorField& U,
+    spray& sm
+)
+:
+    dict_(dict),
+    mesh_(U.mesh()),
+    spray_(sm)
+{}
+
+
+// * * * * * * * * * * * * * * * * Destructor  * * * * * * * * * * * * * * * //
+
+wallModel::~wallModel()
+{}
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+} // End namespace Foam
+
+// ************************************************************************* //

### 2016-09-11 master/code/1.7.x/lowmachSolver/pEqn.H

@@ -1,0 +1,39 @@
+//updates rho after solving for enthalpy
+rho = thermo.rho();
+
+volScalarField rUA = 1.0/UEqn.A();
+
+//Computes U without the pressure contribution
+U = rUA*UEqn.H();
+//Updates phi
+    phi =
+        fvc::interpolate(rho)
+       *(
+            (fvc::interpolate(U) & mesh.Sf())
+          + fvc::ddtPhiCorr(rUA, rho, U, phi)
+        );
+
+for (int nonOrth=0; nonOrth<=nNonOrthCorr; nonOrth++)
+{
+	fvScalarMatrix pdEqn
+	(
+		fvc::ddt(psi,p)
+      + fvc::div(phi)
+	  - fvm::laplacian(rho*rUA, pd)
+	 ==
+		Sevap
+	);
+
+	pdEqn.solve();
+
+	if (nonOrth == nNonOrthCorr)
+	{
+		phi += pdEqn.flux();
+	}
+}
+
+#include "rhoEqn.H"
+#include "compressibleContinuityErrs.H"
+
+U -= rUA*fvc::grad(pd);
+U.correctBoundaryConditions();

### 2016-09-11 master/code/1.7.x/mypdfs/RosinRammler/RosinRammler.C

@@ -1,0 +1,84 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "RosinRammler.H"
+#include "addToRunTimeSelectionTable.H"
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+namespace Foam
+{
+    namespace pdfs
+    {
+        defineTypeNameAndDebug(RosinRammler, 0);
+        addToRunTimeSelectionTable(pdf, RosinRammler, dictionary);
+    }
+}
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+Foam::pdfs::RosinRammler::RosinRammler(const dictionary& dict, Random& rndGen)
+:
+    pdf(typeName, dict, rndGen),
+    minValue_(readScalar(pdfDict_.lookup("minValue"))),
+    maxValue_(readScalar(pdfDict_.lookup("maxValue"))),
+    d_(readScalar(pdfDict_.lookup("d"))),
+    n_(readScalar(pdfDict_.lookup("n")))
+{
+    check();
+}
+
+
+// * * * * * * * * * * * * * * * * Destructor  * * * * * * * * * * * * * * * //
+
+Foam::pdfs::RosinRammler::~RosinRammler()
+{}
+
+
[…]

### 2016-09-11 master/tutorial/html/images.pl

@@ -1,0 +1,24 @@
+# LaTeX2HTML 2008 (1.71)
+# Associate images original text with physical files.
+
+
+$key = q/Psi;MSF=1.6;LFS=11;AAT/;
+$cached_env_img{$key} = q|<IMG
+ WIDTH="18" HEIGHT="15" ALIGN="BOTTOM" BORDER="0"
+ SRC="|."$dir".q|img1.png"
+ ALT="$ \Psi$">|; 
+
+$key = q/h_s;MSF=1.6;LFS=11;AAT/;
+$cached_env_img{$key} = q|<IMG
+ WIDTH="22" HEIGHT="34" ALIGN="MIDDLE" BORDER="0"
+ SRC="|."$dir".q|img3.png"
+ ALT="$ h_s$">|; 
+
+$key = q/displaystylerho=pPsi=frac{p}{RWT},.;MSF=1.6;LFS=11;AAT/;
+$cached_env_img{$key} = q|<IMG
+ WIDTH="139" HEIGHT="47" ALIGN="MIDDLE" BORDER="0"
+ SRC="|."$dir".q|img2.png"
+ ALT="$\displaystyle \rho=p\Psi=\frac{p}{RWT}   .$">|; 
+
+1;
+

### 2016-09-24 master/paper/paper.tex

%% This is file `elsarticle-template-1a-num.tex',
%%
%% Copyright 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%%
%% $Id: elsarticle-template-1a-num.tex 151 2009-10-08 05:18:25Z rishi $
%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1a-num.tex $
%%
\documentclass[preprint,12pt,review]{elsarticle}
\usepackage{amssymb,amsmath}
\usepackage{subfig}
\usepackage[nomarkers,nofiglist]{endfloat}
\usepackage[nodots,nocompress]{numcompress}
%Definitions
\newcommand{\bv}[1]{\mathbf{#1}} 
\newcommand{\req}[1]{#1^{*}}

%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
\usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
%\usepackage{lineno}

%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%
%% \biboptions{comma,round}

% \biboptions{}

%\journal{Journal Title}

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{Eulerian-Lagrangian Simulation of a Turbulent Evaporating Spray using RANS Modeling}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author[]{\textbf{Rodrigo B. Piccinini}\\ \textbf{Marcelo J.S. de Lemos}$^{*}$}
\address{Departamento de Energia - IEME \\
Instituto Tecnol\'{o}gico de Aeron\'{a}utica - ITA\\
12228-900 - S\~{a}o Jos\'{e} dos Campos - SP, Brazil\\
$^{*}$\textit{\textbf{Corresponding author} E-mail: \underline{delemos@ita.br}}}

\begin{abstract}
RANS simulation was applied to a turbulent and evaporating spray of acetone in an air co-flowing stream. An Eulerian-Lagrangian description was used with a liquid phase of point-particles dispersed in a gaseous continuous phase. Governing equations of the continuous phase were simplified with a low Mach number approximation. Turbulence was modeled using k-epsilon model. Submodels for heat and mass inter-phase transfers and Nusselt and Sherwood numbers were given by empirical correlations for convective droplet heating and evaporation. Spray interaction with turbulence was modeled using a random walk procedure. Numerical results were compared to measurements and a reasonable prediction was found for droplet and gas velocities and liquid evaporation.
\end{abstract}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
spray \sep round jet \sep low Mach number \sep lagrangian \sep point-particle method
\end{keyword}

\end{frontmatter}

\begin{tabular}{ll}
\textbf{Nomenclature:} & \\
$C_D$ & doplet drag coefficient \\
$c_p$ & specific heat at constant pressure [$J/kg/K$] \\
$D$ & droplet diameter [$m$]\\
$\tilde{f}$ & favre-average of function $f$\\
$f''$ & deviation of favre-averaged function $\tilde{f}$\\
$\bar{f}$ & time-average of function $f$\\
$f'$ & deviation of time-averaged function $\bar{f}$\\
$\bv{g}$ & vector field of gravity force [$m/s^2$]\\
$\dot{m}_d$ & mass flow rate of liquid injection [$kg/s$]\\
$\dot{m}_{g}$ & gas mass rate [$kg/s$]\\
$\dot{m}_{vapor}$ & acetone vapor mass rate [$kg/s$]\\
$h_d$ & droplet enthalpy [$J/kg$]\\
$h_s$ & gas sensible enthalpy [$J/kg$]\\
$k$ & turbulent kinetic energy [$J/kg$]\\
$\hat{k}$ & jet dimensionless $k$: $\hat{k}=k/(\tilde{U}_{x,0}-\tilde{U}_{cl})^2$ \\
$L_v$ & latent heat of vaporization [$J/kg$]\\
$m_d$ & droplet mass [$kg$]\\
$M$ & Mach number \\
$p_{0}$ & thermodynamic pressure [$Pa$]\\
$p_{2}$ & dynamic pressure [$Pa$]\\
$p$ & pressure [$Pa$]\\
$Pr$ & Prandtl number \\
\end{tabular}

\vspace{25pt}

\begin{tabular}{ll}
$Re_d$ & droplet Reynolds number, $Re=(\rho |\bv{U}|_{slip} D) /\mu$ \\
$S_{hs}$ & spray enthalpy source term [$J/m^3/s$]\\
$S_{m}$ & spray mass source term [$kg/m^3/s$]\\
$S_{mom}$ & spray momentum source term [$kg/m^3/s$]\\
$S_{Y_k}$ & spray source term of species k [$kg/m^3/s$]\\
$Sc$ & Schmidt number \\
$Sh$ & Sherwood number \\
$T_d$ & droplet temperature [$K$]\\
$T$ & gas temperature [$K$]\\
$t$ & time [$s$]\\
$\bv{U}$ & gas velocity vector [$m/s$]\\
$\bv{U}_d$ & droplet velocity vector [$m/s$]\\
$\bv{U}_{slip}$ & droplet velocity relative to the gas: $\bv{U}_{slip}=\bv{U}_d -\bv{U}$  [$m/s$]\\
$\tilde{U}_{x,0}$ & jet centerline velocity \\
$\tilde{U}_{x,cf}$ & co-flow velocity \\
$\hat{U}$ & dimensionless jet velocity, $\hat{U} = (\tilde{U}_x-\tilde{U}_{x,cf})/(\tilde{U}_{x0}-\tilde{U}_{x,cf})$ \\
$y_{1/2}(x)$ & jet half-radius, the radial coordinate for which $\tilde{U}_x (y=y_{1/2}) = 1/2\tilde{U}_{x,0}$\\
$\hat{y}(x,y)$ & dimensionless radial coordinate: $\hat{y}(x,y)=y/y_{1/2}$ \\
$x$ & axial coordinate [$m$]\\
$\bv{x}$ & position vector [$m$] \\
$y$ & radial coordinate [$m$]\\
$Y_k$ & mass fraction of species k\\
$W_k$ & molecular weight of species k [$kg/mol$]\\
\end{tabular}

\vspace{25pt}

\begin{tabular}{ll}
\textbf{Greek Characters:} & \\
$\alpha$ & thermal diffusivity [$m^2/s$]\\
$\gamma$ & ratio of constant pressure and constant volume specific heats \\
$\epsilon$ & dissipation rate of turbulent kinetic energy [$J/kg/s$]\\
$\kappa$ & thermal conductivity [$W/m/K$]\\
$\mu$ & gas dynamic viscosity [$Pa.s$]\\
$\nu$ & gas kinematic viscosity [$m^2 /s$]\\
$\nu_T$ &  turbulent viscosity [$m^2 /s$] \\
$\hat{\nu}_T$ &  jet dimensionless turbulent viscosity: $\hat{\nu}_T = \mu/(\bar{\rho}\tilde{U}_{x,0} y_{1/2})$\\
$\rho$ & gas density [$kg/m^3$]\\
$\rho_d$ & droplet density [$kg/m^3$]\\
$\bv{\tau}$ & viscous stress tensor [$Pa$]\\
$\tau_e$ & droplet evaporation relaxation time [$s$]\\
$\tau_h$ & droplet heating relaxation time [$s$]\\
$\tau_u$ & droplet momentum relaxation time [$s$] \\
$\phi_{v,l}$ & liquid volume fraction
\end{tabular}

\vspace{25pt}

\begin{tabular}{ll}
\textbf{Special Characters:} & \\
cdf & cumulative density function\\
pdf & probability densitiy function\\
pmf & probability mass function\\
SMD & Sauter mean diameter \\
$\mathcal{D}_k$ & binary diffusion coefficient of species k [$m^2/s$]
\end{tabular}

%%
%% Start line numbering here if you want
%%
% \linenumbers

%% main text
\section{Introduction}
\label{intro}
A spray jet is a particular case of a dispersed multiphase flow produced by an unstable liquid jet emerging in a gaseous atmosphere. 
Depending on liquid volumetric fraction, a spray may be classified as dense ($\phi_l > 10^{-3}$) or dilute ($\phi_l \le 10^{-3}$). Dilute sprays are commonly modeled using the so called two-way coupling.  In this coupling approach, interactions among gas and liquid droplets are modeled, but interactions among droplets are neglected.

RANS modeling has been extensively used to spray jets simulations, see \citet{baumgarten2006mixture}, and, more recently, large-eddy (LES) and direct simulations (DNS) have also been applied. Independently on the turbulence modeling, droplet heating and evaporation are often similarly modeled with empirical correlations for Nusselt and Sherwood numbers, such as those of Ranz and Marshall's \citet{ranzmarshall}. This indicates that much of the improvement obtained in LES simulations in the prediction of droplet velocity and evaporation rate, e.g. \citet{bini} and \citet{moin},  was due to the best modeling of turbulent motion and the allowance of gas properties such as temperature ($T$) and species mass fractions ($Y_k$) to deviate from mean values. It certainly provides a more accurate description of droplet environment in a turbulent spray jet. Actually, \citet{bini} has also shown that subgrid modeling of Sherwood number in LES further improved predictions of droplet evaporation.

This scenario suggests the possibility of improving RANS simulations with stochastic submodeling of not only droplet dispersion, but of heat and mass transfers as well. Also, the proposition of RANS submodels would certainly benefit from the results coming from LES and DNS.

This paper presents results of applying RANS modeling to the experiment developed by \citet{chen}. It has found overestimation of jet spread rate using the usual coefficients of k-epsilon model and underprediction of droplet velocity and evaporation rate, specially after the jet core. It suggests that future modification of turbulence model coefficients and stochastic submodels for heat and mass transfers could improve predictions.

\section{Governing Equations}
\label{equations}

\subsection{Continuous Phase}\label{continuous_phase}
The equations for the continuous gaseous phase were derived from the fully compressible formulation, \citet{poinsot2005theoretical}, with additional spray source terms ($S_m$, $S_Y$, $S_{mom}$ and $S_{hs}$). It was applied the low Mach number approximation, see \citet{majda}, and performed the Reynolds and Favre averages for turbulence modeling.

The low Mach number approximation expands all variables in power series of $\xi=\sqrt{\gamma}M^2$ and only collects the equations of zeroth order in $\xi$. This procedure splits pressure in two parcels,
\begin{equation}
p = p_{0}(t) + p_{2}(\bv{x},t)\xi^2 + O\left( \xi^3 \right) \, .
\end{equation}
$p_0$ is called the thermodynamic pressure, it is shown to be function of time only and it is present in the state and energy equations; $p_1$ is not present in the equations for zeroth order terms and $p_2$ is the dynamic pressure, whose gradient is present in the momentum equation.

The continuity equation reads:
\begin{equation}\label{av: mass}
  \frac{\partial \bar{\rho}}{\partial t} + \nabla \cdot \left( \bar{\rho}
\tilde{\bv{U}} \right) =  \bar{S}_m \, .
\end{equation}

The acetone ($Y_{ac}$) conservation equation for binary diffusion and unitary Schmidt number reads:
\begin{equation}\label{av: species}
 \frac{\partial \bar{\rho} \tilde{Y}_{ac} }{\partial t} + \nabla
\cdot\left(\bar{\rho} \tilde{\bv{U}} \tilde{Y}_{ac} \right) = \nabla \cdot \left[ \bar{\rho} \left( \nu + \nu_T \right)
\nabla \tilde{Y}_{ac} \right] +\bar{S}_{Yac} \, ,
\end{equation}
air is treated as an inert species with
\begin{equation}
\tilde{Y}_{air} = 1- \tilde{Y}_{ac} \, .
\end{equation}

The momentum equation reads:
\begin{equation}\label{av: momentum}
 \frac{\partial \bar{\rho} \tilde{\bv{U}} }{\partial t} + \nabla \cdot \left(
\bar{\rho} \tilde{\bv{U}} \tilde{\bv{U}} \right) = -\nabla \bar{p}_2 +  
 \nabla \cdot\overline{\bv{\tau}} +\bar{\rho}\bv{g}+\bar{\bv{S}}_{mom}  \, .
\end{equation}
where $\bv{\tau}$ is the viscous stress tensor with effective viscosity $\mu_{eff}=\mu+\mu_T$.

The energy equation is written in terms of sensible enthalpy with unitary turbulent Prandtl number:
\begin{equation}\label{av: enthalpy}
\frac{\partial \bar{\rho} \tilde{h}_{s}}{\partial t} + \nabla \cdot
\left(\bar{\rho} \tilde{\bv{U}} \tilde{h}_{s} \right) = \nabla \cdot \bar{\bv{J}}_s + \bar{S}_{hs}\, ,
\end{equation} where 
\begin{equation}
\bar{\bv{J}}_s =  \bar{\rho} \left( \alpha + \nu_T\right) \nabla \tilde{h}_s \, .
\end{equation}
The pressure work ($Dp/Dt$ ) and species thermal diffusion were neglected.

State Equation is:
\begin{equation}
  \bar{p}_0 W = R \bar{\rho} \tilde{T} 
\end{equation}

The k-epsilon equations are those presented in \citet{nordin} with $\dot{W}^{s}= 0$, that is, no sink of turbulent kinetic energy was defined here due to the work done by the eddies on the droplets. The model has six parameters whose values are shown in Table \ref{table: kepscoeff}.

\section{Dispersed Phase}\label{dispersed_phase}

The dispersed phase is written in a Lagrangian reference frame. The droplet equations are:
\begin{align}\label{eq: drop_dudt}
\text{Momentum:}&\quad \frac{d\bv{U}_d}{dt} = -\frac{ \bv{U}_{slip}}{\tau_{u}} + \bv{g}  \\
\label{eq: dropmass}
\text{Mass:}&\quad \frac{dm_d}{dt}=-\frac{m_d}{\tau_e} \, , \\
\text{Energy:}&\quad \frac{dT_d}{dt}= \frac{T-T_d}{\tau_h} f - \frac{1}{c_{p,d}} \frac{L_v\left( T_d\right) }{\tau_e}
\end{align}

In the momentum equation, $\tau_u$ is the momentum relaxation time:
\begin{equation}
 \tau_u = \frac{8 m_d}{\pi \rho C_D D^2 | \bv{U}_{slip}|}=\frac{4}{3}
\frac{\rho_d D}{\rho C_D | \bv{U}_{slip}|} \, .
\end{equation}
The drag coefficient ($C_D$) is computed using the expression in \citet{nordin}.
The droplet slip velocity is:
\begin{equation}\label{eq: Urel}
 \bv{U}_{slip} = \bv{U_d} - \tilde{\bv{U}}-\bv{U}'' \, ,
\end{equation}
where $\tilde{\bv{U}}$ is the Favre-averaged gas velocity computed in Equation \eqref{av: momentum}. The fluctuation of gas velocity ($\bv{U}''$) is determined using the random walk procedure of \citet{rourke}. For the k-epsilon model, a characteristic velocity may be defined as:
\begin{equation}
 <U''>= \sqrt{\frac{2}{3} k} \, .
\end{equation}
The magnitude of the velocity fluctuation ($\bv{U}''$) is then sampled from a Gaussian distribution with a variance equal to $<U''>$ and zero mean. Finally, the direction of $\bv{U}''$  is randomly chosen.

The sampling is performed once the time passed from the last sampling is greater than the characteristic time $\tau_{turb}$: the minimum between the eddy lifetime and the time taken to the droplet to cross it, given by the ratio of a turbulent length scale ($l_t = C_{\mu}^{3/4} k^{3/2} / \epsilon $) and the magnitude of the droplet slip velocity:
\begin{equation}
 \tau_{turb} = min \left[ \frac{k}{\epsilon} , \frac{k^{3/2}}{\epsilon} \frac{C_{\mu}^{3/4}}{|\bv{U}_{slip}|}\right] \, .
\end{equation}

The characteristic time scales for heating ($\tau_h$) and evaporation ($\tau_e$) are:
\begin{equation}
 \tau_h=\frac{m_d c_{l,d}}{\pi D \kappa Nu} \, , \quad  \tau_e = \frac{m_d}{\pi D \mathcal{D} Sh \rho_v ln\left(\frac{1-Y_{v,\infty}}{1-Y_{v,s}} \right)} \, .
\end{equation}
The $f$ factor in the energy equation corrects for the heat not transfered to the droplet, but to the surrounding vapor:
\begin{equation}
f =  \frac{z}{e^z-1} \,  \quad \text{and} \quad z= - \frac{c_{p,f} \dot{m_f} B}{ \pi D^2 h_c} \, .
\end{equation}

The Nusselt and Sherwood numbers are computed from the empirical correlations of \citet{ranzmarshall}:
\begin{equation}\label{nueq}
 Nu= 2.0 +0.6 Re^{1/2} Pr^{1/3} \, ,
\end{equation}
\begin{equation}\label{sheq}
 Sh = 2.0 +0.6 Re^{1/2} Sc^{1/3} \, .
\end{equation}
They account for the extra heat and mass transfer due to gas motion around the droplet. All gas properties in Equations \eqref{nueq} and \eqref{sheq} are computed using the film temperature, see \citet{naca}.

The vapor mass fraction in the droplet surrounding ($Y_{v,s}$) is obtained assuming liquid-vapor equilibrium and using Raoult's law:
\begin{equation}
Y_{v,s} = \frac{p_v \left(T_d\right)}{p}\frac{W_v}{W} \, ,
\end{equation}
$p_v \left(T_d\right)$ is the equilibrium vapor pressure at droplet temperature $T_d$.

\section{The Experiment Configuration}
\label{experiment}
\citet{chen} performed at Sydney University a detailed experimental investigation of a turbulent evaporating jet of acetone. Droplet diameter, droplet velocity, droplet number density and liquid volumetric flux were measured using a two-component phase Doppler interferometry (PDI) and acetone vapor mass flux was measured using planar laser-induced fluorescence (PLIF).

The experiment consists of a spray nozzle centered on the exit plane of a wind tunnel with 150mm by 150mm that supplies a co-flowing air stream, see Figure \ref{spray_jet}. 
Inside the nozzle, a pressurized liquid jet is surrounded by a carrier air flow until the nozzle exit. The co-flow has a low turbulence intensity of less than $2\%$, so that the effect on the spray jet turbulence is negligible. The main benefit of this setup is the avoidance of flow recirculation near the nozzle exit, simplifying the boundary conditions for a numerical simulation.

Table \ref{table: exp} summarizes the basic characteristics of the experiment. The droplet velocities in the nozzle exit were measured and the droplet and gas temperatures were calculated from thermal equilibrium assumption. The statistical distribution of droplet diameters in the nozzle was not provided by the experiment and only the Sauter mean diameter (SMD) is known.  A lognormal distribution was then specified with expected value and standard deviation being, respectively, $E = 10\ \mu m$ and $SD=5.5\ \mu m$.

\section{Aspects of the Numerical Solution}
The numerical solution of the equations in Section \ref{equations} was obtained using finite volume method with the OpenFOAM code, more specifically the dieselFOAM solver and dieselSpray library. Minor modifications were made in order to solve for the low Mach number equations and to adapt boundary conditions to the experiment.

\subsection{Domain Discretization (The Numerical Grid)}

Two axisymmetric (2D) orthogonal mesh grids were built: a fine and a coarse grid. OpenFOAM uses the collocated grid arrangement. The fine grid was composed of $14,800$ cells ($74$ cells in the radial direction and $200$ in the axial direction) and the coarse grid was composed of $5,460$ cells ($42$ in the radial direction and $130$ in the axial direction).

Solutions for both grids were obtained for a pure gas flow. In despite of some differences, the coarse grid solution was considered enough accurate and it was used for the results presented in this paper.

Droplets were injected in the domain at a rate of $5 \times 10^5\ droplets/s$.

\subsection{Equation Discretization}

The general transport equation for a scalar intensive property $\phi$ is:
\begin{equation}
 \underbrace{\frac{\partial \rho \phi}{\partial t}}_{\text{temporal derivative}} + \underbrace{\nabla \cdot \left(\rho \bv{U} \phi \right)}_{\text{advection}} - \underbrace{\nabla \cdot \left( \rho \Gamma \nabla \phi \right)}_{\text{diffusion}} = \underbrace{S\left(\phi\right)}_{\text{source}}
\end{equation}

The discretization of each term used the following schemes:
\begin{itemize}
 \item Temporal derivative: Euler implicit method;
 \item Advection term: Gauss theorem is applied to transform the divergence in surface integrals and the upwind scheme is used to interpolate values from cell center to cell faces;
 \item Diffusion term: Gauss theorem for the divergent with linear interpolation for the diffusion coefficient from cell center to cell faces. Central difference is used for the gradient;
 \item Sources: all sprays sources are explicitly handled. The spray is evolved from time step "n" to "n+1" using the gas phase properties at time step "n", and all source terms are computed. The gas is then evolved to time step "n+1".
\end{itemize}

After discretization, the resulting algebraic equations were solved in segregated linear systems using conjugate gradient algorithms.
The pressure-velocity coupling is solved using the PISO algorithm, see \citet{issa}. In the low Mach number formulation, the convective-diffusive pressure equation for compressible flows becomes a Poisson equation as it is for incompressible flows.

\section{Results}

\subsection{Gas and Droplet Velocities and Droplet Dispersion}

An empirical observation of a gas turbulent round jet is that the profiles of $\hat{U}$, $\hat{k}$ and $\hat{\nu}_T$ as function of $\hat{y}$ becomes self-similar (i.e. independent of the axial coordinate) from some distance to the nozzle exit.
Furthermore, there is a linear relation between the half-radius ($y_{1/2}$) and the axial coordinate, that is, the jet sprad rate is constant:
\begin{equation}
S = \frac{d y_{1/2} (x)}{dx} \, .
\end{equation}

\citet{chen} has experimentally shown that self-similarity also occurred for $\hat{U}$ in the presence of the spray jet for $x/D \ge 10$. For arriving at this conclusion, it assumed the velocity measured for the smaller droplets ($D<3\mu m$) as being the gas velocity.

The group of Figures \ref{fig: selfsimilar_profile} shows the same investigation for the gas velocity field obtained in the numerical simulation. The Figure \ref{ssA} shows the profiles of $\hat{U}$ for several axial coordinates. Differently than the experiment, the self-similarity only occurs for $x/D \ge 15$ (and not for $x/D \ge 10$). This suggests that the transition from the jet developing region to the turbulent region is retarded in the simulation.

Figures \ref{ssB} and \ref{ssC} show that the evolution to self-similar profiles is slower for the turbulent quantities $\hat{k}$ and $\hat{\nu}_T$ and one may not say that the self-similarity was achieved inside the domain. \citet{chen} has shown that for $x/D \le 25$ it was also not achieved in the experiment for the root mean square of the axial velocity fluctuation ($<U''_x>$).

One observation concerning the turbulent viscosity ($\hat{\nu}_T$) is that the centerline value seems to evolve to a value of $\hat{\nu}_T = 0.047$, higher than values found in the literature for the turbulent round jet. \citet{pope2000turbulent} reports a self-similar turbulent viscosity at centerline of $\hat{\nu}_T = 0.029$.

The jet spread rate measures the spread of the axial velocity in the radial direction. A value about $S=0.095$ is found on the literature for the pure gas flow \citet{pope2000turbulent}, and it is independent of Reynolds number. For the spray jet, however, \citet{chen} has reported $S=0.0665$. In the simulation it was found a higher value of $S=0.071$, a difference of $23\%$.

The Figure \ref{fig: ssS} shows the indeed different half-radius values found in the simulation and the experiment for the same axial coordinates and Figure \ref{fig: ssP} compares the self-similar profiles of $\hat{U}_x$ from the simulation and the experiment.

The overestimation of the round jet spread rate with the default coefficients of k-epsilon model has been reported previously by different authors. \citet{luppes} has discussed four alternatives to the correction. The modifications consist not only in changing the model coefficients, but also in tunning of boundary conditions.

\subsection{Gas and Droplet Velocities and Droplet Dispersion}

\citet{chen} has presented velocity measurements for droplets of four diameter classes in the spray jet: $D< 5\mu m$, $10\mu m <D<20\mu m$, $20 \mu m < D < 30 \mu m$ and $30 \mu m < D < 40 \mu m$. The importance of separating the data in droplet size classes is that the drag force is dependent on its size.

Figure \ref{fig: Ux_gas} shows in red squares the measured radial profiles of the mean axial velocity for the gas phase ($\tilde{U}_x$) in the axial coordinates of: $x/D=5$, $10$, $15$, $20$ and $25$. The radial profile obtained in the simulation is also shown in a solid black line. It is evident that the underprediction of the gas velocity in the simulation becomes higher in the downstream direction. In the axial coordinate of $x/D=25$, the centerline value in the simulation is about $20\%$ below the experimental value. In fact, \citet{luppes} has reported an error of the same percentual magnitude using the k-epsilon model with the default coefficients.

\citet{chen} has also reported the root mean square of the axial and the radial velocities, respectively $<U''_x>$ and $<U''_y>$. Again, the values for the smaller droplets ($D< 5\ \mu m$) are being used to represent the gas velocity. The same quantities are not provided in the simulation for direct comparison. The turbulence model only solves for the turbulent kinetic energy and its dissipation rate. A characteristic velocity, however, may be defined as $<U''>=\sqrt{2/3 k}$. This quantity is shown in Figure \ref{fig: UUx} with $<U''_x>$ and $<U''_y>$. Again, it is shown radial profiles for different axial coordinates.

Three observations are readily made:  
\begin{itemize}
  \item The radial profile of $<U''_x>$ is similar to that of $<U''_y>$;
  \item The characteristic velocity defined for the k-epsilon model ($<U''>$) agrees well with $<U''_x>$ and $<U''_y>$;
 \item $<U''>$ declines faster than $<U''_x>$ and $<U''_y>$ as the axial coordinate increases.
\end{itemize}

The first observation indicates that the turbulent jet differs from the self-similar profiles found in the literature \citet{pope2000turbulent}, where $<U''_x>$ is about twice as big as $<U''_y>$ in the centerline. It remains the question whether this occurs because the jet is not fully developed or because of some spray influence.

The second observation suggests that the k-epsilon model is appropriate to describe the round jet turbulence, providing an appropriate characteristic velocity. The third observation, however, indicates that the turbulent dissipation is overpredicted.

Returning to the mentioned influence of the discrepancies in the mean axial velocity of the gas phase on the droplet velocities, the lower velocity of the gas increases the droplet slip velocity and causes a larger drag to be felt by them. The obvious conclusion is that the same discrepancy in the gas velocity will be observed in the comparison of experimental and numerical values for the droplet velocities.

Figure \ref{fig: Ux} shows in red squares the radial velocity profile measured  by \citet{chen} for droplets in the class of $10\mu m <D<20\mu m$ in the axial coordinates of: $x/D=5$, $10$, $15$, $20$ and $25$.  For each radial position, the velocity value is the ensemble average of the droplets crossing the laser PDI beams. The droplet velocity obtained in the simulation is also shown as black scattered points representing each of them a different droplet in the computation, no averaging was performed.

It is observed that the computed droplet velocity also becomes systematically lower than the measurements in the downstream direction.

\subsection{Liquid and Vapor Mass Fluxes}

The correct prediction of droplet evaporation is, together with the velocity, of high importance for the applications of the spray simulation. The reasonably low discrepancies found for droplet velocity in Figure \ref{fig: Ux} already indicates that the evaporation model performed well. This is due to the fact that an incorrect modeling of evaporation would lead to an erroneous droplet diameter, further affecting the velocity prediction.

Figure \ref{fig: droplet_flux} shows the numerical and the experimental liquid mass flow rate for different axial coordinates. The computed values are in good agreement with the measurements for $x/D \le 10$. Further downstream, the simulation overpredicts the liquid flow rate. Clearly, this means that droplet evaporation is lower than it should be. In fact, in Figure \ref{fig: vapor_flux}, it is made the comparison of experimental and numerical radial profiles of vapor mass fluxes and the trend is exactly the same: as the axial coordinate increases, less vapor (hence more liquid) is present in the simulation than it is in the experiment.

The underestimation of evaporation rate has also been noticed in a large-eddy simulation performed by \citet{bini}, where it was discussed the effect of a subgrid model to the evaporation. It is said that the lack of a subgrid model affects evaporation mainly in the jet core. In fact, the evaporation model presented in Section \ref{dispersed_phase} only dealt with mean flow properties and transfer coefficients, and the temperature and mass fraction oscillations, ($T''$) and ($Y''_{ac}$), were not taken into account.

Surely, for a large-eddy simulation, the lack of a subgrid model is less severe than it is in RANS modeling, since at least the non-filtered oscillations are present.

It must also be pointed that uncertainties in the gas mean velocity also affect the prediction of droplet heating and evaporation because of the different slip velocity. The correct attempt to improve accuracy of the present computation would be to adjust the turbulence model to first correct the gas mean velocity and see the new evaporation rates. Next, a stochastic subgrid model for the heat and mass transfer coefficients could be studied.

The last studied property of the spray is the droplet Sauter mean diameter (SMD). Figure \ref{fig: SMDprofile} shows the numerical and experimental radial profiles of SMD.

The computed Sauter mean diameter shows a good agreement with measurements, in despite of the uncertainty about the droplet size distribution in the boundary conditions and the higher liquid flow rates in Figure \ref{fig: droplet_flux}. This may also be a consequence of the low variance in droplet diameter in the nozzle injection.

This agreement also confirms that the velocity discrepancy is caused by the anomalous jet spread rate. As discussed previously, if diameters were also in disagreement, the predicted drag force on the droplets would be further incorrect.

\section{Conclusions}

The velocity of both the gas and the liquid phase were underpredicted due to the high turbulent viscosity of the turbulence model. It is suggested in the literature that this discrepancies for a round jet might be diminished by changing the model coefficients and reducing the turbulent length scale in the jet inlet boundary condition.

The evaporation rate was underpredicted after the jet developing region. Turbulence effects on Nusselt and Sherwood numbers were not modeled and this is belivied to decrease evaporation, specially in the jet core, as reported by \citet{bini}.

Disagreements in the prediction of gas velocity may also have contributed to differences in the evaporation rate. The error in liquid mass flow rate at some axial location was about $20\%$.

The shape of the radial profile of vapor mass flux was well predicted. The same can be said to the SMD radial profiles.

Future work could explore improvements in the round jet turbulence modeling and new heating and evaporation models for RANS simulations accounting for fluctuations of flow properties.

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

%% \section{}
%% \label{}

%% References
%%
%% Following citation commands can be used in the body text:
%% Usage of \cite is as follows:
%%   \citet{key}          ==>>  [#]
%%   \cite[chap. 2]{key} ==>>  [#, chap. 2]
%%   \citet{key}         ==>>  Author [#]

%% References with bibTeX database:

\bibliographystyle{model1c-num-names}
\bibliography{../thesis/references.bib}
\nocite{*}
%% Authors are advised to submit their bibtex database files. They are
%% requested to list a bibtex style file in the manuscript if they do
%% not want to use model1a-num-names.bst.

%% References without bibTeX database:

% \begin{thebibliography}{00}

%% \bibitem must have the following form:
%%   \bibitem{key}...
%%

% \bibitem{}

% \end{thebibliography}

\newpage
\listoffigures

\newpage

\begin{table}
\centering
 \begin{tabular}{lc}
\hline
Experiment  Data & \\ \hline
Liquid Phase & Acetone \\
Liquid Flow rate at nozzle exit $(g/min)$  &  $7.0$ \\
Carrier air flow rate $(g/min)$ & $135$ \\
Vapor flux at nozzle exit $(g/min)$ & $5.0$ \\
SMD at nozzle exit $(\mu m)$ & $13.7$ \\
Gas temperature at nozzle exit $(K)$ & $280$ \\ 
Gas jet Reynolds Number $(Re=4 \dot{m}_g/\pi D_{nozzle} \mu_g)$ & $16,300$ \\
\hline
 \end{tabular}
\caption{Basic information about the experiment of \citet{chen}.}
\label{table: exp}
\end{table}

\begin{table}
\centering
 \begin{tabular}{cc}
  \hline
  $C_{\mu}$ & $0.09$ \\
  $C_{\epsilon 1}$ & $1.44$ \\
  $C_{\epsilon 2}$ & $1.92$ \\
  $C_{\epsilon 3}$ & $-0.33$ \\
  $\sigma_k$ & $1$ \\
  $\sigma_{\epsilon}$ & $1.3$ \\ \hline
 \end{tabular}
\caption{k-epsilon model constants.}
\label{table: kepscoeff}
\end{table}

\begin{figure}
 \centering
 \includegraphics[width=0.75\textwidth]{../thesis/figuras/chap3/setup.png}
 % setup.png: 645x675 pixel, 72dpi, 22.75x23.81 cm, bb=
 \caption{Configuration of the spray jet nozzle of \citet{chen}.}
 \label{spray_jet}
\end{figure}

\begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \subfloat[]{\label{ssA}\includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/selfsimilar/selfsimilar_U.png}} &
 \subfloat[]{\label{ssB}\includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/selfsimilar/selfsimilar_k.png}} \\
 \subfloat[]{\label{ssC}\includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/selfsimilar/selfsimilar_nut.png}} &\\
\end{tabular}
 \caption{Numerical results showing the evolution to self-similar profiles of the following gas phase properties: (a) dimensionless mean axial velocity - $\hat{U}_x$, (b) dimensionless turbulent kinetic energy - $\hat{k}$ - and (c) dimensionless turbulent viscosity - $\hat{\nu}_t$. The different curves show the profiles for the correspondent axial coordinate indicated in the legend.}
\label{fig: selfsimilar_profile}
\end{figure}

\begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \subfloat[]{\label{fig: ssS}\includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/selfsimilar/selfsimilar_spread.png}} & \subfloat[]{\label{fig: ssP}\includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/selfsimilar/selfsimilar_num_exp.png}}
\end{tabular}
 \caption{(a) Half-radius as a function of the axial coordinate in the experiment and in the simulation.  (b) Self-similar radial profile of the numerical and the experimental dimensionless axial velocity - $\hat{U}_x$. The measurements were obtained from \citet{chen}.}
\label{fig: selfsimilar_spread}
\end{figure}

\begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/Ux/Ux_gas/Ux_gas5.png} & \includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/Ux/Ux_gas/Ux_gas10.png} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/Ux/Ux_gas/Ux_gas15.png} & \includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/Ux/Ux_gas/Ux_gas20.png} \\
(c) & (d) \\
\includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/Ux/Ux_gas/Ux_gas25.png} &   \\
(e) & \\
\end{tabular}
 \caption{Measured and computed radial profiles of the mean axial velocity of the gas phase ($\tilde{U}_x$). The measurements were obtained from \citet{chen}.}
 \label{fig: Ux_gas}
\end{figure}

\begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/UU/UUx5.png} & \includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/UU/UUx10.png} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/UU/UUx15.png} & \includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/UU/UUx20.png} \\
(c) & (d) \\
\includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/UU/UUx25.png} &  \\
(e) & \\
\end{tabular}
 \caption{Measured radial profiles of the gas phase velocities $<U''_x>$ and $<U''_y>$ and the characteristic velocity defined for the k-epsilon model $<U''>=\sqrt{2/3 k}$ obtained from the simulation. The measurements were obtained from \citet{chen}.}
\label{fig: UUx}
\end{figure}

\begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/Ux/Ux_drops/Ux_x5.png} & \includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/Ux/Ux_drops/Ux_x10.png} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/Ux/Ux_drops/Ux_x15.png} & \includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/Ux/Ux_drops/Ux_x20.png} \\
(c) & (d) \\
\includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/Ux/Ux_drops/Ux_x25.png} &   \\
(e) & \\
\end{tabular}
 \caption{Red squares are the measured radial profile of the mean axial velocity of droplets ($\tilde{U}_{d,x}$) in the size class of $10\mu m <D<20\mu m$. Scattered black dots are the computed velocities for each droplet in the numerical simulation. The measurements were obtained from \citet{chen}.}
 \label{fig: Ux}
\end{figure}

\begin{figure}[!htb]
\centering
  \includegraphics[width=0.6\textwidth]{../thesis/figuras/chap5/massflow/drop_mflux.png}
\caption{Droplet mass flow rate at four axial locations. The measurements were obtained from \citet{chen}.}
\label{fig: droplet_flux}
\end{figure}

\begin{figure}[!htb]
 \centering
\begin{tabular}{cc}
 \includegraphics[width=0.45\textwidth]{../thesis/figuras/chap5/massflow/mvapor5.png} & \includegraphics[width=0.45\textwidth]{../thesis/figuras/chap5/massflow/mvapor10.png} \\
(a) & (b) \\
\includegraphics[width=0.45\textwidth]{../thesis/figuras/chap5/massflow/mvapor15.png} & \includegraphics[width=0.45\textwidth]{../thesis/figuras/chap5/massflow/mvapor20.png} \\
(c) & (d)
\end{tabular}
 \caption{Radial profiles of mean axial mass flux of acetone vapor: $\dot{m}''_{ac}=\rho \bar{Y}_{ac} \tilde{U}_x$. The measurements were obtained from \citet{chen}.}
 \label{fig: vapor_flux}
\end{figure}

\begin{figure}[!htb]
 \centering
\begin{tabular}{c}
 \includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/smd/smd5.png} \\
(a) \\
 \includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/smd/smd15.png} \\
(b) \\
 \includegraphics[width=0.5\textwidth]{../thesis/figuras/chap5/smd/smd25.png} \\
(c) 
\end{tabular}
 \caption{Numerical and experimental radial profiles of Sauter mean diameter (SMD). The measurements were obtained from \citet{chen}.}
 \label{fig: SMDprofile}
\end{figure}

\end{document}

%%
%% End of file `elsarticle-template-1a-num.tex'.

### 2016-09-24 pod_encit/tags/draft/paper.tex
new lib version

\documentclass[10pt,fleqn,a4paper]{article}
\usepackage{encit2010}
\usepackage{amsmath}
\usepackage{amsthm}

\newcommand{\bv}[1]{\mathbf{#1}}
\newcommand{\xp}{x^{\mathbf{\prime}}}
\newtheorem{theorem}{Theorem}

% LEGEND FOR THE GEOMETRIES
% Action = Head 1
% Scania = Head 2
% Sygma = Head 3
% Ducato = Head 4

\begin{document}
\hspace{-8.5mm}
\begin{tabular}{||p{\textwidth}}
\begin{center}
\vspace{-4mm}
\title{ANALYSIS OF STEREOSCOPIC PIV MEASUREMENTS OF IN-CYLINDER FLOWS USING PROPER ORTHOGONAL DECOMPOSITION}
\end{center}
\authors{Rodrigo Piccinini, rodrigo.piccinini@sygmamotors.com.br} \\
\authors{Carla Fernandes, carla.fernandes@sygmamotors.com.br} \\
\authors{Marcelo Andreotti, marcelo.andreotti@sygmamotors.com.br} \\
\institution{Sygma Motors, Av. Cassiano Ricardo, 1306, sala 2, Jardim Alvorada, CEP 12240 - 540, S\~{a}o Jos\'{e} dos Campos - SP - Brasil} \\
\\
\abstract{\textbf{Abstract}

% SHORT ABSTRACT

%In the present work, the Proper Orthogonal Decomposition (POD) was applied in the
% identification of flow coherent structures. The experiment consisted of 
% mounting on an Plexiglas cylinder the intake ports of an internal 
% combustion engine and establishing a steady flow through this system.
%Velocity fields at different cylinder cross sections and for several 
%instants of time were measured by Particle Image Velocimetry (PIV).
%It was shown that the principal eigenvector might contains up to 80\% of the
%time averaged kinetic energy of the flow field.

In the present work, the Proper Orthogonal Decomposition (POD) was applied in the identification of flow coherent structures. For the velocity flow field, stereo Particle Image Velocimetry (PIV) measurements were carried out in a fixed cross section. We have observed that the principal eigenvector contains up to 80$\%$ of the time averaged kinetic energy of the flow field.
}
\\
\\
\keywords{\textbf{Keywords:} proper orthogonal decomposition, stereo PIV, coherent structures, turbulent flow, IC engines.}\\
\end{tabular}

\section{INTRODUCTION}

In the development of cylinder head intake ports for spark-ignited internal combustion engines, the characterization of coherent flow structures generated during induction - the most usual being tumble and swirl motions - is an important issue for combustion acceleration.

The flow field will directly affect the mixture formation and hence the spatial fuel distribution inside the combustion chamber. A spatially non-homogeneous mixture leads to changes in flame behavior in terms of heat release, propagation speed, etc. This behavior is not well understood and has motivated the research reported in numerous studies (Haworth et al., 2000; Jim\'{e}nez et al., 2002; Pasquier et al., 2007; H\'{e}lie and Trouv\'{e}, 2000; Xiong and Roberts, 2002). The characteristics of the combustion depend strongly on local variations of mixture and also on the length-scale of these variations.

In a general way, higher turbulence levels during flame development reduce combustion duration.  A common strategy to achieve these higher turbulence levels consists of disintegrating large flow structures into turbulence near the spark release event. In this way, the kernel formation will depend on the local equivalence ratio and temperature around the spark plug. The flow field might also advect the initial kernel and the local turbulence will influence its evolution to a turbulent flame. The turbulent flame for the premixed thin reaction sheet will grow faster than the laminar flame due to the large scale turbulence ($l_{t}/\delta^{0}_{L} > 1$, where $l_{t}$ is the integral length scale and $\delta^{0}_{L}$ is the laminar flame thickness (Poinsot and Veynante, 2005)). A faster combustion event is desired because it will increase the efficiency of converting chemical energy to piston work. Therefore, the majority of combustion strategies will attempt to create such desirable high turbulence levels near the ignition.

For this purpose, the creation of a strong swirling flow inside the combustion chamber is often studied. The idea is to store a great amount of kinetic energy in large structures during the intake. During the compression, these structures will remain until the ignition event and then they will break into smaller structures increasing the turbulence level. In case of swirling structures not achieved during the intake, the kinetic energy of the flow might dissipate faster.

Furthermore, the turbulent flow is characterized by random and chaotic three-dimensional motion. In this way, the flow will differently affect the formation of large structures cycle-by-cycle, rendering the engine operation difficult. Thus, not only the magnitude of the swirling motion is important, but also its regularity. Therefore, the knowledge of the bulk motion in the cylinder and the turbulence evolution is of great importance for the improvement of engines.

The objective of the present work is the study of the statistical properties of the flow induced by the intake ports in four different cylinder heads using proper orthogonal decomposition (POD). In order to obtain the velocity flow field, the stereo particle image velocimetry (PIV) measurements were carried out in a specific cylinder cross section. This technique enables spatially resolved measurements of the three instantaneous components of the flow velocity field within a very short time and allows the detection of large and small scale spatial structures. Once the velocity field was acquired, the POD was applied in order to know how the flow energy is stored. This statistical analysis demonstrates a promising tool in the design of intake ports.

\section{EXPERIMENTAL SET-UP AND MEASUREMENTS}

The PIV is a non-intrusive optical technique widely used in fluid mechanics to determine the velocity flow field. This is performed by adding small tracer particles (seeding) to the flow. The region under investigation is illuminated at least twice by a thin pulsed laser. The light scattered by the particles is recorded by a high-resolution camera. The velocity field is obtained from the displacement of the particle images between two consecutive light pulses. Standard PIV makes use of a single camera, which allows one to obtain only two components of the velocity vector. In order to obtain all the three velocity components it is required a recording of an additional view of the field (Stereoscopic PIV - SPIV).

The experiments were carried out by means of a 3D-PIV test bench for visualization of intake port generated flow field. The Figure~\ref{fig: bench} presents a schematic set-up of the flow test rig. This Stereoscopic PIV system consists of the follow components: (1) A double pulsed Nd:YAG laser (New Wave Solo XT 120) that outputs a maximum energy of 120\,mJ per pulse with a wavelength of 532\,nm into a beam with diameter of 5\,mm and a pulse duration of 4\,ns. The light sheet was shaped with thickness of about 2\,mm. (2) Two PCO.1600 double shutter CCD cameras with resolution of 1600x1200\,pixel and 14-bit quantization, and two Canon EF 35mm lenses. The two CCD cameras were positioned symmetrically at two sides of the cylinder and inclined at 45$^{\circ}$ with respect to the horizontal measurement plane under the cylinder head face. This arrangement permits the most accurate determination of the out-of-plane displacement (Raffel et al., 2007). The camera bodies were tilted with respect to the optical axis of their lenses. This is done to satisfy the Scheimpflug condition, i.e., to focus over the whole imaged measurement plane, as shown in Fig.~\ref{fig: bench}. (3) A controller (ILA GmbH) guarantees the synchronization between generated laser pulses and image acquisition. (4) A ILATEC 40 seeder operating with Di-Ethyl-Hexyl-Sebacate (DEHS) as the seeding material for generating tracer particles with mean diameter of 1\,$\mu$m. (5) For image acquisition it was used the CamWare software. Stereoscopic calibration and data evaluation were realized by VidPIV 4.6 XP (ILA GmbH) software. (6) A test cylinder head without the piston. The cylinder head is mounted on top of a Plexiglas cylinder with the same inner diameter as the engine bore and 300\,mm length. The test bench is equipped with an universal valve-adjusting device which allows for the valve lift positioning in steps of 1\,mm. %For the flow testing the hard original valve springs are replaced by soft ones.

\begin{figure}[h]
 \centering
 \includegraphics[width=1\textwidth]{./imgs/bench.pdf}
 \caption{Stereoscopic PIV test bench. (Left) Stereo PIV system; (Right) (a) Positioning of the particle feeding, (b) Scheimpflug configuration.}
 \label{fig: bench}
\end{figure}

The first step before starting the PIV experiment is the calibration that determines the metric on the measured plane, which is performed as follows. The horizontal laser sheet was aligned with a flat calibration target placed on top of the cylinder. After imaging the target a Tsai model (Tsai, 1987) was used to mapping pixel to real coordinates for each CCD camera. This model is based on pinhole camera model and describes radial lens distortions using nonlinear expression. Although the Tsai model employs more complex mapping function, it determines the position of CCD cameras for later reconstruction of the three-dimensional vector field. As the laser and the CCD cameras are installed on a three-axis traversing bench, different cross sections of the cylinder can be measured without requiring additional recalibration.

During the experiments, the pressure drop inside the cylinder has been kept constant at 4.9\,kPa by means of a blower bypass. The air flow is measured by a roots meter with simultaneous measurement of air temperature and pressure drop at the metering device.

The PIV measurements were taken on a cross section at distance D (cylinder diameter) bellows cylinder head face. A total of 100 frames pairs were recorded for each CCD camera at a frequency of 8\,Hz with a time interval between laser pulses of 10\,$\mu$s. The statistical sample size (100 frames) showed be enough for a precise flow decomposition.

The tracer particle feeding was displaced close to the intake port of cylinder head and the particle flow has been adjusted until it was seen roughly 10 to 20 tracer particles in a typical 64x64\,pixel interrogation area, as shown in Fig.~\ref{fig: densityimage}.

\begin{figure}[h]
 \centering
 \includegraphics[width=0.4\textwidth]{./imgs/density.pdf}
 \caption{PIV sample image showing the calculated velocity field vectors over a cross section of an in-cylinder. Green square highlights a given interrogation area of 64x64 pixel. White dots are tracer particles.}
 \label{fig: densityimage}
\end{figure}

The PIV images were post-processed using multi-pass correlation to provide higher resolution vector data. A first correlation was started with a 64x64\,pixel interrogation size with 50$\%$ overlap followed by an adaptive cross-correlation with a final 32x32\,pixel interrogation size with 50$\%$ overlap. A median filter was applied in order to filter out outlier vectors. An interpolation algorithm based on a weighted mean technique was employed to replace the filtered velocity vectors. Finally, the perspective vector maps from each CCD camera were combined to obtain the third component velocity and to improve accuracy for the other two components.

\section{PROPER ORTHOGONAL DECOMPOSITION - SNAPSHOT METHOD}

The characterization of coherent structures is an interesting issue in the study of turbulent flows and it is the subject of vast literature (Heywood, 1988; Cosadia et al., 2006; Cazemier, 1997; Bergmann and Cordier, 2003, Hung and Hien, 1998; Boree, Maurel and Bazile, 2002; Li Y. et al., 2004). However, the mathematical definition of such structures is not necessarily standard and might differ depending on context. One possible way is the Proper Orthogonal Decomposition (POD) (Lumley, 1967), which establishes a criterium based on maximum energy content.

For optimizing the operation of an internal combustion engine, as previously mentioned in Section 1, the characterization of flow main structures is fairly important and the POD demonstrates to be an useful tool for their identification. As a favourable point, the efficiency of the decomposition increases with flow inhomogeneity such as those in the strong swirling flows sought in engine application.

The main definitions used in the derivation of POD formulation are summarized below. Details of the theory might be found in Griffel (2002). Let's consider a set of $N$ observations (snapshots) of the flow velocity field denoted by $\left \lbrace \mathbf{U}_{s} : (1 \le s \le N) \in Z, \mathbf{U}_s \in \Omega \right \rbrace$, where $\Omega$ is the flow domain. $\bv{U}_{s}$ is the velocity field, a vector formed by the three velocity components $\bv{U}_{s}(\bv{x})=(u_1 (\bv{x}), u_2 (\bv{x}), u_3(\bv{x}))$. Due to the physical meaning of the velocity field, it is assumed that $\bv{U}_s \in L^2 (\Omega)$, where $L^2(\Omega)$ is the Hilbert space of the square-integrable functions (Riesz and Sz.-Nagy, 1990).

The observations can be the result of experimental measurements or computer simulations. These results may be not necessarily the flow velocity, but any of the flow properties. Furthermore, each observation must be taken in a different instant of time. In the present work, one observation is made up of the full procedure required by the PIV technique (as described in the Section 2) to provide a velocity field in some instant of time. The frames are taken at different times equally spaced and each event is assumed linearly independent.

The proper orthogonal decomposition consists in decomposing the set of observations in an orthonormal basis which ensures an extra condition for its elements (or modes, the nomenclature used hereafter). Thus, the velocity field can be represented by a series expansion in terms of the modes:

\begin{equation}\label{eq: Us}
 \bv{U}_s (\bv{x}) = \sum_{j=1}^{N} b^{s}_{j} \bv{\Phi}_{j}(\bv{x}) \,,
\end{equation}
\\
where $b^{s}_{j}$ is the time-dependent amplitude coefficient and the $\bv{\Phi}_{j}(\bv{x})$ is the set of vectors generating the POD basis.  
Each mode might be understood as a coherent structure.

These functions will be the most similar to the members of $\bv{U}_{s}$ on average if the normalized average projection of $\bv{U}_{s}$ onto $\bv{\Phi}$ is maximum:

\begin{equation}\label{eq: max-cond}
max \left \lbrace \frac{1}{N} \sum_{s=1}^{N} | \left(\bv{U}_{s} , \bv{\Phi}\right) |^2 \right \rbrace \quad \text{with} \quad (\bv{\Phi},\bv{\Phi}) =  \parallel \bv{\Phi} \parallel^2 = 1 \quad \text{and} \quad \bv{\Phi} \in L^2 (\Omega) \,,
\end{equation}
\\
where $(.,.)$ denotes the inner product associated with the norm $\parallel . \parallel$. The inner product of two vector fields $\bv{F}$ and $\bv{G}$ is defined as $(\bv{F},\bv{G}) \equiv \int_{\Omega} {\bv{F} \cdot \bv{G}} dS$, and the norm by $\parallel \bv{F} \parallel = \sqrt{(\bv{F},\bv{F})}$.

If it is made the assumption of linearly independent observations, the $\bv{\Phi}$ functions might be written as a linear combination of them:

\begin{equation}\label{eq: phifunctions}
 \bv{\Phi}_s = \sum_{j=1}^{N} a^{s}_{j} \bv{U}_{j}(\bv{x}) \,.
\end{equation}

The maximisation, Eq.~\eqref{eq: max-cond}, would be solved if all the coefficients $a^{s}_{j}$ in Eq.~\eqref{eq: phifunctions} were determined. For this purpose, let's define an integral operator $\bv{R}$ with kernel $\bv{K}$ given by:

\begin{equation}\label{KandR}
 \bv{K} (\bv{x}, \bv{x}^{\prime}) \equiv \frac{1}{N} \sum_{s=1}^{N} \bv{U}_s (\bv{x}) \bv{U}_s (\bv{\xp}) \quad \text{and} \quad \bv{R} \bv{\Phi} \equiv \int_{\Omega} \bv{K} (\bv{x}, \bv{x}^{\prime}) \cdot \bv{\Phi}(\bv{x}^{\prime}) dS^{\prime} \,,
\end{equation}
\\
where $\bv{K}=\sum_{s=1}^{N} (u_i u_j)_s$ is a tensor of second rank (a three-by-three matrix) and $\bv{R} \bv{\Phi}$ is a vector of three components. Both are composed of $L^2(\Omega)$ functions. As mentioned previously, from the physical nature of the velocity field, the elements of $\bv{K}$ satisfy:

\begin{equation}\label{eq: kernel}
 \int_{\Omega} | k_{ij} (\bv{x},\bv{\xp}) |^{2} dS < \infty \,.
\end{equation}

In this case, for a measure space $\Omega$, the linear integral operator on $L^{2}(\Omega)$ with kernel K is compact if $\int_{\Omega} | K( \bv{x})|^2 \bv{dx} \quad$ converges.

Performing some algebraic manipulation, it is possible to show two properties of $\bv{R}$: non-negativeness and self-adjointness. The follow equation shows the non-negativeness of $\bv{R}$:

\begin{equation}\label{st1}
% (\bv{R}\bv{\Phi},\bv{\Phi}) = \sum_{i=1}^{N} | \left( \bv{\Phi} , \bv{U}_i\right) |^2 \ge 0\\
 \begin{split}
 (\bv{R}\bv{\Phi},\bv{\Phi}) &= \int_{\Omega} \bv{R}\bv{\Phi}(\bv{x}) \bv{\Phi}(\bv{x}) dS \\
 &= \int_{\Omega} \int_{\Omega} \bv{K}(\bv{x},\bv{\xp}) \bv{\Phi}(\bv{\xp}) dS^{\prime} \bv{\Phi}(\bv{x}) dS \\
 &= \frac{1}{N} \sum_{i=1}^{N} \int_{\Omega} \int_{\Omega} \bv{U}_{i}(\bv{x}) \bv{U}_{i}(\bv{\xp}) \bv{\Phi}(\bv{\xp}) dS^{\prime} \bv{\Phi}(\bv{x}) dS \\
 &= \frac{1}{N} \sum_{i=1}^{N} | \left(\bv{U}_{i} , \bv{\Phi}\right) |^2 \ge 0 \,.
 \end{split}
\end{equation}

Now, Equation~\eqref{st2} shows the self-adjointness of $\bv{R}$:

\begin{equation}\label{st2}
 \begin{split}
 (\bv{R}\bv{\Phi},\bv{\Psi}) &= \int_{\Omega} \left (\int_{\Omega} \bv{K}(\bv{x},\bv{\xp}) \cdot \bv{\Phi}(\bv{x}) dS \right) \cdot \bv{\Psi}(\bv{\xp}) dS^{\prime} = 
\int_{\Omega} \left( \int_{\Omega} \bv{K}(\bv{x},\bv{\xp}) \cdot \bv{\Psi}(\bv{\xp}) dS^{\prime} \right) \cdot \bv{\Phi}(\bv{x}) dS \\
 &= (\bv{\Phi},\bv{R}\bv{\Psi}) \quad \text{for all} \bv{\Phi}, \bv{\Psi} \in L^{2}(\Omega) \,.
\end{split}
\end{equation}

With these two properties of $\bv{R}$, the problem of maximizing Eq.~\eqref{eq: max-cond} is thus equivalent to a Rayleigh quotient e.g. if $\lambda$ is the largest eigenvalue of a positive self-adjoint compact operator, $\bv{A}$, on a Hilbert space $L^2 (\Omega)$, $\lambda = max \left\lbrace \bv{R(x)} : \bv{x} \in L^2 (\Omega) \right\rbrace$ where $\bv{R(x)=(x,Ax)/(x,x)}$, then:

\begin{equation}\label{eig1}
 \left(\bv{R} \bv{\Phi} , \bv{\Phi} \right) = \lambda \quad \text{with} \quad \parallel \bv{\Phi} \parallel = 1 \,.
\end{equation}

Thus, the Equation~\eqref{eig1} can be rewritten by the follow way:

\begin{equation}\label{eig2}
 \begin{split}
 \left(\bv{R} \bv{\Phi} , \bv{\Phi} \right) &= \lambda \parallel \bv{\Phi} \parallel \,; \\
 \left(\bv{R} \bv{\Phi} , \bv{\Phi} \right) &= \lambda \left(\bv{\Phi}, \bv{\Phi} \right) \, .
 \end{split}
\end{equation}

As $\lambda$ is a real positive eigenvalue: $\lambda \left(\bv{\Phi}, \bv{\Phi} \right) = \left(\lambda \bv{\Phi}, \bv{\Phi} \right)$. So:

\begin{equation}\label{eig3}
 \left(\bv{R} \bv{\Phi} , \bv{\Phi} \right) - \left(\lambda \bv{\Phi}, \bv{\Phi} \right) = 0 \,.
\end{equation}

Applying the additivity property of an inner product on the Eq.~\eqref{eig3} we have:

\begin{equation}\label{eig4}
 \left(\bv{R} \bv{\Phi} - \lambda \bv{\Phi}, \bv{\Phi} \right) = 0 \,.
\end{equation}

According to Eq.~\eqref{eq: max-cond}, $\bv{\Phi}$ can not be zero because his norm equals one. Therefore:

\begin{equation}\label{eig5}
 \bv{R} \bv{\Phi} = \lambda \bv{\Phi} \,.
\end{equation}

Replacing the Eq.~\eqref{eq: phifunctions} and Eq.~\eqref{KandR} on the Eq.~\eqref{eig5} we obtain:

\begin{equation} \label{eq: substituting} 
\int_{\Omega}  \left[\frac{1}{N} \sum_{i=1}^{N} \bv{U}_i(\bv{x})\bv{U}_i(\bv{x}) \right] \cdot \left[\sum_{j=1}^{N} a^{s}_{j} \bv{U}_{j}(\bv{\xp}) \right] dS^{\prime} = \lambda_{s} \sum_{i=1}^{N} a^{s}_{j} \bv{U}_{j} (\bv{x}) \,.
\end{equation}

Applying the dyadic product property $(\mathbf{u} \mathbf{v}) \cdot \mathbf{w} =  (\mathbf{v} \cdot \mathbf{w}) \mathbf{u}$ in Eq.~\eqref{eq: substituting}:

\begin{equation*}
\sum_{i=1}^{N} \left \lbrace \sum_{j=1}^{N} \left[\int_{\Omega} \frac{1}{N} \bv{U}_i(\bv{x}) \cdot \bv{U}_{j} (\bv{x}) dS^{\prime}  \right] a^{s}_{j} \right \rbrace \bv{U}_i(\bv{x}) = \lambda_{s} \sum_{j=1}^{N} a^{s}_{j} \bv{U}_{j} (\bv{x}) \,.
\end{equation*}

A sufficient condition for the equality to hold is:

\begin{equation}
\left[\frac{1}{N} \sum_{j=1}^{N} \int_{\Omega} \bv{U}_{i}(\bv{x}) \cdot \bv{U}_{j}(\bv{x}) dS^{\prime} \right] a^{s}_{j}= \lambda_{s} a^{s}_{j} \,.
\end{equation}

This expression can be written in matricial form according to the definitions below:

\begin{equation} \label{cov1}
\bv{C}_{N \times N} \bv{A}_{N \times 1} = \lambda \bv{A}_{N \times 1} \,,
\end{equation}

\begin{equation*}
\text{where} \, \, C_{ij}= \frac{1}{N}  \int_{\Omega} \begin{bmatrix}  \bv{U}_i(\bv{x}) \cdot \bv{U}_{j} (\bv{x})  \end{bmatrix}_{N \times N} dS^{\prime} \,.
\end{equation*}

Each eigenvector of Eq.~\eqref{cov1} will provide the $a^{s}_{j}$ coefficients of a POD basis element $\bv{\Phi}_s$, which can be obtained by computing Eq.~\eqref{eq: phifunctions}. Each eigenvalue $\lambda_s$ represents the time average kinetic energy of the correspondent mode. For notation purposes, the modes are ordered by the decreasing eigenvalue, so that the average energy content of $\bv{\Phi}_1$ is greater than $\bv{\Phi}_2$ and so on.

Due to the optimality condition imposed by the POD, Eq.~\eqref{eq: max-cond}, the convergence of the energy spectrum is faster than any other possible decomposition (in the sense of the average and the inner product defined in this paper).

The data provided by the PIV system consists of a regular 2D mesh with measured values positioned at grid nodes. Since the POD analysis will require surface integration, the nodes were grouped in triangular cells created by the Delaunay algorithm. The node values were then linearly interpolated to cell geometric centers and the surface integral of a property over a cell was computed from the following simple approximation,

\begin{equation}
 \int_{\Omega_i} \phi (\bv{x}) dS \approx \phi (\bv{x}_{CG}) S \,.
\end{equation}

After performing the integrations and having the matrix $C$ (Eq.~\eqref{cov1}), the computation of eigenvalues and eigenvectors is carried out with the opensource library SciPy.

\section{RESULTS}

The flow field of four different intake ports, namely \textit{Head 1, 2, 3} and \textit{4}, were measured in the stereoscopic PIV test bench for one valve lift fixed at 12\,mm.  Further, the flow field of Head 1 and 4 were also measured for a valve lift of 2\,mm.

The Figure~\ref{fig: energy_fraction} shows the content of time average energy in modes. The energy fraction content in each mode is presented by Fig.~\ref{fig: energy_fraction}(a). The Figure~\ref{fig: energy_fraction}(b) shows the cumulative energy when the modes are grouped from $\bv{\Phi}_1$ to $\bv{\Phi}_{100}$. For higher valve lifts, the flow field has shown more organized in the sense that more energy is stored in few modes. Only the first mode, $\bv{\Phi}_1$, contains up to 80$\%$ of the total energy (see the zoom of the Fig.~\ref{fig: energy_fraction}(a)). Moreover, from the 30$^{th}$ mode, the energy decreases following an exponential decay. The exponential law is given by $E(k) \approx a_1 e^{a_2 k} ,  k \ge 30$. The exponential constant ($a_2$) is in the interval of $[-0.031,-0.023]$ for all cases.

\begin{figure}[h]
\begin{tabular}{cc}
 \includegraphics[width=0.5\textwidth]{./imgs/energy_spectra.pdf} & \includegraphics[width=0.5\textwidth]{./imgs/cumulative_energy.pdf} \\
(a) & (b)
\end{tabular}
 \caption{(a) Fraction of energy in each mode for all tested heads. The small frame is zooming the interval of $[0,10]$. (b) Cumulative energy, that is, the sum of the contributions from $\bv{\Phi}_1$ to some $\bv{\Phi}_k$.}
 \label{fig: energy_fraction}
\end{figure}

\begin{figure}[h]
 \centering
 \includegraphics[width=0.78\textwidth]{./imgs/scania-modes.png}
 % scania-modes.png: 2196x879 pixel, 72dpi, 77.46x31.01 cm, bb=
 \caption{Velocity fields of Head 2: $\bv{\Phi}_1$ (left), $\bv{\Phi}_{15}$ (center) and $\bv{\Phi}_{30}$ (right).}
 \label{fig: organization}
\end{figure}

\clearpage

The link between energy content and the flow organization is supported by the observation of the mode velocity field presented in Fig.~\ref{fig: organization}. Here, the vector magnitudes are normalized according to the same norm defined in Eq.~\eqref{eq: max-cond}. It is possible to notice that the motion organization decreases with the higher order modes. Furthermore, these modes presents several vortical structures while the first mode contains the majority of swirling structure. Hence, $\bv{\Phi}_1$ have the majority of the angular momentum.

So far, only the time averaged energy has been discussed. However, if the $b^{s}_{j}$ coefficients from Eq.~\eqref{eq: Us} are computed, the energy time history is recovered. The Figure~\ref{fig: energy_hist} presents the energy history for the first mode and the set of the first ten modes. It is seen that the first mode is in general more oscillatory when viewed separately in the gray line.  This is rather consistent since energy transfers between the modes are expected. The energy decrease in the first mode possibly increases the energy of some other mode in the set and the other way around. This energy transfer will reduce the oscillations.

The Figure~\ref{fig: Lz} shows the contribution of each mode to angular momentum around the cylinder axis. It is observed that the first mode contains the main swirling motion while the subsequent modes corresponds to the oscillating motion as shown in Fig.~\ref{fig: organization}. As one could expect, the low valve lift of 2\,mm did not clearly define swirl motion because of great magnitude oscillations.

In order to define which intake port geometry will better perform, in terms of increase the turbulence level close to spark ignition event, not only the energy in the modes most be analyzed but also their respective angular momentum. As showed in Fig.~\ref{fig: energy_hist}, the results (c) and (d) have a similar energy history level in the first mode and for the set of ten modes. It is mean that the Head 2 and 3 distribute the energy in a similar way. This same trend is also observed in terms of energy spectra (see Fig.~\ref{fig: energy_fraction}). However, the angular momentum is quite different as presented in Fig.~\ref{fig: Lz} by the plots (c) and (d). Comparing the results of the first mode velocity field presented by Fig.~\ref{fig: scaniavssygma}, it is noticed that the Head 3 velocity field is composed by two-vortex structure against only one big vortex in the Head 2. This is the main reason in the angular momentum production.

\begin{figure}[h]
 \centering
 \includegraphics[width=0.6\textwidth]{./imgs/scania-vs-sygma.png}
 % scania-vs-sygma.png: 1429x740 pixel, 72dpi, 50.41x26.10 cm, bb=
 \caption{Velocity field of the first mode ($\bv{\Phi}_1$) for the Heads 2 (left) and 3 (right). The orange lines highlight the two vortex structure formed by the Head 3 in contrast to the one big vortex of Head 2.}
 \label{fig: scaniavssygma}
\end{figure}

\begin{figure}[h]
\begin{tabular}{cc}
 \includegraphics[width=0.5\textwidth]{./imgs/energy1.pdf} & \includegraphics[width=0.5\textwidth]{./imgs/energy2.pdf} \\
(a) & (b) \\
 \includegraphics[width=0.5\textwidth]{./imgs/energy3.pdf} & \includegraphics[width=0.5\textwidth]{./imgs/energy4.pdf} \\
(c) & (d) \\
 \includegraphics[width=0.5\textwidth]{./imgs/energy5.pdf} & \includegraphics[width=0.5\textwidth]{./imgs/energy6.pdf} \\
(e) & (f)
\end{tabular}
 \caption{Energy history for $\bv{\Phi}_1$ (gray line) and for $\bv{\Phi}_{1}+...+\bv{\Phi}_{10}$ (black line).}
 \label{fig: energy_hist}
\end{figure}

\begin{figure}[h]
\begin{tabular}{cc}
 \includegraphics[width=0.5\textwidth]{./imgs/Lz_H1L2.pdf} & \includegraphics[width=0.5\textwidth]{./imgs/Lz_H1L12.pdf} \\
 (a) & (b) \\
 \includegraphics[width=0.5\textwidth]{./imgs/Lz_H2L12.pdf} & \includegraphics[width=0.5\textwidth]{./imgs/Lz_H3L12.pdf} \\
(c) & (d) \\
\includegraphics[width=0.5\textwidth]{./imgs/Lz_H4L2.pdf} & \includegraphics[width=0.5\textwidth]{./imgs/Lz_H4L12.pdf} \\
(e) & (f)
\end{tabular}
 \caption{Contributions to angular momentum of the first five modes.}
 \label{fig: Lz}
\end{figure}

\clearpage

\section{CONCLUSION}

In order to better understand how the intake ports geometry influence the flow which enters inside the combustion chamber during the intake phase, a statistical decomposition, like proper orthogonal decomposition, was applied on the three dimensional velocity flow field obtained with a stereoscopic PIV.

The results have shown that the first mode contains up to 80$\%$ of the time average kinetic energy of the flow field and have the major swirling structure. Therefore, the first mode defines the average angular momentum. The other modes, which we can call low energy modes, have shown several vortical structures, as presented in Fig.~\ref{fig: organization}, and have an average angular momentum around zero (see Fig.~\ref{fig: Lz}).

In what concerns the flow regularity, Fig.~\ref{fig: Lz} shows that the first mode basically defines the average angular momentum, which is superposed by the contributions of higher modes with average angular momentum around zero. Looking for the differences from one port to another, $\bv{\Phi}_1$ is much more disturbed by higher modes in the case of Head 3. This implies in having more dispersed swirling structures in the engine when concerning the ensemble of different cycles. It is also notable the regularity of Head 4, for which all oscillations are rather small.

Even if a mean swirl is measured, we have shown that the structure of the mean flow is complex and difficult to interpret. Thus, POD has shown itself to be an important methodology to analyze the flow and an important tool for the design of the cylinder head.

\section{ACKNOWLEDGEMENTS}

The authors gratefully thanks the Vale Solu\c{c}\~{o}es em Energia (VSE) for the laboratory structure installed in S\~{a}o Jos\'{e} dos Campos-SP.

\section{REFERENCES}

\begin{thebibliography}{1}
\bibitem[Bergmann and Cordier, 2003]{bergmann} Bergmann M. and Cordier L., 2003, "Post-Processing of Experimental and Numerical Data", Lecture of Von Karman Institute for Fluid Dynamics.
\bibitem[Boree, Maurel and Bazile, 2002]{tumble} Boree J., Maurel S. and Bazile R., 2002, "Disruption of a compressed vortex", Physics of Fluids.
\bibitem[Cazemier, 1997]{podthesis} Cazemier, W., 1997, "Proper Orthogonal Decomposition and Low Dimensional Models for Turbulent Flows", Ph.D Dissertation.
\bibitem[Cosadia et al., 2006]{cyclic}  Cosadia I., Bor\'{e} J., Charnay G. and Dumont P., 2006, "Cyclic variations of the swirling flow in a Diesel transparent engine", Experiments in Fluids.
\bibitem[Griffel (2002)]{griffel} Griffel, D. H., 2002, "Applied Functional Analysis", Dover Books on Mathematics.
\bibitem[Haworth et al., 2000]{Haworth} Haworth D.C., Cuenot B., Poinsot T.J. and Blint R.J., 2000, "Numerical Simulation of turbulent propane-air combustion with non-homogeneous reactants", Combustion $\&$ Flame 121, pp. 395-417.
\bibitem[H\'{e}lie and Trouv\'{e}, 2000]{Helie} H\'{e}lie J. and Trouv\'{e} A., 2000, "A modified coherent flame model to describe turbulent flame propagation in mixtures with variable composition", Proceedings of the Combustion Institute 28, pp.193-201.
\bibitem[Heywood, 1988]{heywood} Heywood, J. B. , 1988, "Internal Combustion Engine Fundamentals", McGraw-Hill International Editions.
\bibitem[Hung and Hien, 1998]{hung} Hung V. Ly and Hien T. Tran, 1998, "Proper Orthogonal Decomposition for Flow Calculations and Optimal Control in a Horizontal CVD Reactor", North Carolina State University.
\bibitem[Jim\'{e}nez et al., 2002]{Jimenez} Jim\'{e}nez C., Haworth D., Cuenot B. and Poinsot T.J., 2002, "Direct numerical simulation and modeling for lean stratified propane-air flames", Combustion $\&$ Flame 128, pp. 1-21.
\bibitem[Li Y. et al., 2004]{cyclic2} Li Y., Zhao H., Leach B., Ma T. and Ladommatos N., 2004, "Characterization of an in-cylinder flow structure in a high-tumble spark ignition engine", International Journal  of Engine Research.
\bibitem[Lumley, 1967]{Lumley} Lumley J. L., 1967, "The structure of inhomogeneous turbulence", In Yaglom, A.M., and Tatarski, V. I. (eds.), Atmospheric turbulence and wave propagation, Moscow,Nauka, pp.166-178.
\bibitem[Pasquier et al., 2007]{Pasquier} Pasquier N., Lecordier B., Trinit\'{e} M. and Cessou A., 2007, "An experimental investigation of flame propagation through a turbulent stratified mixture", Proceedings of the Combustion Institute 31, pp.1567-1574.
\bibitem[Poinsot and Veynante,2005]{poinsot} Poinsot, T. and Veynante, D., 2005, "Theoretical and Numerical Combustion",  R.T. Edwards.
\bibitem[Raffel et al., 2007]{Raffel} Raffel M., Willert C., Wereley S. and Kompenhans J., 2007, "Particle Image Velocimetry: A Practical Guide", Springer-Verlag Berlin Heidelberg.
\bibitem[Riesz and Sz.-Nagy,1990]{Riesz} Riesz F. and Sz.-Nagy B.,1990, "Functional analysis", New York:Dover.
\bibitem[Tsai, 1987]{Tsai} Tsai R.Y., 1987, "A versatile camera calibration technique for high-accuracy 3D machine vision metrology using off-the-shelf TV cameras and lenses", IEEE J. Robot. Autom.
\bibitem[Xiong and Roberts, 2002]{Xiong} Xiong Y. and Roberts W.L., 2002, "Observations on the interaction between a premixed flame kernel and a vortex of different equivalent ratio", Proceedings of the Combustion Institute 29, pp.1687-1693.

\end{thebibliography}

\end{document}

### 2016-09-24 pod_encit/trunk/paper.tex
new lib version

--- 2016-09-24 pod_encit/tags/draft/paper.tex
@@
-\documentclass[10pt,fleqn,a4paper]{article}
-\usepackage{encit2010}
+\documentclass[12pt,a4paper]{article}
+\usepackage[margin=1in]{geometry}
 \usepackage{amsmath}
 \usepackage{amsthm}
+\usepackage{graphicx}
+\usepackage{setspace}
 
@@
 
-% LEGEND FOR THE GEOMETRIES
-% Action = Head 1
-% Scania = Head 2
-% Sygma = Head 3
-% Ducato = Head 4
+\title{Application of proper orthogonal decomposition to measurements of steady engine head flow}
+\author{Rodrigo Piccinini, Carla Fernandes, Marcelo Andreotti.}
 
 \begin{document}
-\hspace{-8.5mm}
-\begin{tabular}{||p{\textwidth}}
-\begin{center}
-\vspace{-4mm}
-\title{ANALYSIS OF STEREOSCOPIC PIV MEASUREMENTS OF IN-CYLINDER FLOWS USING PROPER ORTHOGONAL DECOMPOSITION}
-\end{center}
-\authors{Rodrigo Piccinini, rodrigo.piccinini@sygmamotors.com.br} \\
-\authors{Carla Fernandes, carla.fernandes@sygmamotors.com.br} \\
-\authors{Marcelo Andreotti, marcelo.andreotti@sygmamotors.com.br} \\
-\institution{Sygma Motors, Av. Cassiano Ricardo, 1306, sala 2, Jardim Alvorada, CEP 12240 - 540, S\~{a}o Jos\'{e} dos Campos - SP - Brasil} \\
-\\
-\abstract{\textbf{Abstract}
+\maketitle
 
-% SHORT ABSTRACT
+\doublespacing
 
-%In the present work, the Proper Orthogonal Decomposition (POD) was applied in the
-% identification of flow coherent structures. The experiment consisted of 
-% mounting on an Plexiglas cylinder the intake ports of an internal 
-% combustion engine and establishing a steady flow through this system.
-%Velocity fields at different cylinder cross sections and for several 
-%instants of time were measured by Particle Image Velocimetry (PIV).
-%It was shown that the principal eigenvector might contains up to 80\% of the
-%time averaged kinetic energy of the flow field.
+\abstract{The proper orthogonal decomposition (POD) is applied to particle image velocimetry (PIV) datasets of an engine static flow bench.
+Different flow fields are produced changing intake port design and flow snapshots are taken at a fixed position downstream port exit.
+It is observed that a swirl motion is mostly contained in the first POD mode and that the swirl motion steadiness depends on the amount of kinetic energy in the first mode.
+All higher POD modes showed pairs of counter rotating vortices with zero mean angular momentum.
+}
 
-In the present work, the Proper Orthogonal Decomposition (POD) was applied in the identification of flow coherent structures.
-For the velocity flow field, stereo Particle Image Velocimetry (PIV) measurements were carried out in a fixed cross section.
-We have observed that the principal eigenvector contains up to 80$\%$ of the time averaged kinetic energy of the flow field.
-}
-\\
-\\
-\keywords{\textbf{Keywords:} proper orthogonal decomposition, stereo PIV, coherent structures, turbulent flow, IC engines.}\\
-\end{tabular}
+\section{Introduction}
 
-\section{INTRODUCTION}
+Characterization and understanding of flow field inside internal combustion engines is of primary importance to improvements in engine efficiency and predictability.
+Flow field directly affects mixture formation and turbulence level, both important factors dictating flame front advance.
 
-In the development of cylinder head intake ports for spark-ignited internal combustion engines, the characterization of coherent flow structures generated during induction - the most usual being tumble and swirl motions - is an important issue for combustion acceleration.
-
-The flow field will directly affect the mixture formation and hence the spatial fuel distribution inside the combustion chamber.
-A spatially non-homogeneous mixture leads to changes in flame behavior in terms of heat release, propagation speed, etc.
-This behavior is not well understood and has motivated the research reported in numerous studies (Haworth et al., 2000; Jim\'{e}nez et al., 2002; Pasquier et al., 2007; H\'{e}lie and Trouv\'{e}, 2000; Xiong and Roberts, 2002).
-The characteristics of the combustion depend strongly on local variations of mixture and also on the length-scale of these variations.
-
-In a general way, higher turbulence levels during flame development reduce combustion duration.
-A common strategy to achieve these higher turbulence levels consists of disintegrating large flow structures into turbulence near the spark release event.
-In this way, the kernel formation will depend on the local equivalence ratio and temperature around the spark plug.
-The flow field might also advect the initial kernel and the local turbulence will influence its evolution to a turbulent flame.
-The turbulent flame for the premixed thin reaction sheet will grow faster than the laminar flame due to the large scale turbulence ($l_{t}/\delta^{0}_{L} > 1$, where $l_{t}$ is the integral length scale and $\delta^{0}_{L}$ is the laminar flame thickness (Poinsot and Veynante, 2005)).
-A faster combustion event is desired because it will increase the efficiency of converting chemical energy to piston work.
-Therefore, the majority of combustion strategies will attempt to create such desirable high turbulence levels near the ignition.
-
-For this purpose, the creation of a strong swirling flow inside the combustion chamber is often studied.
-The idea is to store a great amount of kinetic energy in large structures during the intake.
-During the compression, these structures will remain until the ignition event and then they will break into smaller structures increasing the turbulence level.
-In case of swirling structures not achieved during the intake, the kinetic energy of the flow might dissipate faster.
-
-Furthermore, the turbulent flow is characterized by random and chaotic three-dimensional motion.
-In this way, the flow will differently affect the formation of large structures cycle-by-cycle, rendering the engine operation difficult.
-Thus, not only the magnitude of the swirling motion is important, but also its regularity.
-Therefore, the knowledge of the bulk motion in the cylinder and the turbulence evolution is of great importance for the improvement of engines.
+In the development of intake ports for spark-ignited internal combustion engines, the characterization of coherent flow structures generated during intake - most usual ones being tumble and swirl motions - is an important issue for combustion acceleration.
 
 The objective of the present work is the study of the statistical properties of the flow induced by the intake ports in four different cylinder heads using proper orthogonal decomposition (POD).
-In order to obtain the velocity flow field, the stereo particle image velocimetry (PIV) measurements were carried out in a specific cylinder cross section.
+In order to obtain the velocity flow field, the stereo particle image velocimetry (PIV) was performed in a specific cylinder cross section.
 This technique enables spatially resolved measurements of the three instantaneous components of the flow velocity field within a very short time and allows the detection of large and small scale spatial structures.
-Once the velocity field was acquired, the POD was applied in order to know how the flow energy is stored.
-This statistical analysis demonstrates a promising tool in the design of intake ports.
+Once the velocity field was acquired, the POD was applied in order to know how the flow energy was stored.
+This statistical analysis suggests to be a promising tool in the design of intake ports.
 
@@
 This arrangement permits the most accurate determination of the out-of-plane displacement (Raffel et al., 2007).
-The camera bodies were tilted with respect to the optical axis of their lenses.
+The camera bodies were tilted with respect to the optical axis of
+their lenses.
 This is done to satisfy the Scheimpflug condition, i.e., to focus over the whole imaged measurement plane, as shown in Fig.~\ref{fig: bench}.
@@
 
-The characterization of coherent structures is an interesting issue in the study of turbulent flows and it is the subject of vast literature (Heywood, 1988; Cosadia et al., 2006; Cazemier, 1997; Bergmann and Cordier, 2003, Hung and Hien, 1998; Boree, Maurel and Bazile, 2002; Li Y.
+Characterization of coherent structures is an interesting issue in the study of turbulent flows and it is the subject of vast literature (Heywood, 1988; Cosadia et al., 2006; Cazemier, 1997; Bergmann and Cordier, 2003, Hung and Hien, 1998; Boree, Maurel and Bazile, 2002; Li Y.
 et al., 2004).
@@
 Furthermore, each observation must be taken in a different instant of time.
-In the present work, one observation is made up of the full procedure required by the PIV technique (as described in the Section 2) to provide a velocity field in some instant of time.
+In the present work, one observation is made up of the full procedure required by the PIV technique (as described in the Section 2) to provide the velocity field in some instant of time.
 The frames are taken at different times equally spaced and each event is assumed linearly independent.
@@
 
-According to Eq.~\eqref{eq: max-cond}, $\bv{\Phi}$ can not be zero because his norm equals one. Therefore:
+According to Eq.~\eqref{eq: max-cond}, $\bv{\Phi}$ can not be zero because its norm equals one. Therefore:
 
@@
 
-After performing the integrations and having the matrix $C$ (Eq.~\eqref{cov1}), the computation of eigenvalues and eigenvectors is carried out with the opensource library SciPy.
+After performing the integrations and having the matrix $C$ (Eq.~\eqref{cov1}), the computation of eigenvalues and eigenvectors is performed with the opensource library SciPy.
 
@@
 The Figure~\ref{fig: energy_fraction}(b) shows the cumulative energy when the modes are grouped from $\bv{\Phi}_1$ to $\bv{\Phi}_{100}$.
-For higher valve lifts, the flow field has shown more organized in the sense that more energy is stored in few modes.
+For higher valve lifts, the flow field is more organized in the sense that more energy is stored in fewer modes.
 Only the first mode, $\bv{\Phi}_1$, contains up to 80$\%$ of the total energy (see the zoom of the Fig.~\ref{fig: energy_fraction}(a)).
@@
 
-\begin{figure}[h]
+\begin{figure}
  \centering
@@
 
-\clearpage
-
 The link between energy content and the flow organization is supported by the observation of the mode velocity field presented in Fig.~\ref{fig: organization}.
 Here, the vector magnitudes are normalized according to the same norm defined in Eq.~\eqref{eq: max-cond}.
-It is possible to notice that the motion organization decreases with the higher order modes.
-Furthermore, these modes presents several vortical structures while the first mode contains the majority of swirling structure.
-Hence, $\bv{\Phi}_1$ have the majority of the angular momentum.
+It is possible to notice that the motion organization decreases in the higher order modes.
+Furthermore, these modes presents several vortical structures while the first mode contains one major swirling structure.
+It will be shown later that $\bv{\Phi}_1$ contains the greatest amount of the angular momentum.
 
@@
 However, if the $b^{s}_{j}$ coefficients from Eq.~\eqref{eq: Us} are computed, the energy time history is recovered.
-The Figure~\ref{fig: energy_hist} presents the energy history for the first mode and the set of the first ten modes.
+The Figure~\ref{fig: energy_hist} presents the energy history for the first mode and the set of the first ten modes ($\bv{\Phi}_1 + \bv{\Phi}_2 + ...
++\bv{\Phi}_{10}$).
 It is seen that the first mode is in general more oscillatory when viewed separately in the gray line.
-This is rather consistent since energy transfers between the modes are expected.
+This is rather consistent since energy transfers between the modes are expected from the nonlinearity of the Navier-Stokes equations.
 The energy decrease in the first mode possibly increases the energy of some other mode in the set and the other way around.
-This energy transfer will reduce the oscillations.
+This energy transfer reduces the oscillations of the set.
 
-The Figure~\ref{fig: Lz} shows the contribution of each mode to angular momentum around the cylinder axis.
-It is observed that the first mode contains the main swirling motion while the subsequent modes corresponds to the oscillating motion as shown in Fig.~\ref{fig: organization}.
-As one could expect, the low valve lift of 2\,mm did not clearly define swirl motion because of great magnitude oscillations.
+The Figure~\ref{fig: Lz} shows the contribution of each mode to the angular momentum around the cylinder axis.
+It is observed that the first mode contains the main swirling motion and the subsequent modes correspond to the oscillations shown in Fig.~\ref{fig: organization}.
+As one could expect, the low valve lift of 2\,mm did not clearly define a swirling motion because of the great magnitude oscillations.
 
-In order to define which intake port geometry will better perform, in terms of increase the turbulence level close to spark ignition event, not only the energy in the modes most be analyzed but also their respective angular momentum.
 As showed in Fig.~\ref{fig: energy_hist}, the results (c) and (d) have a similar energy history level in the first mode and for the set of ten modes.
-It is mean that the Head 2 and 3 distribute the energy in a similar way.
-This same trend is also observed in terms of energy spectra (see Fig.~\ref{fig: energy_fraction}).
+It means that the Heads 2 and 3 distribute the energy in a similar way.
+This same trend is also observed in the  energy spectra (see Fig.~\ref{fig: energy_fraction}).
 However, the angular momentum is quite different as presented in Fig.~\ref{fig: Lz} by the plots (c) and (d).
-Comparing the results of the first mode velocity field presented by Fig.~\ref{fig: scaniavssygma}, it is noticed that the Head 3 velocity field is composed by two-vortex structure against only one big vortex in the Head 2.
-This is the main reason in the angular momentum production.
+Comparing the results of the first mode velocity field presented by Fig.~\ref{fig: scaniavssygma}, we see that the Head 3 velocity field is composed by a two-vortex structure against only one big vortex in the Head 2.
+This would explain the different angular momenta.
 
@@
 
-In order to better understand how the intake ports geometry influence the flow which enters inside the combustion chamber during the intake phase, a statistical decomposition, like proper orthogonal decomposition, was applied on the three dimensional velocity flow field obtained with a stereoscopic PIV.
+A mathematical tool called Proper Orthogonal Decomposition was used along with the experimental technique of stereoscopic PIV in the study of the in-cylinder flow generated by the intake ports of an internal combustion engine.
 
 The results have shown that the first mode contains up to 80$\%$ of the time average kinetic energy of the flow field and have the major swirling structure.
-Therefore, the first mode defines the average angular momentum.
-The other modes, which we can call low energy modes, have shown several vortical structures, as presented in Fig.~\ref{fig: organization}, and have an average angular momentum around zero (see Fig.~\ref{fig: Lz}).
-
 In what concerns the flow regularity, Fig.~\ref{fig: Lz} shows that the first mode basically defines the average angular momentum, which is superposed by the contributions of higher modes with average angular momentum around zero.
@@
 Even if a mean swirl is measured, we have shown that the structure of the mean flow is complex and difficult to interpret.
-Thus, POD has shown itself to be an important methodology to analyze the flow and an important tool for the design of the cylinder head.
+Thus, POD has shown itself to be an important methodology to analyze the flow and an important tool in the design of the cylinder head.
 

### 2016-09-24 lowmachspray/fe32/lowmachLib/injector/commonRailInjector/commonRailInjector.C

@@ -1,0 +1,389 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "commonRailInjector.H"
+#include "addToRunTimeSelectionTable.H"
+#include "Random.H"
+#include "mathematicalConstants.H"
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+namespace Foam
+{
+
+defineTypeNameAndDebug(commonRailInjector, 0);
+
+addToRunTimeSelectionTable
+(
+    injectorType,
+    commonRailInjector,
+    dictionary
+);
+}
+// * * * * * * * * * * * * * Private Member Functions  * * * * * * * * * * * //
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+// Construct from components
+Foam::commonRailInjector::commonRailInjector
+(
+    const Foam::Time& t,
+    const Foam::dictionary& dict
+)
+:
+    injectorType(t, dict),
+    propsDict_(dict.subDict(typeName + "Props")),
+    position_(propsDict_.lookup("position")),
+    direction_(propsDict_.lookup("direction")),
[…]

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/Chomiak.H

@@ -1,0 +1,1 @@
+../spraySubModels/injectorModel/Chomiak/Chomiak.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/RanzMarshall.H

@@ -1,0 +1,1 @@
+../spraySubModels/heatTransferModel/RanzMarshall/RanzMarshall.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/RutlandFlashBoil.H

@@ -1,0 +1,1 @@
+../spraySubModels/evaporationModel/RutlandFlashBoil/RutlandFlashBoil.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/TAB.C

@@ -1,0 +1,1 @@
+../spraySubModels/breakupModel/TAB/TAB.C

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/blobsSheetAtomization.H

@@ -1,0 +1,1 @@
+../spraySubModels/atomizationModel/blobsSheetAtomization/blobsSheetAtomization.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/boundaryTreatment.H

@@ -1,0 +1,1 @@
+../parcel/boundaryTreatment.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/breakupModel.H

@@ -1,0 +1,1 @@
+../spraySubModels/breakupModel/breakupModel/breakupModel.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/collisionModel.H

@@ -1,0 +1,1 @@
+../spraySubModels/collisionModel/collisionModel/collisionModel.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/constInjector.H

@@ -1,0 +1,1 @@
+../spraySubModels/injectorModel/constant/constInjector.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/definedHollowCone.H

@@ -1,0 +1,1 @@
+../spraySubModels/injectorModel/definedHollowCone/definedHollowCone.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/dispersionModel.C

@@ -1,0 +1,1 @@
+../spraySubModels/dispersionModel/dispersionModel/dispersionModel.C

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/dragModel.C

@@ -1,0 +1,1 @@
+../spraySubModels/dragModel/dragModel/dragModel.C

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/gradientDispersionRAS.H

@@ -1,0 +1,1 @@
+../spraySubModels/dispersionModel/gradientDispersionRAS/gradientDispersionRAS.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/hollowCone.H

@@ -1,0 +1,1 @@
+../spraySubModels/injectorModel/hollowCone/hollowCone.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/injectorI.H

@@ -1,0 +1,1 @@
+../injector/injector/injectorI.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/injectorType.C

@@ -1,0 +1,1 @@
+../injector/injectorType/injectorType.C

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/multiHoleInjector.C

@@ -1,0 +1,1 @@
+../injector/multiHoleInjector/multiHoleInjector.C

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/noAtomization.C

@@ -1,0 +1,1 @@
+../spraySubModels/atomizationModel/noAtomization/noAtomization.C

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/noBreakup.C

@@ -1,0 +1,1 @@
+../spraySubModels/breakupModel/noBreakup/noBreakup.C

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/noHeatTransfer.H

@@ -1,0 +1,1 @@
+../spraySubModels/heatTransferModel/noHeatTransfer/noHeatTransfer.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/parcel.C

@@ -1,0 +1,1 @@
+../parcel/parcel.C

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/parcel.H

@@ -1,0 +1,1 @@
+../parcel/parcel.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/parcelIO.C

@@ -1,0 +1,1 @@
+../parcel/parcelIO.C

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/pressureSwirlInjector.H

@@ -1,0 +1,1 @@
+../spraySubModels/injectorModel/pressureSwirl/pressureSwirlInjector.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/reitzKHRT.H

@@ -1,0 +1,1 @@
+../spraySubModels/breakupModel/reitzKHRT/reitzKHRT.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/removeParcel.C

@@ -1,0 +1,1 @@
+../spraySubModels/wallModel/removeParcel/removeParcel.C

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/spray.H

@@ -1,0 +1,1 @@
+../spray/spray.H

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/sprayFunctions.C

@@ -1,0 +1,1 @@
+../spray/sprayFunctions.C

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/sprayInject.C

@@ -1,0 +1,1 @@
+../spray/sprayInject.C

### 2016-09-24 lowmachspray/fe32/lowmachLib/lnInclude/stochasticDispersionRAS.C

@@ -1,0 +1,1 @@
+../spraySubModels/dispersionModel/stochasticDispersionRAS/stochasticDispersionRAS.C

### 2016-09-24 lowmachspray/fe32/lowmachLib/parcel/setRelaxationTimes.C

@@ -1,0 +1,299 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "parcel.H"
+
+#include "spray.H"
+#include "dragModel.H"
+#include "evaporationModel.H"
+#include "heatTransferModel.H"
+#include "basicMultiComponentMixture.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * * Member Functions  * * * * * * * * * * * * * //
+
+void parcel::setRelaxationTimes
+(
+    label celli,
+    scalar& tauMomentum,
+    scalarField& tauEvaporation,
+    scalar& tauHeatTransfer,
+    scalarField& tauBoiling,
+    const spray& sDB,
+    const scalar rho,
+    const vector& Up,
+    const scalar temperature,
+    const scalar pressure,
+    const scalarField& Yfg,
+    const scalarField& m0,
+    const scalar dt
+)
+{
+
+    const liquidMixture& fuels = sDB.fuels();
+
+    scalar mCell = rho*sDB.mesh().V()[cell()];
+    scalarField mfg(Yfg*mCell);
+
+    label Ns = sDB.composition().Y().size();
+    label Nf = fuels.components().size();
+
+    // Tf is based on the 1/3 rule
+    scalar Tf  = T() + (temperature - T())/3.0;
+
[…]

### 2016-09-24 lowmachspray/fe32/lowmachLib/spray/sprayI.H

@@ -1,0 +1,384 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * * Member Functions  * * * * * * * * * * * * * //
+
+inline const Time& spray::runTime() const
+{
+    return runTime_;
+}
+
+
+inline const fvMesh& spray::mesh() const
+{
+    return mesh_;
+}
+
+
+inline const volVectorField& spray::U() const
+{
+    return U_;
+}
+
+
+inline const volScalarField& spray::rho() const
+{
+    return rho_;
+}
+
+
+inline const volScalarField& spray::p() const
+{
+    return p_;
+}
+
+
+inline const volScalarField& spray::T() const
+{
+    return T_;
+}
+
+
+inline PtrList<injector>& spray::injectors()
+{
+    return injectors_;
+}
+
+
+inline const PtrList<injector>& spray::injectors() const
+{
+    return injectors_;
+}
+
+
+inline const atomizationModel& spray::atomization() const
+{
+    return atomization_;
+}
+
+
+inline const breakupModel& spray::breakup() const
+{
+    return breakupModel_;
+}
+
+
[…]

### 2016-09-24 lowmachspray/fe32/lowmachLib/spraySubModels/atomizationModel/atomizationModel/atomizationModel.C

@@ -1,0 +1,64 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "error.H"
+#include "atomizationModel.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+defineTypeNameAndDebug(atomizationModel, 0);
+
+defineRunTimeSelectionTable(atomizationModel, dictionary);
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+// Construct from components
+atomizationModel::atomizationModel
+(
+    const dictionary& dict,
+    spray& sm
+)
+:
+    dict_(dict),
+    spray_(sm),
+    rndGen_(sm.rndGen())
+{}
+
+
+// * * * * * * * * * * * * * * * * Destructor    * * * * * * * * * * * * * * //
+
+atomizationModel::~atomizationModel()
+{}
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+} // End namespace Foam
+
+// ************************************************************************* //

### 2016-09-24 lowmachspray/fe32/lowmachLib/spraySubModels/breakupModel/SHF/SHF.C

@@ -1,0 +1,286 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "SHF.H"
+#include "addToRunTimeSelectionTable.H"
+#include "mathematicalConstants.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+defineTypeNameAndDebug(SHF, 0);
+
+addToRunTimeSelectionTable
+(
+    breakupModel,
+    SHF,
+    dictionary
+);
+
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+// Construct from components
+SHF::SHF
+(
+    const dictionary& dict,
+    spray& sm
+)
+:
+    breakupModel(dict, sm),
+    coeffsDict_(dict.subDict(typeName + "Coeffs")),
+    g_(sm.g()),
+    rndGen_(sm.rndGen()),
[…]

### 2016-09-24 lowmachspray/fe32/lowmachLib/spraySubModels/collisionModel/noCollision/noCollision.H

@@ -1,0 +1,87 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::noCollision
+
+Description
+    No collisions between parcels
+    All parcels move independently of eachother
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef noCollision_H
+#define noCollision_H
+
+#include "collisionModel.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+/*---------------------------------------------------------------------------*\
+                           Class noCollision Declaration
+\*---------------------------------------------------------------------------*/
+
+class noCollision
+:
+    public collisionModel
+{
+
+public:
+
+    //- Runtime type information
+    TypeName("off");
+
+
+    // Constructors
+
+        //- Construct from components
+        noCollision
+        (
+            const dictionary& dict,
+            spray& sm,
+            Random& rndGen
+        );
+
+    // Destructor
+
+        ~noCollision();
+
+
+    // Member Functions
+
+        void collideParcels(const scalar dt) const;
+};
+
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+} // End namespace Foam
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+#endif
+
+// ************************************************************************* //

### 2016-09-24 lowmachspray/fe32/lowmachLib/spraySubModels/dispersionModel/noDispersion/noDispersion.H

@@ -1,0 +1,86 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::noDispersion
+
+Description
+    No turbulent dispersion of the parcels
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef noDispersion_H
+#define noDispersion_H
+
+#include "dispersionModel.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+/*---------------------------------------------------------------------------*\
+                           Class noDispersion Declaration
+\*---------------------------------------------------------------------------*/
+
+class noDispersion
+:
+    public dispersionModel
+{
+
+public:
+
+    //- Runtime type information
+    TypeName("off");
+
+
+    // Constructors
+
+        //- Construct from components
+        noDispersion
+        (
+            const dictionary& dict,
+            spray& sm
+        );
+
+
+    // Destructor
+
+        ~noDispersion();
+
+
+    // Member Functions
+
+    void disperseParcels() const;
+};
+
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+} // End namespace Foam
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+#endif
+
+// ************************************************************************* //

### 2016-09-24 lowmachspray/fe32/lowmachLib/spraySubModels/dispersionModel/stochasticDispersionRAS/stochasticDispersionRAS.C

@@ -1,0 +1,141 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "error.H"
+
+#include "stochasticDispersionRAS.H"
+#include "addToRunTimeSelectionTable.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+defineTypeNameAndDebug(stochasticDispersionRAS, 0);
+
+addToRunTimeSelectionTable
+(
+    dispersionModel,
+    stochasticDispersionRAS,
+    dictionary
+);
+
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+// Construct from components
+stochasticDispersionRAS::stochasticDispersionRAS
+(
+    const dictionary& dict,
+    spray& sm
+)
+:
+    dispersionRASModel(dict, sm)
+{}
+
+
+// * * * * * * * * * * * * * * * * Destructor  * * * * * * * * * * * * * * * //
+
+stochasticDispersionRAS::~stochasticDispersionRAS()
+{}
+
+
+// * * * * * * * * * * * * * * * Member Functions  * * * * * * * * * * * * * //
+
+void stochasticDispersionRAS::disperseParcels() const
+{
+
[…]

### 2016-09-24 lowmachspray/fe32/lowmachLib/spraySubModels/evaporationModel/RutlandFlashBoil/RutlandFlashBoil.C

@@ -1,0 +1,362 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "error.H"
+
+#include "RutlandFlashBoil.H"
+#include "addToRunTimeSelectionTable.H"
+#include "mathematicalConstants.H"
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+defineTypeNameAndDebug(RutlandFlashBoil, 0);
+
+addToRunTimeSelectionTable
+(
+    evaporationModel,
+    RutlandFlashBoil,
+    dictionary
+);
+
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+// Construct from dictionary
+RutlandFlashBoil::RutlandFlashBoil
+(
+    const dictionary& dict
+)
+:
+    evaporationModel(dict),
+    evapDict_(dict.subDict(typeName + "Coeffs")),
+    preReScFactor_(readScalar(evapDict_.lookup("preReScFactor"))),
+    ReExponent_(readScalar(evapDict_.lookup("ReExponent"))),
[…]

### 2016-09-24 lowmachspray/fe32/lowmachLib/spraySubModels/heatTransferModel/noHeatTransfer/noHeatTransfer.H

@@ -1,0 +1,106 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::noHeatTransfer
+
+Description
+    no heat transfer. return a negative value to prevent
+    the liquid being heated up.
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef noHeatTransfer_H
+#define noHeatTransfer_H
+
+#include "heatTransferModel.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+/*---------------------------------------------------------------------------*\
+                           Class noHeatTransfer Declaration
+\*---------------------------------------------------------------------------*/
+
+class noHeatTransfer
+:
+    public heatTransferModel
+{
+
+public:
+
+    //- Runtime type information
+        TypeName("off");
+
+
+    // Constructors
+
+        //- Construct from dictionary
+        noHeatTransfer
+        (
+            const dictionary& dict
+        );
+
+
+    // Destructor
+
+        ~noHeatTransfer();
+
+
+    // Member Functions
+
+        bool heatTransfer() const;
+
+        //- Nusselt Number
+        scalar Nu
+        (
+            const scalar ReynoldsNumber,
+            const scalar PrandtlNumber
+        ) const;
+
+        scalar relaxationTime
+        (
+            const scalar liquidDensity,
+            const scalar diameter,
[…]

### 2016-09-24 lowmachspray/fe32/lowmachLib/spraySubModels/wallModel/reflectParcel/reflectParcel.C

@@ -1,0 +1,195 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "reflectParcel.H"
+#include "addToRunTimeSelectionTable.H"
+#include "wallPolyPatch.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+defineTypeNameAndDebug(reflectParcel, 0);
+
+addToRunTimeSelectionTable
+(
+    wallModel,
+    reflectParcel,
+    dictionary
+);
+
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+// Construct from components
+reflectParcel::reflectParcel
+(
+    const dictionary& dict,
+    const volVectorField& U,
+    spray& sm
+)
+:
+    wallModel(dict, U, sm),
+    U_(U),
+    coeffsDict_(dict.subDict(typeName + "Coeffs")),
+    elasticity_(readScalar(coeffsDict_.lookup("elasticity")))
+{}
+
+
+// * * * * * * * * * * * * * * * * Destructor  * * * * * * * * * * * * * * * //
+
+reflectParcel::~reflectParcel()
+{}
+
+
[…]

### 2016-09-24 lowmachspray/fe32/lowmachLib/spraySubModels/wallModel/reflectParcel/reflectParcel.H

@@ -1,0 +1,107 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::reflectParcel
+
+Description
+    The parcels are reflected back from the wall in an inelastic collision.
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef reflectParcel_H
+#define reflectParcel_H
+
+#include "wallModel.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+/*---------------------------------------------------------------------------*\
+                           Class reflectParcel Declaration
+\*---------------------------------------------------------------------------*/
+
+class reflectParcel
+:
+    public wallModel
+{
+
+private:
+    // Private data
+
+        const volVectorField& U_;
+
+        dictionary coeffsDict_;
+
+    // model constants
+
+    //- Elasticity constant v^2_out = elasticity * v^2_in
+        scalar elasticity_;
+
+public:
+
+    //- Runtime type information
+    TypeName("reflect");
+
+
+    // Constructors
+
+        //- Construct from components
+        reflectParcel
+        (
+            const dictionary& dict,
+            const volVectorField& U,
+            spray& sm
+        );
+
+
+    // Destructor
+
+        virtual ~reflectParcel();
+
+
+    // Member Functions
+
+        //- Return true if parcel is to be kept, and false if it is to be
+        // removed
[…]

### 2016-09-24 lowmachspray/fe32/lowmachLib/spraySubModels/wallModel/removeParcel/removeParcel.C

@@ -1,0 +1,84 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "removeParcel.H"
+#include "addToRunTimeSelectionTable.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+defineTypeNameAndDebug(removeParcel, 0);
+
+addToRunTimeSelectionTable
+(
+    wallModel,
+    removeParcel,
+    dictionary
+);
+
+// * * * * * * * * * * * * * Private Member Functions  * * * * * * * * * * * //
+
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+// Construct from components
+removeParcel::removeParcel
+(
+    const dictionary& dict,
+    const volVectorField& U,
+    spray& sm
+)
+:
+    wallModel(dict, U, sm)
+{}
+
+
+// * * * * * * * * * * * * * * * * Destructor  * * * * * * * * * * * * * * * //
+
+removeParcel::~removeParcel()
+{}
+
+
+// * * * * * * * * * * * * * * * Member Functions  * * * * * * * * * * * * * //
+
+bool removeParcel::wallTreatment
+(
+    parcel&,
+    const label facei
+) const
+{
+    return false;
+}
+
+
[…]

### 2016-09-24 lowmachspray/fe32/myLiquids/aC3H6O/aC3H6O.C

@@ -1,0 +1,149 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "aC3H6O.H"
+#include "addToRunTimeSelectionTable.H"
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+namespace Foam
+{
+    defineTypeNameAndDebug(aC3H6O, 0);
+    addToRunTimeSelectionTable(liquid, aC3H6O,);
+    addToRunTimeSelectionTable(liquid, aC3H6O, Istream);
+}
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+Foam::aC3H6O::aC3H6O()
+:
+    liquid
+    (
+        60.096,
+        508.31,
+        4.7643e+6,
+        0.22013,
+        0.248,
+        185.28,
+        3.20e-2,
+        355.41,
+        5.5372e-30,
+        0.6689,
+        2.3575e+4
+    ),
+    rho_(70.91328, 0.26475, 508.31, 0.243),
[…]

### 2016-09-24 lowmachspray/fe32/myLiquids/aC3H6O/aC3H6O.H

@@ -1,0 +1,208 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::aC3H6O
+
+Description
+    iso-propanol
+
+SourceFiles
+    aC3H6O.C
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef aC3H6O_H
+#define aC3H6O_H
+
+#include "liquid.H"
+#include "NSRDSfunc0.H"
+#include "NSRDSfunc1.H"
+#include "NSRDSfunc2.H"
+#include "NSRDSfunc3.H"
+#include "NSRDSfunc4.H"
+#include "NSRDSfunc5.H"
+#include "NSRDSfunc6.H"
+#include "NSRDSfunc7.H"
+#include "NSRDSfunc14.H"
+
+#include "cpFunc.H"
+#include "rhoFunc.H"
+#include "pvFunc.H"
+#include "DFunc.H"
+#include "muFunc.H"
+#include "KFunc.H"
+
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+/*---------------------------------------------------------------------------*\
+                           Class aC3H6O Declaration
+\*---------------------------------------------------------------------------*/
+
+class aC3H6O
+:
+    public liquid
+{
+    // Private data
+
+        rhoFunc rho_;
+        pvFunc pv_;
[…]

### 2016-09-24 lowmachspray/fe32/myLiquids/aC3H6O/lnInclude/aC3H6OI.H

@@ -1,0 +1,1 @@
+../aC3H6OI.H

### 2016-09-24 lowmachspray/fe32/myLiquids/myFunctions/cpFunc/cpFunc.H

@@ -1,0 +1,149 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::cpFunc
+
+Description
+    NSRDS function number 101
+
+    Source:
+    @verbatim
+                      NSRDS - AICHE
+                 Data Compilation Tables
+                    of Properties of
+                     Pure Compounds
+
+        Design Institute for Physical Property Data
+          American Institute of Chemical Engineers
+                  345 East 47th Street
+                New York, New York 10017
+
+         National Standard Reference Data System
+         American Institute of Chemical Engineers
+
+          T.E. Daubert       -       R.P. Danner
+
+            Department of Chemical Engineering
+            The Pennsylvania State University
+                University Park, PA 16802
+    @endverbatim
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef cpFunc_H
+#define cpFunc_H
+
+#include "thermophysicalFunction.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+
+/*---------------------------------------------------------------------------*\
+                           Class cpFunc Declaration
+\*---------------------------------------------------------------------------*/
+
+class cpFunc
+:
+    public thermophysicalFunction
+{
+    // Private data
+
+        // NSRDS function 101 coefficients
[…]

### 2016-09-24 lowmachspray/fe32/myLiquids/myFunctions/lnInclude/DFunc.C

@@ -1,0 +1,1 @@
+../DFunc/DFunc.C

### 2016-09-24 lowmachspray/fe32/myLiquids/myFunctions/lnInclude/rhoFunc.H

@@ -1,0 +1,1 @@
+../rhoFunc/rhoFunc.H

### 2016-09-24 lowmachspray/fe32/mypdfs/exponential/exponential.C

@@ -1,0 +1,81 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "exponential.H"
+#include "addToRunTimeSelectionTable.H"
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+namespace Foam
+{
+    namespace pdfs
+    {
+        defineTypeNameAndDebug(exponential, 0);
+        addToRunTimeSelectionTable(pdf, exponential, dictionary);
+    }
+}
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+Foam::pdfs::exponential::exponential(const dictionary& dict, Random& rndGen)
+:
+    pdf(typeName, dict, rndGen),
+    minValue_(readScalar(pdfDict_.lookup("minValue"))),
+    maxValue_(readScalar(pdfDict_.lookup("maxValue"))),
+    lambda_(readScalar(pdfDict_.lookup("lambda")))
+{
+    check();
+}
+
+
+// * * * * * * * * * * * * * * * * Destructor  * * * * * * * * * * * * * * * //
+
+Foam::pdfs::exponential::~exponential()
+{}
+
+
+// * * * * * * * * * * * * * * * Member Functions  * * * * * * * * * * * * * //
+
+Foam::scalar Foam::pdfs::exponential::sample() const
+{
[…]

### 2016-09-24 lowmachspray/fe32/mypdfs/normal/normal.C

@@ -1,0 +1,121 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+\*---------------------------------------------------------------------------*/
+
+#include "normal.H"
+#include "addToRunTimeSelectionTable.H"
+#include "mathematicalConstants.H"
+
+// * * * * * * * * * * * * * * Static Data Members * * * * * * * * * * * * * //
+
+namespace Foam
+{
+    namespace pdfs
+    {
+        defineTypeNameAndDebug(normal, 0);
+        addToRunTimeSelectionTable(pdf, normal, dictionary);
+    }
+}
+
+// * * * * * * * * * * * * * * * * Constructors  * * * * * * * * * * * * * * //
+
+Foam::pdfs::normal::normal(const dictionary& dict, Random& rndGen)
+:
+    pdf(typeName, dict, rndGen),
+    minValue_(readScalar(pdfDict_.lookup("minValue"))),
+    maxValue_(readScalar(pdfDict_.lookup("maxValue"))),
+    expectation_(readScalar(pdfDict_.lookup("expectation"))),
+    variance_(readScalar(pdfDict_.lookup("variance"))),
+    a_(0.147)
+{
+    if (minValue_ < 0)
+    {
+        FatalErrorIn("normal::normal(const dictionary&, Random&)")
+            << "Minimum value must be greater than zero. "
[…]

### 2016-09-24 lowmachspray/fe32/mypdfs/uniform/uniform.H

@@ -1,0 +1,109 @@
+/*---------------------------------------------------------------------------*\
+  =========                 |
+  \\      /  F ield         | OpenFOAM: The Open Source CFD Toolbox
+   \\    /   O peration     |
+    \\  /    A nd           | Copyright (C) 1991-2010 OpenCFD Ltd.
+     \\/     M anipulation  |
+-------------------------------------------------------------------------------
+License
+    This file is part of OpenFOAM.
+
+    OpenFOAM is free software: you can redistribute it and/or modify it
+    under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    OpenFOAM is distributed in the hope that it will be useful, but WITHOUT
+    ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
+    FITNESS FOR A PARTICULAR PURPOSE.  See the GNU General Public License
+    for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with OpenFOAM.  If not, see <http://www.gnu.org/licenses/>.
+
+Class
+    Foam::uniform
+
+Description
+    uniform pdf
+
+SourceFiles
+    uniform.C
+
+\*---------------------------------------------------------------------------*/
+
+#ifndef uniform_H
+#define uniform_H
+
+#include "pdf.H"
+
+// * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * * //
+
+namespace Foam
+{
+namespace pdfs
+{
+
+/*---------------------------------------------------------------------------*\
+                           Class uniform Declaration
+\*---------------------------------------------------------------------------*/
+
+class uniform
+:
+    public pdf
+{
+    // Private data
+
+        //- Distribution minimum
+        scalar minValue_;
+
+        //- Distribution maximum
+        scalar maxValue_;
+
+        //- Distribution range
+        scalar range_;
+
+
+public:
+
+    //- Runtime type information
+    TypeName("uniform");
+
+
+    // Constructors
+
+        //- Construct from components
+        uniform
+        (
+            const dictionary& dict,
+            Random& rndGen
+        );
+
+
+    //- Destructor
+    virtual ~uniform();
+
+
+    // Member Functions
+
+         //- Sample the pdf
+        virtual scalar sample() const;
+
+        //- Return the minimum value
+        virtual scalar minValue() const;
+
+        //- Return the maximum value
[…]

### 2016-09-24 pod_encit/trunk/python/andreotti/dados/Lz.py

@@ -1,0 +1,23 @@
+#!/usr/bin/env python
+from pylab import *
+
+def main(outputdir,n):
+	b=loadtxt(outputdir+'../b.dat')
+	modeLz=zeros(b.shape[1])
+	Lz=zeros([b.shape[0],n])
+	for i in range(n):
+		modeLz[i]=loadtxt(outputdir+'Lz.'+str(i)+'.csv',skiprows=1,usecols=[0],delimiter=',')
+	
+	for i in range(n):
+		Lz[:,i]=modeLz[i]*b[:,i]
+	
+	savetxt(outputdir+'../Lz.dat',Lz)
+
+outputdir='/home/piccinini/pod_encit/trunk/python/andreotti/dados/dir_close_esq_open/pods/Lz/'
+main(outputdir,10)
+
+#outputdir='/home/piccinini/pod_encit/trunk/python/andreotti/dados/duas_abertas/pods/Lz/'
+#main(outputdir,10)
+
+#outputdir='/home/piccinini/pod_encit/trunk/python/andreotti/dados/duas_meio_abertas/pods/Lz/'
+#main(outputdir,10)

### 2016-09-24 pod_encit/trunk/python/sygma/snaps/plt2vtk.py

@@ -1,0 +1,18 @@
+#!/usr/bin/env python
+import visit_writer
+from pylab import *
+import glob
+
+files=glob.glob('sygma/pods/*.plt')
+
+for file in files:
+    x=loadtxt(file, skiprows=4)
+    pts=[]
+    vec=[]
+    for i in range(len(x)):
+        if sqrt(x[i,0]**2+x[i,1]**2) <= 64:
+            pts=pts+[x[i,0],x[i,1],x[i,2]]
+            vec=vec+[x[i,3],x[i,4],x[i,5]]
+
+    vars = (("data", 3, 1, vec), ("ptsvec", 3, 1, pts))
+    visit_writer.WritePointMesh(file[:-4]+'.vtk', 1, pts, vars)
